### 2015-05-08 ms.Rmd
Setup paper to build via remake

---
title: "TREE: A package for modelling plant TRait Ecology and Evolution"
author: "Daniel Falster, Rich FitzJohn, Mark Westoby"
affiliation: Biological sciences, Macquarie University NSW 2109
output: pdf_document
csl: downloads/style.csl
bibliography: refs.bib
---

# Abstract

Population dynamics in forest communities are strongly size-structured: larger
plants shade smaller plants while also expending proportionately more energy
on building and maintaining woody stems. Although the importance of size-
structure for demography is widely recognised, many mechanistic models either
omit it entirely, or include only coarse approximations. In this paper we
outline an extensible framework and package (called TREE, for TRait Ecology
and Evolution) for modelling plant demography, driven by a mechanistic model
of plant function, across the entire life cycle. At its core, TREE is an
individual-based model, wherein plants are modelled as a system of coupled
differential equations. Individuals from multiple species can be grown in
isolation, in patches of competing plants, or in  metapopulations under a
prevailing disturbance regime. The dynamics within patches of competing plants
are resolved using either 1) a stochastic simulator, or 2) a deterministic
approximation of their density. The latter is achieved via a novel extension
of the Escalator Boxcar Train technique. The combined effects of trait-, size-
and patch-structured dynamics can then be integrated into a population level
estimate of reproductive fitness. The core of TREE is programmed in C++ with
wrappers allowing both interactive and scripted use from R. Moreover, we allow
for alternative physiologies and hyper-parametrisation on the basis of plant
functional traits. A detailed test suite is provided to ensure accuracy. We
discuss how the methods outlined can be used to address fundamental questions
on how functional traits influence the growth of individual plants, whole patches
and assembly of ecological communities.

# Introduction

Problem is that only one dominant vegetation model includes size structure, even
though competition for light widely acknowledged as most fundamental of processes.
Methods for studying size-structured metapopulations poorly developed, relying mostly
on simulation.
Means that unable to answer many questions, particular about traits, where need to
evaluate affect across entire life cycle, and taking into account effects on growth,
fecundity and mortality.

What does tree model do

- Individual plants growing
  - estimate plant growth rates in relation to size, light env, traits
- Plants competing in a patch (stochastic, deterministic)
  -  multi-species self-thinning, successional transitions
- Trait, Size and Patch structured vegetation
  - emergent behaviours: vegetation (e.g. Falster et al 2011), demography
- Fitness in metapopulation
  - community assembly, evolution, adaptation
- Evolution, community assembly, e.g. successional diversity

Series of nested uses

- has multiple possible spin-offs, we highlight some ways of extended
- Each step down is just one way of extending thing above

# The methods

# The approach

Briefly about design.

- Extension of @Falster-2011. - separate
- Major new piece of software (lines), designed with..... 1300 R code, 7000 cpp
- New method for solving PDEs
  + estimate density using modified version of EBT, integration along characteristics.
  + not sure if better than original EBT but simpler
  + at present don't have cohort mean implemented, but could easily enough [RF ~
  2 days] --> would
  need this is wanted to make any claims about one method being better
- stochastic and deterministic versions
- multiple models physiological models possible
- hyper-parametrisation with respect to traits
- Possible to drive core from C++ (with minimal modification), and thereby interfacing other languages

Differences to EBT

- density via point estimates rather than averages across cohorts
- In original EBT, total number of individuals constrained through formation of a
cohort,, but only via first-order approximation. Here estimates of n is exact,
but could be error in total number of individuals.
- adaptive spacing of cohorts

# Use cases

Examples of biological problems solved using the software / How would other people use the model (evolutionary simulations, demographic, etc)

- consider a small code example, expanded on in the vignette

## Demographics

Example from growth trajectories

## Scaling from trait to patch and vegetation

Successional dynamics

Emergent mortality, growth, size distributions

## Community assembly

- Fundamental versus realised niche
- Niche differentiation and evolved neutrality
- Evolutionary responses to environmental change

# Closing comments

Lists of unexplored research questions

- questions about complementarity

List of possible extensions

- other resources
- physiology

\newpage

# Figures

\begin{figure}[ht]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
\caption{\textbf{Schematic.} Similar to Figure 1 of Falster 2011.
\label{fig:schematic}}
\end{figure}

\newpage

\begin{figure}[ht]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
\caption{\textbf{Growth trajectories.} something biological to point at, possibly stochastic and deterministic overlaid.
\label{fig:trajectories}}
\end{figure}

\newpage

# Supplementary material

# References

### 2015-05-08 R/figures.R

@@ -1,0 +1,6 @@
+
+empty_box <- function() {
+  par(oma=c(0,0,0,0), mar=rep(0.1,4))
+  plot(1,1, ann=FALSE, axes=FALSE, type='n')
+  box()
+}

### 2015-05-08 R/ms.R

@@ -1,0 +1,6 @@
+
+pandoc_build <- function(file, template, engine="xelatex" ){
+  args <- list(sprintf('--template=%s',template), "--listings", sprintf('--latex-engine=%s', engine))
+  pandoc_convert(file, output= paste0(tools::file_path_sans_ext(file), ".pdf"),
+    citeproc = TRUE, options = args, verbose = TRUE)
+}

### 2015-05-12 tree_demography.Rmd
Import existing documentation from tree
For #7

traitecoevo/tree/blob/c3567bdbdee7af9ebf37b3698dfa126cde994a60/doc/EBT.md
traitecoevo/tree/blob/c3567bdbdee7af9ebf37b3698dfa126cde994a60/doc/growth_model.Rmd

---
title: "Modelling demography in the TREE package"
author: "Daniel Falster, Rich FitzJohn"
affiliation: Biological sciences, Macquarie University NSW 2109
output: pdf_document
csl: downloads/style.csl
bibliography: refs.bib
---

# To do

- Table of units, symbols, subscripts
- Add equations for germination limit
- clean up text at start
- formula for EBT cohorts
- formula for error in EBT

\newpage

# System dynamics

A system is physiologically or size structured whenever the basic rates of individuals (growth, mortality, reproduction) vary as functions of their size. These rates may also vary through environment, which can include feedbacks dependent on the current population state (e.g. density dependence).  Theory exists for tracing the development of a size- structured based on a partial differential equation (PDE) describing changes in the size distribution of individuals at any given time (describing the density function of individuals, rather than location of specific individuals). PSPs are formulated such that spatial distribution is neglected; only large populations are considered. Consequently the dynamics are entirely deterministic. These assumptions ensure similar individuals experience a homogeneous environment and are thus treated equally, and that inherently stochastic processes such as mortality can be reliably approximated by a deterministic description [makes use of the law of large numbers]. Individuals enter at the lower size class according to the rate of offspring production. The continuous nature of the system, with appropriate feedbacks, then ensures a dynamical solution to evolving size-structure including the problem of self-thinning. The dynamics of the population is given by the cumulative actions of its individuals.

## Patch- and size-structured population dynamics

Consider a population living in a large area of habitat where:

- Disturbances (such as fires, storms, landslides, floods, or disease outbreaks) strike patches of the habitat on a stochastic basis, killing individuals within affected patches;
- Individuals compete for resources within patches, but the spatial scale of competitive interaction means interactions among individuals in adjacent patches are negligible;
- There is high connectivity via dispersal between all patches in the habitat, allowing empty patches to be quickly re-colonised.

Such a system can be modelled as a metapopulation (sometimes called metacommunity for multiple species). Assuming patches are large, and there are many of them, the dynamics of the metapopulation can be captured deterministically using two PDEs (Kohyama 1993; deRoos 1997). Together these PDEs account for the combined influences of size-dependent variation in an individual's vital rates and disturbances on the size distribution of individuals. Below we outline equations suitable for modelling the equilibrium behaviour of such metapopulations.

The first PDE accounts for the effects of disturbances on the distribution of patch ages (defined as time since last disturbance) in the landscape (Von Foerster 1959; McKendrick 1926).

Let $p(a)$  be the frequency-density of patches age $a$  and let $\gamma(a)$ be the age-dependent probability that a patch of age $a$ is transformed into a patch of age 0 by a disturbance event. Here we consider situations where the age structure has reached an equilibrium state, which causes the PDE to reduce to an ordinary differential equation (ODE) with respect to patch age. (See appendix for [non-equilibrium case and derivation](#pdeDeriv-age)). The dynamics of $p$ are given by

\begin{equation} \frac{\partial}{\partial a} p(a)  = -\gamma(a) \, p(a) , \end{equation}
with boundary condition
\begin{equation}  p(0)  = \int_0^\infty \gamma(a) \, p(a) \; {\rm d} a. \end{equation}

This leads to a solution for the equilibrium distribution of patch-ages
\begin{equation} p(a) = p(0) \exp \left( \int_0^a \gamma(a^\prime) \; {\rm d} a^\prime \right),\end{equation}
where
\begin{equation}
\frac1{p(0)} = \left( \int_0^\infty \exp \left( \int_0^{a^{\prime\prime}} - \gamma(a^\prime) \; {\rm d} a^\prime \right) d a^{\prime\prime} \right),\end{equation}
 corresponds to the average lifespan of a patch.

The second PDE accounts for the development of competitive hierarchies within patches recovering from a disturbance (Sinko \& Streifer 1967; Hara 1984; Metz \& Diekmann 1986; de Roos 1997). Here we only consider cases where dispersal is restricted to newborns and where disturbances remove all established individuals within a patch. Let $n(x,m,a)$ represent the density of individuals with traits $x$ and mass $m$ in a patch of age $a$. For sufficiently large populations, the dynamics of $n$ are then given by the PDE

\begin{equation} \label{eq:PDE} \frac{\partial }{\partial a} n(x,m,a)= -d(x,m, a) \; n(x,m,a)-\frac{\partial }{\partial m} \left[g(x,m,a) \; n(x,m,a)\right].  \end{equation}
(See appendix for [derivation of PDE approximating size structured dynamics](#pdeDeriv-size).)

The functions $g(x,m,a)$, $d(x,m,a)$ and $f(x,m,a)$ give the growth, death rate, and fecundity rates of individuals with traits $x$ and mass $m$ in a patch of age $a$. In principle, the arguments of these functions should be written as $g(x,m,E(x,\cdot,a))$, $d(x,m,E(x,\cdot,a))$ and $f(x,m,E(x,\cdot,a))$, where $E(x,\cdot,a)$ is the distribution of resources with respect to size $m$ in a patch age $a$ occupied by residents with traits $x$, and $E(x,m,a)$ gives the resources available to an individual of size $m$. To simplify notation throughout the remainder of the document we will drop the $E(x,\cdot,a)$ from $g, d,$ and $f$, and instead simply refer to patch age $a$, on the understanding by patch age $a$, we mean the structure of a patch age $a$ with given residents.

Eq. $\ref{eq:PDE}$ has two boundary conditions. The first,

\begin{equation}  \label{eq:BC1} \frac{n(x,m_0,a)g(x,m_0 , a)}{\pi_1(x,m_0 ,a)} = \int_0^{\infty} p(a)  \int_0^{\infty}  \pi_0f(x,m,a) \; n(x,m,a)\,{\rm d} m \, {\rm d} a, \end{equation}

links the flux of newly arriving offspring across the lower bound $(m_0)$ of the size distribution within each patch (LHS), to the rate at which newly arriving offspring are produced in the metapopulation (RHS). The latter quantity, denoted $y_x$, serves as a useful measure for the abundance of the $x$ strategy. $\pi_0$  and $\pi_1$ give probabilities of offspring surviving dispersal and successfully establishing in a new patch, respectively. In this example  $\pi_1$ is dependent on resource availability while $\pi_0$ is a constant.

The second boundary condition of Eq. $\ref{eq:PDE}$ gives the size distribution for patches immediately after a disturbance, $n\left(x,m,0\right)$. Since we are assuming disturbances remove all individuals within a patch,

\begin{equation} \label{eq:BC2} n\left(x,m,0\right) =0. \end{equation}

A convenient feature of Eqs. $\ref{eq:PDE}$ - $\ref{eq:BC2}$ is that the dynamics of a single patch scale up to give the dynamics of the entire metapopulation. Note that the rate offspring arrive from the disperser pool, $y_x$, is constant for a metapopulation at equilibrium. Combined with the assumption that all patches have the same initial (empty) size distribution, this feature ensures that all patches show the same temporal behaviour, the only difference between them being the ages at which they are disturbed.

To model the temporal dynamics of an archetypal patch, we need only a value for $y_x$. The numerical challenge is therefore to find a value of $y_x$ that balances production and arrival of dispersers:

\begin{equation} \label{eq:solution} \int_0^\infty p(a) \int_0^\infty \pi_0 f(x,m,a) \; n(x,m,a) \, {\rm d} m \, {\rm d} a = y_x\end{equation}

Solving for $y_x$ is a straightforward one-dimensional root finding problem. Calculation of the LHS of Eq. $\ref{eq:solution}$, however, requires detailed information about $n(x,m,a)$, which in turn, requires a separate numerical method capable of solving the PDE in Eq. $\ref{eq:PDE}$.

### Environmental feedback

Individuals are assumed to interact  via their influence on the availability of resources. A competition kernel is required, giving the influence of an individual on the availability of resources for another individual. To get the total influence of all individuals on the target individual requires integration of the competition kernel over the density distribution $n$. If there are $N$ different species or types in the system, then their combined effects must be summed.

The approach outlined is very general and can be tailored to suit a range of different systems. To do so we need to describe:
\begin{itemize}
 \item the influence of an individual's size (state), its traits (type), and the availability of resources (environment) on the vital rates of growth, mortality and reproduction
 \item the dependence of resource availability on size-distribution of individuals within each species and their traits.
\end{itemize}

For plants, where we are interested in the effect of shading from other individuals on growth, mortality and fecundity, we can define a function for the amount of light at height $z$ in a patch of age $a$:

\begin{equation} \label{eq:light}
E(z,a) = \exp \left(-c_{ext}  \sum_{i=1}^{N} \int_{0}^{\infty} \phi(m) \, L(z, h(m)) \, n(x_i, m,a) \, dm \right), \end{equation}
where $\phi(m)$ is total leaf area and $L(z, h(m))$ is fraction of this leaf area held above height $z$ for plants size $m$, and $c_{ext}$ is the light extinction coefficient.

## Invasion fitness

Let us now consider how we can estimate the fitness of a rare individual with trats $x^\prime$ growing in the environment of a resident community with traits $x$. We will focus on the phenotypic components of fitness -- i.e. the consequences of a given set of traits for growth, fecundity and mortality -- only, taking into account the non-linear effects of competition on individual success, but ignoring the underlying genetic basis for the trait determination. We also adhere to the standard conventions in such analyses in assuming that the mutant is sufficiently rare to have a negligible effect on the environment where it is growing.

Invasion fitness is most correctly defined as the long-term per capita growth rate of a rare mutant population growing in the environment determined by the resident strategy (Metz et al. 1992). Calculating per-capita growth rates, however, can be particularly challenging in a structured metapopulation model (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001). As an alternative measure of fitness, we can use the basic reproduction ratio, which gives the expected number of new dispersers arising from a single dispersal event. Evolutionary inferences made using the basic reproduction ratio will be similar to those made using per-capita growth rates for metapopulations at demographic equilibrium (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001).

Let $R\left(x^\prime,x\right)$ be the basic reproduction ratio of individuals with traits $x^\prime$ growing in the competitive environment of the resident traits $x$. Recalling that patches of age $a$ have density $p(a)$ in the landscape, it follows that any seed of $x^\prime$ has probability $p(a)$ of landing in a patch age $a$. The basic reproduction ratio for individuals with traits $x^\prime$ is then:
\begin{equation} \label{eq:InvFit}
R\left(x^\prime,x\right)=\int _0^{\infty }p\left(a\right)\tilde{R}\left(x^\prime,a, \infty \right)\; {\rm d}a ,
\end{equation}
where $\tilde{R}\left(x^\prime,a_0,a \right)$ is the expected number of dispersing offspring produced by a single disperser arriving in a patch of age $a_0$ up until age $a$ (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001). $\tilde{R}\left(x^\prime,a,\infty\right)$ is calculated by integrating an individual's fecundity over the expected lifetime the patch, taking into account competitive shading from residents with traits $x$, the individual's probability of surviving, and its traits via the equation:

\begin{equation} \label{eq:tildeR} \tilde{R}(x^\prime, a_0, a) =\int_{a_0}^{a}  \pi_0f(x^\prime, m(x^\prime, a_0, a^\prime), a^\prime) \, S_{\rm I} (x^\prime, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime.\end{equation}

Here,
\begin{equation} \label{eq:size} m(x^\prime, a_0, a) = m_0 + \int_{a_0}^{a} g(x^\prime, m(x^\prime, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime \end{equation}

and

\begin{equation} \label{eq:survivalIndividual} S_{\rm I} (x^\prime, a_0, a) = \pi_1 (x^\prime,m_0, a_0) \, \exp\left(- \int_{a_0}^{a} d(x,m_i(x^\prime, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right) \end{equation}

are the size $m$ and survival $S_{\rm I}$ of dispersers that arrived in a patch of age $a_0$ at age $a$, while

\begin{equation} \label{eq:survivalPatch} S_{\rm P} (a_0,a) = \exp\left(-\int_{a_0}^{a} \gamma(a^\prime) \, {\rm d} a^\prime \right) \end{equation}
gives the probability that the patch remains undisturbed from $a_0$ to $a$.

The notational complexity required for a general, non-linear solution might obscure an otherwise simple point: Eqs. $\ref{eq:size}$ - \ref{eq:survivalPatch} are simply integrals of the growth and mortality functions, $\frac{\rm d}{{\rm d}a} m =g\left(x^\prime,m,a\right)$ and $\frac{\rm d}{{\rm d}a} S_{\rm I} =d\left(x^\prime,m,a\right) \; {\rm \; }S_{\rm I}$, with initial conditions $m\left(x^\prime,0,0\right)=m_0$ and $S_{\rm I} \left(x^\prime,0,0\right)= \pi_1 (x^\prime,m_0, a_0)$, the survival through germination.

## Emergent properties of resident community

Summary statistics of the metapopulation are obtained by integrating over the density distribution, weighting by patch abundance $p(a)$. Firstly, the total number of individuals in the metapopulation is given by
\begin{equation}\hat{n}(x) = \int_{0}^{\infty} \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a \, {\rm d}m,\end{equation}
and the average density of plants size $m$ by
\begin{equation}\bar{n}(x,m) = \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a.\end{equation}

Average values for other quantities can also be calculated. Let $\phi(x, m, a)$ be a demographic quantity of interest, such growth rate, mortality rate, light environment, or size. The average value of $\phi$ across the metapopulation is
	\begin{equation}\hat{\phi}(x) = \frac1{\hat{n}(x) }\int_{0}^{\infty} \int_{0}^{\infty} p(a) \; n(x,m,a) \phi(x,m,a) \, {\rm d}a \, {\rm d}m,\end{equation}
while the average  value of $\phi$ for plants size $m$ is
 \begin{equation}\bar{\phi}(x,m) = \frac1{\bar{n}(x,m) }\int_{0}^{\infty}p(a)  n(x,m,a) \phi(x,m,a)\, {\rm d}a.\end{equation}

 When calculating average mortality rate, a choice must be made as to whether mortality due patch disturbance is included. Non-disturbance mortality is obtained by setting $\phi(x,m, a) = d(x,m,a)$, while total mortality due to growth processes and disturbance is obtained by setting $\phi(x,m, a) = d(x,m,a)+ \gamma(a)\Phi(x).$)

(TODO: check this next section
One might be interested in the frequency distribution of values as well as means. What we really want is the quartile function giving $Pr(\phi(m,a,t) < \phi_0)$.....rest commented out for now...

<!-- The quartile function corresponds to the inverse of the cumulative distribution function. Even though we have a distn of $\phi$ wrt to $a$, I don't know whether this can be converted to a cdf. Therefore, use a resampling approach to obtain quartile.
\begin{enumerate}
	\item Interpolate to get a distn of $p(a)  n(m,a,t)$ and $\phi(m,a,t)$ with respect to a (this step also needed for mean above)
	\item Using Von Nueman Rejection Algorithm, randomly sample individuals from the probability distribution given by $f = p(a) \; n(m,a,t)$ [Generate 2D uniform numbers in range [$(0, max(f)), 0, max(a))$]. Using a spline fitted to true values of $a \& f$, estimate $f$ for the randomly generated $a$ values -> accept all samples where $f_{rand} < f_{spline}$]
	\item Use a spline to estimate $\phi$ for individuals accepted by rejection algorithm
	\item calculate quartiles empirically based on observed distns
\end{enumerate} -->

## Converting density from one size unit to another

Population density is explicitly modelled in relation to a given size unit (Eq. $\ref{eq:PDE}$). But what if we want to express density in relation to another size unit? A relation between the two can be derived by noting that the total number of individuals within a given size range must be equal. So let's say density is expressed in units of size $m$, but we want density in units of size $h$. First we require a one-to-one function which h for a given $m$: $h = \hat{h}(m)$. Then the following must hold
\begin{equation} \label{eq:n_conversion} \int_{m_1}^{m_2} n(x,m,a) \; \textrm{d}m =  \int_{\hat{h}(m_1)}^{\hat{h}(m_2)} n^\prime(x,m,a) \; \textrm{d}h  \end{equation}

For very small size intervals, this equation is equivalent to
\begin{equation} \left(m_2- m_1 \right) \; n(x,m_1,a) = \left( \hat{h}(m_2) - \hat{h}(m_1)\right) \; n^\prime(x, \hat{h}(m_1),a). \end{equation}

Rearranging gives
\begin{equation}  n^\prime(x, \hat{h}(m_1),a) = n(x, m_1,a) \; \frac{m_2- m_1}{\hat{h}(m_2) - \hat{h}(m_1)}  \end{equation}

Noting that the second term on the RHS is simply the definition of $\frac{\delta m}{\delta h}$ evaluated at $m_1$, we have
\begin{equation} \label{eq:n_conversion2} n^\prime(x, h, a) = n(x, m,a) \; \frac{\delta m}{\delta h}. \end{equation}

\newpage

# Approximating system dynamics using the escalator boxcar train (EBT)

The EBT solves the PDE describing development of $n(x,m,a)$ (Eq. $\ref{eq:PDE}$) by approximating the density function with a collection of cohorts spanning the size spectrum (de Roos 1988, 1997). Following a disturbance, a series of cohorts are introduced into each patch. These cohorts are then transported along the characteristics of Eq. $\ref{eq:PDE}$ according to the individual-level growth function. (Characteristics are curves along which Eq. $\ref{eq:PDE}$  becomes an ordinary differential equation (ODE); biologically, these are the growth trajectories of individuals, provided they do not die). The numerical technique consists of approximating changes to the number of individuals within each cohort by a system of closed-form linear ODEs. Eq. $\ref{eq:PDE}$ is thereby reduced to a family of ODEs, which can be stepped using an appropriate ODE solver with an adaptive step size (Press 1995).

Below we consider two alternative implementations of the EBT. Both require that we introduce a series of cohorts into a patch, and that these be transported along the characterstics of the Eq. $\ref{eq:PDE}$ (see figure).

<!-- ![Cohorts in the EBT](figs/EBT.png) -->

Let $\Omega = \left[m_0,m_+ \right)$ represent the entire state-space attainable by any individual and let the interior of $\Omega$ be subdivided into $k$ sub-domains  $$\Omega_i(a) = \left[m_{i=1}(a),m_i(a) \right) $$ , with $i=1,\dots, k$ and $m_0 < m_1<	\ldots<m_k = m_+$. If we allow these sub-domains to be transported through time along the characteristics of Eq. $\ref{eq:PDE}$ , they will be closed to transport processes across their upper and lower boundaries.

## First method for approximating $n(x,m,a)$, based on modelling of cohort boundaries

In the EBT, the cohort (or sub-domain) boundaries are defined by a series of time points where one cohort is ended, and another is initialised. As the patch ages, the width of all cohorts varies, as they are transported up along the size spectrum (see figure). It is helpful to think of these boundaries as representing individuals that germinated exactly at the age defining the cohort boundary. The growth trajectory of that individual then defines the trajectory of the cohort boundary.

In the derivation of the original EBT, equations were given for tracking the size of cohort boundaries, but these were not essential for the numerical approximation of systems dynamics. However, it is possible to approximate the entire system by only following the cohort boundaries.

### Size

From equation  $\ref{eq:size}$, the boundary develops as

\begin{equation} \label{eq:boundSize} m_i(x, a_0, a) = m_0 + \int_{a_0}^{a} g(x,m_i(x, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime.\end{equation}

*Numerical technique:* Equation $\ref{eq:boundSize}$ can be expressed as an initial-value ODE problem (IVP) and solved using an ODE stepper:
$$\frac{dy}{dt} = g(x,y,t),$$
$$ y(0) = m_0.$$

### Survival
From equation  $\ref{eq:survivalIndividual}$, the survival of the individuals at the boundary individuals is

\begin{equation} \label{eq:boundSurv} S_{\rm I} (x, a_0, a) =  \pi_1 (x^\prime,m_0, a_0)  \exp\left(- \int_{a_0}^{a} d(x,m_i(x, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right).\end{equation}

*Numerical technique:* To solve equations $\ref{eq:boundSurv}$ we solve the IVP problem:
$$\frac{dy}{dt} = d(x,m_i(a^\prime), a^\prime ),$$
$$ y(0) = - \ln\left(\pi_1 (x^\prime,m_0, a_0)\right) .$$

Survival is then
$$ S_{\rm I} (x, a_0, a) = \exp\left(- y(a) \right).$$

### Density of individuals

The density of individuals at the boundary can be solved by integrating along characteristics ([see Appendix for details](#IntegrationAlongCharacteristics)) to give:

\begin{equation} \label{eq:boundN} \begin{array}{ll}  n(x, m, a) & =n(x,m_0 ,a_0) \times \\
& \exp \left(-\int _{a_0}^{a} \left[\frac{\partial g(x,m(x, a_0, a^\prime),a^\prime)}{\partial m} +d(x,m(x, a_0, a^\prime),a^\prime)\right] {\rm d}a^\prime \right),\end{array} \end{equation}

where
\begin{equation} \label{eq:boundN2}  n(x,m_0 ,a_0)  = \left\{ \begin{array}{ll}   \frac{y_x  \pi_1 (x^\prime,m_0, a_0) }{ g(x,m_0, a_0) }  & \textrm{if } g(x,m_0, a_0) > 0 \\
0 & \textrm{otherwise.}  \end{array} \right. \end{equation}

*Numerical technique:* To solve Eq. $\ref{eq:boundN}$ we need to solve the IVP:
$$\frac{dy}{dt} = \frac{\partial g(x,m_i(a^\prime), }{\partial m} +d(x,m_i(a^\prime),a^\prime),$$
$$ y(0) = -\ln\left(n(x,m_0 ,a_0) \, \pi_1 (x^\prime,m_0, a_0) \right).$$

Density is then given by

$$n(x,m_0 ,a_0) =\exp(-y(a)).$$

*Test case:* In cases where the environment and seed rain are constant and time-independent (i.e $g(x,m,a) = g(x,m)$, $d(x,m,a) = d(x,m)$  for all $a_0, a^\prime$),  it is possible to further simplify Eq. \ref{eq:boundN} to:

\begin{equation}\label{eq:boundN3}
n(x,m,a)=\frac{y_x}{ g(x,m) } \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x , m^\prime) }{g(x, m^\prime)} \, \textrm{d}m^\prime \bigg). \end{equation}
(for a proof, see \cite{de_roos_gentle_1997} page 153). Note also, that the integral on the RHS of $\ref{eq:boundN3}$ is simply the survival of individuals from $a_0 \rightarrow a$ describd in equation $\ref{eq:boundSurv}$
\begin{equation} \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x, m^\prime)}{g(x, m^\prime)}  \textrm{d} m^\prime \bigg) = \exp \bigg(-\int _{a_{0} }^{a}d(x, m(a^\prime)) \textrm{d}{a^\prime} \bigg) = S_{\rm I} (x, a_0, a).\end{equation}

Thus,
\begin{equation}\label{eq:boundN4}
n(x,m,a)=\frac{y_x}{ g(x,m) } S_{\rm I} (x, a_0, a). \end{equation}
For some functions of $g$ and $d$, equation $\ref{eq:boundN4}$ might yield an analytical solution to density.

### Seed production

The lifetime seed production of boundary individuals  is calculated according to Eq. $\ref{eq:tildeR}$ as
\begin{equation}\begin{array}{ll}\tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\\end{array}\end{equation}
where $S_{\rm I}$ is individual survival (defined above) and
$S_{\rm P}$ is calculated as in Eq. $\ref{eq:survivalPatch}$.

*Numerical technique:* To solve equation \ref{eq:tildeR} we need solve the IVP:
$$\frac{dy}{dt} = \pi_0 \,f(x, m_i(a^\prime), a^\prime) \, S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime),$$
$$ y(0) = 0.$$

### Invasion fitness

To scale up seed production for the metapopulation need to integrate Eq. $\ref{eq:boundR}$ across all possible seed-arrival times (as defined in Eq. \ref{eq:InvFit}):
\begin{equation} R(x) = \int_0^{\infty}  p(a) \, \tilde{R}(x, a, \infty) \, {\rm d}a. \end{equation}

*Numerical technique:*  Note that we have individual's introduced at a series of patch ages, and an estimate for $\tilde{R}(x, a, \infty)$ for each of these (see Eq. $\ref{eq:boundR}$). These points can then be integrated using a quadrature routine, e.g.

### Environmental feedback

The estimated density function  (Eq. \ref{eq:boundN4}) is used to calculate the amount of shading on each individual from other individuals in the patch, as described in Eq. \ref{eq:light}.

*Numerical technique:*  $E(z,a)$ is estimated by integrating  eq. \ref{eq:light} using the numerical estimate for $n(x,m,a)$ obtained from Eq. \ref{eq:boundN4} via trapezoidal rule with uneven grid (Eq. \ref{trap_uneven}), taking the cohort boundaries as the knots in the integration.

## Second method for approximating $n(x,m,a)$: based on modelling dynamics within cohorts

The original EBT, derived by de Roos (1997, 1988), proceeded by approximating the first and second moments of the density distribution $n \left( x,m,a \right)$ within sub-domains $\Omega_i$, represented by $\lambda_i$ and $\mu_i$, these being the total number and mean size of individuals within $\Omega_i$ respectively.

Under this assumption, the rates of change $\frac{\rm d} {{\rm d}a} \lambda_i$ and $\frac{\rm d} {{\rm d}a} \mu_i$ can be approximated given by two ODEs, which can be approximated by first-order closed-form solutions. These approximations form the centrepiece of the original EBT, and are derived as follows.

### First moment: Number of individuals

*Definition:*
\begin{equation} \label{eq:lambda}  \lambda_i = \int_{\Omega_i} n(x,m,a) \; {\rm d}m \end{equation}

\begin{equation} \label{eq:dlambda} \frac{{\rm d} }{{\rm d}a} \lambda_i = -\int_{\Omega_i} d(x,m,E) \; n(x,m,a) \; {\rm d}m \end{equation}

*Numerical technique:* The rate of change in $\lambda_i$ is approximated by

\begin{equation} \label{eq:dlambda_approx} \frac{{\rm d} }{{\rm d}a} \lambda_i =  d(x,\mu_i,E) \lambda_i \end{equation}, with $\lambda_i$ estimated by integrating this quantity with an ODE stepper. The approximation for \ref{eq:dlambda_approx} is only accurate to second order (for details on derivation see [appendix](#DeriveEBTapproximation)).

### Second moment: Average size

*Definition:*
\begin{equation}  \label{eq:mu} \mu_i = \frac1{\lambda_i}\int_{\Omega_i} m \; n(x,m,a) \; {\rm d}m  \end{equation}

\begin{equation} \label{eq:dmu} \frac{{\rm d} }{{\rm d}a} \mu_i = \frac1{\lambda_i} \int_{\Omega_i} \left[ g(x,m,E) - ( m-\mu_i ) d(x,m,a) \right] n(x,m,a) \; {\rm d}m  \end{equation}

*Numerical technique:* The rate of change in $\mu_i$ is approximated by
\begin{equation}  \label{eq:dmu_approx} \frac{{\rm d} }{{\rm d}a} \mu_i =  g(x,\mu_i,E), \end{equation}

with $\mu_i$ estimated by integrating this quantity with an ODE stepper. The approximation for \ref{eq:dmu_approx} is only accurate to second order (for details on derivation see [appendix](#DeriveEBTapproximation)).

### Density of individuals

*Definition:* $n(x,m,a)$ for $m \in \Omega_i$

*Numerical technique:* The EBT technique solves the PDE by approximating the changing size distribution $n(x,m,a)$ with a series of point masses with position and amplitude given by $\lambda_i$ and $\mu_i$. If we let $\delta(z)$ be a function for a point mass of size 1 localised at $z=0$, then

\begin{equation} \label{eq:n_delta}  n(x,m,a) \approx = \sum_{i=1}^{k} \lambda_i(a) \; \delta(m-\mu_i(a)). \end{equation}

The density function $n(x,m,a)$ within each sub-domain can also be estimated from its first moment using a discontinuous, piecewise-linear function:

\begin{equation} n_i \left( x,m,a \right) \approx \frac{\lambda_i}{m_i-m_{i-1}}, \end{equation}

although this requires that the position of the cohort boundaries are also known (see previous section for equations).

### Seed production
The lifetime seed production of individuals located at the cohort boundary is calculated according to Eq. $\ref{eq:tildeR}$ as
\begin{equation}\label{eq:boundR}\begin{array}{ll}\tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\\end{array}\end{equation}
where $S_{\rm I}$ is individual survival (defined above) and
$S_{\rm P}$ is calculated as in Eq. $\ref{eq:survivalPatch}$. Using the EBT approximation, the total number of sees produced within a cohort including individuals born in the interval $(a_1, a_2)$ and up to age $a$ is given by

\begin{equation} \int_{a_1}^{a_2} \tilde{R}(x, a_0, a) \; {\rm d}a_0 = f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) S_{\rm P} (a_0,a^\prime) \; {\rm d}a^\prime.  \end{equation}

Thus the total number of seeds produced by individuals within the cohort is given by

\begin{equation} \label{eq:cohR} \int_{a_1}^{\infty}  \int_{a_1}^{a_2} \tilde{R}(x, a_0, a) \; {\rm d}a_0 {\rm d}a = \int_{a_1}^{\infty} f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) \, S_{\rm P} (a_1,a^\prime).\end{equation}

*Numerical technique:* To solve equation $\ref{eq:cohR}$ we need solve the IVP:
$$\frac{dy}{dt} = f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) S_{\rm P} (a_1,a^\prime),$$
$$ y(a_1) = 0.$$

### Dynamics of the boundary cohort

A slightly modified set of equations is needed to track the moments of the smallest cohort, whose lower bound is fixed at $m_0$. Specifically, a new quantity $\Pi$, representing the total accumulated mass within this boundary cohort, is monitored in place of the second moment, which would otherwise be undefined whenever $\lambda _i=0$. Our approximations for the boundary cohort differ slightly to those of de Roos (1997, 1988) in that we Taylor expand around the centre of mass for the cohort instead of the lower boundary. Because relative growth rates often decline quickly with increasing size, use of lower boundary can overestimate the growth rate of the cohort.

#### Number of individuals

*Definition:*
\begin{equation} \lambda_1 = \int_{\Omega_1} n(x,m,a) \; {\rm d}m \end{equation}

\begin{equation} \frac{{\rm d} }{{\rm d}a} \lambda_1 = y_x \pi_0 \pi_1(x,m_0, a) - \int_{\Omega_1} d(x,m,a) \; n(x,m,a) \; {\rm d}m  \end{equation}
(TODO: I $\pi_0$ in right place here? Should appear in seed production, not inflow).

*Numerical technique:* $\frac{{\rm d} }{{\rm d}a} \lambda_1 = y_x - d(x,\mu_1,a) \lambda_1$

#### New mass

*Definition:* $\Pi_1 = \int_{\Omega_1} (m-m_0) \; n(x,m,a) \; {\rm d}m$

*Exact:* $\frac{{\rm d} }{{\rm d}a} \Pi_1 = \int_{\Omega_1} \left[ g(x,m,E) - ( m-\mu0) d(x,m,a) \right] n(x,m,a) \; {\rm d}m$

*Numerical technique:* $\frac{{\rm d} }{{\rm d}a} \Pi_1 =
 g(x,\mu_1,a) \lambda_1 - \Pi_1  d(x,\mu_1,a)$

#### Average size

*Definition:*
\begin{equation}  \mu_1 = \left\{ \begin{array}{ll} \frac1{\lambda_i}\int_{\Omega_i} m \; n(x,m,a) \; {\rm d}m   & \textrm{if } \lambda_1 > 0 \\
m_0 & \textrm{otherwise}  \end{array} \right. \end{equation}

\begin{equation} \mu_1 = \left\{ \begin{array}{ll}  m_0 + \frac{\Pi_1}{\lambda_1}  & \textrm{if } \lambda_1 > 0 \\
m_0 & \textrm{otherwise}  \end{array} \right. \end{equation}

*Numerical technique:* NA, use exact.

### Environmental feedback

The estimated density function (Eq. \ref{eq:n_delta}) is used to calculate the amount of shading on each individual from other individuals in the patch, as described in Eq. \ref{eq:light}.

*Numerical technique:*  The approximation of  $n(x,m,a)$ by a series of delta peaks  in Eq. \ref {eq:n_delta} allows integration in Eq. \ref{eq:light} to be approximated as a sum over cohorts:

\begin{equation} \label{eq:light_deltasum}
E(z,a) = \exp \left(-c_{ext}  \sum_{n=1}^{N} \sum_{\Omega_i} \phi(\mu_{n,i}) \, L(z, h(\mu_{n,i})) \lambda_{n,i} \right), \end{equation}

where $\mu_{n,i}, \lambda_{n,i}$ are the average size and number of individuals within the i$^{th}$ cohort of species $n$ (Eqs. \ref{eq:mu}, \label{eq:lambda} respectively).

## Controlling error in the EBT

Errors in the EBT approximation to $n$ can arise from two sources: (i) poor spacing of cohorts in the size dimension, and (ii) when stepping cohorts through time. Error of the latter type is effectively controlled using a suitable ODE solver and is not considered further here. However, there is no existing method to control error arising from poor spacing of cohorts. Poor cohort spacing introduces error both because the equations governing cohort dynamics are only accurate up to second order (Table 4-1), and because any integrals over the size distribution (e.g. total offspring production, competitive effect) may be poorly resolved.

TODO: describe approaches used.

\newpage

# Appendices

## Derivation of PDE describing age-structured dynamics {#pdeDeriv-age}

Consider patches of habitat which are subject to some intermittent disturbance and where the age of a patch is corresponds to the time since the last disturbance. Let $p(a,t)$ be the frequency-density of patches age $a$ at time $t$ and let $\gamma(a)$ be the age-dependent probability that a patch of age $a$ is transformed into a patch of age $0$ through disturbance. Then according to the Von Foerster equation for age-structured population dynamics (von_foerster_remarks_1959), the dynamics of $p(a,t)$ are given by
$$ \frac{\partial }{\partial t} p(a,t)=-\frac{\partial }{\partial a} p(a,t)-\gamma(a,t)p(a,t),$$
with boundary condition
$$ p(0,t)=\int^{\infty}_{0}\gamma(a,t)p(a,t)\,{\rm d}a.$$

The frequency of patches with $a < x$ is given by $\int_{0}^{x}p(a,t) \, {\rm d}a$, with $\int_{0}^{\infty} p(a,t) \, {\rm d}a =1$. If $\frac{\partial}{\partial t}\gamma(a,t)=0$, then $p(a)$ will approach an equilibrium solution given by
$$p(a) = p(0) \Pi(a),$$
where
$$\Pi(a) = \exp \left( \int_{0}^{a}-\gamma(\tau)\,{\rm d}\tau\right)$$
is the probability that a local population will remain undisturbed for at least $a$ years (patch survival function),
and
$$p(0) = \frac1{ \int_{0}^{\infty}\Pi(a) \,{\rm d}a}$$
is the frequency-density of patches age 0. The rate of disturbance for patches age $a$ is given by $\frac{\partial (1-\Pi(a))}{\partial a} = \frac{-\partial \Pi(a)}{\partial a}$,
 while the expected lifetime for patches is $\int_0^\infty - a \frac{\partial}{\partial a} \Pi(a) \, {\rm d} a = \int_0^\infty \Pi(a) \, {\rm d} a = \frac1{p(0)}$ (2$^{nd}$ step made using integration by parts).

An equilibrium patch age distribution may be achieved under a variety of conditions, for example if $\gamma(a,t)$ depends on patch age but the this probability is constant over time. The probability of disturbance may also depend on features of the vegetation (rather than age *per se*), in which case  an equilibrium is still possible, provided the vegetation is also assumed to be at equilibrium.

### Exponential distribution
If the probability of patch disturbance is constant with respect to patch age ($=\lambda$), then rates at which patches age $a$ are disturbed follow an exponential distribution: $-\partial \Pi(a)/ \partial a = \lambda e^{-\lambda a}$. The patch age distribution is then given by:
$$ \Pi(a) = \exp\left(-\lambda a\right), p(0) = \lambda.$$

### Weibull distribution
If the probability of patch disturbance changes as a function of time, with $\gamma(a) = \lambda \psi  a^{\psi-1}$, then rates at which patches age $a$ are disturbed follow a Weibull distribution: $-\partial \Pi(a)/ \partial a = \lambda \psi a^{\psi -1}e^{-\lambda a^\psi}$. Values of $\psi>1$ imply probability of disturbance increases with patch age; $\psi<1$ implies probability of disturbance decreases with age. When $\psi=1$ we obtain the exponential distribution, a special case of the Weibull. The Weibull distribution results in following for the patch age distribution:
$$\Pi(a) = e^{-\lambda a^\psi}, p(0) =  \frac{\psi \lambda^{\frac1{\psi}}}{\Gamma\left(\frac1{\psi}\right)},$$
where $\Gamma(x)$ is the gamma function $\left(\Gamma(x) = \int_{0}^{\infty}e^{-t}t^{x-1}\, dt\right)$. We can also specify the distribution by its mean return time $\bar{a}  = \frac1{p(0)}$.  Then, calculate relevant value for $\lambda =  \left(\frac{\Gamma\left(\frac1{\psi}\right)}{\psi \bar{a}}\right)^{\psi}$.

### Variable distributions
The probability of patch disturbance might also be vary with properties of the vegetation. In this case, we cannot prescribe a known distribution to $p(a)$, it must be solved numerically. Patch survival can be calculated numerically as
$$\Pi(a) = \exp \left( \int_{0}^{a}-\gamma(\tau)\,{\rm d}\tau\right).$$
For calculations of fitness, we want to integrate over some fraction $q$ of the patch age distribution (i.e. ignoring the long tail of the distributions). Thus we want to find the point  $x$ which satisfies $\int_{0}^{x} p(a)\,{\rm d}a =q$.Locating $x$ requires knowledge of $p(0)$, which in turn requires us to approximate the tail of the integral  $\int_{0}^{\infty} \Pi(a)\,{\rm d}a$. For $a > x$, let $\Pi(a)$ be approximated by
$$\Pi(a) \approx \Pi(x) \exp\left( - (x-a) \gamma(x)\right).$$ Then
$$p(0)^{-1} =\int_{0}^{\infty} \Pi(a)\,{\rm d}a \approx \int_{0}^{x} \Pi(a)\,{\rm d}a + \int_{x}^{\infty } \Pi(a)\,{\rm d}a$$
$$ = \int_{0}^{x} \Pi(a)\,{\rm d}a + \frac{\Pi(x)}{\gamma(x)}. $$
Substituting into $\int_{0}^{x} p(a)\,{\rm d}a =q$, we obtain
$$q=  p(0) \int_{0}^{x} \Pi(a) \, {\rm d}a = \frac{\int_{0}^{x} \Pi(a) \, {\rm d}a}{\int_{0}^{x} \Pi(a)\,{\rm d}a + \frac{\Pi(x)}{\gamma(x)}}$$
$$\Rightarrow  \frac{\Pi(x)}{\gamma(x)} = \frac{1-q}{q} \int_{0}^{x} \Pi(a) \, {\rm d}a.$$
Thus by monitoring $\Pi(x), \gamma(x),  \int_{0}^{x} \Pi(a) \, {\rm d}a$ we can evaluate when a sufficient range of the patch age distribution has been incorporated.

\newpage

## Derivation of PDE describing size-structured dynamics {#pdeDeriv-size}

For general notes on structured population and analysis of data see (Metz and Diekmann 1986, Taljapurkar and Caswell 1997, Diekmann et al. 1998, Diekmann et al. 2001)

### 1-dimensional case

To model the population we use a PDE describing the dynamics for a thin slice $\Delta m$. Assuming that all rates are constant within the interval $\Delta m$, the total number of individuals within the interval spanned by $[m-0.5\Delta m,m+0.5\Delta m)$ is $n(m,a)\Delta m$ . The flux of individuals in and out of the size class can be expressed as
\begin{equation}\begin{array}{ll} J(m,a)=&g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)-g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a) \\ &-d (m,a) \; n(m,a)\Delta m\\ \end{array}.\end{equation}
The first two terms describe the flux in and out of the size class through growth; the last term describes lossed due through mortality. The change in number of individuals within the interval across a time step \textit{$\Delta $t} is thus:
\begin{equation}\begin{array}{ll} n(m,a+\Delta a)\Delta m-n(m,a)\Delta m= &g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)\Delta a \\ &-g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a)\Delta a\\&-d (m,a) \; n(m,a)\Delta m\Delta a. \end{array}\end{equation}
Rearranging,
\begin{equation}\begin{array}{ll} \frac{n(m,a+\Delta a)-n(m,a)}{\Delta a} = &-d (m,a) \; n(m,a) \\
&-\frac{g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a)-g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)}{\Delta m}.\end{array}\end{equation}
The LHS is corresponds to the derivative of $n$ as $\Delta a\to 0$. For thin slices, $\Delta m\sim 0$, we obtain
\begin{equation} \label{eq:PDE-app}
\frac{\partial }{\partial t} n(m,a)=-d (m,a) \; n(m,a)-\frac{\partial }{\partial m} (g(m,a) \; n(m,a)).
\end{equation}

To complete the model, the PDE must be supplemented with boundary conditions that specify the density at the lower end of $n(m_{0},a)$ for all $t$ as well as the the initial distribution when $t=0$, $n(m,0)$.  The former is derived by integrating the PDE with respect to $m$ over the interval $(m_{0},m_{\infty })$, yielding

\begin{equation}\frac{\partial }{\partial t} \int _{m_{0} }^{m_{\infty } }n(m,a) \partial m=g(m_{0} ,a) \; n(m_{0} ,a)-g(m_{\infty } ,a) \; n(m_{\infty } ,a)-\int _{m_{0} }^{m_{\infty } }d (m,a) \; n(m,a) \partial m.\end{equation}

The LHS of this relationship is evidently the rate of change of total numbers of individual in the population, while the right-hand-term is the total population death rate. Further, $n(m_{\infty } ,a)=0$. Thus to balance total births and deaths, $g(m_{0} ,a) \; n(m_{0} ,a)$ must equal the birth rate $B(x, n, E,a)$. Thus the boundary condition is given by
\begin{equation}g(m_{0} ,a) \; n(m_{0} ,a)=B(x, n, E,a).\end{equation}

### Extension to a population structured by 2 size-dimensions

If there is more than one dimension in the i-state, then a corresponding PDE can be derived. Consider a system where individuals have two state dimensions: $m_1,m_2$. Then the total number of individuals in the bin spanned by intervals $[m_1-\Delta m_1,m_1+\Delta m_1)$,  $[m_2-\Delta m_2,m_2+\Delta m_2)$ is $n(m_1,m_2,a)\Delta m_1, \Delta m_2$ . The flux of individuals in and out of the size class can be expressed as
\begin{equation}\begin{array}{ll} J(m_1,m_2,a)=
&+g(m_1-0.5 \Delta m_1,m_2,a) \; n(m_1-0.5 \Delta m_1,m_2,a)\Delta m_2\\
&-g(m_1+0.5 \Delta m_1,m_2,a) \; n(m_1+0.5 \Delta m_1,m_2,a)\Delta m_2 \\
&+g(m_1,m_2-0.5 \Delta m_2,a) \; n(m_1,m_2-0.5 \Delta m_2,a)\Delta m_1\\
&-g(m_1,m_2+0.5 \Delta m_2,a) \; n(m_1,m_2+0.5 \Delta m_2,a)\Delta m_1 \\
&-d(m_1,m_2,a) \; n(m_1,m_2,a)\Delta m_1 \Delta m_2. \end{array}\end{equation}
The change in number of individuals within the interval across a time step \textit{$\Delta $t} is then:
\begin{equation}\begin{array}{ll} n(m_1,m_2,a+\Delta a)\Delta m_1\Delta m_2&-n(m_1,m_2,a)\Delta m_1\Delta m_2=\\
&+g(m_1-0.5 \Delta m_1,m_2,a) \; n(m_1-0.5 \Delta m_1,m_2,a)\Delta m_2 \Delta a\\
&-g(m_1+0.5 \Delta m_1,m_2,a) \; n(m_1+0.5 \Delta m_1,m_2,a)\Delta m_2 \Delta a \\
&+g(m_1,m_2-0.5 \Delta m_2,a) \; n(m_1,m_2-0.5 \Delta m_2,a)\Delta m_1 \Delta a\\
&-g(m_1,m_2+0.5 \Delta m_2,a) \; n(m_1,m_2+0.5 \Delta m_2,a)\Delta m_1 \Delta a\\
&-d(m_1,m_2,a) \; n(m_1,m_2,a)\Delta m_1 \Delta m_2 \Delta a. \end{array}\end{equation}

Rearranging and taking the limits as $\Delta a, \Delta m_1, \Delta m_2 \to 0$ yields the PDE for the 2D system
\begin{equation}\begin{array}{ll} \frac{\partial }{\partial t} n(m_1,m_2,a)=&-\frac{\partial }{\partial m_1} (g(m_1,m_2,a) \; n(m_1,m_2,a)) -\frac{\partial }{\partial m_2} (g(m_1,m_2,a) \; n(m_1,m_2,a))
\\&-d (m_1,m_2,a) \; n(m_1,m_2,a).\end{array}\end{equation}

\newpage

## Exact solutions: Integration along characteristics {#IntegrationAlongCharacteristics}

Characteristics are the projections of an individual's trajectory in i-space and time, provided they do not die. By integrating along characteristics, the PDE can be solved to yield the a general solution for the density of individuals with time of birth $t_{b} > 0$:
\begin{equation}\label{eq:n1}
n(m,t)=n(m_0 , t_b) \exp \bigg(-\int _{t_b}^{t} \left[\frac{\partial g(m(\theta ),E(\theta ))}{\partial m} +d(m(\theta ),E(\theta ))\right] d\theta \bigg)
\end{equation}

This states that the density $n$ at a specific time is a product of density at the origin adjusted for changes through development and mortality. Using the second boundary condition above, $n(m_{0} ,t_{g})$ can be replaced by $B(x, n, E,t_{b}) / g(m_{0} ,E(t_{b}))$. Note that density decreases through time because of mortality (as per typical survival equation), but that growth also contributes to changes in density. If growth is slowing with size, (i.e. $\frac{\partial g(m(\theta ),E(\theta ))}{\partial m}  <0$), then density increases because the characteristics converge on one another. Eq. \ref{eq:n1} can be solved (numerically or analytically) to give density with full ecological feedback and in changing environemnts. \\

In cases where the environment is constant and time-independent ($E(t)=E_c$), it is possible to further simplify Eq. \ref{eq:n1} to give (see \cite{de_roos_gentle_1997} page 153 for proof):

\begin{equation}\label{eq:n2}
n(m,t)=\frac{B(x, n, E_c,t_{b})}{g(m,E_{c} )} \exp \bigg(-\int _{m_{b} }^{m}\frac{d(\xi ,E_{c} )}{g(\xi ,E_{c} )}  d\xi \bigg). \end{equation}

Equation \ref{eq:n2} can be used to obtain in cases where environment is constant, either externally imposed in which case an analytic solution may be possible or when Eq. \ref{eq:PDE-app} is at equilibrium $\bigg(\frac{\partial n(m, t)}{\partial t} =0\bigg)$. It is necessary to know the origin of the characteristic for individuals of size $m$ at time $t$ to model their density.

To this end, it is useful to specify a function $\tau(m ,m_{b}, Ec )$ for the age of an individual of size $m$ such that $t_{b} =t-\tau (m,m_{b} ,E_{c} )$.  Note also, $\exp \bigg(-\int _{m_{b} }^{m}\frac{d(\xi ,E_{c} )}{g(\xi ,E_{c} )}  d\xi \bigg)$ is equal to the overall survival to size $m$, i.e.
\begin{equation}S(m(t))= \exp \bigg(-\int _{m_{b} }^{m}\frac{d(\xi ,E_{c} )}{g(\xi ,E_{c} )}  d\xi \bigg) = \exp \bigg(-\int _{t_{b} }^{t}d(m(\tau) ,E_{c} ) \; {\rm d}\tau \bigg).\end{equation}

Using equation \ref{eq:n2} it is also possible to show that in populations at demographic equilibrium and without disturbance, the lifetime seed output of an average individual should be 1.0.  The seed rain for a community is given by
\begin{equation}R = \int_{m_{b}}^{m_{max}} f(m, E_{c}) \; n(m)\, dm.\end{equation}
Substituting from above we have
\begin{equation}R =  B(x, n, E_c,t_{b})  \int_{m_{b}}^{m_{max}} \frac{f(m, E_{c})}{g(m,E_{c} )} S(m) \, dm.\end{equation}
Note that $m(t) = m_{0} + \int_{t_{b}}^{t} g(m(\tau))\,{\rm d}\tau$ and $dm(t) = g(m(t)) dt$. This allows us to rexpress the integral above as
\begin{equation}R =  B(x, n, E_c,t_{b})  \int_{t_{b}}^{t_{max}} f(m(t), E_{c}) \; S(m(t)) \, dt.\end{equation}
The term in the integral gives the lifetime seed production of a single individual.  At demographic equilibrium $R =  B(x, n, E_c,t_{b})$ (seed rain = seed production), so
\begin{equation} \int_{t_{b}}^{t_{max}} f(m(t), E_{c}) \; S(m(t)) \, dt =1\end{equation}.

\newpage

## Method for deriving cohort approximations in the EBT {#DeriveEBTapproximation}

Given a function $\phi  \left( m \right)$, the integral $\int_{m_{i-1}}^{m_i}\phi  \left( m \right) n \left( x,m,a \right) \; {\rm d}m$ can be approximated by first Taylor expanding $\phi  \left( m \right)$ around $\mu_i$. Substituting into the integral, we obtain

\begin{equation} \label{eq:taylor1}
	\begin{array}{l}
		{\int_{m_{i-1}}^{m_i}\phi  \left( m \right) n \left( x,m,a \right) \; {\rm d}m } \\
		{{\rm \; \; \; \; \; \; \; \; \; \; \; \; \; }= \int_{m_{i-1}}^{m_i}\left[\phi  \left( \mu_i \right) +\phi ' \left( \mu_i \right)  \left( m-\mu_i \right) +\frac{\phi '' \left( \mu_i \right) }2  \left( m-\mu_i \right) ^2 +...\right]n \left( x,m,a \right) \; {\rm d}m } \\
		{{\rm \; \; \; \; \; \; \; \; \; \; \; \; \; =}\phi  \left( \mu_i \right) \; {\rm \; }\lambda_i+\int_{m_{i-1}}^{m_i}\frac{\phi '' \left( \mu_i \right) }2  \left( m-\mu_i \right) ^2 n \left( x,m,a \right) \; {\rm d}m +...}
	\end{array}
\end{equation}

Discarding the square and higher terms, gives the following first-order approximation,

\begin{equation} \label{eq:taylor2}
	\int_{m_{i-1}}^{m_i}\phi  \left( m \right) n \left( x,m,a \right) \; {\rm d}m {\rm =}\phi  \left( \mu_i \right) \; {\rm \; }\lambda_i,
\end{equation}

with an estimate for the truncation error of Eq. $\ref{eq:taylor2}$ given by the first discarded term from Eq. $\ref{eq:taylor1}$:

\begin{equation} \label{eq:error1}
	\epsilon_i = 0.5 \phi''(\mu_i)\int_{m_{i-1}}^{m_i} (m-\mu_i)^2n(x,m, a) \; {\rm d}m.
\end{equation}

\newpage

## Expressing integrals as an ODE {#writeIntegralAsOdeIvp}

There exist different techniques for numerically integrating a function  \cite{press_numerical_1995}. Adaptive techniques are of particular interest because they allow resolution to increased only where it is needed, thereby greatly reducing the amount of computational effort required for given accuracy. Most adaptive integration techniques fall into two broad groups:

1. Those taking points along domain in any order
2. Those that proceed along the domain in directional manner.

Second class includes ODE problems, but worth noting that ODE methods can be applied to any integral, as
\begin{equation}I=\int_{a}^{b}f(x) \; {\rm d}x\end{equation}
is equivalent to solving for the value $I\equiv y(b)$ the differential equation $\frac{dy}{dx}=f(x)$ with boundary  condition $y(a)=0.$ \cite{press_numerical_1995}.

## Changing base variables for integrations

Total photosynthesis is calculated by integrating assimilation over the leaf area of the plant using the simplified trapezoidal rule for regularly spaced intervals. The formula for total photosynthesis for a plant size $m$ as described in the model is given by
\begin{equation} \frac{m}{\phi}  \int_{0}^{h(m)}A(I(z)) \, l\left(\frac{z}{h(m)}\right) \, dz,\end{equation}
where $x\in [0,1]$ is the relative height in the canopy (1 gives top) and $l(x)$ is the density of leaf area at height $x$. However, rather than integrating this function directly, we can reformulate it to give an integral directly over leaf area, so that computational effort is spent on resolving changes in light availability with height, rather than resolving the shape of the leaf area distribution itself. To achieve this we require the integral of $l(x)$: $L(x) = 1-x^{\eta}(2-x^{\eta})$ for $0\leq x\leq1$,  which gives the cumulative fraction of leaf area above height $x$, and its inverse function,  $L^{-1}(u) = (1-\sqrt{u})^{\frac1{\eta}}$, which gives the relative height $x$ which has a fraction $u$ of total leaf area above it. Substituting into the above integral for total photosynthesis, we obtain
\begin{equation} \frac{m}{\phi}  \int_{0}^1A\big(I(h(m)L^{-1}(u))\big) \, du.\end{equation}

The formula for trapezoidal integration with unevenly grid is:
\begin{equation} \label{trap_uneven} \int_{x_1}^{x_{n}}f(x)dx \approx 0.5 \sum_{i=0}^{i=N-1} \bigg[ \left(f(x_{i}) + f(x_{i+1})\right)\left( x_{i+1}- x_{i}\right)\bigg].\end{equation}

For regularly spaced intervals we have the simplified version
\begin{equation} \label{trap_even} \int_{x_1}^{x_{n}}f(x)dx \approx \Delta x \bigg[0.5 f_{0}+f_2+\ldots+ f_{N-1}+0.5 f_{N}\bigg] .\end{equation}

### 2015-05-12 tree_physiology.Rmd
Import existing documentation from tree
For #7

traitecoevo/tree/blob/c3567bdbdee7af9ebf37b3698dfa126cde994a60/doc/EBT.md
traitecoevo/tree/blob/c3567bdbdee7af9ebf37b3698dfa126cde994a60/doc/growth_model.Rmd

---
title: "Modelling plant physiology in the TREE package"
author: "Daniel Falster, Rich FitzJohn"
affiliation: Biological sciences, Macquarie University NSW 2109
output: pdf_document
csl: downloads/style.csl
bibliography: refs.bib
---

\newcommand{\ud}{\mathrm{d}}
\newcommand{\tm}{\textrm}

```{r, results="hide", echo=FALSE, message=FALSE}
library(methods)
library(tree)
s <- new(Strategy)
tree_p <- s$parameters
```

# Equations

For the sake of brevity, dependencies of functions are shown only in the Symbol column. The variable $x$ refers to a vector of four traits that are varied in the model: $x = (\phi, \rho, h_m, s)$. Subscripts for size variables are: l = leaves, s = sapwood, b = bark and phloem, h = heartwood, r = fine roots. All mass measurements are in terms of dry mass.

## Traits

### Leaf mass per unit area (LMA)

Symbol: $\varphi$

Units: kg m$^{-2}$

Description: Mass cost for constructing a unit of leaf area

Default value: $\varepsilon$ = `r tree_p[["lma"]]`

### Stem tissue density
Symbol: $\rho$

Units: kg m$^{-3}$

Description: Mass cost for constructing a unit of stem volume

Default value: $\rho$ = `r tree_p[["rho"]]`

### Leaf nitrogen content
Symbol: $N$

Units: kg m$^{-2}$

Description: leaf nitrogen per unit leaf area

Default value: $N$ = `r tree_p[["n_area"]]`

### Height at maturation

Symbol: $h_m$

Units: m

Description: Height at which plants allocate 50% of surplus energy to reproduction.

Default value: $h_m$ = `r tree_p[["hmat"]]`

### Seed size

Symbol: $s$

Units: kg

Description: Mass of individual seeds dispersing from parent.

Default value: $s$ = `r tree_p[["s"]]`

## Individual size

### Height

Symbol: $h(x, a)$

Units: m

Description: Height of plant with traits $x$ at and age $a$. Height is the key size variable tracked during plant growth. With exception of any variable combination involving heartwood area or heartwood mass, the remaining size metrics are tied directly to height, so knowing height means you also know the other variables.

Determination: Plants begin with initial height at germination $h_{0}$, obtained by finding a value that satisfies the equation
\begin{equation} s = m_\tm{t}(x, h_0(x)). \end{equation}

Height growth rate $g(x, h(x, a^\prime))$ is calculated as

\begin{equation} \label{eq:dhdt}
\frac{\textrm{d}h}{\textrm{d}t}=\frac{\textrm{d}h}{\textrm{d}a_\textrm{l}}
\times \frac{\textrm{d}a_\textrm{l}}{\textrm{d}m_\textrm{t}}
\times \frac{\textrm{d}m_\textrm{t}}{\textrm{d}B}
\times \frac{\textrm{d}B}{\textrm{d}t}.
\end{equation}

Each element on the RHS of eq. \ref{eq:dhdt} is also a function of current traits and height. Height at given age is then obtained by integrating eq. \ref{eq:dhdt} with the initial conditions:
\begin{equation} h(x, 0) = h_0(x), \end{equation}
such that
\begin{equation} h(x, a) = h_0(x) + \int_{0}^{a} g(x, h(x, a^\prime)) \, \rm{d}a^\prime. \end{equation}

### Leaf area

Symbol: $a_\tm{l}(h)$

Units: m$^2$

Description: Total leaf area of plant with traits $x$ and height $h$

Determination: Leaf area is calculated directly from plant height as
\begin{equation} a_\tm{l} = \alpha_1 \, h^{\beta_1}. \end{equation}

Useful derivatives: The growth rate of leaf area is given by
\begin{equation} \label{eq:dhdt}
\frac{\textrm{d}a_\textrm{l}}{\textrm{d}t} = \frac{\textrm{d}h}{\textrm{d}a_\textrm{l}}
\times \frac{\textrm{d}a_\textrm{l}}{\textrm{d}m_\textrm{t}}
\times \frac{\textrm{d}m_\textrm{t}}{\textrm{d}B}
\times \frac{\textrm{d}B}{\textrm{d}t}
 \end{equation}

To solve equation \ref{eq:dhdt} , the following derivative is also required:

\begin{equation} \frac{\ud h}{\ud a_\tm{l}} = -\beta_1\big(\frac{a_\tm{l}}{\alpha_1}\big)^{-(\beta_1 + 1)}. \end{equation}

Parameter values: $\alpha_1$ = `r tree_p[["a1"]]`, $\beta_1$ = `r tree_p[["B1"]]`

### Area of sapwood

Symbol: $a_{ss}(x, a_\tm{l})$

Units: $m^2$

Description: Area of live sapwood in stem at base of a plant with traits $x$ and leaf area $a_\tm{l}$.

Determination: Sapwood area is assumed to be in functional balance with leaf area (i.e. the pipe model), such that:
\begin{equation} a_\tm{ss} = \theta^{-1} \, a_\tm{l}. \end{equation}

Useful derivatives:
\begin{equation} \frac{\ud a_\tm{ss}}{\ud a_\tm{l}} = \theta^{-1} \end{equation}

\begin{equation} \frac{\ud a_\tm{ss}}{\ud t} = \theta^{-1} \, \frac{\ud a_\tm{l}}{\ud t} \end{equation}

Parameter values: $\theta$ = `r tree_p[["theta"]]`

### Mass of sapwood

Symbol: $m_{ss}(x, h)$

Units: kg

Description: Total mass of sapwood in a plant with traits $x$ and height $h$.

Determination: Calculated from leaf area and height, by XXXX.

\begin{equation} m_\tm{ss} = \theta^{-1} \, \rho \, \eta_c \, a_\tm{l}(x,h) \, h \end{equation}

Useful derivatives:
\begin{equation} \frac{\ud m_\tm{ss}}{\ud a_\tm{l}} = \theta^{-1} \, \rho \, \eta_c \, \big( h + a_\tm{l} \, \frac{\ud h}{\ud a_\tm{l}} \big) \end{equation}

Parameter values: $\eta$ = `r tree_p[["eta"]]`

### Area of bark

Symbol: $a_{sb}(x, a_\tm{l})$

Units: kg

Description: Area of live bark area in stem at base of a plant with traits $x$ and leaf area $a_\tm{l}$.

Determination: Bark cross-sectional area is treated analogously to sapwood area. We assume a constant ratio $b$ between bark and sapwood area so that:

\begin{equation} a_\tm{sb} = b \, \theta^{-1} \, a_\tm{l} \end{equation}

Useful derivatives:
\begin{equation} \frac{\ud a_\tm{sb}}{\ud t} = b \, \theta^{-1} \, \frac{\ud a_\tm{l}}{\ud t} \end{equation}

\begin{equation} \frac{\textrm{d}a_\textrm{sb}}{\textrm{d}t}=b \, \theta^{-1} \, \frac{\textrm{d}a_\textrm{l}}{\textrm{d}t} \end{equation}

Parameter values: $b$ = `r tree_p[["b"]]`

### Mass of bark

Symbol: $m_{sb}(x, a_\tm{l})$

Units: kg

Description: Total mass of bark in a plant with traits $x$ and height $h$.

Determination: As for $m_{ss}$, calculated from leaf area and height such that:
\begin{equation} m_\tm{sb} = \theta^{-1} \, \rho \, \eta_c \, a_\tm{l} \, h \end{equation}

Useful derivatives:
\begin{equation} \frac{\ud m_\tm{sb}}{\ud a_\tm{l}} = \theta^{-1} \, \rho \, \eta_c \, \big( h + a_\tm{l} \, \frac{\ud h}{\ud a_\tm{l}} \big) \end{equation}

\begin{equation} \frac{\textrm{d}m_\textrm{b}}{\textrm{d}a_\textrm{l}}=b \, \theta^{-1} \, \rho \, \eta_c\big( h + a_\textrm{l} \, \frac{\textrm{d}h}{\textrm{d}A} \big) \end{equation}

### Area of heartwood

Symbol: $a_\tm{sh}(x, h, a)$

Units: kg

Description: Area of dead sapwood in stem at base of a plant with traits $x$, height $h$ and age $a$.

Determination: Heartwood area is created by the turnover of sapwood. As such area of heartwood is obtained by integrating the IVP with initial conditions:
\begin{equation} a_\tm{sh} = 0, \end{equation}
and rates of change
\begin{equation} \frac{\ud a_\tm{sh}}{\ud t} = k_\tm{s} \, a_\tm{ss}(x, a_\tm{l}(h)). \end{equation}
Thus,
\begin{equation} a_\tm{sh} = \int_0^t \frac{\ud a_\tm{sh}}{\ud t}(t^\prime) \, dt^\prime. \end{equation}

Parameter values: $k_\tm{s}$ = `r tree_p[["k_s0"]]`

### Mass of heartwood

Symbol: $m_\tm{sh}(x, h, a)$

Units: kg

Description: Mass of dead sapwood in stem for a plant with traits $x$, height $h$ and age $a$.

Determination: Heartwood mass is created by the turnover of sapwood. As such mass of heartwood is obtained by integrating the IVP with initial conditions:
\begin{equation} m_\tm{sh} = 0, \end{equation}
and rates of change
\begin{equation} \frac{\ud m_\tm{sh}}{\ud t} = k_\tm{s} \, m_\tm{ss}(x,h). \end{equation}
Thus,
\begin{equation} m_\tm{sh} = \int_0^t \frac{\ud m_\tm{sh}}{\ud t}(t^\prime) \, dt^\prime. \end{equation}

### Mass of fine roots

Symbol: $m_\tm{r}(x, a_\tm{l})$

Units: kg

Description: Mass of fine roots on a plant with traits $x$ and leaf area $a_l$.

Determination: root mass is assumed to be in functional balance with leaf area (i.e. the pipe model), such that:
\begin{equation} m_\tm{r} = \alpha_3 a_\tm{l}(h) \end{equation}

Useful derivatives:
\begin{equation} \frac{\ud m_\tm{r}}{\ud a_\tm{l}} = \alpha_3 \end{equation}
\begin{equation} \frac{\ud m_\tm{r}}{\ud t} = \alpha_3 \frac{\ud a_\tm{l}}{\ud t} \end{equation}

Parameter values: $\alpha_3$ = `r tree_p[["a3"]]`

### Total live mass

Symbol: $m_\tm{t}(x, h, a)$

Units: kg

Description: Total mass of live tissue on a plant with traits $x$ and height $h$.

Determination: Calculated from sum of mass in leaves, sapwood, bark and root.

\begin{equation} m_\tm{t} = m_\tm{l} + m_\tm{ss} + m_\tm{sb} + m_\tm{r} \end{equation}

## Mass production

All rates are per plant.

### Net production

Symbol: $\frac{\textrm{d}B}{\textrm{d}t}(x, h, E)$

Units: kg yr$^{-1}$

Description: Rate of biomass production, after replacing parts lost via turnover, for plant with traits $x$, height $h$ and growing in environment $E$.

Determination: Calculated as the difference between income (photosynthesis) and losses (respiration and turnover):
\begin{equation} \frac{\textrm{d}B}{\textrm{d}t} = \underbrace{\strut c_{\tm{bio}} Y}_\textrm{yield} \big(\underbrace{\strut \, A_\tm{p}\left(x, h, E_a\right)}_\textrm{photosynthesis} - \underbrace{\strut \, R(x, h)}_\textrm{respiration}\big) - \underbrace{\strut T(x, h)}_\textrm{turnover}. \end{equation}

Parameter values: $c_{\tm{bio}}$ = `r tree_p[["c_bio"]]`, $Y$ = `r tree_p[["Y"]]`

### Total photosynthesis (= gross primary production)

Symbol: $A_\tm{p}(x, h_\tm{l}, E)$

Units: mol yr$^{-1}$

Description:

Determination: Calculated by integrating assimilation per unit leaf area over the plant:

\begin{equation} A_\tm{p} = a_\tm{l}(h) \int_0^{h} A(x, E_a(z)) \, q(z, h) \; \ud z \end{equation}

### Photosynthesis per leaf area

Symbol: $A(x, E)$

Units: mol yr$^{-1}$

Description: $A(x, E)$ is the gross annual CO2 assimilation per unit leaf area at canopy openness $E$ for a leaf with traits $x$.

Determination: Calculated by using the approximation, which was derived by integrating instantaneous rates of assimilation (described by a rectangular hyperbola) over the diurnal solar cycles throughout the year:

\begin{equation} A = \frac{c_\tm{p1}}{c_\tm{p2}+ E} \end{equation}

Parameter values: $c_\tm{p2}$ = `r tree_p[["c_p1"]]`, $c_\tm{p2}$ = `r tree_p[["c_p2"]]`

### Maintenance respiration

Symbol: $R(x, h)$

Units: mol yr$^{-1}$

Description:

Determination: Calculated from sum of respiration in leaves, sapwood, bark and root. Leaf respiration is calculated per unit area, while sapwood and bark respiration are calculated per unit volume.

\begin{equation} R = c_\tm{R,l} \, N \, a_\tm{l} + c_\tm{R,ss} \, \frac{m_\tm{ss}}{\rho} + c_\tm{R,sb} \, \frac{m_\tm{sb}}{\rho}  +  c_\tm{R,r} \, m_\tm{r} \end{equation}

Parameter values: $c_\tm{R,l}$ = `r tree_p[["c_Rl"]]` mol CO2 / kgN / yr, $c_\tm{R,ss}$ = `r tree_p[["c_Rs"]]` mol CO2 / m3 / yr, $c_\tm{R,sb}$ = `r tree_p[["c_Rb"]]` mol CO2 / m3 / yr, $c_\tm{R,r}$ = `r tree_p[["c_Rr"]]` mol CO2 / kg / yr.

### Total turnover

Symbol: $T(x, h)$

Units: kg yr$^{-1}$

Description:

Determination:
\begin{equation} T = \sum_\textrm{i=l,ss,sb,r}{k_i \, m_i} \end{equation}

### Leaf turnover rate

Symbol: $k_\tm{l}$

Units: yr$^{-1}$

Description: Rate at which leaf mass dies and is lost.

Determination: Assumed proportional to leaf mass per area, as per scaling relationship observed in large datasets:

\begin{equation} k_\tm{l} = k_\tm{l0} \left(\frac{\phi}{\phi_0}\right)^{-\beta_4} \end{equation}

The relationship is centred around lma of $\phi_0$. Thus  $k_\tm{l0}$ is the leaf turnover rate for a leaf with $\phi = \phi_0$.
Parameter values: $k_\tm{l0}$ = `r tree_p[["k_l0"]]`, $\beta_4$ = `r tree_p[["B4"]]`, $\phi_0$ = `r tree_p[["lma"]]`

### Sapwood turnover rate

Symbol: $k_\tm{ss}$

Units: yr$^{-1}$

Description: Rate at which sapwood mass dies and becomes heartwood. Relationship below allows for this value to be set

\begin{equation} k_\tm{ss} = k_\tm{s0} \left(\frac{\rho}{\rho_0}\right)^{-\beta_5} \end{equation}

The relationship is centred around wood density of $\phi_0$. Thus  $k_\tm{s0}$ is the leaf turnover rate for a sapwood with $\rho = \rho_0$. Setting $\beta_5=0$ should recover the previous model, where tunrover was independent of wood density.

Parameter values: $k_\tm{s0}$ = `r tree_p[["a5"]]`, $\beta_5$ = `r tree_p[["B5"]]`, $\rho_0$ = `r tree_p[["rho"]]`

### Bark turnover rate

Symbol: $k_\tm{sb}$

Units: yr$^{-1}$

Description: Rate at which bark mass dies and is lost.

Determination: Set to be same as sapwood turnover rate
\begin{equation} k_\tm{sb} = k_\tm{ss} \end{equation}

Parameter values: $k_\tm{sb}$ = `r tree_p[["k_b"]]`,

### Root turnover rate

Symbol: $k_\tm{r}$

Units: yr$^{-1}$

Description: Rate at which root mass dies and is lost.

Parameter values: $k_\tm{r}$ = `r tree_p[["k_r"]]`,

### Reproductive allocation

Symbol: $r(x, h)$

Units: Dimensionless (0 to 1)

Determination: Fraction of surplus production ($allocated to reproduction

Determination:
\begin{equation} r = c_{\tm{r}, 1}\left(1 + \exp\left(c_{\tm{r}, 2} \left(1-\frac{h(x, m_l)}{h_m}\right)\right)\right)^{-1} \end{equation}

Parameter values: $c_{\tm{r}, 1}$ = `r tree_p[["c_r1"]]`, $c_{\tm{r}, 2}$ = `r tree_p[["c_r2"]]`

### Leaf deployment rate
Symbol: $\frac{\ud a_\tm{l}}{\ud m_\tm{t}} (x, h)$

Units: Dimensionless (0 to 1)

Determination: $\textrm{d}a_\textrm{l} / \textrm{d}m_\textrm{t}$ accounts for the cost of deploying a unit of leaf area, including construction of the leaf itself and various support structures. As such, $\textrm{d}a_\textrm{l} / \textrm{d}m_\textrm{t}$ can itself be decomposed into a sum of construction costs per unit leaf area:

\begin{equation} \label{eq:daldmt}
\frac{\textrm{d}a_\textrm{l}}{\textrm{d}m_\textrm{t}}
= \big(\phi
 + \frac{\textrm{d}m_\textrm{s}}{\textrm{d}a_\textrm{l}} + \frac{\textrm{d}m_\textrm{b}}{\textrm{d}a_\textrm{l}} + \frac{\textrm{d}m_\textrm{r}}{\textrm{d}a_\textrm{l}}\big)^{-1}.
 \end{equation}

### Fraction of production allocated to growth

Symbol: $\frac{\textrm{d}m_\textrm{t}}{\textrm{d}B}$

Units: Dimensionless (0 to 1)

Determination:
\begin{equation} \frac{\textrm{d}m_\textrm{t}}{\textrm{d}B} = 1- r(x, h) \end{equation}

### Fecundity

Symbol: $f(x, h, E)$

Units: yr$^{-1}$

Description: Rate of offspring production for plant with traits $x$, height $h$ and growing in environment $E$.

Determination:
\begin{equation} f = \frac{r(x, h) \;\frac{\textrm{d}B}{\textrm{d}t}(x, h, E)}{s + c_{\tm{acc}} \left(\frac{s}{s_0}\right)^{B7}} \end{equation}
if $\frac{\textrm{d}B}{\textrm{d}t}(x, h_0(x), E)>0$, otherwise 0

Parameter values: $c_{\tm{acc}}$ = `r tree_p[["c_acc"]]`, $B_7$ = `r tree_p[["B7"]]`

### Probability density of leaf area at height z for an individual of height h

Symbol: $q(z, h)$

Units: m$^{-1}$

Description: Fraction of plant's leaf area at height $z$, for plant of height $h$.

Determination:
\begin{equation} q = 2\eta(1-z^\eta h^{-\eta}) z^{\eta-1} h^{-\eta} \end{equation}
if $z \leq h$, otherwise 0.

## Mortality

### Survival of seedlings during germination

Symbol: $\Pi_1(x, m_{\tm{l}, 0}, E)$

Units: Dimensionless (0 to 1)

Description: Gives probability of plant suriving germination. Is a function of biomass growth rate for seedlings, such that seedlings unable to grow die.

Determination:
\begin{equation} \Pi_1 = \left(1 + \left(\frac{c^2_{\tm{s}, 0}}{\frac{\frac{\textrm{d}B}{\textrm{d}t}(x, h_0(x), E)}{a_\tm{l}(x, h_0(x))}} \right)^{2} \right)^{-1} \end{equation}
if $\frac{\textrm{d}B}{\textrm{d}t}(x, h_0(x), E) > 0$, otherwise 0

Parameter values: $c_{\tm{s}, 0}$ = `r tree_p[["c_s0"]]`

### Mortality rate

Symbol: $d_\tm{t}(x, h, E)$

Units: yr$^{-1}$

Description: Instantaneous mortality rate for a plant with traits $x$, height $h$ and growing in environment $E$. Given as sum of growth-dependent and growth-independent hazards.

Determination:
\begin{equation} d_\tm{t} = d_\tm{i}(x, h) + d_\tm{d}(x, h, E) \end{equation}

### Growth-independent mortality rate

Symbol: $d_\tm{i}(x, h)$

Units: yr$^{-1}$

Description: Mortality rate for a plant with traits $x$ and height $h$ due to growth-independent factors. Set to potentially vary with wood density and height. Setting $-c_{\ud 1}>0$ means species with lower wood density have higher mortality.  Setting $\beta_6=0$ means effect varies with wood density but is independent of $h$, which is similar to the model in @Falster-2011 when $-c_{\ud 1} = 1.3$. When $\beta_6 >0$ the effect of low wood density increases with height.

Determination:

\begin{equation}
d_\tm{i} = c_{\ud 0}\left(\frac{\rho}{\rho_0}\right)^{-c_\tm{d1}}  h^{\beta_6}
\end{equation}

Parameter values: $c_{\ud 0}$ = `r tree_p[["c_d0"]]`, $\rho_0$ = `r tree_p[["rho"]]`, $c_{\ud 1}$ = `r tree_p[["c_d1"]]`, $\beta_6$ = `r tree_p[["B6"]]`

### Growth-dependent mortality rate

Symbol: $d_\tm{d}(x, h)$

Units: yr$^{-1}$

Description: Mortality rate for a plant with traits $x$, height $h$ and growing in environment $E$ due to growth-dependent factors.

Determination:
\begin{equation} d_\tm{d} = c_{\ud 2} \exp\left(-c_{\ud 3} \frac{\frac{\textrm{d}B}{\textrm{d}t}(x, h, E)}{a_\tm{l}(h)} \right) \end{equation}

Parameter values: $c_{\ud 2}$ = `r tree_p[["c_d2"]]`, $c_{\ud 3}$ = `r tree_p[["c_d3"]]`

\newpage

# Appendix 1: Moving from leaf mass to height as key variable

The focus on leaf mass as the key variable in @Falster-2011 is fairly arbitrary.
Instead, consider a change of variables to use "height" as the key
size variable. This is useful because we need to know about plant
height elsewhere, and it's a common measuring stick that is fairly
independent of the key traits.

To convert from leaf mass ($m_l$) to height ($h$), use equations 3 and
4 in @Falster-2011:

\begin{equation}
h = \alpha_1 \left(\frac{m_l}{\phi}\right)^{\beta_1}
 \end{equation}

where $\phi$ is the leaf mass per unit area (\textsc{lma}) and
$\alpha_1$ and $\beta_1$ are scaling parameters relating leaf area
($m_l/\phi$) to height.

What we need for the main calculations is the derivative of height $h$
with respect to time. Because $m_l$ is a function of time, we have

\begin{equation}
\frac{\ud h}{\ud t} = \frac{\ud}{\ud t}[m_l(t)]
\frac{\alpha_1 \beta_1}{\phi} \left(\frac{m_l(t)}{\phi}\right)^{\beta_1 - 1}
 \end{equation}

where $\frac{\ud}{\ud t}[m_l(t)]$ comes from equation 19:

\begin{equation}
\frac{\ud}{\ud t}[m_l(t)] = (1 - r)P\frac{\ud m_l}{\ud m_t}
 \end{equation}

where $P$ is total production, $r$ is allocation to reproduction and
$\ud m_l/\ud m_t$ is the fraction of whole plant growth that is leaf.

So overall, we have

\begin{equation}
\frac{\ud h}{\ud t} =
\frac{\ud m_t}{\ud t} \frac{\ud m_l}{\ud m_t} \frac{\ud h}{\ud m_l}
 \end{equation}

---

Here, I'm just explicitly writing down the derivations of the pieces
needed for this equation, and for the equivalent equation used in the
R version of the growth model.

From equation (18) we have

\begin{equation}
\frac{\ud m_l}{\ud m_t} =
\frac{1}{1 +
	\frac{\ud m_s}{\ud m_l} +
	\frac{\ud m_b}{\ud m_l} +
	\frac{\ud m_h}{\ud m_l} +
	\frac{\ud m_r}{\ud m_l}}
 \end{equation}

(because $m_t = m_l + m_s + m_b + mh + m_r$, and $\ud m_l/\ud m_t = 1
/(\ud m_t / \ud m_l)$).

Using appendix S2 of Falster et al. (2010).

\begin{equation}
\frac{\ud m_s}{\ud m_l} = \frac{\ud}{\ud m_l}\left[
\frac{\rho \eta_c}{\theta} \omega(m_l) h(m_l)\right] =
\frac{\ud}{\ud m_l}\left[
\frac{\rho \eta_c}{\theta} \frac{m_l}{\phi}
\alpha_1\left(\frac{m_l}{\phi}\right)^{\beta_1}\right] =
\frac{\ud}{\ud m_l}\left[
\frac{\rho \eta_c}{\theta}
\alpha_1\left(\frac{m_l}{\phi}\right)^{\beta_1 + 1}\right] =
% r n / t (m/p) a (m/p) ^ b
% D[r n / t (m/p) a (m/p) ^ b, m] // Simplify
\frac{\rho \eta_c}{\theta\phi}\alpha_1(\beta_1 + 1)
\left(\frac{m_l}{\phi}\right)^{\beta_1}
 \end{equation}
and
\begin{equation}
\frac{\ud m_b}{\ud m_l} = \frac{\ud}{\ud m_l}[b m_s] =
b \frac{\ud m_s}{\ud m_l} =
b \frac{\rho \eta_c}{\theta\phi}\alpha_1(\beta_1 + 1)
\left(\frac{m_l}{\phi}\right)^{\beta_1}
 \end{equation}
and
\begin{equation}
\frac{\ud m_h}{\ud m_l} =
\frac{\ud}{\ud m_l}\left[
\rho\eta_c\alpha_2 \omega(m_l)^{\beta_2}
\right] =
\frac{\ud}{\ud m_l}\left[
\rho\eta_c\alpha_2 \left(\frac{m_l}{\phi}\right)^{\beta_2}
\right] =
\frac{\rho\eta_c\alpha_2}{\phi}\beta_2
\left(\frac{m_l}{\phi}\right)^{\beta_2-1} =
% D[r n a (m / p)^b, m]
% D[r n a (m / p)^b, m] // Simplify
\frac{\rho\eta_c\alpha_2}{m_l}\beta_2
\left(\frac{m_l}{\phi}\right)^{\beta_2}
 \end{equation}
and
\begin{equation}
\frac{\ud m_r}{\ud m_l} =
\frac{\ud}{\ud m_l}\left[\alpha_3 \omega(m_l)\right] =
\frac{\ud}{\ud m_l}\left[\alpha_3 \frac{m_l}{\phi}\right] =
\frac{\alpha_3}{\phi}
 \end{equation}

So, the denomininator of $\ud m_l / \ud m_t$ looks like:

\begin{equation}
1 +
% d m_s / d m_l
\frac{\rho \eta_c}{\theta\phi}\alpha_1(\beta_1 + 1)
\left(\frac{m_l}{\phi}\right)^{\beta_1} +
% d m_b / d m_l
b \frac{\rho \eta_c}{\theta\phi}\alpha_1(\beta_1 + 1)
\left(\frac{m_l}{\phi}\right)^{\beta_1} +
% d m_h / d m_l
\frac{\rho\eta_c\alpha_2}{m_l}\beta_2
\left(\frac{m_l}{\phi}\right)^{\beta_2} +
% d m_r / d m_l
\frac{\alpha_3}{\phi}
 \end{equation}

\begin{equation}
1 +
% d m_s / d m_l + d m_b / d m_l
(1 + b) \frac{\rho \eta_c}{\theta\phi}\alpha_1(\beta_1 + 1)
\left(\frac{m_l}{\phi}\right)^{\beta_1} +
% d m_h / d m_l
\frac{\rho\eta_c\alpha_2}{m_l}\beta_2
\left(\frac{m_l}{\phi}\right)^{\beta_2} +
% d m_r / d m_l
\frac{\alpha_3}{\phi}
 \end{equation}

Repeating this with the version in the R verison, working in terms of
leaf area.
\begin{equation}
\frac{\ud \omega}{\ud m_t} =
\frac{1}{
	\frac{\ud m_l}{\ud \omega} +
	\frac{\ud m_s}{\ud \omega} +
	\frac{\ud m_b}{\ud \omega} +
	\frac{\ud m_h}{\ud \omega} +
	\frac{\ud m_r}{\ud \omega}}
 \end{equation}

\begin{equation}
\frac{\ud m_l}{\ud \omega} =
\frac{\ud}{\ud \omega}\left[\omega\phi\right] =
\phi
 \end{equation}
and
\begin{equation}
\frac{\ud m_s}{\ud \omega} =
\frac{\ud}{\ud \omega}\left[
\frac{\rho \eta_c}{\theta} \omega h(\omega)\right] =
\frac{\ud}{\ud \omega}\left[
\frac{\rho \eta_c}{\theta} \alpha_1 \omega^{\beta_1 + 1}\right] =
\frac{\rho \eta_c}{\theta}\alpha_1(\beta_1 + 1)\omega^{\beta_1}
 \end{equation}
and
\begin{equation}
\frac{\ud m_b}{\ud \omega} =
\frac{\ud}{\ud \omega}\left[b m_s\right] =
b \frac{\ud m_s}{\ud \omega} =
b \frac{\rho \eta_c}{\theta}\alpha_1(\beta_1 + 1)\omega^{\beta_1}
 \end{equation}
and
\begin{equation}
\frac{\ud m_h}{\ud \omega} =
\frac{\ud}{\ud \omega}\left[\rho\eta_c\alpha_2 \omega^{\beta_2}
\right] =
\rho\eta_c\alpha_2\beta_2\omega^{\beta_2 - 1}
 \end{equation}
and
\begin{equation}
\frac{\ud m_r}{\ud \omega} =
\frac{\ud}{\ud \omega}\left[\alpha_3 \omega\right] = \alpha_3
 \end{equation}

Other useful derivatives:

Because
\begin{equation}
h = \alpha_1 \left(\frac{m_l}{\phi}\right)^{\beta_1}
\qquad
m_l = \phi\left(\frac{h}{\alpha_1}\right)^{1/\beta_1}
 \end{equation}

we have

\begin{equation}
\frac{\ud h}{\ud m_l} = \frac{\alpha_1 \beta_1}{\phi}
\left(\frac{m_l}{\phi}\right)^{\beta_1 - 1}
\qquad
\frac{\ud m_l}{\ud h} =
\frac{\phi}{\alpha_1 \beta_1}\left(\frac{h}{\alpha_1}\right)^
{\frac{1}{\beta_1} - 1}
 \end{equation}

# References

### 2015-05-12 tree_demography.Rmd
Clean up vignettes (add refs, remove superfluous material). #7

@@
 
-# To do
-
-- Table of units, symbols, subscripts
-- Add equations for germination limit
-- clean up text at start
-- formula for EBT cohorts
-- formula for error in EBT
-
-\newpage
-
 # System dynamics
-
-A system is physiologically or size structured whenever the basic rates of individuals (growth, mortality, reproduction) vary as functions of their size.
-These rates may also vary through environment, which can include feedbacks dependent on the current population state (e.g.
-density dependence).
-Theory exists for tracing the development of a size- structured based on a partial differential equation (PDE) describing changes in the size distribution of individuals at any given time (describing the density function of individuals, rather than location of specific individuals).
-PSPs are formulated such that spatial distribution is neglected; only large populations are considered.
-Consequently the dynamics are entirely deterministic.
-These assumptions ensure similar individuals experience a homogeneous environment and are thus treated equally, and that inherently stochastic processes such as mortality can be reliably approximated by a deterministic description [makes use of the law of large numbers].
-Individuals enter at the lower size class according to the rate of offspring production.
-The continuous nature of the system, with appropriate feedbacks, then ensures a dynamical solution to evolving size-structure including the problem of self-thinning.
-The dynamics of the population is given by the cumulative actions of its individuals.
-
-## Patch- and size-structured population dynamics
 
@@
 Such a system can be modelled as a metapopulation (sometimes called metacommunity for multiple species).
-Assuming patches are large, and there are many of them, the dynamics of the metapopulation can be captured deterministically using two PDEs (Kohyama 1993; deRoos 1997).
+Assuming patches are large, and there are many of them, the dynamics of the metapopulation can be captured deterministically using two PDEs [@Kohyama-1993; @Deroos-1997; @Moorcroft-2001].
 Together these PDEs account for the combined influences of size-dependent variation in an individual's vital rates and disturbances on the size distribution of individuals.
@@
 
-The first PDE accounts for the effects of disturbances on the distribution of patch ages (defined as time since last disturbance) in the landscape (Von Foerster 1959; McKendrick 1926).
+The first PDE accounts for the effects of disturbances on the distribution of patch ages (defined as time since last disturbance) in the landscape [@Vonfoerster-1959; @Mckendrick-1926].
 
@@
 
-\begin{equation} \frac{\partial}{\partial a} p(a)  = -\gamma(a) \, p(a) , \end{equation}
+\begin{equation} \frac{\partial}{\partial a} p(a)  = -\gamma(a) \, p(a) ,
+\end{equation}
 with boundary condition
-\begin{equation}  p(0)  = \int_0^\infty \gamma(a) \, p(a) \; {\rm d} a. \end{equation}
+\begin{equation}  p(0)  = \int_0^\infty \gamma(a) \, p(a) \; {\rm d} a.
+\end{equation}
 
 This leads to a solution for the equilibrium distribution of patch-ages
-\begin{equation} p(a) = p(0) \exp \left( \int_0^a \gamma(a^\prime) \; {\rm d} a^\prime \right),\end{equation}
+\begin{equation} p(a) = p(0) \exp \left( \int_0^a \gamma(a^\prime) \; {\rm d} a^\prime \right),
+\end{equation}
 where
 \begin{equation}
-\frac1{p(0)} = \left( \int_0^\infty \exp \left( \int_0^{a^{\prime\prime}} - \gamma(a^\prime) \; {\rm d} a^\prime \right) d a^{\prime\prime} \right),\end{equation}
+  \frac{p(0)} = \left( \int_0^\infty \exp \left( \int_0^{a^{\prime\prime}} - \gamma(a^\prime) \; {\rm d} a^\prime \right) d a^{\prime\prime} \right),
+\end{equation}
  corresponds to the average lifespan of a patch.
 
-The second PDE accounts for the development of competitive hierarchies within patches recovering from a disturbance (Sinko \& Streifer 1967; Hara 1984; Metz \& Diekmann 1986; de Roos 1997).
+The second PDE accounts for the development of competitive hierarchies within patches recovering from a disturbance [@Sinko-1967;  @Hara-1984; @Metz-1986; @Deroos-1997].
 Here we only consider cases where dispersal is restricted to newborns and where disturbances remove all established individuals within a patch.
@@
 
-\begin{equation} \label{eq:PDE} \frac{\partial }{\partial a} n(x,m,a)= -d(x,m, a) \; n(x,m,a)-\frac{\partial }{\partial m} \left[g(x,m,a) \; n(x,m,a)\right].  \end{equation}
+\begin{equation} \label{eq:PDE} \frac{\partial }{\partial a} n(x,m,a)= -d(x,m, a) \; n(x,m,a)-\frac{\partial }{\partial m} \left[g(x,m,a) \; n(x,m,a)\right].
+\end{equation}
 (See appendix for [derivation of PDE approximating size structured dynamics](#pdeDeriv-size).)
@@
 
-\begin{equation}  \label{eq:BC1} \frac{n(x,m_0,a)g(x,m_0 , a)}{\pi_1(x,m_0 ,a)} = \int_0^{\infty} p(a)  \int_0^{\infty}  \pi_0f(x,m,a) \; n(x,m,a)\,{\rm d} m \, {\rm d} a, \end{equation}
+\begin{equation}  \label{eq:BC1} \frac{n(x,m_0,a)g(x,m_0 , a)}{\pi_1(x,m_0 ,a)} = \int_0^{\infty} p(a)  \int_0^{\infty}  \pi_0f(x,m,a) \; n(x,m,a)\,{\rm d} m \, {\rm d} a,
+\end{equation}
 
@@
 
-\begin{equation} \label{eq:BC2} n\left(x,m,0\right) =0. \end{equation}
+\begin{equation} \label{eq:BC2} n\left(x,m,0\right) =0.
+\end{equation}
 
@@
 
-### Environmental feedback
+## Environmental feedback
 
 Individuals are assumed to interact  via their influence on the availability of resources.
-A competition kernel is required, giving the influence of an individual on the availability of resources for another individual.
-To get the total influence of all individuals on the target individual requires integration of the competition kernel over the density distribution $n$.
-If there are $N$ different species or types in the system, then their combined effects must be summed.
-
-The approach outlined is very general and can be tailored to suit a range of different systems. To do so we need to describe:
-\begin{itemize}
- \item the influence of an individual's size (state), its traits (type), and the availability of resources (environment) on the vital rates of growth, mortality and reproduction
- \item the dependence of resource availability on size-distribution of individuals within each species and their traits.
-\end{itemize}
-
 For plants, where we are interested in the effect of shading from other individuals on growth, mortality and fecundity, we can define a function for the amount of light at height $z$ in a patch of age $a$:
@@
 \begin{equation} \label{eq:light}
-E(z,a) = \exp \left(-c_{ext}  \sum_{i=1}^{N} \int_{0}^{\infty} \phi(m) \, L(z, h(m)) \, n(x_i, m,a) \, dm \right), \end{equation}
+  E(z,a) = \exp \left(-c_{ext}  \sum_{i=1}^{N} \int_{0}^{\infty} \phi(m) \, L(z, h(m)) \, n(x_i, m,a) \, dm \right),
+\end{equation}
 where $\phi(m)$ is total leaf area and $L(z, h(m))$ is fraction of this leaf area held above height $z$ for plants size $m$, and $c_{ext}$ is the light extinction coefficient.
@@
 
-Invasion fitness is most correctly defined as the long-term per capita growth rate of a rare mutant population growing in the environment determined by the resident strategy (Metz et al.
-1992).
-Calculating per-capita growth rates, however, can be particularly challenging in a structured metapopulation model (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001).
+Invasion fitness is most correctly defined as the long-term per capita growth rate of a rare mutant population growing in the environment determined by the resident strategy [@Metz-1992].
+Calculating per-capita growth rates, however, can be particularly challenging in a structured metapopulation model [@Gyllenberg-2001; @Metz-2001].
 As an alternative measure of fitness, we can use the basic reproduction ratio, which gives the expected number of new dispersers arising from a single dispersal event.
-Evolutionary inferences made using the basic reproduction ratio will be similar to those made using per-capita growth rates for metapopulations at demographic equilibrium (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001).
+Evolutionary inferences made using the basic reproduction ratio will be similar to those made using per-capita growth rates for metapopulations at demographic equilibrium [@Gyllenberg-2001; @Metz-2001].
 
@@
 \end{equation}
-where $\tilde{R}\left(x^\prime,a_0,a \right)$ is the expected number of dispersing offspring produced by a single disperser arriving in a patch of age $a_0$ up until age $a$ (Gyllenberg \& Metz 2001; Metz \& Gyllenberg 2001).
+where $\tilde{R}\left(x^\prime,a_0,a \right)$ is the expected number of dispersing offspring produced by a single disperser arriving in a patch of age $a_0$ up until age $a$ [@Gyllenberg-2001; @Metz-2001].
 $\tilde{R}\left(x^\prime,a,\infty\right)$ is calculated by integrating an individual's fecundity over the expected lifetime the patch, taking into account competitive shading from residents with traits $x$, the individual's probability of surviving, and its traits via the equation:
 
-\begin{equation} \label{eq:tildeR} \tilde{R}(x^\prime, a_0, a) =\int_{a_0}^{a}  \pi_0f(x^\prime, m(x^\prime, a_0, a^\prime), a^\prime) \, S_{\rm I} (x^\prime, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime.\end{equation}
+\begin{equation} \label{eq:tildeR}
+  \tilde{R}(x^\prime, a_0, a) =\int_{a_0}^{a}  \pi_0f(x^\prime, m(x^\prime, a_0, a^\prime), a^\prime) \, S_{\rm I} (x^\prime, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime.
+\end{equation}
 
 Here,
-\begin{equation} \label{eq:size} m(x^\prime, a_0, a) = m_0 + \int_{a_0}^{a} g(x^\prime, m(x^\prime, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime \end{equation}
+\begin{equation} \label{eq:size}
+  m(x^\prime, a_0, a) = m_0 + \int_{a_0}^{a} g(x^\prime, m(x^\prime, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime
+\end{equation}
 
@@
 
-\begin{equation} \label{eq:survivalIndividual} S_{\rm I} (x^\prime, a_0, a) = \pi_1 (x^\prime,m_0, a_0) \, \exp\left(- \int_{a_0}^{a} d(x,m_i(x^\prime, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right) \end{equation}
+\begin{equation} \label{eq:survivalIndividual}
+  S_{\rm I} (x^\prime, a_0, a) = \pi_1 (x^\prime,m_0, a_0) \, \exp\left(- \int_{a_0}^{a} d(x,m_i(x^\prime, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right)
+\end{equation}
 
@@
 
-\begin{equation} \label{eq:survivalPatch} S_{\rm P} (a_0,a) = \exp\left(-\int_{a_0}^{a} \gamma(a^\prime) \, {\rm d} a^\prime \right) \end{equation}
+\begin{equation} \label{eq:survivalPatch}
+  S_{\rm P} (a_0,a) = \exp\left(-\int_{a_0}^{a} \gamma(a^\prime) \, {\rm d} a^\prime \right)
+\end{equation}
 gives the probability that the patch remains undisturbed from $a_0$ to $a$.
@@
 Firstly, the total number of individuals in the metapopulation is given by
-\begin{equation}\hat{n}(x) = \int_{0}^{\infty} \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a \, {\rm d}m,\end{equation}
+\begin{equation}
+  \hat{n}(x) = \int_{0}^{\infty} \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a \, {\rm d}m,
+\end{equation}
 and the average density of plants size $m$ by
-\begin{equation}\bar{n}(x,m) = \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a.\end{equation}
+\begin{equation}\bar{n}(x,m) = \int_{0}^{\infty}p(a) \; n(x,m,a) \, {\rm d}a.
+\end{equation}
 
@@
 The average value of $\phi$ across the metapopulation is
-	\begin{equation}\hat{\phi}(x) = \frac1{\hat{n}(x) }\int_{0}^{\infty} \int_{0}^{\infty} p(a) \; n(x,m,a) \phi(x,m,a) \, {\rm d}a \, {\rm d}m,\end{equation}
+\begin{equation}\hat{\phi}(x) = \frac1{\hat{n}(x) }\int_{0}^{\infty} \int_{0}^{\infty} p(a) \; n(x,m,a) \phi(x,m,a) \, {\rm d}a \, {\rm d}m,
+\end{equation}
 while the average  value of $\phi$ for plants size $m$ is
- \begin{equation}\bar{\phi}(x,m) = \frac1{\bar{n}(x,m) }\int_{0}^{\infty}p(a)  n(x,m,a) \phi(x,m,a)\, {\rm d}a.\end{equation}
+\begin{equation}\bar{\phi}(x,m) = \frac1{\bar{n}(x,m) }\int_{0}^{\infty}p(a)  n(x,m,a) \phi(x,m,a)\, {\rm d}a.
+\end{equation}
 
@@
 
-## Converting density from one size unit to another
-
-Population density is explicitly modelled in relation to a given size unit (Eq.
-$\ref{eq:PDE}$).
-But what if we want to express density in relation to another size unit?
-A relation between the two can be derived by noting that the total number of individuals within a given size range must be equal.
-So let's say density is expressed in units of size $m$, but we want density in units of size $h$.
-First we require a one-to-one function which h for a given $m$: $h = \hat{h}(m)$.
-Then the following must hold
-\begin{equation} \label{eq:n_conversion} \int_{m_1}^{m_2} n(x,m,a) \; \textrm{d}m =  \int_{\hat{h}(m_1)}^{\hat{h}(m_2)} n^\prime(x,m,a) \; \textrm{d}h  \end{equation}
-
-For very small size intervals, this equation is equivalent to
-\begin{equation} \left(m_2- m_1 \right) \; n(x,m_1,a) = \left( \hat{h}(m_2) - \hat{h}(m_1)\right) \; n^\prime(x, \hat{h}(m_1),a). \end{equation}
-
-Rearranging gives
-\begin{equation}  n^\prime(x, \hat{h}(m_1),a) = n(x, m_1,a) \; \frac{m_2- m_1}{\hat{h}(m_2) - \hat{h}(m_1)}  \end{equation}
-
-Noting that the second term on the RHS is simply the definition of $\frac{\delta m}{\delta h}$ evaluated at $m_1$, we have
-\begin{equation} \label{eq:n_conversion2} n^\prime(x, h, a) = n(x, m,a) \; \frac{\delta m}{\delta h}. \end{equation}
-
-\newpage
-
 # Approximating system dynamics using the escalator boxcar train (EBT)
 
+Our approach for modelling solving size-structured population dynamics is based on the Escalator Boxcar Train technique  (EBT) [@Deroos-1997; @Deroos-1992; @Deroos-1988].
 The EBT solves the PDE describing development of $n(x,m,a)$ (Eq.
-$\ref{eq:PDE}$) by approximating the density function with a collection of cohorts spanning the size spectrum (de Roos 1988, 1997).
+$\ref{eq:PDE}$) by approximating the density function with a collection of cohorts spanning the size spectrum.
 Following a disturbance, a series of cohorts are introduced into each patch.
@@
 $\ref{eq:PDE}$  becomes an ordinary differential equation (ODE); biologically, these are the growth trajectories of individuals, provided they do not die).
-The numerical technique consists of approximating changes to the number of individuals within each cohort by a system of closed-form linear ODEs.
+
+The original EBT [@Deroos-1997; @Deroos-1992; @Deroos-1988] proceeds by approximating the first and second moments of the density distribution $n \left( x,m,a \right)$ within sub-domains $\Omega_i$, represented by $\lambda_i$ and $\mu_i$, these being the total number and mean size of individuals within $\Omega_i$ respectively.
+Under this assumption, the rates of change $\frac{\rm d} {{\rm d}a} \lambda_i$ and $\frac{\rm d} {{\rm d}a} \mu_i$ can be approximated given by two ODEs, which can in turn be approximated by first-order closed-form solutions.
 Eq.
 $\ref{eq:PDE}$ is thereby reduced to a family of ODEs, which can be stepped using an appropriate ODE solver with an adaptive step size (Press 1995).
+The size distribution $n(x,m,a)$ is then approximated by a series of point masses with position and amplitude given by $\lambda_i$ and $\mu_i$.
 
-Below we consider two alternative implementations of the EBT.
-Both require that we introduce a series of cohorts into a patch, and that these be transported along the characterstics of the Eq.
-$\ref{eq:PDE}$ (see figure).
-
-<!-- ![Cohorts in the EBT](figs/EBT.png) -->
+In the current implementation we use a different approach for approximating the size distribution $n(x,m,a)$.
+Instead of tracking the first and second moments of the size-distribution within cohorts, we track a point mass estimate of $n$ along the characteristics corresponding to cohort boundary.
+By integrating along characteristics, the PDE can be solved to yield the a general solution for the density of individuals with time of birth $t_{b} > 0$:
+\begin{equation}\label{eq:n1}
+  n(m,t)=n(m_0 , t_b) \exp \bigg(-\int _{t_b}^{t} \left[\frac{\partial g(m(\theta ),E(\theta ))}{\partial m} +d(m(\theta ),E(\theta ))\right] d\theta \bigg)
+\end{equation}
+Eq.
+\ref{eq:n1} states that the density $n$ at a specific time is a product of density at the origin adjusted for changes through development and mortality.
+Using the second boundary condition above, $n(m_{0} ,t_{g})$ can be replaced by $B(x, n, E,t_{b}) / g(m_{0} ,E(t_{b}))$.
+Note that density decreases through time because of mortality (as per typical survival equation), but that growth also contributes to changes in density.
+If growth is slowing with size, (i.e.
+$\frac{\partial g(m(\theta ),E(\theta ))}{\partial m}  <0$), then density increases because the characteristics converge on one another.
+In cases where the environment is constant and time-independent, it may be possible to further simplify Eq.
+\ref{eq:n1} to give an analytical solution  for $n$ (see [@Deroos-1997} page 153 for proof).
+In most cases involving interesting ecological feedback, however, eq.
+\ref{eq:n1} must be solved numerically.
 
@@
 $\ref{eq:PDE}$ , they will be closed to transport processes across their upper and lower boundaries.
-
-## First method for approximating $n(x,m,a)$, based on modelling of cohort boundaries
 
@@
 
-### Size
+## Size
 
 From equation  $\ref{eq:size}$, the boundary develops as
-
-\begin{equation} \label{eq:boundSize} m_i(x, a_0, a) = m_0 + \int_{a_0}^{a} g(x,m_i(x, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime.\end{equation}
+\begin{equation} \label{eq:boundSize}
+  m_i(x, a_0, a) = m_0 + \int_{a_0}^{a} g(x,m_i(x, a_0, a^\prime),a^\prime) \, \rm{d}a^\prime.
+\end{equation}
 
@@
 
-### Survival
+## Survival
 From equation  $\ref{eq:survivalIndividual}$, the survival of the individuals at the boundary individuals is
 
-\begin{equation} \label{eq:boundSurv} S_{\rm I} (x, a_0, a) =  \pi_1 (x^\prime,m_0, a_0)  \exp\left(- \int_{a_0}^{a} d(x,m_i(x, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right).\end{equation}
+\begin{equation} \label{eq:boundSurv} S_{\rm I} (x, a_0, a) =  \pi_1 (x^\prime,m_0, a_0)  \exp\left(- \int_{a_0}^{a} d(x,m_i(x, a_0 ,a^\prime), a^\prime) \, {\rm d} a^\prime \right).
+\end{equation}
 
@@
 
-### Density of individuals
+## Density of individuals
 
@@
 
-\begin{equation} \label{eq:boundN} \begin{array}{ll}  n(x, m, a) & =n(x,m_0 ,a_0) \times \\
-& \exp \left(-\int _{a_0}^{a} \left[\frac{\partial g(x,m(x, a_0, a^\prime),a^\prime)}{\partial m} +d(x,m(x, a_0, a^\prime),a^\prime)\right] {\rm d}a^\prime \right),\end{array} \end{equation}
+\begin{equation} \label{eq:boundN}
+  \begin{array}{ll}
+  n(x, m, a) & =n(x,m_0 ,a_0) \times \\
+  & \exp \left(-\int _{a_0}^{a} \left[\frac{\partial g(x,m(x, a_0, a^\prime),a^\prime)}{\partial m} +d(x,m(x, a_0, a^\prime),a^\prime)\right] {\rm d}a^\prime \right),
+  \end{array}
+\end{equation}
 
 where
-\begin{equation} \label{eq:boundN2}  n(x,m_0 ,a_0)  = \left\{ \begin{array}{ll}   \frac{y_x  \pi_1 (x^\prime,m_0, a_0) }{ g(x,m_0, a_0) }  & \textrm{if } g(x,m_0, a_0) > 0 \\
-0 & \textrm{otherwise.}  \end{array} \right. \end{equation}
+\begin{equation} \label{eq:boundN2}
+  n(x,m_0 ,a_0)  = \left\{
+  \begin{array}{ll}   \frac{y_x  \pi_1 (x^\prime,m_0, a_0) }{ g(x,m_0, a_0) }  & \textrm{if } g(x,m_0, a_0) > 0 \\
+  0 & \textrm{otherwise.}
+  \end{array} \right.
+\end{equation}
 
@@
 \begin{equation}\label{eq:boundN3}
-n(x,m,a)=\frac{y_x}{ g(x,m) } \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x , m^\prime) }{g(x, m^\prime)} \, \textrm{d}m^\prime \bigg). \end{equation}
-(for a proof, see \cite{de_roos_gentle_1997} page 153).
+	n(x,m,a)=\frac{y_x}{ g(x,m) } \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x , m^\prime) }{g(x, m^\prime)} \, \textrm{d}m^\prime \bigg).
+\end{equation}
+(for a proof, see [@Deroos-1997] page 153).
 Note also, that the integral on the RHS of $\ref{eq:boundN3}$ is simply the survival of individuals from $a_0 \rightarrow a$ describd in equation $\ref{eq:boundSurv}$
-\begin{equation} \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x, m^\prime)}{g(x, m^\prime)}  \textrm{d} m^\prime \bigg) = \exp \bigg(-\int _{a_{0} }^{a}d(x, m(a^\prime)) \textrm{d}{a^\prime} \bigg) = S_{\rm I} (x, a_0, a).\end{equation}
+\begin{equation} \exp \bigg(-\int _{m_{0} }^{m}\frac{d(x, m^\prime)}{g(x, m^\prime)}  \textrm{d} m^\prime \bigg) = \exp \bigg(-\int _{a_{0} }^{a}d(x, m(a^\prime)) \textrm{d}{a^\prime} \bigg) = S_{\rm I} (x, a_0, a).
+\end{equation}
 
@@
 \begin{equation}\label{eq:boundN4}
-n(x,m,a)=\frac{y_x}{ g(x,m) } S_{\rm I} (x, a_0, a). \end{equation}
+	n(x,m,a)=\frac{y_x}{ g(x,m) } S_{\rm I} (x, a_0, a).
+\end{equation}
 For some functions of $g$ and $d$, equation $\ref{eq:boundN4}$ might yield an analytical solution to density.
 
-### Seed production
+## Seed production
 
 The lifetime seed production of boundary individuals  is calculated according to Eq. $\ref{eq:tildeR}$ as
-\begin{equation}\begin{array}{ll}\tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\\end{array}\end{equation}
+\begin{equation}
+  \begin{array}{ll}
+  \tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\
+  \end{array}
+\end{equation}
 where $S_{\rm I}$ is individual survival (defined above) and
@@
 
-### Invasion fitness
+## Invasion fitness
 
 To scale up seed production for the metapopulation need to integrate Eq. $\ref{eq:boundR}$ across all possible seed-arrival times (as defined in Eq. \ref{eq:InvFit}):
-\begin{equation} R(x) = \int_0^{\infty}  p(a) \, \tilde{R}(x, a, \infty) \, {\rm d}a. \end{equation}
+\begin{equation} R(x) = \int_0^{\infty}  p(a) \, \tilde{R}(x, a, \infty) \, {\rm d}a.
+\end{equation}
 
@@
 
-### Environmental feedback
+## Environmental feedback
 
@@
 
-## Second method for approximating $n(x,m,a)$: based on modelling dynamics within cohorts
-
-The original EBT, derived by de Roos (1997, 1988), proceeded by approximating the first and second moments of the density distribution $n \left( x,m,a \right)$ within sub-domains $\Omega_i$, represented by $\lambda_i$ and $\mu_i$, these being the total number and mean size of individuals within $\Omega_i$ respectively.
-
-Under this assumption, the rates of change $\frac{\rm d} {{\rm d}a} \lambda_i$ and $\frac{\rm d} {{\rm d}a} \mu_i$ can be approximated given by two ODEs, which can be approximated by first-order closed-form solutions.
-These approximations form the centrepiece of the original EBT, and are derived as follows.
-
-### First moment: Number of individuals
-
-*Definition:*
-\begin{equation} \label{eq:lambda}  \lambda_i = \int_{\Omega_i} n(x,m,a) \; {\rm d}m \end{equation}
-
-\begin{equation} \label{eq:dlambda} \frac{{\rm d} }{{\rm d}a} \lambda_i = -\int_{\Omega_i} d(x,m,E) \; n(x,m,a) \; {\rm d}m \end{equation}
-
-*Numerical technique:* The rate of change in $\lambda_i$ is approximated by
-
-\begin{equation} \label{eq:dlambda_approx} \frac{{\rm d} }{{\rm d}a} \lambda_i =  d(x,\mu_i,E) \lambda_i \end{equation}, with $\lambda_i$ estimated by integrating this quantity with an ODE stepper.
-The approximation for \ref{eq:dlambda_approx} is only accurate to second order (for details on derivation see [appendix](#DeriveEBTapproximation)).
-
-### Second moment: Average size
-
-*Definition:*
-\begin{equation}  \label{eq:mu} \mu_i = \frac1{\lambda_i}\int_{\Omega_i} m \; n(x,m,a) \; {\rm d}m  \end{equation}
-
-\begin{equation} \label{eq:dmu} \frac{{\rm d} }{{\rm d}a} \mu_i = \frac1{\lambda_i} \int_{\Omega_i} \left[ g(x,m,E) - ( m-\mu_i ) d(x,m,a) \right] n(x,m,a) \; {\rm d}m  \end{equation}
-
-*Numerical technique:* The rate of change in $\mu_i$ is approximated by
-\begin{equation}  \label{eq:dmu_approx} \frac{{\rm d} }{{\rm d}a} \mu_i =  g(x,\mu_i,E), \end{equation}
-
-with $\mu_i$ estimated by integrating this quantity with an ODE stepper.
-The approximation for \ref{eq:dmu_approx} is only accurate to second order (for details on derivation see [appendix](#DeriveEBTapproximation)).
-
-### Density of individuals
-
-*Definition:* $n(x,m,a)$ for $m \in \Omega_i$
-
-*Numerical technique:* The EBT technique solves the PDE by approximating the changing size distribution $n(x,m,a)$ with a series of point masses with position and amplitude given by $\lambda_i$ and $\mu_i$.
-If we let $\delta(z)$ be a function for a point mass of size 1 localised at $z=0$, then
-
-\begin{equation} \label{eq:n_delta}  n(x,m,a) \approx = \sum_{i=1}^{k} \lambda_i(a) \; \delta(m-\mu_i(a)). \end{equation}
-
-The density function $n(x,m,a)$ within each sub-domain can also be estimated from its first moment using a discontinuous, piecewise-linear function:
-
-\begin{equation} n_i \left( x,m,a \right) \approx \frac{\lambda_i}{m_i-m_{i-1}}, \end{equation}
-
-although this requires that the position of the cohort boundaries are also known (see previous section for equations).
-
-### Seed production
-The lifetime seed production of individuals located at the cohort boundary is calculated according to Eq. $\ref{eq:tildeR}$ as
-\begin{equation}\label{eq:boundR}\begin{array}{ll}\tilde{R}(x, a_0, \infty) = \int_{a_0}^{\infty}  &\pi_0 \, f(x, m_i( a^\prime), a^\prime)\times\\ \, &S_{\rm I} (x, a_0, a^\prime) \, S_{\rm P} (a_0,a^\prime)  {\rm d} a^\prime,\\\end{array}\end{equation}
-where $S_{\rm I}$ is individual survival (defined above) and
-$S_{\rm P}$ is calculated as in Eq.
-$\ref{eq:survivalPatch}$.
-Using the EBT approximation, the total number of sees produced within a cohort including individuals born in the interval $(a_1, a_2)$ and up to age $a$ is given by
-
-\begin{equation} \int_{a_1}^{a_2} \tilde{R}(x, a_0, a) \; {\rm d}a_0 = f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) S_{\rm P} (a_0,a^\prime) \; {\rm d}a^\prime.  \end{equation}
-
-Thus the total number of seeds produced by individuals within the cohort is given by
-
-\begin{equation} \label{eq:cohR} \int_{a_1}^{\infty}  \int_{a_1}^{a_2} \tilde{R}(x, a_0, a) \; {\rm d}a_0 {\rm d}a = \int_{a_1}^{\infty} f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) \, S_{\rm P} (a_1,a^\prime).\end{equation}
-
-*Numerical technique:* To solve equation $\ref{eq:cohR}$ we need solve the IVP:
-$$\frac{dy}{dt} = f(x, \mu(x, a^\prime), a^\prime) \lambda(x, a^\prime) S_{\rm P} (a_1,a^\prime),$$
-$$ y(a_1) = 0.$$
-
-### Dynamics of the boundary cohort
-
-A slightly modified set of equations is needed to track the moments of the smallest cohort, whose lower bound is fixed at $m_0$.
-Specifically, a new quantity $\Pi$, representing the total accumulated mass within this boundary cohort, is monitored in place of the second moment, which would otherwise be undefined whenever $\lambda _i=0$.
-Our approximations for the boundary cohort differ slightly to those of de Roos (1997, 1988) in that we Taylor expand around the centre of mass for the cohort instead of the lower boundary.
-Because relative growth rates often decline quickly with increasing size, use of lower boundary can overestimate the growth rate of the cohort.
-
-#### Number of individuals
-
-*Definition:*
-\begin{equation} \lambda_1 = \int_{\Omega_1} n(x,m,a) \; {\rm d}m \end{equation}
-
-\begin{equation} \frac{{\rm d} }{{\rm d}a} \lambda_1 = y_x \pi_0 \pi_1(x,m_0, a) - \int_{\Omega_1} d(x,m,a) \; n(x,m,a) \; {\rm d}m  \end{equation}
-(TODO: I $\pi_0$ in right place here? Should appear in seed production, not inflow).
-
-*Numerical technique:* $\frac{{\rm d} }{{\rm d}a} \lambda_1 = y_x - d(x,\mu_1,a) \lambda_1$
-
-#### New mass
-
-*Definition:* $\Pi_1 = \int_{\Omega_1} (m-m_0) \; n(x,m,a) \; {\rm d}m$
-
-*Exact:* $\frac{{\rm d} }{{\rm d}a} \Pi_1 = \int_{\Omega_1} \left[ g(x,m,E) - ( m-\mu0) d(x,m,a) \right] n(x,m,a) \; {\rm d}m$
-
-*Numerical technique:* $\frac{{\rm d} }{{\rm d}a} \Pi_1 =
- g(x,\mu_1,a) \lambda_1 - \Pi_1  d(x,\mu_1,a)$
-
-#### Average size
-
-*Definition:*
-\begin{equation}  \mu_1 = \left\{ \begin{array}{ll} \frac1{\lambda_i}\int_{\Omega_i} m \; n(x,m,a) \; {\rm d}m   & \textrm{if } \lambda_1 > 0 \\
-m_0 & \textrm{otherwise}  \end{array} \right. \end{equation}
-
-\begin{equation} \mu_1 = \left\{ \begin{array}{ll}  m_0 + \frac{\Pi_1}{\lambda_1}  & \textrm{if } \lambda_1 > 0 \\
-m_0 & \textrm{otherwise}  \end{array} \right. \end{equation}
-
-*Numerical technique:* NA, use exact.
-
-### Environmental feedback
-
-The estimated density function (Eq. \ref{eq:n_delta}) is used to calculate the amount of shading on each individual from other individuals in the patch, as described in Eq. \ref{eq:light}.
-
-*Numerical technique:*  The approximation of  $n(x,m,a)$ by a series of delta peaks  in Eq. \ref {eq:n_delta} allows integration in Eq. \ref{eq:light} to be approximated as a sum over cohorts:
-
-\begin{equation} \label{eq:light_deltasum}
-E(z,a) = \exp \left(-c_{ext}  \sum_{n=1}^{N} \sum_{\Omega_i} \phi(\mu_{n,i}) \, L(z, h(\mu_{n,i})) \lambda_{n,i} \right), \end{equation}
-
-where $\mu_{n,i}, \lambda_{n,i}$ are the average size and number of individuals within the i$^{th}$ cohort of species $n$ (Eqs. \ref{eq:mu}, \label{eq:lambda} respectively).
-
-## Controlling error in the EBT
+# Controlling error in the EBT
 
@@
 However, there is no existing method to control error arising from poor spacing of cohorts.
-Poor cohort spacing introduces error both because the equations governing cohort dynamics are only accurate up to second order (Table 4-1), and because any integrals over the size distribution (e.g.
+Poor cohort spacing introduces error both because the equations governing cohort dynamics are only accurate up to second order, and because any integrals over the size distribution (e.g.
 total offspring production, competitive effect) may be poorly resolved.
@@
 Let $p(a,t)$ be the frequency-density of patches age $a$ at time $t$ and let $\gamma(a)$ be the age-dependent probability that a patch of age $a$ is transformed into a patch of age $0$ through disturbance.
-Then according to the Von Foerster equation for age-structured population dynamics (von_foerster_remarks_1959), the dynamics of $p(a,t)$ are given by
+Then according to the Von Foerster equation for age-structured population dynamics [@Vonfoerster-1959], the dynamics of $p(a,t)$ are given by
 $$ \frac{\partial }{\partial t} p(a,t)=-\frac{\partial }{\partial a} p(a,t)-\gamma(a,t)p(a,t),$$
@@
 
-For general notes on structured population and analysis of data see (Metz and Diekmann 1986, Taljapurkar and Caswell 1997, Diekmann et al. 1998, Diekmann et al. 2001)
-
 ### 1-dimensional case
@@
 The flux of individuals in and out of the size class can be expressed as
-\begin{equation}\begin{array}{ll} J(m,a)=&g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)-g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a) \\ &-d (m,a) \; n(m,a)\Delta m\\ \end{array}.\end{equation}
+\begin{equation}\begin{array}{ll} J(m,a)=&g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)-g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a) \\ &-d (m,a) \; n(m,a)\Delta m\\ \end{array}.
+\end{equation}
 The first two terms describe the flux in and out of the size class through growth; the last term describes lossed due through mortality.
 The change in number of individuals within the interval across a time step \textit{$\Delta $t} is thus:
-\begin{equation}\begin{array}{ll} n(m,a+\Delta a)\Delta m-n(m,a)\Delta m= &g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)\Delta a \\ &-g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a)\Delta a\\&-d (m,a) \; n(m,a)\Delta m\Delta a.
-\end{array}\end{equation}
+\begin{equation}
+  \begin{array}{ll} n(m,a+\Delta a)\Delta m-n(m,a)\Delta m= &g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)\Delta a \\ &-g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a)\Delta a\\&-d (m,a) \; n(m,a)\Delta m\Delta a.
+  \end{array}
+\end{equation}
 Rearranging,
-\begin{equation}\begin{array}{ll} \frac{n(m,a+\Delta a)-n(m,a)}{\Delta a} = &-d (m,a) \; n(m,a) \\
-&-\frac{g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a)-g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)}{\Delta m}.\end{array}\end{equation}
+\begin{equation}
+  \begin{array}{ll}
+  \frac{n(m,a+\Delta a)-n(m,a)}{\Delta a} = &-d (m,a) \; n(m,a) \\
+  &-\frac{g(m+0.5 \Delta m,a) \; n(m+0.5 \Delta m,a)-g(m-0.5 \Delta m,a) \; n(m-0.5 \Delta m,a)}{\Delta m}.
+  \end{array}
+\end{equation}
 The LHS is corresponds to the derivative of $n$ as $\Delta a\to 0$. For thin slices, $\Delta m\sim 0$, we obtain
@@
 
-\begin{equation}\frac{\partial }{\partial t} \int _{m_{0} }^{m_{\infty } }n(m,a) \partial m=g(m_{0} ,a) \; n(m_{0} ,a)-g(m_{\infty } ,a) \; n(m_{\infty } ,a)-\int _{m_{0} }^{m_{\infty } }d (m,a) \; n(m,a) \partial m.\end{equation}
+\begin{equation}\frac{\partial }{\partial t} \int _{m_{0} }^{m_{\infty } }n(m,a) \partial m=g(m_{0} ,a) \; n(m_{0} ,a)-g(m_{\infty } ,a) \; n(m_{\infty } ,a)-\int _{m_{0} }^{m_{\infty } }d (m,a) \; n(m,a) \partial m.
+\end{equation}
 
@@
 Thus the boundary condition is given by
-\begin{equation}g(m_{0} ,a) \; n(m_{0} ,a)=B(x, n, E,a).\end{equation}
+\begin{equation}g(m_{0} ,a) \; n(m_{0} ,a)=B(x, n, E,a).
+\end{equation}
 
@@
 The flux of individuals in and out of the size class can be expressed as
-\begin{equation}\begin{array}{ll} J(m_1,m_2,a)=
+\begin{equation}
+  \begin{array}{ll}
+  J(m_1,m_2,a)=
   &+g(m_1-0.5 \Delta m_1,m_2,a) \; n(m_1-0.5 \Delta m_1,m_2,a)\Delta m_2\\
@@
   The change in number of individuals within the interval across a time step \textit{$\Delta $t} is then:
-\begin{equation}\begin{array}{ll} n(m_1,m_2,a+\Delta a)\Delta m_1\Delta m_2&-n(m_1,m_2,a)\Delta m_1\Delta m_2=\\
+\begin{equation}
+  \begin{array}{ll}
+  n(m_1,m_2,a+\Delta a)\Delta m_1\Delta m_2&-n(m_1,m_2,a)\Delta m_1\Delta m_2=\\
   &+g(m_1-0.5 \Delta m_1,m_2,a) \; n(m_1-0.5 \Delta m_1,m_2,a)\Delta m_2 \Delta a\\
@@
   &-g(m_1,m_2+0.5 \Delta m_2,a) \; n(m_1,m_2+0.5 \Delta m_2,a)\Delta m_1 \Delta a\\
-&-d(m_1,m_2,a) \; n(m_1,m_2,a)\Delta m_1 \Delta m_2 \Delta a. \end{array}\end{equation}
-
-Rearranging and taking the limits as $\Delta a, \Delta m_1, \Delta m_2 \to 0$ yields the PDE for the 2D system
-\begin{equation}\begin{array}{ll} \frac{\partial }{\partial t} n(m_1,m_2,a)=&-\frac{\partial }{\partial m_1} (g(m_1,m_2,a) \; n(m_1,m_2,a)) -\frac{\partial }{\partial m_2} (g(m_1,m_2,a) \; n(m_1,m_2,a))
-\\&-d (m_1,m_2,a) \; n(m_1,m_2,a).\end{array}\end{equation}
-
-\newpage
-
-## Exact solutions: Integration along characteristics {#IntegrationAlongCharacteristics}
-
-Characteristics are the projections of an individual's trajectory in i-space and time, provided they do not die.
-By integrating along characteristics, the PDE can be solved to yield the a general solution for the density of individuals with time of birth $t_{b} > 0$:
-\begin{equation}\label{eq:n1}
-n(m,t)=n(m_0 , t_b) \exp \bigg(-\int _{t_b}^{t} \left[\frac{\partial g(m(\theta ),E(\theta ))}{\partial m} +d(m(\theta ),E(\theta ))\right] d\theta \bigg)
-\end{equation}
-
-This states that the density $n$ at a specific time is a product of density at the origin adjusted for changes through development and mortality.
-Using the second boundary condition above, $n(m_{0} ,t_{g})$ can be replaced by $B(x, n, E,t_{b}) / g(m_{0} ,E(t_{b}))$.
-Note that density decreases through time because of mortality (as per typical survival equation), but that growth also contributes to changes in density.
-If growth is slowing with size, (i.e.
-$\frac{\partial g(m(\theta ),E(\theta ))}{\partial m}  <0$), then density increases because the characteristics converge on one another.
-Eq.
-\ref{eq:n1} can be solved (numerically or analytically) to give density with full ecological feedback and in changing environemnts.
-\\
-
-In cases where the environment is constant and time-independent ($E(t)=E_c$), it is possible to further simplify Eq. \ref{eq:n1} to give (see \cite{de_roos_gentle_1997} page 153 for proof):
-
-\begin{equation}\label{eq:n2}
-n(m,t)=\frac{B(x, n, E_c,t_{b})}{g(m,E_{c} )} \exp \bigg(-\int _{m_{b} }^{m}\frac{d(\xi ,E_{c} )}{g(\xi ,E_{c} )}  d\xi \bigg). \end{equation}
-
-Equation \ref{eq:n2} can be used to obtain in cases where environment is constant, either externally imposed in which case an analytic solution may be possible or when Eq.
-\ref{eq:PDE-app} is at equilibrium $\bigg(\frac{\partial n(m, t)}{\partial t} =0\bigg)$.
-It is necessary to know the origin of the characteristic for individuals of size $m$ at time $t$ to model their density.
-
-To this end, it is useful to specify a function $\tau(m ,m_{b}, Ec )$ for the age of an individual of size $m$ such that $t_{b} =t-\tau (m,m_{b} ,E_{c} )$.
-Note also, $\exp \bigg(-\int _{m_{b} }^{m}\frac{d(\xi ,E_{c} )}{g(\xi ,E_{c} )}  d\xi \bigg)$ is equal to the overall survival to size $m$, i.e.
-\begin{equation}S(m(t))= \exp \bigg(-\int _{m_{b} }^{m}\frac{d(\xi ,E_{c} )}{g(\xi ,E_{c} )}  d\xi \bigg) = \exp \bigg(-\int _{t_{b} }^{t}d(m(\tau) ,E_{c} ) \; {\rm d}\tau \bigg).\end{equation}
-
-Using equation \ref{eq:n2} it is also possible to show that in populations at demographic equilibrium and without disturbance, the lifetime seed output of an average individual should be 1.0.
-The seed rain for a community is given by
-\begin{equation}R = \int_{m_{b}}^{m_{max}} f(m, E_{c}) \; n(m)\, dm.\end{equation}
-Substituting from above we have
-\begin{equation}R =  B(x, n, E_c,t_{b})  \int_{m_{b}}^{m_{max}} \frac{f(m, E_{c})}{g(m,E_{c} )} S(m) \, dm.\end{equation}
-Note that $m(t) = m_{0} + \int_{t_{b}}^{t} g(m(\tau))\,{\rm d}\tau$ and $dm(t) = g(m(t)) dt$. This allows us to rexpress the integral above as
-\begin{equation}R =  B(x, n, E_c,t_{b})  \int_{t_{b}}^{t_{max}} f(m(t), E_{c}) \; S(m(t)) \, dt.\end{equation}
-The term in the integral gives the lifetime seed production of a single individual.  At demographic equilibrium $R =  B(x, n, E_c,t_{b})$ (seed rain = seed production), so
-\begin{equation} \int_{t_{b}}^{t_{max}} f(m(t), E_{c}) \; S(m(t)) \, dt =1\end{equation}.
-
-\newpage
-
-## Method for deriving cohort approximations in the EBT {#DeriveEBTapproximation}
-
-Given a function $\phi  \left( m \right)$, the integral $\int_{m_{i-1}}^{m_i}\phi  \left( m \right) n \left( x,m,a \right) \; {\rm d}m$ can be approximated by first Taylor expanding $\phi  \left( m \right)$ around $\mu_i$.
-Substituting into the integral, we obtain
-
-\begin{equation} \label{eq:taylor1}
-	\begin{array}{l}
-		{\int_{m_{i-1}}^{m_i}\phi  \left( m \right) n \left( x,m,a \right) \; {\rm d}m } \\
-		{{\rm \; \; \; \; \; \; \; \; \; \; \; \; \; }= \int_{m_{i-1}}^{m_i}\left[\phi  \left( \mu_i \right) +\phi ' \left( \mu_i \right)  \left( m-\mu_i \right) +\frac{\phi '' \left( \mu_i \right) }2  \left( m-\mu_i \right) ^2 +...\right]n \left( x,m,a \right) \; {\rm d}m } \\
-		{{\rm \; \; \; \; \; \; \; \; \; \; \; \; \; =}\phi  \left( \mu_i \right) \; {\rm \; }\lambda_i+\int_{m_{i-1}}^{m_i}\frac{\phi '' \left( \mu_i \right) }2  \left( m-\mu_i \right) ^2 n \left( x,m,a \right) \; {\rm d}m +...}
+  &-d(m_1,m_2,a) \; n(m_1,m_2,a)\Delta m_1 \Delta m_2 \Delta a.
   \end{array}
@@
 
-Discarding the square and higher terms, gives the following first-order approximation,
-
-\begin{equation} \label{eq:taylor2}
-	\int_{m_{i-1}}^{m_i}\phi  \left( m \right) n \left( x,m,a \right) \; {\rm d}m {\rm =}\phi  \left( \mu_i \right) \; {\rm \; }\lambda_i,
-\end{equation}
-
-with an estimate for the truncation error of Eq. $\ref{eq:taylor2}$ given by the first discarded term from Eq. $\ref{eq:taylor1}$:
-
-\begin{equation} \label{eq:error1}
-	\epsilon_i = 0.5 \phi''(\mu_i)\int_{m_{i-1}}^{m_i} (m-\mu_i)^2n(x,m, a) \; {\rm d}m.
+Rearranging and taking the limits as $\Delta a, \Delta m_1, \Delta m_2 \to 0$ yields the PDE for the 2D system
+\begin{equation}
+  \begin{array}{ll} \frac{\partial }{\partial t} n(m_1,m_2,a)=&-\frac{\partial }{\partial m_1} (g(m_1,m_2,a) \; n(m_1,m_2,a)) -\frac{\partial }{\partial m_2} (g(m_1,m_2,a) \; n(m_1,m_2,a)) \\
+  &-d (m_1,m_2,a) \; n(m_1,m_2,a).
+  \end{array}
 \end{equation}
@@
 
+## Converting density from one size unit to another
+
+Population density is explicitly modelled in relation to a given size unit (Eq.
+$\ref{eq:PDE}$).
+But what if we want to express density in relation to another size unit?
+A relation between the two can be derived by noting that the total number of individuals within a given size range must be equal.
+So let's say density is expressed in units of size $m$, but we want density in units of size $h$.
+First we require a one-to-one function which h for a given $m$: $h = \hat{h}(m)$.
+Then the following must hold
+\begin{equation} \label{eq:n_conversion} \int_{m_1}^{m_2} n(x,m,a) \; \textrm{d}m =  \int_{\hat{h}(m_1)}^{\hat{h}(m_2)} n^\prime(x,m,a) \; \textrm{d}h
+\end{equation}
+
+For very small size intervals, this equation is equivalent to
+\begin{equation} \left(m_2- m_1 \right) \; n(x,m_1,a) = \left( \hat{h}(m_2) - \hat{h}(m_1)\right) \; n^\prime(x, \hat{h}(m_1),a).
+\end{equation}
+
+Rearranging gives
+\begin{equation}  n^\prime(x, \hat{h}(m_1),a) = n(x, m_1,a) \; \frac{m_2- m_1}{\hat{h}(m_2) - \hat{h}(m_1)}
+\end{equation}
+
+Noting that the second term on the RHS is simply the definition of $\frac{\delta m}{\delta h}$ evaluated at $m_1$, we have
+\begin{equation} \label{eq:n_conversion2} n^\prime(x, h, a) = n(x, m,a) \; \frac{\delta m}{\delta h}.
+\end{equation}
+
 ## Expressing integrals as an ODE {#writeIntegralAsOdeIvp}
 
-There exist different techniques for numerically integrating a function  \cite{press_numerical_1995}.
+There exist different techniques for numerically integrating a function  [@Press-1995].
 Adaptive techniques are of particular interest because they allow resolution to increased only where it is needed, thereby greatly reducing the amount of computational effort required for given accuracy.
@@
 Second class includes ODE problems, but worth noting that ODE methods can be applied to any integral, as
-\begin{equation}I=\int_{a}^{b}f(x) \; {\rm d}x\end{equation}
-is equivalent to solving for the value $I\equiv y(b)$ the differential equation $\frac{dy}{dx}=f(x)$ with boundary  condition $y(a)=0.$ \cite{press_numerical_1995}.
+\begin{equation}
+  I=\int_{a}^{b}f(x) \; {\rm d}x
+\end{equation}
+is equivalent to solving for the value $I\equiv y(b)$ the differential equation $\frac{dy}{dx}=f(x)$ with boundary  condition $y(a)=0.$ [@Press-1995].
 
@@
 The formula for total photosynthesis for a plant size $m$ as described in the model is given by
-\begin{equation} \frac{m}{\phi}  \int_{0}^{h(m)}A(I(z)) \, l\left(\frac{z}{h(m)}\right) \, dz,\end{equation}
+\begin{equation}
+  \frac{m}{\phi}  \int_{0}^{h(m)}A(I(z)) \, l\left(\frac{z}{h(m)}\right) \, dz,
+\end{equation}
 where $x\in [0,1]$ is the relative height in the canopy (1 gives top) and $l(x)$ is the density of leaf area at height $x$.
@@
 Substituting into the above integral for total photosynthesis, we obtain
-\begin{equation} \frac{m}{\phi}  \int_{0}^1A\big(I(h(m)L^{-1}(u))\big) \, du.\end{equation}
+\begin{equation}
+  \frac{m}{\phi}  \int_{0}^1A\big(I(h(m)L^{-1}(u))\big) \, du.
+\end{equation}
 
 The formula for trapezoidal integration with unevenly grid is:
-\begin{equation} \label{trap_uneven} \int_{x_1}^{x_{n}}f(x)dx \approx 0.5 \sum_{i=0}^{i=N-1} \bigg[ \left(f(x_{i}) + f(x_{i+1})\right)\left( x_{i+1}- x_{i}\right)\bigg].\end{equation}
+\begin{equation} \label{trap_uneven}
+  \int_{x_1}^{x_{n}}f(x)dx \approx 0.5 \sum_{i=0}^{i=N-1} \bigg[ \left(f(x_{i}) + f(x_{i+1})\right)\left( x_{i+1}- x_{i}\right)\bigg].
+\end{equation}
 
 For regularly spaced intervals we have the simplified version
-\begin{equation} \label{trap_even} \int_{x_1}^{x_{n}}f(x)dx \approx \Delta x \bigg[0.5 f_{0}+f_2+\ldots+ f_{N-1}+0.5 f_{N}\bigg] .\end{equation}
+\begin{equation} \label{trap_even}
+  \int_{x_1}^{x_{n}}f(x)dx \approx \Delta x \bigg[0.5 f_{0}+f_2+\ldots+ f_{N-1}+0.5 f_{N}\bigg] .
+\end{equation}

### 2015-05-14 tree_demography.Rmd
Section on metapopulation and fitness

@@
 
-Let us now consider how we can estimate the fitness of a rare individual with trats $x^\prime$ growing in the environment of a resident community with traits $x$.
+Let us now consider how we can estimate the fitness of a rare individual with traits $x^\prime$ growing in the environment of a resident community with traits $x$.
 We will focus on the phenotypic components of fitness -- i.e.

### 2015-05-14 R/figures_plant.R
Draft version of plant level figure

@@ -1,0 +1,129 @@
+## Eventually, this will be part of a vignette showing how to drive
+## `tree` at the plant level.  Unfortunately, there's a lot of
+## book-keeping to keep track of, so it's not that nice at this
+## point.
+
+## First, growth trajectories of plants
+default_parameters <- function() {
+  p <- ebt_base_parameters()
+  p$strategy_default <- FFW16_Strategy(hmat = 30.0)
+  p
+}
+
+## This ensures that the plants are up-to-date with the environment
+## (that's not guaranteed by the running function) and extracts a ton
+## of data about them.
+extract_plant_info <- function(p, env) {
+  p$compute_vars_phys(env)
+  p$compute_vars_growth()
+  x <- unlist(p$internals)
+
+  ## Relative growth methods (= d(y)/dt / y)
+  v <- c("height", "area_stem", "diameter_stem", "mass_above_ground")
+  x[sprintf("%s_dt_relative", v)] <- x[sprintf("%s_dt", v)] / x[v]
+  x
+}
+
+## Small wrapper around grow_plant_to_size that collects information
+## about plants.
+run_plant_to_heights <- function(heights, strategy, env,
+                                 time_max=100) {
+  res <- grow_plant_to_size(FFW16_PlantPlus(strategy),
+                            heights, "height", env,
+                            time_max, warn=FALSE, filter=TRUE)
+  res$info <- do.call("rbind", lapply(res$plant, extract_plant_info, env))
+  res
+}
+
+## Given vectors of heights and openness, lets do it all.
+height_dt_openness <- function(height, openness, strategy) {
+  pl <- FFW16_PlantPlus(strategy)
+  pl$height <- height
+  min <- lcp_whole_plant(pl)
+  openness <- c(min, openness[openness > min])
+  f <- function(x) {
+    env <- fixed_environment(x)
+    pl$compute_vars_phys(env)
+    pl$internals$height_dt
+  }
+  cbind(openness=openness,
+        height_dt=sapply(openness, f))
+}
+
+figure_plant <- function() {
[…]

### 2015-05-15 R/figures.R
Rework text based on comments

@@ -1,6 +1,0 @@
-
-empty_box <- function() {
-  par(oma=c(0,0,0,0), mar=rep(0.1,4))
-  plot(1,1, ann=FALSE, axes=FALSE, type='n')
-  box()
-}

### 2015-05-15 R/figures_fitness.R

@@ -61,7 +61,7 @@ figure_fitness_landscape <- function(data) {
   plot(lma[i], w$second1[i], type="l", lty=2,
        log="x", xlim=xlim, ylim=ylim, ylab="Fitness", las=1)
   label_panel(2)
-  mtext("Leaf mass per unit area (lma; g / m2)", 1, 3, cex=1)
+  mtext("Leaf mass per unit area (LMA; kg / m2)", 1, 3, cex=1)
   lines(lma[i], w$second2[i], lty=3)
   lines(lma[i], w$second3[i], lty=3)
   lines(lma, w$attractor2)

### 2015-05-15 R/utils.R

@@ -1,3 +1,9 @@
+empty_box <- function() {
+  par(oma=c(0,0,0,0), mar=rep(0.1,4))
+  plot(1,1, ann=FALSE, axes=FALSE, type='n')
+  box()
+}
+
 label <- function(text, px=0.03, py=NULL, ..., adj=c(0, 1)) {
   if (is.null(py)) {
     fin <- par("fin")

### 2015-05-19 ms.tex
Swicth to latex

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{graphicx}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[verbose,tmargin=2cm,bmargin=2cm,lmargin=2.5cm,rmargin=2.5cm]{geometry}

\usepackage{color}
\definecolor{grey}{rgb}{0.5, 0.5, 0.5}

\usepackage[unicode=true]{hyperref}
\urlstyle{same}  % don't use monospace font for urls
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Daniel Falster, Rich FitzJohn, Mark Westoby},
            pdftitle={TREE: A package for modelling plant TRait Ecology and Evolution},
            colorlinks=true,
            citecolor=grey,
            urlcolor=grey,
            linkcolor=grey,
            pdfborder={0 0 0}}

\usepackage{lineno}
\linespread{2}

\usepackage[colorinlistoftodos]{todonotes}

\usepackage{authblk}
\title{TREE: A package for modelling plant TRait Ecology}
\author[1]{Daniel S. Falster}
\author[1]{Richard G. FitzJohn}
\author[1]{Mark Westoby}
\affil[1]{Dept of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}
\date{}

\usepackage{natbib}
\bibliographystyle{mee}

\usepackage[title,titletoc,toc]{appendix}

\begin{document}
\maketitle

A manuscript in consideration as a research paper for publication in MEE
as part of the Special Feature Demography beyond the Population

\listoftodos

\linenumbers

\section{Abstract}\label{abstract}

\emph{(max 350 words; rework into 4 points)}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Population dynamics in forests are strongly size-structured. Larger
  plants shade smaller plants. They also expend proportionately more
  energy on their woody stems. Although the importance of size-
  structure for demography is widely recognised, many mechanistic models
  either omit it entirely, or include only coarse approximations.
\item
  Here we introduce the TREE package (TRait Ecology and Evolution), an
  extensible framework for modelling plant demography across the entire
  life cycle via coupled differential equations. At its core, TREE is an
  individual-based mechanistic model where plant function is mediated by
  traits, Individuals from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
  disturbance regime. Dynamics within patches of competing plants are
  resolved using novel extensions of the Escalator Boxcar Train
  technique. Combined effects of trait-, size- and patch-structured
  dynamics are integrated into population level estimates of
  reproductive fitness.
\item
  TREE is presented as an open source \texttt{R} package and is
  available at
  \href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
  While accessed from R, the core routines in TREE are written in C++.
  The package provides for alternative physiologies and for
  hyper-parameterisation on the basis of plant functional traits. A
  detailed test suite is provided to ensure accuracy.
\item
  We provide worked examples illustrating TREE's use to study the
  influence of functional traits on growth of individual plants, of
  whole stands and on assembly of ecological communities.
\end{enumerate}

\textbf{Key-words}: demography, emergent property, fitness, growth,
physiology, metapopulation, mortality, reproduction, size-structure,
trade-off

\section{Introduction}\label{introduction}

Plant growth and demography are fundamentally structured by species
traits and by individual size. Both influence dynamics over time-scales
ranging from instantaneous physiological effects to long-term
evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As an
individual increases its leaf area, its potential to generate
photosynthate also rises. But for larger individuals, increasing
fractions of photosynthetic income must also be diverted towards
building and maintaining support tissues rather than leaves
\citep{Givnish-1988, Enquist-1999}. A larger fraction of income is also
allocated to reproduction \citep{Thomas-2011}. Consequently at
population level, rates of growth and mortality change with size
\citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
a species also interact with size to shape its demography. Leaf
construction costs and wood density affect growth rates
\citep{Wright-2010}. Other traits such as seed size and height at
maturation determine start and end points of ontogenetic trajectories
\citep{Westoby-2002}. Over longer time-frames, natural selection favours
traits making plants more competitive such as faster growth, larger
seeds, taller height \citep{Falster-2003}. Selective forces thus
potentially amplify size-related effects on demography. Accounting for
the effects of size and traits in models therefore seems like a key step
in developing theory for the structure and dynamics of the biosphere
\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular notable
phenomena such as successional dynamics, self thinning, trait evolution
and species coexistence.

Although the importance of size-structure for vegetation function and
demography has been widely recognised for a long time
\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001, Kohyama-1993, Enquist-1999, Pacala-1996, Coomes-2007},
most large-scale mechanistic models of plant growth either omit it
entirely, or include only coarse approximations
\citep{Cramer-2001, Sitch-2003}. The likely reason is the computational
challenge of modelling size-structured dynamics. During the 1980s and
early 1990s, size-structured ``gap'' models were widely used
\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
parameterised in terms of growth and demography, and consequently did a
poor job of simulating carbon fluxes. The next generation of models
focussed more on physiology and, presumably for computational reasons,
sacrificed details of demography in order to achieve global coverage
\citep[eg.][]{Haxeltine-1996}. This history has brought us to the
current perplexing situation where all but a couple
\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the leading vegetation
models lack size-structure
\citep{Cramer-2001, Dekauwe-2014, Sitch-2003, Kelley-2013}. These models
are focused on modelling fluxes carbon and water, but say little about
demographic behaviour, and consequently are also unable to account for
any non-linear demographic feedbacks that may arise.

Interestingly, most models used in theoretical ecology to explore
questions about species coexistence also lack size-structured dynamics
\citep[e.g.][]{MacArthur-1967, Levin-1974, Leimar-2013, Geritz-1995, Calcagno-2006, Tilman-1985}.
It is common to assume competitive interactions are influenced by
size-related traits such as adult and seed size, but detailed
size-structured demography is rarely considered, at least for plant
models \citetext{\citealp[for animal
examples,][]{Deroos-1988}; \citealp[see][]{Deroos-1992}}. Thus it has
been difficult to connect mechanisms described in these simplistic
models with field data and real traits, which are almost always strongly
size-structured.

In this note, we describe the \texttt{TREE} package for R
\citep{R-2015}, a mechanistic framework for studying the effects of size
structure and trait variation on the demography of individual plants, of
patches of competing plants, and of meta-populations structured by a
prevailing disturbance regime. The package came about as a way to extend
previous work investigating the effect of traits on vegetation
properties \citep{Falster-2011} and the mechanisms facilitating
coexistence of trait mixtures \citep{Falster-2015}. These studies
provided the foundation methods needed; here we provide robust and
extensible implementations. We describe the general approach of the
package and then a series of use cases, focussing at different levels of
ecological organisation. Each section provides a short description of
the required methods, design considerations, and then some worked
examples illustrating potential use cases and design features. Detailed
technical documents are provided as supplementary information, and updated
versions of technical documents are also available within the TREE
package itself. The use cases highlighted reflect our own interests in
understanding trait-based demography and community assembly. Yet we
expect the methods provided to be useful in other contexts. For example,
prior to TREE there was no R package that allowed one to simply grow a
plant over time, based on well-understood physiology. Surely this is one
of the simplest demographic actions one might hope for, with diverse
potential applications in ecology, evolution, forestry, agriculture, and
vegetation modelling. We hope the mechanistic platform provided here
will complement the increasingly sophisticated statistical methods now
available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.

\section{The methods}\label{the-methods}

TREE is a mechanistic model, meaning the dynamics of the system arise
from rules specifying on how individuals grow and interact. In ecology,
there is a long history of using simple deterministic models \citep[such
as Lotka-Volterra population dynamics,
e.g.][]{MacArthur-1967, Leimar-2013} to understand biological phenomena.
TREE was developed with this style of analysis in mind, while also
allowing for a richer set of ecological dynamics than is possible in
population models lacking size structure. The package implements methods
for physiological, population, and adaptive dynamics (Fig.
\ref{fig:schematic}) following methods described by \citet{Falster-2011}
and \citet{Falster-2015}. The core rules in TREE are about the
short-term physiological functioning of an individual plant and how
these are influenced by its traits, size and light environment. Dynamics
at higher levels of organisation then arise as emergent properties,
driven by growth physiology, competition for light and disturbance (Fig.
\ref{fig:schematic}). Demographic phenomena can be studied at the three
levels of individual plants, stands of competing plants, and entire
metapopulations.

\section{Effects of size, trait and environment on demography of
individual
plants}\label{effects-of-size-trait-and-environment-on-demography-of-individual-plants}

The core of TREE is a model for a plant's physiological strategy as
specified by its traits (Fig. \ref{fig:schematic}a). This sub-model
estimates rate of biomass production for a plant, given its size, light
environment, and the supplied physiological constants. Assimilation is
estimated from total leaf area and the light distribution across the
plant's canopy. Costs of tissue respiration and turnover are then
subtracted. The remaining biomass is then allocated between growth and
reproduction. The strategy model thus includes a list of physiological
rules and associated parameters.

The core job of the physiological model is to take size, light
environment and parameters as inputs, and to return growth rate,
mortality and fecundity as outputs, in other words to move from
individual growth to demography. However additional quantities are also
available as output, such as total assimilation and the respiration,
turnover, and allocation rates for different tissues.

The default physiological model used in TREE largely reflects that
presented in Falster-2011, but with extensions allowing for diameter
growth also to be estimated. To properly model trait variation,
different parameters in the model are linked via assumed trade-offs. For
example, the trait LMA is used to estimate the rate of leaf turnover,
based on widely observed scaling relationship from \citet{Wright-2004}.

\subsection{Design features}\label{design-features}

TREE includes a default physiological model (fully described and derived
in the Appendix \ref{sec:FFW16}). However a feature of the package is that alternative
physiologies can be used, and the parameters of each model can be
altered. Trait differences are accounted for by altering relevant
parameters. This example shows how to establish a new plant with height
of 5 m, and estimate its physiological and demographic rates in the
given light environment.

\begin{verbatim}
# Example illustrating use of individual plant, all the outputs from plant internals
\end{verbatim}

\subsection{Use cases and examples}\label{use-cases-and-examples}

\todo[inline]{these aren't demographic rates - they're
physiological perphaps? Height growth rate, allocation, WPLCP, etc are
properties of individuals here. I've updated but we can discuss.}
\todo[inline]{allocation to replacement costs are not shown in
the figure at all, so this section is quite confused. I've reworded but
you'll want to go through and tweak. The graph shows ``fraction of live
mass that is X'' which is different to what is talked about.}

With above functionality, TREE can be used to estimate essential
physiological rates for individual plants (Fig. \ref{fig:plant}). The
function \texttt{grow\_plant\_to\_size} takes a given strategy and light
environment and grows the plant, producing a trajectory of size over
time (Fig. \ref{fig:plant}a). This is achieved by integrating an
Ordinary Differential Equation (ODE) of size-dependent growth rate (see
Appendix \ref{sec:demography} for more
details). As the plant grows, allocation to different types varies, and
the composition of the plant changes to become more stemmy and less
leafy (Fig. \ref{fig:plant}b). This in turn affects maintenance and
turnover costs and the growth rate of plants (Fig. \ref{fig:plant}c).
This shift alone generates a distinctive hump-shaped pattern of absolute
growth with size, and equally the widely regarded decline in relative
growth rate with size from birth onwards.

By varying the light environment and measuring growth rate, the whole
plant light compensation point (WPLCP) can be computed (Fig.
\ref{fig:plant}d). The WPLCP is increasingly regarded as the most useful
measure of a strategy's shade tolerance
\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, WPLCP
increases with plant size, due to increased costs of building and
maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, WPLCP
decreases with LMA, because high LMA species have lower leaf turnover
\citep{Baltzer-2007, Lusk-2013}.

\section{Plants competing in a patch}\label{plants-competing-in-a-patch}

Within patches of competing plants, competition for light generates
strong non-linear feedbacks on growth, survival and reproduction. In the
\texttt{FFW16} physiological model, we consider only the effect of
shading on rates of biomass production, although competition for other
resources such as nitrogen or water could also be considered with
suitable extensions.

Modelling the development of a competitive hierarchy within a patch
requires that both the initial size distribution and the inflow of new
recruits be specified. We have primarily been interested in dynamics of
a patch recovering from disturbance, and have therefore started with an
empty patch and constant flow of seeds from a global seed rain (Fig.
\ref{fig:schematic}b).

\subsection{Design features}\label{design-features-1}

When modelling a patch, we are interested in modelling changes in the
size-density-distribution \(n(h,a)\) over time, i.e.~the density of
individuals with height \(h\) within a patch of age \(a\). We assume
that patches are large, and are vertically but not horizontally
structured. Thus, shading in TREE is modelled as a mean field with
respect to spatial layout. Similar assumptions are found in many models
simulating size-structured dynamics
\citep{Moorcroft-2001, Huston-1987, Smith-2014}. Under these
assumptions, the dynamics of \(n\) behave deterministically and can be
approximated via a Partial Differential Equation (PDE) (see Appendix
\ref{sec:demography}). The same PDE has been shown to capture average behaviour
across a large number of small patches \citep{Moorcroft-2001}. Ideally
one would also consider full spatial interactions within patches,
however such models are very computationally demanding, and also
impossible to solve in a deterministic fashion \citep{Pacala-1996}.
Moreover, it remains unclear whether resolving spatial details within a
patch would provide further ecological detail, in part because the
process of competitive thinning tends to break down spatial clusters
\citep{Strigul-2008}.

Our approach for modelling solving size-structured population dynamics
is based on the Escalator Boxcar Train technique (EBT)
\citep{Deroos-1997, Deroos-1992, Deroos-1988}. The EBT solves the PDE
describing development of \(n\) by approximating the density function
with a collection of cohorts spanning the size spectrum. Following a
disturbance, a series of cohorts are introduced into each patch. These
cohorts are then transported along the characteristic equations of the
PDE. Biologically these are the growth trajectories of individuals,
conditioned on them surviving. Fig. \ref{fig:patch}a illustrates the
trajectories of individuals within a patch recovering from disturbance.
Fig. \ref{fig:patch}b shows the same trajectories, but with shading
indicating the density of plants at that size. (see also Fig.
\ref{fig:schematic}b).

We introduce two extensions to the EBT: i) a new method for estimating
the size-density distribution, and ii) a technique for handling strong
size-asymmetric competitive feedbacks, such as occur under strong
competition for light. The original EBT
\citep{Deroos-1997, Deroos-1992, Deroos-1988} proceeds by approximating
the first and second moments of the density distribution \(n\) within
each cohort. In TREE we use a different but from a theoretical
perspective equally valid approach. Instead of tracking first and second
moments of the size-distribution within cohorts, we track a point mass
estimate of \(n\) along the trajectories corresponding to the boundary
of each cohort. We found this approach preferable because it does not
require us to maintain a separate set of equations for the smallest
cohort \citep{Deroos-1997}.

The extension for handling strong size-asymmetric competition involves
adaptive refining of time points at which new cohorts are introduced
into the population. Under competition, the growth trajectories of
individuals born at similar times can diverge substantially over time
(Fig. \ref{fig:patch}a). This is can become a problem later during patch
development because much of the population can be located between
widely-spaced cohorts, delivering low numerical precision. To maintain
numerical precision we use an iterative algorithm to adaptively refine
the cohort spacing until the trajectories of adjacent cohorts remain
adequately resolved throughout development of the patch (Fig.
\ref{fig:characteristics}b; see Appendix\ref{sec:demography} for details).

\subsection{Use cases and examples}\label{use-cases-and-examples-1}

Modelling of size-structured dynamics via solving of a deterministic PDE
enables multiple complex non-linear demographic phenomena to be
investigated.

Firstly, TREE can simulate patterns of stand development after
disturbance, including self-thinning behaviours and successional
replacement. Whereas previous approaches for modelling self-thinning
were limited to a single species
\citep[e.g.][]{Barnes-2004, Coomes-2007}, TREE accommodates multiple
different species with different traits (Fig. \ref{fig:schematic}b; Fig.
\ref{fig:patch}).

Second, TREE allows for the effects of competition and succession on
productivity and biomass accumulation within a stand to be investigated
\citep{Falster-2011}. Leaf area cover and rates of biomass production
tend to vary with stand age. Such changes might come about via a
combination of physiological and structural changes
\citep{Binkley-2002, Smith-2001, Ogawa-2010, Coomes-2007}. TREE enables
the effects of competitive feedbacks to be incorporated, in addition to
any intrinsic physiological factors. Moreover, TREE allows for the
contributions of different species to be mapped (Fig. \ref{fig:patch}c).

\section{Trait, size and patch structured
vegetation}\label{trait-size-and-patch-structured-vegetation}

Most vegetation is subject to a disturbance regime, such that patches of
the landscape are cleared at some rate, by fire, cyclone, landslide, or
disease outbreak
\citep{Connell-1978, White-1979, Chambers-2013, Bormann-1979, Clark-1991, Coomes-2007}.
The vegetation thus comprises a collection of patches differing in time
since last disturbance and linked via seed dispersal (Fig.
\ref{fig:schematic}a). Such a system is often refereed to as a
structured metapopulation \citep{Gyllenberg-2001}.

\subsection{Design features}\label{design-features-2}

Having simulated a single patch, it is straightforward to scale up from
a single patch to an entire metapopulation, at least for the situation
of metapopulations at an equilibrium mix of patch ages and where
disturbances remove all established vegetation. The frequency-density
\(p(a)\) of patches age \(a\) in the landscape is needed. Assuming there
are a large number of patches means the dynamics of \(p\) behave
deterministically and can be approximated via a second PDE
\citep{Vonfoerster-1959, Mckendrick-1926} (see Appendix
\ref{sec:demography} for
details). Scaling from patch to metapopulation is then achieved by
weighting the temporal dynamics of a single patch by \(p(a)\), the
relative abundance of patches age \(a\) in the metapopulation.

\todo[inline]{This should be largely dealt with now, but vignette
needed.MW: Not clear to me (and the fig doesn't exist yet) --
isn't the seed rain an output property? -- what's it at equilibrium
with? (besides just being stable because the age dist of patches has
stabilized)} The main numerical challenge in stepping from single patch
to metapopulation is to identify the equilibrium seed rain of the
metapopulation. For a given input seed rain, the model produces an
output seed rain. The equilibrium seed rain is the vector of seed rain
where input seed rain equals output seed rain. Finding this equilibrium
requires finding a root of a multidimensional function, subject to the
constraints that this root is stable (e.g., the trivial equilibrium of
zero seed rain always exists, but is often not stable).

\subsection{Use cases and examples}\label{use-cases-and-examples-2}

As previously noted \citep{Kohyama-1993, Moorcroft-2001, Falster-2011} a
feature of the size-structured metapopulation concept implemented here
is that it reconciles equilibrium and non-equilibrium approaches to
modelling ecological dynamics
\citep{Levin-1974, Bormann-1979, Connell-1978, Coomes-2007}. An
equilibrium may be approached at the level of metapopulation, meaning
the seed rain and size structure across the entire metapopulation is
approximately stable. Yet the structure of vegetation within individual
patches is constantly in flux. Patches continue to age and their
residents continue to grow until they are disturbed.

Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
us to close the demographic feedback loop and thereby create a
self-regulating system. The equilibrium seed rain for multi-species
communities is approached demographically, incorporating the combined
effects of the model's physiological rules, species traits, competitive
interactions and the prevailing disturbance regime on the
size-distributions of plants within a metapopulation of patches. Unlike
some vegetation models, there is no need to set the density of seeds or
seedlings, these arise as outputs from the model. Characteristics of the
vegetation such as total leaf area, biomass, productivity and allocation
to different plant parts arise as emergent properties (Fig.
\ref{fig:patch}). Similarly, we can observe emergent patterns for
size-distribution, growth and mortality (Fig. \ref{fig:emergent}).

\section{Invasion fitness}\label{invasion-fitness}

With the extension to metapopulation, we are modelling demography across
the entire life cycle and are thus in a position to estimate the fitness
of different types. Invasion fitness quantifies the ability of a rare
individual with particular traits to establish in a community of
established residents at their equilibrium densities. Invasion fitness
is ideally calculated as the long-term per capita growth rate of the new
type; however, in structured metapopulations, the most convenient
indicator of per-capita growth rate is given by the logarithm of the
basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
reproduction ratio \(R\) is simply the total seed output of the new
type, averaged across the metapopulation. The new type can invade when
\(R > 1\).

\subsection{Use cases and examples}\label{use-cases-and-examples-3}

With fitness defined, it becomes possible to model adaptative dynamics
and community assembly \citep{Dieckmann-2007, Geritz-1998}. This is best
thought of as occurring on a fitness landscape, a surface with fitness
plotted against one or more species traits (Fig \ref{fig:fitness}). The
reproductive success of individuals possessing a particular trait value
is measured in the size- and patch-structured metapopulation, and
depends on both the traits and densities of other species in the current
resident community. The reason this is interesting is that under
frequency- and density-dependence, evolution tends to favour different
trait mixtures compared to those predicted by models that optimise some
feature of the community, such as total biomass or carbon production or
seed production in the absence of competition
\citep{Falster-2003, Dieckmann-2007}.

\todo[inline]{this now comes too close to the caption; reworded
to say something a little broader here.}

Fitness landscapes will be positive in regions that allow invasion, and
zero for species that are at equilibrium. The slope of the fitness
landscape at a resident indicates the direction of selection; a negative
slope (dashed line in Fig. \ref{fig:fitness}a) indicates selection for
smaller trait values. Following the fitness gradient, traits may
encounter evolutionary branching points (\todo[inline]{ dfalster-  please add appropriate adaptive dynamics reference
here}; at this point there is no directional selection (solid line in
figure \ref{fig:fitness}a) but invasion is possible both above and below
it in trait space. The distortions of the fitness landscape as species
are introduce show the strong frequency dependent selection that emerges
from the model through the demographic interactions.

\todo[inline]{dfalster - please finesse/delete that last
line.}

Adding in additional species further ``draws down'' the fitness
landscape. Depending on where additional species are added, the first
species may be selected on, and further invasion may or may not be
possible (Fig. \ref{fig:fitness}b, dotted lines). It is often possible
to find a configuration where fitness is negative everywhere (solid line
in Fig. \ref{fig:fitness}b) except for the zero points where species are
present. This community is immune to invasion from any other strategies.

Identifying conditions promoting trait diversity is a valuable use-case
for TREE. Coexisting plant species differ in a range of physiological
traits, yet the conditions allowing for these different types to coexist
in face of competition for a common set of resources remain largely
unknown. The simple example here can be extended to multiple dimensions
and can be used to investigate how trade-offs facilitate coexistence
\citep{Falster-2015}

\section{Implementation details}\label{implementation-details}

TREE is written in C++ and R, using the package \texttt{Rcpp}
\citep{Eddelbuettel-2011, Eddelbuettel-2013} to bridge between the two
languages. Due to the computationally intensive nature of the model, the
core physiological and demographic components are in C++ to maximise
speed. We used templated types allow the demographic and evolutionary
components to be driven by alternative physiological models while
retaining the same interface at higher levels. Despite mostly being
programmed in C++, all components of the model, down to individual
plants, can be extracted and interacted with dynamically in R, using the
\texttt{RcppR6} package \citep{RcppR6}.

TREE makes use of much existing software, including the programming
language C++; the R computational environment \citep{R-2015}, the R
packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
\texttt{R6}\citep{Chang-2014}, and \texttt{testhat}
\citep{Wickham-2011}; and the Boost Library for C++
\citep{Schaling-2014} via the R package \texttt{BH}
\citep{Eddelbuettel-2015}. Source code is hosted at
\href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
Installation instructions are available at this homepage, and binary
releases will also be available there.

\section{Closing comments}\label{closing-comments}

The use-cases described here arise from our interest in the assembly of
trait mixtures in communities. They are nested at different scales
within that question. However a variety of other uses would be possible,
and also extensions to address other facets of community assembly.

Rather than provide a model with a few routines, we have developed an
extensible framework that we hope can be used to answer a number of
questions.

The physiology of growth could be modified, for example to account for
rising atmospheric CO2. - (MW) my suggestion here would be more to
emphasize what you've been saying about using it for growth of
individuals or of stands, rather than the future directions we are
interested in

\begin{itemize}
\item
  Compare to IPM approach
\item
  mathematics similar, difference is source of variability
\end{itemize}

\section{Acknowledgements}\label{acknowledgements}

DS Falster was supported by an ARC discovery grant (DP110102086). RG
FitzJohn was supported by the Science and Industry Endowment Fund
(RP04-174). M Westoby was supported by a fellowship from Australian
Research Council.

\clearpage
\bibliography{refs.bib}

\clearpage

\section{Figures}\label{figures}

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/schematic}
\caption{\textbf{Overview of processes included in TREE package,
including physiological dynamics, population dynamics, and evolutionary
dynamics.} \textbf{a,} An individual's vital rates are jointly
determined by its light environment, height, and traits. \textbf{b,} A
metapopulation consists of a distribution of patches linked by seed
dispersal. Disturbances occasionally remove all vegetation within a
patch. Competitive hierarchies within developing patches are modelled by
tracking the height distribution of individuals across multiple species
(distinguished by colours) as patches age after a disturbance. The
intensity of shading indicates the density of individuals at a given
height. \textbf{c,} The traits of the resident species determine the
shading environment across the metapopulation, which in turn determines
fitness landscapes. Resident traits adjust through directional selection
up the fitness landscape and through the introduction of new species
where the fitness landscape is positive. Adapted from Falster *et al.* (2015).
\label{fig:schematic}}
\end{figure}

\newpage

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/plant.pdf}
\caption{\textbf{Growth trajectories.} Individual plant height growth with different LMA values and light environments.  (a) Lower light environments flatten growth trajectories, and different species traits have nonlinear interactions with the light level.  (b) Over time, the fraction of living tissue switches from leaf towards sapwood, with high LMA species having relatively more mass in leaf than low LMA species.  (c) Growth rates peak around 5m high, but this position varies with both trait and light level (not that this is the derivative of panel (a)).  (d) Growth rates vary with plant size (seedling and mature shown) and with traits; growth rate differences are more pronounced in low light.  Whole plant light compensation points emerge from the model as the point where growth rate is zero (x intercept).  The light level for panels (a) and (c) is indicated with dots.  Throughout "low LMA" is 0.1 and solid lines, "high LMA" is 0.267 and dashed.
\label{fig:plant}}
\end{figure}

\newpage

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/patch.pdf}
\caption{Cohorts of two species growing within a competitive environment.  Each line represents a "cohort"; some number of individuals that share a birth time.  Red lines are the low LMA species from figure \ref{fig:plant}, blue lines are the high LMA species.  The initial growth rate advantage of the low LMA species (see Figure \ref{fig:plant} c) means it quickly overtops the high LMA species, suppressing its growth.  Self-thining leaves space for the high LMA species to invade.  (a) all cohorts, equally weighted; the spacing is generated by an adaptive refinement algorithm.  (b) Cohorts shaded by by their density.  (c) The leaf area index at ground level for each species.  The dotted line is the total leaf area index (sum across both species).}
\label{fig:patch}
\end{figure}

\newpage

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
\caption{\textbf{Emergent size-distributions across metapopulation}
\label{fig:emergent}}
\end{figure}

\newpage

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/fitness.pdf}
\caption{Example fitness landscape for LMA, showing potential for stable coexistence of multiple types.  Fitness is the log of the population growth rate - at zero species are at equilibrium, above zero they will increase in density.  (a) Landscae generated by a single species.  At LMA of 0.1 (the dashed line) there is directional selection for smaller LMA.  At LMA of 0.08 (solid line) there is a branching point (inset).  (b) Landscape generated by two species, holing the first at LMA of 0.08. Introducing a species at the maximum fitness value from panel (a) reduces the region of positive fitness considerably (dashed line).  Dotted lines show subsequent invasions and replacements of the second species, and the solid line shows a stable fitness landscape where both species exist on a peak in the fitness landscape.  At this point no futher invasion is possible.}
\label{fig:fitness}
\end{figure}

\clearpage

\begin{appendices}\label{sec:appendices}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

\setcounter{figure}{0}
\setcounter{table}{0}

\section{Default physiological model for TREE}\label{sec:FFW16}

See attached file \url{tree_physiology.pdf}

\section{Modelling demography in the TREE package}\label{sec:demography}

see attached file \url{tree_demography.pdf}

\section{Supplementary figures}\label{supplementary-figures}

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
\caption{\textbf{Adpative cohort spacing solves problem of diverging characteristics}}
\label{fig:characteristics}
\end{figure}

\newpage

\begin{figure}[h!]
\centering
\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
\caption{\textbf{Approach for solving equilibrium seed rain across the metapopulation \todo[inline]{RGF: I favour referring to a package vignette here - I've started rebuilding one that we used to have}}}
\label{fig:seed_rain}
\end{figure}

\end{appendices}
\end{document}

### 2015-05-19 R/ms.R

@@ -1,4 +1,3 @@
-
 pandoc_build <- function(file, template, toc=FALSE, engine="xelatex" ){
   args <- list(sprintf('--template=%s',template), "--listings", sprintf('--latex-engine=%s', engine))
   if(toc){
@@ -6,4 +5,8 @@ pandoc_build <- function(file, template, toc=FALSE, engine="xelatex" ){
   }
   pandoc_convert(file, output= paste0(tools::file_path_sans_ext(file), ".pdf"),
     citeproc = TRUE, options = args, verbose = TRUE)
+}
+
+ms_deps_timestamp <- function(filename) {
+  writeLines(as.character(Sys.time()), filename)
 }

### 2015-05-19 R/figures_plant.R
Minor improvements to plant figure
See #10

@@ -84,9 +84,9 @@ figure_plant <- function() {
   ## at the moment they run off the RHS because they're run to a
   ## common *height*.
 
-  cols_light <- c("black", "red")
-  cols_height <- c("grey60", "black")
-  cols_part <- c("black", "blue")
+  cols_light <- c("black", "#e34a33")
+  cols_height <- c("#31a354", "black")
+  cols_part <- c("black", "#045a8d")
 
   op <- par(mfrow=c(2, 2), mar=c(3.6, 4.1, .5, .5), mgp=c(2.2, 1, 0))
   on.exit(par(op))
@@ -130,8 +130,10 @@ figure_plant <- function() {
     cbind(h=x$info[, "height"], dhdt=x$info[, "height_dt"])
   }
   dhdt_h <- lapply(data, f)
-  ylim <- range(sapply(dhdt_h, function(x) range(x[, 2])), 0)
-  plot(dhdt_h[["1_1"]], ylim=ylim, lty=1, col=cols_light[[1]],
+  ymax_dhdt <- max(sapply(dhdt_h, function(x) max(x[, 2])), 0)
+  ylim_dhdt <- c(-0.05 * ymax_dhdt, ymax_dhdt)
+
+  plot(dhdt_h[["1_1"]], ylim=ylim_dhdt, lty=1, col=cols_light[[1]],
        type="l", las=1, xlab="Height (m)", ylab="d height / d t")
   lines(dhdt_h[["2_1"]], type="l", lty=2, col=cols_light[[1]])
   lines(dhdt_h[["1_2"]], type="l", lty=1, col=cols_light[[2]])
@@ -142,7 +144,7 @@ figure_plant <- function() {
          bty="n")
 
   ## Panel d: wplcp
-  plot(NA, xlim=c(0, 1), ylim=ylim, las=1,
+  plot(NA, xlim=c(0, 1), ylim=ylim_dhdt, las=1,
        xlab="Canopy openness (%)", ylab="d height / d t")
   label_panel(4)
   for (i in seq_along(d1)) {
@@ -152,6 +154,20 @@ figure_plant <- function() {
     lines(d2[[i]], lty=2, col=cols_height[[i]])
   }
   points(c(env1$canopy_openness(0), env2$canopy_openness(0)),
-         c(0, 0), pch=19, col=1:2)
+         c(0, 0), pch="|", col=cols_light)
[…]

### 2015-05-19 ms.tex
Tidy draft for pre-submission feedback

@@
 
-\usepackage[verbose,tmargin=2cm,bmargin=2cm,lmargin=2.5cm,rmargin=2.5cm]{geometry}
+\usepackage[verbose,tmargin=2cm,bmargin=2cm,lmargin=3.5cm,rmargin=2.5cm]{geometry}
 
@@
             pdfauthor={Daniel Falster, Rich FitzJohn, Mark Westoby},
-            pdftitle={TREE: A package for modelling plant TRait Ecology and Evolution},
+            pdftitle={\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution},
             colorlinks=true,
@@
 \usepackage{authblk}
-\title{TREE: A package for modelling plant TRait Ecology}
+\title{\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution}
 \author[1]{Daniel S. Falster}
@@
 
-\usepackage{natbib}
+\usepackage[sort&compress]{natbib}
 \bibliographystyle{mee}
@@
 
-A manuscript in consideration as a research paper for publication in MEE
-as part of the Special Feature Demography beyond the Population
-
-\listoftodos
+\noindent A manuscript in consideration as a research paper for publication in MEE
+as part of the Special Feature \emph{Demography beyond the Population}.
 
@@
 \section{Abstract}\label{abstract}
-
-\emph{(max 350 words; rework into 4 points)}
-
 \begin{enumerate}
@@
 \item
-  Population dynamics in forests are strongly size-structured. Larger
-  plants shade smaller plants. They also expend proportionately more
-  energy on their woody stems. Although the importance of size-
+  Population dynamics in forests are strongly size-structured:
+  larger plants shade smaller plants while also expending
+  proportionately more energy on building and maintaining woody stems.
+  Although the importance of size-
   structure for demography is widely recognised, many mechanistic models
@@
 \item
-  Here we introduce the TREE package (TRait Ecology and Evolution), an
+  Here we introduce the \texttt{TREE} package (TRait Ecology and Evolution), an
   extensible framework for modelling plant demography across the entire
-  life cycle via coupled differential equations. At its core, TREE is an
+  life cycle via coupled differential equations. At its core, \texttt{TREE} is an
   individual-based mechanistic model where plant function is mediated by
-  traits, Individuals from multiple species can be grown in isolation,
+  traits. Individuals from multiple species can be grown in isolation,
   in patches of competing plants, or in metapopulations under a
@@
 \item
-  TREE is presented as an open source \texttt{R} package and is
+  \texttt{TREE} is presented as an open source \texttt{R} package and is
   available at
   \href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
-  While accessed from R, the core routines in TREE are written in C++.
+  While accessed from R, the core routines in \texttt{TREE} are written in C++.
   The package provides for alternative physiologies and for
@@
 \item
-  We provide worked examples illustrating TREE's use to study the
+  We provide worked examples illustrating \texttt{TREE}'s use to study the
   influence of functional traits on growth of individual plants, of
-  whole stands and on assembly of ecological communities.
+  whole patches and on assembly of ecological communities.
 \end{enumerate}
@@
 physiology, metapopulation, mortality, reproduction, size-structure,
-trade-off
+tradeoff
 
@@
 
-Plant growth and demography are fundamentally structured by species
-traits and by individual size. Both influence dynamics over time-scales
-ranging from instantaneous physiological effects to long-term
-evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As an
-individual increases its leaf area, its potential to generate
-photosynthate also rises. But for larger individuals, increasing
+Plant growth and demography are fundamentally size- and trait- structured,
+influencing dynamics over time-scales ranging from instantaneous physiological
+effects to long-term evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As
+an individual increases its leaf area, its potential to generate
+photosynthetic also rises. Yet for larger individuals,  increasing
 fractions of photosynthetic income must also be diverted towards
 building and maintaining support tissues rather than leaves
-\citep{Givnish-1988, Enquist-1999}. A larger fraction of income is also
+\citep{Givnish-1988, Enquist-2007}. A larger fraction of income is also
 allocated to reproduction \citep{Thomas-2011}. Consequently at
@@
 demography has been widely recognised for a long time
-\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001, Kohyama-1993, Enquist-1999, Pacala-1996, Coomes-2007},
+\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
+Kohyama-1993, Enquist-2007, Pacala-1996, Coomes-2007},
 most large-scale mechanistic models of plant growth either omit it
 entirely, or include only coarse approximations
-\citep{Cramer-2001, Sitch-2003}. The likely reason is the computational
+\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
+reason is the computational
 challenge of modelling size-structured dynamics. During the 1980s and
@@
 sacrificed details of demography in order to achieve global coverage
-\citep[eg.][]{Haxeltine-1996}. This history has brought us to the
+\citep{Haxeltine-1996}. This history has brought us to the
 current perplexing situation where all but a couple
-\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the leading vegetation
-models lack size-structure
-\citep{Cramer-2001, Dekauwe-2014, Sitch-2003, Kelley-2013}. These models
+\citep{Moorcroft-2001, Smith-2014} of the leading vegetation
+models lack size-structure. These models
 are focused on modelling fluxes carbon and water, but say little about
@@
 questions about species coexistence also lack size-structured dynamics
-\citep[e.g.][]{MacArthur-1967, Levin-1974, Leimar-2013, Geritz-1995, Calcagno-2006, Tilman-1985}.
+\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013, Levin-1974,MacArthur-1967,Tilman-1985}.
 It is common to assume competitive interactions are influenced by
@@
 size-structured demography is rarely considered, at least for plant
-models \citetext{\citealp[for animal
-examples,][]{Deroos-1988}; \citealp[see][]{Deroos-1992}}. Thus it has
+models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Thus it has
 been difficult to connect mechanisms described in these simplistic
@@
 technical documents are provided as supplementary information, and updated
-versions of technical documents are also available within the TREE
+versions of technical documents are also available within the \texttt{TREE}
 package itself. The use cases highlighted reflect our own interests in
@@
 expect the methods provided to be useful in other contexts. For example,
-prior to TREE there was no R package that allowed one to simply grow a
+prior to \texttt{TREE} there was no \texttt{R} package that allowed one to simply grow a
 plant over time, based on well-understood physiology. Surely this is one
@@
 
-\section{The methods}\label{the-methods}
+\section{The approach}
 
-TREE is a mechanistic model, meaning the dynamics of the system arise
+\texttt{TREE} is a mechanistic model, meaning the dynamics of the system arise
 from rules specifying on how individuals grow and interact. In ecology,
-there is a long history of using simple deterministic models \citep[such
-as Lotka-Volterra population dynamics,
-e.g.][]{MacArthur-1967, Leimar-2013} to understand biological phenomena.
-TREE was developed with this style of analysis in mind, while also
+there is a long history of using simple deterministic models
+to understand biological phenomena, such as Lotka-Volterra population dynamics
+\citep{MacArthur-1967, Leimar-2013} .
+\texttt{TREE} was developed with this style of analysis in mind, while also
 allowing for a richer set of ecological dynamics than is possible in
@@
 \ref{fig:schematic}) following methods described by \citet{Falster-2011}
-and \citet{Falster-2015}. The core rules in TREE are about the
+and \citet{Falster-2015}. The core rules in \texttt{TREE} are about the
 short-term physiological functioning of an individual plant and how
@@
 
-\section{Effects of size, trait and environment on demography of
-individual
-plants}\label{effects-of-size-trait-and-environment-on-demography-of-individual-plants}
+\section{Effects of size, trait and environment on demography of individual plants}
 
-The core of TREE is a model for a plant's physiological strategy as
+The core of \texttt{TREE} is a model for a plant's physiological strategy as
 specified by its traits (Fig. \ref{fig:schematic}a). This sub-model
@@
 
-The default physiological model used in TREE largely reflects that
-presented in Falster-2011, but with extensions allowing for diameter
+The default physiological model used in \texttt{TREE} largely reflects that
+presented in \citet{Falster-2011}, but with extensions allowing for diameter
 growth also to be estimated. To properly model trait variation,
-different parameters in the model are linked via assumed trade-offs. For
+different parameters in the model are linked via assumed tradeoffs. For
 example, the trait LMA is used to estimate the rate of leaf turnover,
@@
 
-\subsection{Design features}\label{design-features}
+\subsection{Design features}
 
-TREE includes a default physiological model (fully described and derived
+\texttt{TREE} includes a default physiological model (fully described and derived
 in the Appendix \ref{sec:FFW16}). However a feature of the package is that alternative
@@
 altered. Trait differences are accounted for by altering relevant
-parameters. This example shows how to establish a new plant with height
-of 5 m, and estimate its physiological and demographic rates in the
-given light environment.
+parameters (see Appendix \ref{sec:parameters}).
 
-\begin{verbatim}
-# Example illustrating use of individual plant, all the outputs from plant internals
-\end{verbatim}
+\subsection{Use cases and examples}
 
-\subsection{Use cases and examples}\label{use-cases-and-examples}
-
-\todo[inline]{these aren't demographic rates - they're
-physiological perphaps? Height growth rate, allocation, WPLCP, etc are
-properties of individuals here. I've updated but we can discuss.}
-\todo[inline]{allocation to replacement costs are not shown in
-the figure at all, so this section is quite confused. I've reworded but
-you'll want to go through and tweak. The graph shows ``fraction of live
-mass that is X'' which is different to what is talked about.}
-
-With above functionality, TREE can be used to estimate essential
+With above functionality, \texttt{TREE} can be used to estimate essential
 physiological rates for individual plants (Fig. \ref{fig:plant}). The
@@
 details). As the plant grows, allocation to different types varies, and
-the composition of the plant changes to become more stemmy and less
-leafy (Fig. \ref{fig:plant}b). This in turn affects maintenance and
+the composition of the plant changes to include more stem and less
+leaf (Fig. \ref{fig:plant}b). This in turn affects maintenance and
 turnover costs and the growth rate of plants (Fig. \ref{fig:plant}c).
-This shift alone generates a distinctive hump-shaped pattern of absolute
-growth with size, and equally the widely regarded decline in relative
-growth rate with size from birth onwards.
+This shift in allocation alone generates a distinctive hump-shaped pattern of absolute
+growth with size \citep{King-2011}, and equally the widely regarded
+decline in relative growth rate with size from birth onwards
+\citep{Enquist-2007}.
 
@@
 
-\section{Plants competing in a patch}\label{plants-competing-in-a-patch}
+\section{Plants competing in a patch}
 
@@
 
-\subsection{Design features}\label{design-features-1}
+\subsection{Design features}
 
@@
 that patches are large, and are vertically but not horizontally
-structured. Thus, shading in TREE is modelled as a mean field with
+structured. Thus, shading in \texttt{TREE} is modelled as a mean field with
 respect to spatial layout. Similar assumptions are found in many models
@@
 PDE. Biologically these are the growth trajectories of individuals,
-conditioned on them surviving. Fig. \ref{fig:patch}a illustrates the
-trajectories of individuals within a patch recovering from disturbance.
-Fig. \ref{fig:patch}b shows the same trajectories, but with shading
-indicating the density of plants at that size. (see also Fig.
-\ref{fig:schematic}b).
+conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
+growth and mortality combine to alter the density of individuals along
+the trajectories (Fig. \ref{fig:patch}b).
 
@@
 the first and second moments of the density distribution \(n\) within
-each cohort. In TREE we use a different but from a theoretical
-perspective equally valid approach. Instead of tracking first and second
+each cohort. In \texttt{TREE} we use a different -- but from a theoretical
+perspective equally valid --  approach. Instead of tracking first and second
 moments of the size-distribution within cohorts, we track a point mass
-estimate of \(n\) along the trajectories corresponding to the boundary
+estimate of \(n\) along the trajectory for the boundary
 of each cohort. We found this approach preferable because it does not
@@
 adaptive refining of time points at which new cohorts are introduced
-into the population. Under competition, the growth trajectories of
+into the population. In the original EBT, cohorts were introduced at a
+fixed time interval. Under competition, the growth trajectories of
 individuals born at similar times can diverge substantially over time
-(Fig. \ref{fig:patch}a). This is can become a problem later during patch
+(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography}).
+This is can become a problem later during patch
 development because much of the population can be located between
 widely-spaced cohorts, delivering low numerical precision. To maintain
-numerical precision we use an iterative algorithm to adaptively refine
+numerical precision we therefore use an iterative algorithm to adaptively refine
 the cohort spacing until the trajectories of adjacent cohorts remain
-adequately resolved throughout development of the patch (Fig.
-\ref{fig:characteristics}b; see Appendix\ref{sec:demography} for details).
+adequately resolved throughout development of the patch (see
+Appendix \ref{sec:demography} for details). This refinement causes an uneven
+distribution of cohort introduction times across patch age (Fig.
+\ref{fig:patch}a).
 
-\subsection{Use cases and examples}\label{use-cases-and-examples-1}
+\subsection{Use cases and examples}
 
@@
 
-Firstly, TREE can simulate patterns of stand development after
+First, \texttt{TREE} can simulate patterns of patch development after
 disturbance, including self-thinning behaviours and successional
@@
 were limited to a single species
-\citep[e.g.][]{Barnes-2004, Coomes-2007}, TREE accommodates multiple
+\citep[e.g.][]{Barnes-2004, Coomes-2007}, \texttt{TREE} accommodates multiple
 different species with different traits (Fig. \ref{fig:schematic}b; Fig.
@@
 
-Second, TREE allows for the effects of competition and succession on
-productivity and biomass accumulation within a stand to be investigated
+Second, \texttt{TREE} allows for the effects of competition and succession on
+productivity and biomass accumulation within a patch to be investigated
 \citep{Falster-2011}. Leaf area cover and rates of biomass production
-tend to vary with stand age. Such changes might come about via a
+tend to vary with patch age. Such changes might come about via a
 combination of physiological and structural changes
-\citep{Binkley-2002, Smith-2001, Ogawa-2010, Coomes-2007}. TREE enables
+\citep{Binkley-2002, Smith-2001, Ogawa-2010, Coomes-2007}. \texttt{TREE} enables
 the effects of competitive feedbacks to be incorporated, in addition to
-any intrinsic physiological factors. Moreover, TREE allows for the
+any intrinsic physiological factors. Moreover, \texttt{TREE} allows for the
 contributions of different species to be mapped (Fig. \ref{fig:patch}c).
 
-\section{Trait, size and patch structured
-vegetation}\label{trait-size-and-patch-structured-vegetation}
+While many vegetation models still lack size structure, the importance of
+size-structure for modelling phenomena such as allocation and production is
+being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
+
+\section{Trait, size and patch structured vegetation}
 
@@
 since last disturbance and linked via seed dispersal (Fig.
-\ref{fig:schematic}a). Such a system is often refereed to as a
+\ref{fig:schematic}b). Such a system is often refereed to as a
 structured metapopulation \citep{Gyllenberg-2001}.
 
-\subsection{Design features}\label{design-features-2}
+\subsection{Design features}
 
@@
 
-\todo[inline]{This should be largely dealt with now, but vignette
-needed.MW: Not clear to me (and the fig doesn't exist yet) --
-isn't the seed rain an output property? -- what's it at equilibrium
-with? (besides just being stable because the age dist of patches has
-stabilized)} The main numerical challenge in stepping from single patch
+The main numerical challenge in stepping from single patch
 to metapopulation is to identify the equilibrium seed rain of the
@@
 constraints that this root is stable (e.g., the trivial equilibrium of
-zero seed rain always exists, but is often not stable).
+zero seed rain always exists, but is often not stable) (Appendix \ref{sec:fitness});
 
-\subsection{Use cases and examples}\label{use-cases-and-examples-2}
+\subsection{Use cases and examples}
 
@@
 to different plant parts arise as emergent properties (Fig.
-\ref{fig:patch}). Similarly, we can observe emergent patterns for
-size-distribution, growth and mortality (Fig. \ref{fig:emergent}).
+\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
+growth rate with respect to size emerge at both patch scale and scale of
+metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
+physiological model produces a variety of relationships, dependent on patch
+age. These emergent phenomena might then be compared to the observed
+distributions in large forest tracts \citep{Muller-2006}.
 
-\section{Invasion fitness}\label{invasion-fitness}
+\section{Invasion fitness}
 
@@
 individual with particular traits to establish in a community of
-established residents at their equilibrium densities. Invasion fitness
+established residents at their equilibrium densities\citep{Metz-1992}.
+Invasion fitness
 is ideally calculated as the long-term per capita growth rate of the new
-type; however, in structured metapopulations, the most convenient
+type; however, in structured metapopulations, a more convenient
 indicator of per-capita growth rate is given by the logarithm of the
@@
 
-\subsection{Use cases and examples}\label{use-cases-and-examples-3}
+\subsection{Use cases and examples}
 
-With fitness defined, it becomes possible to model adaptative dynamics
+With fitness defined, it becomes possible to model adaptive dynamics
 and community assembly \citep{Dieckmann-2007, Geritz-1998}. This is best
 thought of as occurring on a fitness landscape, a surface with fitness
-plotted against one or more species traits (Fig \ref{fig:fitness}). The
+plotted against one or more species traits (Fig. \ref{fig:fitness}). The
 reproductive success of individuals possessing a particular trait value
@@
 
-\todo[inline]{this now comes too close to the caption; reworded
-to say something a little broader here.}
-
 Fitness landscapes will be positive in regions that allow invasion, and
@@
 slope (dashed line in Fig. \ref{fig:fitness}a) indicates selection for
-smaller trait values. Following the fitness gradient, traits may
-encounter evolutionary branching points (\todo[inline]{ dfalster-  please add appropriate adaptive dynamics reference
-here}; at this point there is no directional selection (solid line in
+smaller trait values. Fitness landscapes are constantly reshaped
+as species are introduced and evolve, reflecting the non-linear
+demographic feedbacks that occur under frequency-dependent evolution
+ \citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient, can lead to an
+evolutionary branching point \citep{Geritz-1998,Dieckmann-2007};
+at this point there is no directional selection (solid line in
 figure \ref{fig:fitness}a) but invasion is possible both above and below
-it in trait space. The distortions of the fitness landscape as species
-are introduce show the strong frequency dependent selection that emerges
-from the model through the demographic interactions.
-
-\todo[inline]{dfalster - please finesse/delete that last
-line.}
-
-Adding in additional species further ``draws down'' the fitness
-landscape. Depending on where additional species are added, the first
-species may be selected on, and further invasion may or may not be
-possible (Fig. \ref{fig:fitness}b, dotted lines). It is often possible
+it in trait space. It is often possible
 to find a configuration where fitness is negative everywhere (solid line
 in Fig. \ref{fig:fitness}b) except for the zero points where species are
-present. This community is immune to invasion from any other strategies.
+present. This community is immune to invasion from any other strategies
+\citep{Geritz-1998}.
 
 Identifying conditions promoting trait diversity is a valuable use-case
-for TREE. Coexisting plant species differ in a range of physiological
+for \texttt{TREE}. Coexisting plant species differ in a range of physiological
 traits, yet the conditions allowing for these different types to coexist
@@
 unknown. The simple example here can be extended to multiple dimensions
-and can be used to investigate how trade-offs facilitate coexistence
+and can be used to investigate how tradeoffs facilitate coexistence
 \citep{Falster-2015}
 
-\section{Implementation details}\label{implementation-details}
+\section{Implementation details}
 
-TREE is written in C++ and R, using the package \texttt{Rcpp}
+\texttt{TREE} is written in C++ and \texttt{R}, using the package \texttt{Rcpp}
 \citep{Eddelbuettel-2011, Eddelbuettel-2013} to bridge between the two
@@
 core physiological and demographic components are in C++ to maximise
-speed. We used templated types allow the demographic and evolutionary
+speed. We used templated types that allow the demographic and evolutionary
 components to be driven by alternative physiological models while
@@
 
-TREE makes use of much existing software, including the programming
-language C++; the R computational environment \citep{R-2015}, the R
+\texttt{TREE} makes use of much existing software, including the programming
+language C++; the \texttt{R} computational environment \citep{R-2015}, the \texttt{R}
 packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
@@
 \citep{Wickham-2011}; and the Boost Library for C++
-\citep{Schaling-2014} via the R package \texttt{BH}
+\citep{Schaling-2014} via the \texttt{R} package \texttt{BH}
 \citep{Eddelbuettel-2015}. Source code is hosted at
@@
 
-\section{Closing comments}\label{closing-comments}
+The current parametrisation of the physiological model is based on
+\citet{Falster-2011}. These values were drawn from available literature (see
+Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
+readily available. Providing updated parameter sets is trivial (see Appendix
+\ref{sec:parameters}). Ideally, parameters would be inferred through a combination of
+independent estimates and constraining outputs from \texttt{TREE} against diverse data sources
+\citep{Lebauer-2012, Keenan-2013}.
 
-The use-cases described here arise from our interest in the assembly of
-trait mixtures in communities. They are nested at different scales
-within that question. However a variety of other uses would be possible,
-and also extensions to address other facets of community assembly.
+\section{Closing comments}
 
-Rather than provide a model with a few routines, we have developed an
-extensible framework that we hope can be used to answer a number of
-questions.
+The \texttt{TREE} package makes it easier for researchers and forest practitioners
+alike to investigate a variety of non-linear demographic phenomena.  Rather
+than provide a model with a few routines, we have developed an extensible
+framework that we hope can be used to answer a number of questions at a
+variety of scales. The specific use-cases highlighted here arise from our
+interest in the assembly of trait mixtures in communities. However a variety
+of other uses would be possible, and also extensions to address other facets
+of community assembly. For example,  the physiology of growth could be
+modified to account for rising atmospheric CO2 or the effects of competition
+for multiple resources could likewise be investigated.
 
-The physiology of growth could be modified, for example to account for
-rising atmospheric CO2. - (MW) my suggestion here would be more to
-emphasize what you've been saying about using it for growth of
-individuals or of stands, rather than the future directions we are
-interested in
+It is hoped that the approach here can be used to inform both future empirical
+research and model development in neighbouring areas. As noted above, most
+large-scale mechanistic models of plant growth do not explicitly model size
+structure, so remain unable to investigate demographic phenomena
+\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. At the same time,
+much attention is paid to detailed leaf-level physiology. \texttt{TREE} aims to bridge
+this gap that emerged in the model landscape, where detailed physiology and
+size-structured dynamics combine to drive forest structure and function via
+traits.
 
-\begin{itemize}
-\item
-  Compare to IPM approach
-\item
-  mathematics similar, difference is source of variability
-\end{itemize}
-
-\section{Acknowledgements}\label{acknowledgements}
+\section{Acknowledgements}
 
@@
 \clearpage
-\bibliography{refs.bib}
+\bibliography{refs}
 
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{output/schematic}
-\caption{\textbf{Overview of processes included in TREE package,
-including physiological dynamics, population dynamics, and evolutionary
-dynamics.} \textbf{a,} An individual's vital rates are jointly
-determined by its light environment, height, and traits. \textbf{b,} A
-metapopulation consists of a distribution of patches linked by seed
-dispersal. Disturbances occasionally remove all vegetation within a
-patch. Competitive hierarchies within developing patches are modelled by
-tracking the height distribution of individuals across multiple species
-(distinguished by colours) as patches age after a disturbance. The
-intensity of shading indicates the density of individuals at a given
-height. \textbf{c,} The traits of the resident species determine the
-shading environment across the metapopulation, which in turn determines
-fitness landscapes. Resident traits adjust through directional selection
-up the fitness landscape and through the introduction of new species
-where the fitness landscape is positive. Adapted from Falster *et al.* (2015).
-\label{fig:schematic}}
+
+\caption{Processes modelled within \texttt{TREE} include physiological, population, and
+evolutionary dynamics.
+Dynamics across all three levels are driven by the
+physiological sub model determining an individual's physiological and
+demographic rates on the basis of its traits, size and light environment (a).
+(b) Competitive hierarchies  are modelled by tracking the height distribution of
+individuals across within a "patch" after a disturbance. The intensity
+of shading indicates the density of individuals at a given height for
+different species (distinguished by colours). A size- and patch-structured
+metapopulation consists of a distribution of patches linked by seed dispersal.
+Disturbances occasionally remove all vegetation within a patch, resetting the
+patch. (c) The traits of the resident species determine the shading environment
+across the metapopulation, which in turn determines the fitness landscape.
+Resident traits adjust through directional selection up the fitness landscape
+and through the introduction of new species. Adapted from
+\citet{Falster-2011}, \citet{Falster-2015}. }
+
+\label{fig:schematic}
 \end{figure}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{output/plant.pdf}
-\caption{\textbf{Growth trajectories.} Individual plant height growth with different LMA values and light environments.
-(a) Lower light environments flatten growth trajectories, and different species traits have nonlinear interactions with the light level.
-(b) Over time, the fraction of living tissue switches from leaf towards sapwood, with high LMA species having relatively more mass in leaf than low LMA species.
-(c) Growth rates peak around 5m high, but this position varies with both trait and light level (not that this is the derivative of panel (a)).
-(d) Growth rates vary with plant size (seedling and mature shown) and with traits; growth rate differences are more pronounced in low light.
-Whole plant light compensation points emerge from the model as the point where growth rate is zero (x intercept).
-The light level for panels (a) and (c) is indicated with dots.
-Throughout "low LMA" is 0.1 and solid lines, "high LMA" is 0.267 and dashed.
-\label{fig:plant}}
+\caption{Physiological dynamics for individual plants vary with size, trait
+and light environment. (a) Growth trajectories for plants differing in light
+environment and trait (LMA, low = 0.1, solid line; high =  0.267, dashed
+line). Lower light environments slow trajectories and interact non-linearly
+with traits. (b) Over time, the fraction of living tissue switches from leaf
+towards sapwood, with high LMA species having relatively more mass in leaf
+than low LMA species. (c) Growth rates peak around 5m high, but this position
+varies with both trait and light level (note that this is the derivative of
+panel (a)). (d) Growth rates vary with plant size (seedling and mature shown)
+and with traits; growth rate differences are more pronounced in high light.
+Whole plant light compensation points emerge from the model as the point where
+growth rate is zero (x intercept).  The light level used in panels (a) and (c)
+is indicated with dots.}
+\label{fig:plant}
 \end{figure}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{output/patch.pdf}
-\caption{Cohorts of two species growing within a competitive environment.
-Each line represents a "cohort"; some number of individuals that share a birth time.
-Red lines are the low LMA species from figure \ref{fig:plant}, blue lines are the high LMA species.
-The initial growth rate advantage of the low LMA species (see Figure \ref{fig:plant} c) means it quickly overtops the high LMA species, suppressing its growth.
-Self-thining leaves space for the high LMA species to invade.
-(a) all cohorts, equally weighted; the spacing is generated by an adaptive refinement algorithm.
-(b) Cohorts shaded by by their density.
-(c) The leaf area index at ground level for each species.
-The dotted line is the total leaf area index (sum across both species).}
+\caption{Population dynamics for two species competing within a patch. Lines
+in panels (a) and (b) show the growth trajectories for individuals on the
+boundary of successive "cohorts" for each species. A single cohort includes
+all individuals born within a given time period. Red lines are the low LMA
+species from figure \ref{fig:plant}, blue lines are the high LMA species.  The
+initial growth rate advantage of the low LMA species (see Figure
+\ref{fig:plant} c) means it quickly over tops the high LMA species,
+suppressing its growth.  Self-thinning leaves space for the high LMA species
+to establish below the canopy of the fast species. Panel (a) shows all
+trajectories equally weighted; the spacing is generated by an adaptive
+refinement algorithm.  In panel (b) trajectories are shaded by the density of
+individuals at given size and patch age.  (c) The leaf area index at ground
+level for each species.  The dotted line is the total leaf area index (sum
+across both species).}
 \label{fig:patch}
@@
 \centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
-\caption{\textbf{Emergent size-distributions across metapopulation}
-\label{fig:emergent}}
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/emergent.pdf}
+\caption{Emergent size-distributions across a metapopulation.
+Grey lines show patterns within individual patches differing in time since disturbance (age), with blue lines indicating patches of a specific age.
+Panels show (a) patterns of abundance, (b) leaf area and (c) height growth rate against size.
+Note that units of panels (a) and (b) are per unit height and per unit ground area.
+}
+\label{fig:emergent}
 \end{figure}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{output/fitness.pdf}
-\caption{Example fitness landscape for LMA, showing potential for stable coexistence of multiple types.
-Fitness is the log of the population growth rate - at zero species are at equilibrium, above zero they will increase in density.
-(a) Landscae generated by a single species.
-At LMA of 0.1 (the dashed line) there is directional selection for smaller LMA.
-At LMA of 0.08 (solid line) there is a branching point (inset).
-(b) Landscape generated by two species, holing the first at LMA of 0.08.
-Introducing a species at the maximum fitness value from panel (a) reduces the region of positive fitness considerably (dashed line).
-Dotted lines show subsequent invasions and replacements of the second species, and the solid line shows a stable fitness landscape where both species exist on a peak in the fitness landscape.
-At this point no futher invasion is possible.}
+\caption{Example fitness landscape for LMA, showing potential for stable
+coexistence of multiple types.  Fitness is the log of the population growth
+rate - at zero species are at equilibrium, above zero they will increase in
+density.  (a) Landscape generated by a single species.  At LMA of 0.1 (the
+dashed line) there is directional selection for smaller LMA.  At LMA of 0.08
+(solid line) there is a branching point (see inset for more detail).  (b)
+Landscape generated by two species, holing the first at LMA of 0.08.
+Introducing a species at the maximum fitness value from panel (a) reduces the
+region of positive fitness considerably (dashed line).  Dotted lines show
+subsequent invasions and replacements of the second species, and the solid
+line shows a stable fitness landscape where both species exist on a peak in
+the fitness landscape.  At this point no further invasion is possible.}
 \label{fig:fitness}
@@
 \begin{appendices}\label{sec:appendices}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thetable}{S\arabic{table}}
 
-\setcounter{figure}{0}
-\setcounter{table}{0}
-
-\section{Default physiological model for TREE}\label{sec:FFW16}
+\section{Default physiological model for \texttt{TREE}}\label{sec:FFW16}
 
@@
 
-\section{Modelling demography in the TREE package}\label{sec:demography}
+\section{Modelling demography in the \texttt{TREE} package}\label{sec:demography}
 
-see attached file \url{tree_demography.pdf}
+See attached file \url{tree_demography.pdf}
 
-\section{Supplementary figures}\label{supplementary-figures}
+\section{Solving for equilibrium seed rain}\label{sec:seed}
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
-\caption{\textbf{Adpative cohort spacing solves problem of diverging characteristics}}
-\label{fig:characteristics}
-\end{figure}
+See attached file \url{vignette_equilibrium.R}
 
-\newpage
+\section{Calculating fitness in the \texttt{TREE} package}\label{sec:fitness}
 
-\begin{figure}[h!]
-\centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/empty.pdf}
-\caption{\textbf{Approach for solving equilibrium seed rain across the metapopulation \todo[inline]{RGF: I favour referring to a package vignette here - I've started rebuilding one that we used to have}}}
-\label{fig:seed_rain}
-\end{figure}
+See attached file \url{vignette_fitness.R}
+
+\section{Modifying parameters of physiological model}\label{sec:parameters}
+
+See attached file \url{vignette_parameters.R}
 

### 2015-05-19 vignettes/vignette_equilibrium.R

@@ -16,7 +16,9 @@ cobweb <- function(m, ...) {
   lines(rep(m[,1], each=2), c(t(m)), ...)
 }
 
-## Try to establish what the equilubrium seed rain is.
+## Try to establish what the equilibrium seed rain is.
+
+## Set model parameters
 p0 <- ebt_base_parameters()
 p <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
 
@@ -49,14 +51,14 @@ cobweb(approach, pch=19, cex=.5, type="o")
 
 ## # 2: Near the equilibrium point:
 
-## Then, in the vinicity of the root we should look at what the curve
+## Then, in the vicinity of the root we should look at what the curve
 ## actually looks like, without adaptive refinement.
 dr <- 2 # range of input to vary (plus and minus this many seeds)
 seed_rain_in <- seq(p_eq$seed_rain - dr,
                     p_eq$seed_rain + dr, length.out=31)
 seed_rain_out <- unlist(mclapply(seed_rain_in, run, p_eq))
 
-## Here is input seeds vs output seeds:
+## Here is input seeds vs. output seeds:
 ##+ seeds_in_seeds_out
 plot(seed_rain_in, seed_rain_out, xlab="Incoming seed rain",
      ylab="Outgoing seed rain", las=1, type="l", asp=5)

### 2015-05-19 vignettes/vignette_fitness.R

@@ -20,7 +20,8 @@ w0 <- fitness_landscape(lma, p0)
 plot(lma, w0, type="l", log="x", las=1, ylab="Fitness (empty environment)")
 abline(h=0, col="grey")
 
-## Any trait value along this point can persist:
+## Any trait value along this point can persist, so start with random sample
+## (Set seed for random number generator so that get same results when rerun)
 set.seed(5)
 lma1 <- sort(sample(lma, 4))
 
@@ -40,7 +41,7 @@ matplot(lma, w1, lty=1, type="l", log="x", ylim=c(-5, max(w1)))
 abline(h=0, col="grey")
 points(lma1, rep(0, 4), col=1:4, pch=19)
 
-## For this system, there is an evolutionary branching point around
+## For this system, there is an evolutionary attractor around
 ## lma 0.08:
 lma_b <- 0.08
 p1b <- add_eq(lma_b, p0)

### 2015-05-25 R/ms.R
Emergent properties vignette and html/pdf build system

@@ -7,6 +7,38 @@ pandoc_build <- function(file, template, toc=FALSE, engine="xelatex" ){
     citeproc = TRUE, options = args, verbose = TRUE)
 }
 
+## This works around the (frankly) weird misbehaviour of
+## rmarkdown::pandoc_convert's wd argument.  I may be misusing it, but
+## it seems basically not to work to me.  So instead we'll do the
+## chdir ourselves.
+pandoc_build2 <- function(file, template=NULL, format="pdf") {
+  if (!grep("\\.md$", file)) {
+    stop("Expected a markdown input")
+  }
+
+  if (!is.null(template)) {
+    template <- normalizePath(template, mustWork=TRUE)
+  }
+  owd <- setwd(dirname(file))
+  on.exit(setwd(owd))
+
+  file_local <- basename(file)
+  file_out   <- sub("\\.md$", paste0(".", format), file_local)
+
+  ## Simplify the call a bit, if possible.
+  if (is.null(template)) {
+    args <- NULL
+  } else {
+    wd <- getwd()
+    if (substr(template, 1, nchar(wd)) == wd) {
+      template <- sub("^/+", "",
+                      substr(template, nchar(wd) + 1, nchar(template)))
+    }
+    args <- list(sprintf("--template=%s", template))
+  }
+  pandoc_convert(file_local, output=file_out, options=args, verbose=FALSE)
+}
+
 ms_deps_timestamp <- function(filename) {
   writeLines(as.character(Sys.time()), filename)
 }

### 2015-05-25 vignettes/src/emergent.R

@@ -1,0 +1,114 @@
+## ---
+## title: "tree: A package for modelling plant TRait Ecology & Evolution: _patch level emergent properties_"
+## ---
+
+## The aim here is to use tree to investigate dynamics of plants
+## within a single patch, but to focus on patch level properties,
+## rather than properties of plants within the patch (though I guess
+## that the leaf area index calculations in vignette:patch are also
+## patch level properties).
+library(tree)
+
+p0 <- ebt_base_parameters()
+p0$control$equilibrium_nsteps <- 30
+p0$control$equilibrium_solver_name <- "hybrid"
+p0$disturbance_mean_interval <- 30.0
+
+## Working with a single species at equilibrium
+p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
+p1_eq <- equilibrium_seed_rain(p1)
+data1 <- run_ebt_collect(p1_eq)
+
+## There was a bit of hassle in vignette:patch in plotting all
+## trajectories along with a couple of lines corresponding to focal
+## years.  We're going to do the same thing here:
+closest <- function(t, time) {
+  which.min(abs(time - t))
+}
+last <- function(x) {
+  x[[length(x)]]
+}
+
+times <- c(5, 10, 20, 40, last(data1$time))
+i <- vapply(times, closest, integer(1), data1$time)
+blues <- c("#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6",
+           "#4292C6", "#2171B5", "#08519C", "#08306B")
+cols <- colorRampPalette(blues[-(1:2)])(length(i))
+
+height      <- t(data1$species[[1]]["height", , ])
+log_density <- t(data1$species[[1]]["log_density", , ])
+
+## As with height in vignette:patch, the density is a matrix of time
+## and cohort identity.  However, to show the profiles for a given
+## time slice using `matplot` it is convenient to transpose this
+## matrix (`matplot` works column-by-column so we want each column to
+## to be the state of the system at a given time)
+density <- exp(log_density)
+matplot(height, density, type="l", lty=1,
+        col=make_transparent("black", 0.15),
+        xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
+        log="y")
+
+## Note that the densities here can be *extremely* low, and yet
+## cohorts continue to grow in size; this is the "atto-fox problem",
+## thou here we drop down as low as `r formatC(signif(min(density, na.rm=TRUE) / 1e-24, 1))`
[…]

### 2015-05-25 vignettes/src/equilibrium.R

@@ -1,3 +1,7 @@
+## ---
+## title: "tree: A package for modelling plant TRait Ecology & Evolution: _finding demographic equilibrium_"
+## ---
+
 library(tree)
 library(parallel)
 

### 2015-05-25 vignettes/src/fitness.R

@@ -1,3 +1,7 @@
+## ---
+## title: "tree: A package for modelling plant TRait Ecology & Evolution: _fitness_"
+## ---
+
 library(tree)
 
 ## Start by setting a few parameters; this is the base set of

### 2015-05-25 vignettes/src/patch.R

@@ -1,4 +1,6 @@
-## # Patch level dynamics
+## ---
+## title: "tree: A package for modelling plant TRait Ecology & Evolution: _patch level dynamics_"
+## ---
 
 ## The aim here is to use tree to investigate dynamics of plants
 ## within a single patch.

### 2015-05-25 vignettes/src/plant.R

@@ -1,3 +1,7 @@
+## ---
+## title: "tree: A package for modelling plant TRait Ecology & Evolution: _plant level traits_"
+## ---
+
 ## This vignette covers the sort of analysis in the plant figure in
 ## the manuscript.  It also shows some of the features of working
 ## with plant components of the model.
@@ -27,7 +31,7 @@ pl$mortality
 pl$ode_state
 
 ## To compute rates of change for these variables we need a light
-## environment.  The function \code{fixed_environment} creates an
+## environment.  The function `fixed_environment` creates an
 ## environment that has the same light level (here 100%) at all
 ## heights.  The plant *does not affect* this light environment.
 env <- fixed_environment(1.0)
@@ -87,9 +91,8 @@ pl$height # also 1!
 ## height.
 ##
 ## It's possible to directly integrate the equations exposed by the
-## plant, using the \code{ode_state} field, \code{compute_vars_phys}
-## method and \code{ode_rates} field.  For example, for use with
-## \code{deSolve}:
+## plant, using the `ode_state` field, `compute_vars_phys` method and
+## `ode_rates` field.  For example, for use with `deSolve`:
 derivs <- function(t, y, plant, env) {
   plant$ode_state <- y
   plant$compute_vars_phys(env)

### 2015-05-27 R/tables.R
Convert vignettes to latex

@@ -1,0 +1,12 @@
+
+table_tree_parameters <- function(filename, dest){
+
+  tree_parameters <- read.csv(filename, stringsAsFactors=FALSE)
+  x <- xtable(tree_parameters,
+              hline.after=c(1),
+              align='lp{5cm}lll')
+  y <- print(x, sanitize.text.function=as.character,
+             include.rownames=FALSE, floating=FALSE,
+             print.results=FALSE)
+  cat(y, file=dest)
+}

### 2015-05-27 ms.tex
Update formatting to my usual template

@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{graphicx}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-
-\usepackage[verbose,tmargin=2cm,bmargin=2cm,lmargin=3.5cm,rmargin=2.5cm]{geometry}
-
-\usepackage{color}
-\definecolor{grey}{rgb}{0.5, 0.5, 0.5}
-
-\usepackage[unicode=true]{hyperref}
-\urlstyle{same}  % don't use monospace font for urls
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={Daniel Falster, Rich FitzJohn, Mark Westoby},
-            pdftitle={\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution},
-            colorlinks=true,
-            citecolor=grey,
-            urlcolor=grey,
-            linkcolor=grey,
-            pdfborder={0 0 0}}
-
-\usepackage{lineno}
-\linespread{2}
+\documentclass[a4paper,11pt]{article}
+\usepackage[osf]{mathpazo}
+\usepackage{ms}
+\usepackage[sort&compress]{natbib}
+\raggedright
 
@@
 
-\usepackage{authblk}
 \title{\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution}
-\author[1]{Daniel S. Falster}
-\author[1]{Richard G. FitzJohn}
-\author[1]{Mark Westoby}
-\affil[1]{Dept of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}
+\author{Daniel S. Falster, Richard G. FitzJohn, and Mark Westoby}
+\affiliation{
+Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
+Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
+A manuscript in consideration as a research paper for
+publication in MEE as part of the Special Feature \emph{Demography
+  beyond the Population}.}
 \date{}
 
-\usepackage[sort&compress]{natbib}
 \bibliographystyle{mee}
@@
 
+\mstype{Applications Note}
+\runninghead{TREE}
+% TODO: need more keywords here
+\keywords{demography}
+
 \begin{document}
-\maketitle
-
-\noindent A manuscript in consideration as a research paper for publication in MEE
-as part of the Special Feature \emph{Demography beyond the Population}.
-
+\mstitlepage
+\noindent
+\parindent=1.5em
+\addtolength{\parskip}{.3em}
+\doublespacing
 \linenumbers
-
-\section{Abstract}\label{abstract}
+\section{Summary}\label{abstract}
 \begin{enumerate}

### 2015-05-27 R/tables.R
Bold some rows of table

@@ -1,8 +1,13 @@
 
 table_tree_parameters <- function(filename, dest){
 
-  tree_parameters <- read.csv(filename, stringsAsFactors=FALSE)
-  x <- xtable(tree_parameters,
+  xx <- read.csv(filename, stringsAsFactors=FALSE)
+
+  to_bold <- which(xx$Symbol =="" )
+
+  xx$Description[to_bold] <- sprintf("\\textbf{%s}", xx$Description[to_bold])
+
+  x <- xtable(xx,
               hline.after=c(1),
               align='lp{5cm}lll')
   y <- print(x, sanitize.text.function=as.character,

### 2015-05-27 R/tables.R
Tweaks to physiology and demography vignettes
For #9, #8

Germination survival
Rename pi_0 -> S_D, pi_1 -> S_G
Spacing
Harmonise variable names

@@ -9,7 +9,7 @@ table_tree_parameters <- function(filename, dest){
 
   x <- xtable(xx,
               hline.after=c(1),
-              align='lp{5cm}lll')
+              align='lp{7cm}lll')
   y <- print(x, sanitize.text.function=as.character,
              include.rownames=FALSE, floating=FALSE,
              print.results=FALSE)

### 2015-05-27 R/ms.R
Allow for xelatex in build

@@ -34,17 +34,20 @@ pandoc_build <- function(file, template=NULL, format="pdf", ...) {
 ## TODO: options for saying what command is being run
 latex_build <- function(filename, bibliography=NULL,
                         chdir=TRUE, interaction="nonstopmode",
-                        max_attempts=5L, clean=FALSE) {
+                        max_attempts=5L, clean=FALSE, engine="pdflatex") {
   if (chdir && dirname(filename) != "") {
     owd <- setwd(dirname(filename))
     on.exit(setwd(owd))
     filename <- basename(filename)
   }
 
-  res <- run_latex(filename, interaction)
+  res <- run_latex(filename, interaction, engine)
+  if(engine=="xelatex") {
+      res <- run_latex(filename, interaction, engine)    
+  }
   if (!is.null(bibliography)) {
     run_bibtex(filename)
-    res <- run_latex(filename, interaction)
+    res <- run_latex(filename, interaction, engine)
   }
 
   pat <- c("Rerun to get cross-references right",
@@ -54,7 +57,7 @@ latex_build <- function(filename, bibliography=NULL,
   }
   for (i in seq_len(max_attempts)) {
     if (any(vapply(pat, isin, logical(1), res))) {
-      res <- run_latex(filename, interaction)
+      res <- run_latex(filename, interaction, engine)
     } else {
       break
     }
@@ -75,11 +78,11 @@ latex_clean <- function(filename) {
   file.remove(aux[file.exists(aux)])
 }
 
-run_latex <- function(filename, interaction="nonstopmode") {
+run_latex <- function(filename, interaction="nonstopmode", engine="pdflatex") {
   args <- c(paste0("-interaction=", interaction),
             "-halt-on-error",
             filename)
-  callr::run_system(Sys_which("pdflatex"), args)
+  callr::run_system(Sys_which(engine), args)
 }
 
 run_bibtex <- function(filename) {

### 2015-05-28 ms.tex
update schematic
For #14
Build from parts via tikz
Change fonts
Feedback from James

@@
   extensible framework for modelling plant demography across the entire
-  life cycle via coupled differential equations. At its core, \texttt{TREE} is an
-  individual-based mechanistic model where plant function is mediated by
+  life cycle and in structured metapopulations, via coupled differential equations.
+  At its core, \texttt{TREE} is an
+  individual-based mechanistic model where plant physiology and demography is mediated by
   traits. Individuals from multiple species can be grown in isolation,
@@
   hyper-parameterisation on the basis of plant functional traits. A
-  detailed test suite is provided to ensure accuracy.
+  detailed test suite is provided to ensure correct behaviour of the code.
 \item
@@
 an individual increases its leaf area, its potential to generate
-photosynthetic also rises. Yet for larger individuals,  increasing
-fractions of photosynthetic income must also be diverted towards
+photosynthate also rises. Yet as individuals get larger, an increasing
+fraction of photosynthate must be diverted towards
 building and maintaining support tissues rather than leaves
-\citep{Givnish-1988, Enquist-2007}. A larger fraction of income is also
-allocated to reproduction \citep{Thomas-2011}. Consequently at
-population level, rates of growth and mortality change with size
+\citep{Givnish-1988, Enquist-2007}, and reproduction \citep{Thomas-2011}. Consequently 
+rates of growth and mortality change with individual size
 \citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
@@
 \citep{Westoby-2002}. Over longer time-frames, natural selection favours
-traits making plants more competitive such as faster growth, larger
-seeds, taller height \citep{Falster-2003}. Selective forces thus
-potentially amplify size-related effects on demography. Accounting for
-the effects of size and traits in models therefore seems like a key step
-in developing theory for the structure and dynamics of the biosphere
-\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular notable
+traits that provide a competitive advantage, such as faster growth, larger
+seeds or taller height \citep{Falster-2003}.
+Consequently, selective forces may amplify size-related effects on demography.
+Accounting for the effects of size and traits in models is therefore fundamental to developing theory on the structure and dynamics of the biosphere
+\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular for notable
 phenomena such as successional dynamics, self thinning, trait evolution
@@
 Although the importance of size-structure for vegetation function and
-demography has been widely recognised for a long time
+demography has been long been recognised
 \citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
@@
 most large-scale mechanistic models of plant growth either omit it
-entirely, or include only coarse approximations
+entirely, or only include coarse approximations
 \citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
@@
 parameterised in terms of growth and demography, and consequently did a
-poor job of simulating carbon fluxes. The next generation of models
+poor job of simulating carbon fluxes \todo{ref}. The next generation of models
 focussed more on physiology and, presumably for computational reasons,
@@
 current perplexing situation where all but a couple
-\citep{Moorcroft-2001, Smith-2014} of the leading vegetation
-models lack size-structure. These models
+\citep{Moorcroft-2001, Smith-2014} of the most popular vegetation
+models lack size-structure. Instead, most models
 are focused on modelling fluxes carbon and water, but say little about
@@
 It is common to assume competitive interactions are influenced by
-size-related traits such as adult and seed size, but detailed
+size-related traits such as adult size and seed size, but detailed
 size-structured demography is rarely considered, at least for plant
-models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Thus it has
+models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has
 been difficult to connect mechanisms described in these simplistic
@@
 coexistence of trait mixtures \citep{Falster-2015}. These studies
-provided the foundation methods needed; here we provide robust and
+provided the foundation methods needed \todo{needed for what.. its hinted above... maybe rearrange the sentences} here we provide robust and
 extensible implementations. We describe the general approach of the
@@
 \texttt{TREE} is a mechanistic model, meaning the dynamics of the system arise
-from rules specifying on how individuals grow and interact. In ecology,
+from rules specifying how individuals grow and interact. In ecology,
 there is a long history of using simple deterministic models
@@
 \ref{fig:schematic}). Demographic phenomena can be studied at the three
-levels of individual plants, stands of competing plants, and entire
+levels: individual plants, stands of competing plants, and entire
 metapopulations.
@@
 estimates rate of biomass production for a plant, given its size, light
-environment, and the supplied physiological constants. Assimilation is
+environment, and the supplied physiological constants. Gross photosynthetic 
+income is
 estimated from total leaf area and the light distribution across the
@@
 The core job of the physiological model is to take size, light
-environment and parameters as inputs, and to return growth rate,
+environment and physiological parameters as inputs, and to return growth rate,
 mortality and fecundity as outputs, in other words to move from
@@
 growth also to be estimated. To properly model trait variation,
-different parameters in the model are linked via assumed tradeoffs. For
+different parameters (such as rates of leaf turnover, respiration, photosynthesis 
+per leaf area) in the model are linked via assumed tradeoffs. For
 example, the trait LMA is used to estimate the rate of leaf turnover,
@@
 Appendix \ref{sec:demography} for more
-details). As the plant grows, allocation to different types varies, and
+details). As the plant grows, allocation to different tissue types varies, and
 the composition of the plant changes to include more stem and less
@@
 plant light compensation point (WPLCP) can be computed (Fig.
-\ref{fig:plant}d). The WPLCP is increasingly regarded as the most useful
+\ref{fig:plant}d). The WPLCP is the light level at which the plant 
+stops growing, and is increasingly regarded as the most useful
 measure of a strategy's shade tolerance
@@
 across a large number of small patches \citep{Moorcroft-2001}. Ideally
-one would also consider full spatial interactions within patches,
+one would also consider spatial interactions within patches,
 however such models are very computationally demanding, and also
@@
 
-Our approach for modelling solving size-structured population dynamics
+Our approach for modelling size-structured population dynamics
 is based on the Escalator Boxcar Train technique (EBT)
@@
 distribution of cohort introduction times across patch age (Fig.
-\ref{fig:patch}a).
+\ref{fig:patch}a), generally with tighter spacing of cohorts in 
+younger patches.
 
@@
 \citep[e.g.][]{Barnes-2004, Coomes-2007}, \texttt{TREE} accommodates multiple
-different species with different traits (Fig. \ref{fig:schematic}b; Fig.
+ species with different traits (Fig. \ref{fig:schematic}b; Fig.
 \ref{fig:patch}).
@@
 disturbances remove all established vegetation. The frequency-density
-\(p(a)\) of patches age \(a\) in the landscape is needed. Assuming there
+\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
 are a large number of patches means the dynamics of \(p\) behave
@@
 weighting the temporal dynamics of a single patch by \(p(a)\), the
-relative abundance of patches age \(a\) in the metapopulation.
+relative abundance of patches aged \(a\) in the metapopulation.
 
@@
 to metapopulation is to identify the equilibrium seed rain of the
-metapopulation. For a given input seed rain, the model produces an
-output seed rain. The equilibrium seed rain is the vector of seed rain
+metapopulation. For a given seed rain input, the model produces a seed rain output. 
+The equilibrium seed rain is the vector of seed rain
 where input seed rain equals output seed rain. Finding this equilibrium
@@
 vegetation such as total leaf area, biomass, productivity and allocation
-to different plant parts arise as emergent properties (Fig.
+to different plant parts also arise as emergent properties (Fig.
 \ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
@@
 individual with particular traits to establish in a community of
-established residents at their equilibrium densities\citep{Metz-1992}.
+established residents at their equilibrium densities \citep{Metz-1992}.
 Invasion fitness
@@
 The current parametrisation of the physiological model is based on
-\citet{Falster-2011}. These values were drawn from available literature (see
+\citet{Falster-2011}. Parameter values were drawn from available literature (see
 Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
@@
 \centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/schematic}
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/schematic}
 
@@
 \centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/plant.pdf}
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
 \caption{Physiological dynamics for individual plants vary with size, trait
@@
 growth rate is zero (x intercept).  The light level used in panels (a) and (c)
-is indicated with dots.}
+is indicated with dots.} [Could you put delta symbols on the y axis rather than d?
+Also there is alot going in these figures with 4 lines each.
+You could make it easier for the reader by adding the dashed lines to the keys]
 \label{fig:plant}
@@
 \centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/patch.pdf}
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
 \caption{Population dynamics for two species competing within a patch. Lines
-in panels (a) and (b) show the growth trajectories for individuals on the
-boundary of successive "cohorts" for each species. A single cohort includes
-all individuals born within a given time period. Red lines are the low LMA
+in panels (a) same growth trajectories for individuals on the
+boundary of successive "cohorts" for each species, using a schedule of 
+introduction times generated by an adaptive refinement algorithm.
+ Red lines are the low LMA
 species from figure \ref{fig:plant}, blue lines are the high LMA species.  The
@@
 suppressing its growth.  Self-thinning leaves space for the high LMA species
-to establish below the canopy of the fast species. Panel (a) shows all
-trajectories equally weighted; the spacing is generated by an adaptive
-refinement algorithm.  In panel (b) trajectories are shaded by the density of
-individuals at given size and patch age.  (c) The leaf area index at ground
+to establish below the canopy of the fast species. Panel (b) shows the same
+trajectories as in panel (a), but with shading indicating the density $n$ of
+individuals at given size and patch age (light = low density, dark = high density).  
+(c) The leaf area index at ground
 level for each species.  The dotted line is the total leaf area index (sum
@@
 \centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/emergent.pdf}
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/emergent.pdf}
 \caption{Emergent size-distributions across a metapopulation.
@@
 Panels show (a) patterns of abundance, (b) leaf area and (c) height growth rate against size.
-Note that units of panels (a) and (b) are per unit height and per unit ground area.
-}
+Note that units of panels (a) and (b) are per unit height and per unit ground area, respectively.}
 \label{fig:emergent}
@@
 \centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/fitness.pdf}
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
 \caption{Example fitness landscape for LMA, showing potential for stable

### 2015-05-28 R/figures_emergent.R
Rename from tree -> plant for the vignettes

@@ -29,10 +29,10 @@ figure_emergent <- function(data) {
   ## have leaf areas for all the species.
   tmp <- lapply(seq_along(data$time), ebt_patch, data)
   xxx <- lapply(tmp, function(x) x$species[[1]]$area_leafs)
-  leaf_area <- do.call("cbind", tree:::pad_matrix(xxx))
+  leaf_area <- do.call("cbind", plant:::pad_matrix(xxx))
 
   xxx <- lapply(tmp, function(x) matrix(x$species[[1]]$ode_rates, 4)[1, ])
-  growth_rate <- do.call("cbind", tree:::pad_matrix(xxx))
+  growth_rate <- do.call("cbind", plant:::pad_matrix(xxx))
 
   op <- par(mfrow=c(3, 1), mar=c(2.1, 4.1, .5, .5), oma=c(2, 0, 0, 0))
   on.exit(par(op))

### 2015-05-28 R/figures_plant.R

@@ -1,8 +1,3 @@
-## Eventually, this will be part of a vignette showing how to drive
-## `tree` at the plant level.  Unfortunately, there's a lot of
-## book-keeping to keep track of, so it's not that nice at this
-## point.
-
 ## First, growth trajectories of plants
 default_parameters <- function() {
   p <- ebt_base_parameters()

### 2015-05-28 R/tables.R

@@ -1,5 +1,4 @@
-
-table_tree_parameters <- function(filename, dest){
+table_plant_parameters <- function(filename, dest){
 
   xx <- read.csv(filename, stringsAsFactors=FALSE)
 

### 2015-05-28 vignettes/src/cohort_spacing.R

@@ -1,8 +1,12 @@
+## ---
+## title: "plant: A package for modelling forest trait ecology & evolution: _cohort spacing algorithm_"
+## ---
+
 ## This vignette shows some details of cohort splitting.  It's
 ## probably not very interesting to most people.  It also uses a lot
-## of non-exported, non-documented functions from tree so you'll see a
-## lot of `tree:::` prefixes.
-library(tree)
+## of non-exported, non-documented functions from plant so you'll see
+## a lot of `plant:::` prefixes.
+library(plant)
 library(parallel)
 
 p0 <- ebt_base_parameters()
@@ -68,7 +72,7 @@ run_with_insert <- function(i, p, t) {
   run_with_times(p, insert_time(i, t))
 }
 
-## The internal function `tree:::run_ebt_error` runs the EBT and
+## The internal function `plant:::run_ebt_error` runs the EBT and
 ## computes errors as the integration proceeds and helps shed some
 ## light.
 i <- seq_len(length(t1) - 1)
@@ -85,7 +89,7 @@ abline(h=0, col="grey")
 plot(tm, res - sr_1, xlab="Time (years)", ylab="Seed rain", las=1)
 abline(h=0, col="grey")
 
-ebt <- tree:::FFW16_EBT(p)
+ebt <- plant:::FFW16_EBT(p)
 ebt$run()
 
 ## This is the contribution of different cohorts to see rain (x axis
@@ -116,7 +120,7 @@ dat2 <- run_ebt_collect(p2)
 ## and red in regions where the the light environment is darker in the
 ## refined environment.
 f <- function(e, h) {
-  i <- tree:::Interpolator()
+  i <- plant:::Interpolator()
   i$init(e[, 1], e[, 2])
   y <- rep(1, length(h))
   j <- h < i$max
@@ -139,7 +143,7 @@ image(dat1$time, h, dy, xlab="Time (years)", ylab="Height (m)", las=1,
 ## Because the differences are mostly manifest in the leaf area, we
 ## monitor error in both the leaf area and in fitness for all cohorts:
 ## (black line indicates the cohort identified as problematic above)
-dat <- tree:::run_ebt_error(p)
+dat <- plant:::run_ebt_error(p)
 image(dat1$time, dat1$time, dat$err$lai[[1]],
       xlab="Patch age (years)", ylab="Introduction time (years)", las=1)
 abline(h=t1[j[[1]]])

### 2015-05-28 vignettes/src/emergent.R

@@ -1,13 +1,13 @@
 ## ---
-## title: "tree: A package for modelling plant TRait Ecology & Evolution: _patch level emergent properties_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _patch level emergent properties_"
 ## ---
 
-## The aim here is to use tree to investigate dynamics of plants
+## The aim here is to use plant to investigate dynamics of plants
 ## within a single patch, but to focus on patch level properties,
 ## rather than properties of plants within the patch (though I guess
 ## that the leaf area index calculations in vignette:patch are also
 ## patch level properties).
-library(tree)
+library(plant)
 
 p0 <- ebt_base_parameters()
 p0$control$equilibrium_nsteps <- 30
@@ -82,10 +82,10 @@ text(height[1, i] + strwidth("x"), density[1, i],
 ## It's also possible see where the leaf area in the patch is coming
 ## from; a profile of leaf area with respect to height.  Again, this
 ## requires reconstructing the patches, and using an unexported
-## function from `tree` to put this into a matrix:
+## function from `plant` to put this into a matrix:
 patch <- lapply(seq_along(data1$time), ebt_patch, data1)
 leaf_area <- lapply(patch, function(x) x$species[[1]]$area_leafs)
-leaf_area <- do.call("cbind", tree:::pad_matrix(leaf_area))
+leaf_area <- do.call("cbind", plant:::pad_matrix(leaf_area))
 
 matplot(height, leaf_area, type="l", lty=1, col="lightgrey",
         xlim=xlim, xlab="Height (m)",
@@ -98,7 +98,7 @@ text(height[1, i] + strwidth("x"), leaf_area[1, i],
 
 growth_rate <- lapply(patch, function(x)
                       matrix(x$species[[1]]$ode_rates, 4)[1, ])
-growth_rate <- do.call("cbind", tree:::pad_matrix(growth_rate))
+growth_rate <- do.call("cbind", plant:::pad_matrix(growth_rate))
 
 ## Finally, we can see where height growth rate is concentrated in the
 ## population.  This differs from the profile in vignette:plant

### 2015-05-28 vignettes/src/equilibrium.R

@@ -1,8 +1,8 @@
 ## ---
-## title: "tree: A package for modelling plant TRait Ecology & Evolution: _finding demographic equilibrium_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _finding demographic equilibrium_"
 ## ---
 
-library(tree)
+library(plant)
 library(parallel)
 
 run <- function(seed_rain_in, p) {
@@ -111,7 +111,7 @@ abline(v=p2_eq$seed_rain, col=1:2, lty=3)
 ## multidimenensional root finding is difficult; even knowning that
 ## there are roots is not straightforward, much less proving that
 ## we've converged on the "correct" root (for example [0,0] is a root
-## in this case but that root is not stable).  tree uses some
+## in this case but that root is not stable).  plant uses some
 ## heuristics to try to ensure that the root returned is an attracting
 ## point but sequentially applying rounds of iteration and nonlinear
 ## root finding algorithms, as well as rescaling seed rains to repel

### 2015-05-28 vignettes/src/fitness.R

@@ -1,8 +1,8 @@
 ## ---
-## title: "tree: A package for modelling plant TRait Ecology & Evolution: _fitness_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _fitness_"
 ## ---
 
-library(tree)
+library(plant)
 
 ## Start by setting a few parameters; this is the base set of
 ## parameters we'll use.

### 2015-05-28 vignettes/src/patch.R

@@ -1,10 +1,10 @@
 ## ---
-## title: "tree: A package for modelling plant TRait Ecology & Evolution: _patch level dynamics_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _patch level dynamics_"
 ## ---
 
-## The aim here is to use tree to investigate dynamics of plants
+## The aim here is to use plant to investigate dynamics of plants
 ## within a single patch.
-library(tree)
+library(plant)
 
 p0 <- ebt_base_parameters()
 p0$control$equilibrium_nsteps <- 30

### 2015-05-28 vignettes/src/plant.R

@@ -1,11 +1,11 @@
 ## ---
-## title: "tree: A package for modelling plant TRait Ecology & Evolution: _plant level traits_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _plant level properties_"
 ## ---
 
 ## This vignette covers the sort of analysis in the plant figure in
 ## the manuscript.  It also shows some of the features of working
 ## with plant components of the model.
-library(tree)
+library(plant)
 
 ## Plants are constructed with the `FFW16_Plant` function.  That
 ## function takes as its only argument a "strategy" object; the
@@ -70,7 +70,7 @@ zapsmall(unlist(pp$internals))
 
 ## This PlantPlus object also includes diameter and heartwood as two
 ## more variables for the ODE system (this might move into Plant
-## soon -- see [this issue](https://github.com/traitecoevo/tree/issues/139)).
+## soon -- see [this issue](https://github.com/traitecoevo/plant/issues/139)).
 pp$ode_names
 
 ## Plants are a type of *reference object*.  They are different to
@@ -131,7 +131,7 @@ plot(height ~ tt, res$state, type="l", las=1,
      xlab="Time (years)", ylab="Height (m)")
 points(height ~ time, yy, col="grey", cex=.5)
 
-## Completing the set, `tree` also provides a function for growing
+## Completing the set, `plant` also provides a function for growing
 ## plants to a particular size; `grow_plant_to_size`.  This takes
 ## *any* size measurement in the plant and can grow the plant to that
 ## size.  So, for height:
@@ -164,7 +164,7 @@ mass <- seq_log(pl$internals$mass_above_ground + 1e-8, 1000, length.out=21)
 res_mass <- grow_plant_to_size(pl, mass, "mass_above_ground", env,
                                time_max=100, warn=FALSE)
 ## (this seems on the low side - see [this
-## issue](https://github.com/traitecoevo/tree/issues/142)).
+## issue](https://github.com/traitecoevo/plant/issues/142)).
 plot(res_mass$time, mass, type="o", pch=19, las=1, xlab="Time (years)")
 
 ## With all these bits in place, let's look at growth trajectories of

### 2015-05-28 ms.tex
Move keywords up
More acknowledgements

@@
 \usepackage{ms}
-\usepackage[sort&compress]{natbib}
+\usepackage[]{natbib}
 \raggedright
+
+\newcommand{\plant}{\textsc{plant}}
 
@@
 
-\title{\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution}
+\title{{\plant}: A package for modelling forest trait ecology \& evolution}
 \author{Daniel S. Falster, Richard G. FitzJohn, and Mark Westoby}
@@
 \mstype{Applications Note}
-\runninghead{TREE}
-% TODO: need more keywords here
-\keywords{demography}
+\runninghead{plant}
+\keywords{demography, emergent, fitness, growth,
+physiology, metapopulation, mortality, reproduction, size-structure,
+tradeoff}
 
@@
 \item
-  Here we introduce the \texttt{TREE} package (TRait Ecology and Evolution), an
-  extensible framework for modelling plant demography across the entire
-  life cycle and in structured metapopulations, via coupled differential equations.
-  At its core, \texttt{TREE} is an
+  Here we introduce the {\plant} package, an
+  extensible framework for modelling trait-based demography, ecology and evolution 
+  in vegetation, via coupled differential equations.
+  At its core, {\plant} is an
   individual-based mechanistic model where plant physiology and demography is mediated by
@@
   disturbance regime. Dynamics within patches of competing plants are
-  resolved using novel extensions of the Escalator Boxcar Train
+  resolved using novel extensions of the ``Escalator Boxcar Train''
   technique. Combined effects of trait-, size- and patch-structured
@@
 \item
-  \texttt{TREE} is presented as an open source \texttt{R} package and is
-  available at
-  \href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
-  While accessed from R, the core routines in \texttt{TREE} are written in C++.
+  {\plant} is an open source R package and is available at
+  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
+  While accessed from R, the core routines in {\plant} are written in C++.
   The package provides for alternative physiologies and for
-  hyper-parameterisation on the basis of plant functional traits. A
+  hyper-parameterisation to capture trade-offs among parameters. A
   detailed test suite is provided to ensure correct behaviour of the code.
 \item
-  We provide worked examples illustrating \texttt{TREE}'s use to study the
+  We provide worked examples illustrating {\plant}'s use to study the
   influence of functional traits on growth of individual plants, of
@@
 
-\textbf{Key-words}: demography, emergent property, fitness, growth,
-physiology, metapopulation, mortality, reproduction, size-structure,
-tradeoff
-
 \section{Introduction}\label{introduction}
 
-Plant growth and demography are fundamentally size- and trait- structured,
-influencing dynamics over time-scales ranging from instantaneous physiological
-effects to long-term evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As
-an individual increases its leaf area, its potential to generate
-photosynthate also rises. Yet as individuals get larger, an increasing
-fraction of photosynthate must be diverted towards
-building and maintaining support tissues rather than leaves
-\citep{Givnish-1988, Enquist-2007}, and reproduction \citep{Thomas-2011}. Consequently 
-rates of growth and mortality change with individual size
-\citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
-a species also interact with size to shape its demography. Leaf
-construction costs and wood density affect growth rates
-\citep{Wright-2010}. Other traits such as seed size and height at
+Plant growth and demography are fundamentally size- and trait-
+structured, influencing dynamics over time-scales ranging from
+instantaneous physiological effects to long-term evolutionary outcomes
+\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
+area, it increases its potential to generate photosynthate.  In
+contrast, as individuals grow, they must use increasing fractions of
+this photosynthetic income to building and maintaining support tissues
+rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
+reproduction \citep{Thomas-2011}. Consequently rates of growth and
+mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
+%
+The patterns of changing growth rate vary by plant traits; leaf
+construction costs and wood density affect growth rates and rates of
+tissue replacement \citep{Wright-2010}, while seed size and height at
 maturation determine start and end points of ontogenetic trajectories
-\citep{Westoby-2002}. Over longer time-frames, natural selection favours
-traits that provide a competitive advantage, such as faster growth, larger
-seeds or taller height \citep{Falster-2003}.
-Consequently, selective forces may amplify size-related effects on demography.
-Accounting for the effects of size and traits in models is therefore fundamental to developing theory on the structure and dynamics of the biosphere
-\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular for notable
-phenomena such as successional dynamics, self thinning, trait evolution
-and species coexistence.
+\citep{Westoby-2002}
+%
+Strong feedbacks emerge between individuals within a forest due to
+competition for light and other resources, such that the growth rate
+of one individual depends on the size and traits of nearby
+individuals.  This makes it difficult to predict demographic processes
+without modelling both individual growth and the interactions between
+individuals.
+% TODO[RGF]: link back things about natural selection, successional
+% dynamics, self thinning, trait evolution and species coexistance.
 
@@
 demography has been long been recognised
-\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
-Kohyama-1993, Enquist-2007, Pacala-1996, Coomes-2007},
+\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987},
 most large-scale mechanistic models of plant growth either omit it
 entirely, or only include coarse approximations
-\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
+\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. The likely
 reason is the computational
-challenge of modelling size-structured dynamics. During the 1980s and
+challenge of modelling size-structured dynamics. 
+% TODO[RGF]: I don't see the link with the above?  Can this be
+% shortened/tightened?  It's too long for a 3k word paper
+During the 1980s and
 early 1990s, size-structured ``gap'' models were widely used
-\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
-parameterised in terms of growth and demography, and consequently did a
-poor job of simulating carbon fluxes \todo{ref}. The next generation of models
-focussed more on physiology and, presumably for computational reasons,
+\citep[e.g.][]{Shugart-1980,Huston-1987}. These gap models were
+parameterised in terms of growth using statistical functions; this meant 
+that data had to be collected for each species, limiting widespread application. 
+The next generation of models focussed on establishing core physiology for 
+a limited number of functional types and
 sacrificed details of demography in order to achieve global coverage
-\citep{Haxeltine-1996}. This history has brought us to the
-current perplexing situation where all but a couple
-\citep{Moorcroft-2001, Smith-2014} of the most popular vegetation
+\citep{Woodward-1995,Haxeltine-1996}. This history has brought us to the
+current situation where all but a couple
+\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
 models lack size-structure. Instead, most models
 are focused on modelling fluxes carbon and water, but say little about
-demographic behaviour, and consequently are also unable to account for
+demographic behaviour; consequently they are also unable to account for
 any non-linear demographic feedbacks that may arise.
+% While many vegetation models still lack size structure, the importance of
+% size-structure for modelling phenomena such as allocation and production is
+% being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
 
-Interestingly, most models used in theoretical ecology to explore
+Similarly, most models used in theoretical ecology to explore
 questions about species coexistence also lack size-structured dynamics
-\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013, Levin-1974,MacArthur-1967,Tilman-1985}.
-It is common to assume competitive interactions are influenced by
-size-related traits such as adult size and seed size, but detailed
-size-structured demography is rarely considered, at least for plant
-models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has
-been difficult to connect mechanisms described in these simplistic
-models with field data and real traits, which are almost always strongly
-size-structured.
+\citep[e.g.][]{MacArthur-1967,Levin-1974,Tilman-1985,Geritz-1995, 
+Calcagno-2006, Leimar-2013}.  It is common to assume
+competitive interactions are influenced by size-related traits such as
+adult size and seed size, but detailed size-structured demography is
+rarely considered, at least for plant models \cite[for animal
+examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
+difficult to connect mechanisms described in these simplistic models
+with type of data often collected in the field, such as physiological
+leaf traits and demographic data on growth and mortality in long-term
+monitoring plots \citep{Adler-2013}.
 
-In this note, we describe the \texttt{TREE} package for R
-\citep{R-2015}, a mechanistic framework for studying the effects of size
-structure and trait variation on the demography of individual plants, of
-patches of competing plants, and of meta-populations structured by a
-prevailing disturbance regime. The package came about as a way to extend
-previous work investigating the effect of traits on vegetation
-properties \citep{Falster-2011} and the mechanisms facilitating
-coexistence of trait mixtures \citep{Falster-2015}. These studies
-provided the foundation methods needed \todo{needed for what.. its hinted above... maybe rearrange the sentences} here we provide robust and
-extensible implementations. We describe the general approach of the
-package and then a series of use cases, focussing at different levels of
-ecological organisation. Each section provides a short description of
-the required methods, design considerations, and then some worked
-examples illustrating potential use cases and design features. Detailed
-technical documents are provided as supplementary information, and updated
-versions of technical documents are also available within the \texttt{TREE}
-package itself. The use cases highlighted reflect our own interests in
-understanding trait-based demography and community assembly. Yet we
-expect the methods provided to be useful in other contexts. For example,
-prior to \texttt{TREE} there was no \texttt{R} package that allowed one to simply grow a
-plant over time, based on well-understood physiology. Surely this is one
-of the simplest demographic actions one might hope for, with diverse
-potential applications in ecology, evolution, forestry, agriculture, and
-vegetation modelling. We hope the mechanistic platform provided here
-will complement the increasingly sophisticated statistical methods now
-available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.
+In this note, we introduce the {\plant} package for R \citep{R-2015};
+a mechanistic framework for studying the effects of size structure and
+trait variation on the demography of individual plants, of patches of
+competing plants and of meta-populations structured by a prevailing
+disturbance regime.
+%
+The {\plant} package builds on much existing work investigating the
+effect of traits on vegetation properties and the mechanisms
+facilitating coexistence of trait mixtures
+\citep{Kohyama-1993,Deroos-1997,Moorcroft-2001,Falster-2011,Falster-2015}.
+
+% TODO[RGF]: restructure here - I'm getting lost
+{\plant} provides a platform for investigating how morphological and
+physiological traits influence growth rate and shade tolerance, thus
+potentially giving a theoretical foundation for observed empirical
+patterns \citep{Baltzer-2007,Wright-2010}.
+
+% NOTE[RGF]: I dropped the appeal to novelty (growing an individual
+% plant) because I don't find them very convincing.
+While the applications highlighted reflect our own interests in understanding
+trait-based demography and community assembly, we expect the methods provided
+to be useful in variety of contexts. . Being driven by traits,
+the model can be extended to potentially very many species. Users can also 
+add their own physiological models. Individual plants
+can then be  situated in a patch with competitors and the demography of a
+competing population  investigated, thereby capturing phenomena such as multi-
+species self-thinning,  successional transitions and age-related changes in
+vegetation productivity. {\plant}  models the emergent behaviour of
+patches deterministically, based on a continual  size-density distribution. As
+such, it is not required to specify a particular patch size.  Finally,
+{\plant} allows for emergent phenomena across a metapopulation of patches
+to  be investigated, thereby providing a workable mechanism to scale from
+individual plant to  vegetation, incorporating the non-linear effects of size
+and traits on vegetation outcomes. 
+% To strigul?
+In  this sense {\plant} is similar to
+the \textsc{Ecosystem Demography} \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
+\citep{Smith-2014} models. However, whereas these other models were designed
+to primarily capture large-scale regional biogeochemistry driven by climatic
+variables,  {\plant} was designed so that the rules driving the
+demography of individual trees, patches and metapopulations could be
+manipulated and studied in detail. The three models are therefore highly 
+complimentary, they also connect with the increasingly sophisticated statistical 
+methods now available to analyse size-structured data
+\citep[e.g.][]{Metcalf-2013}.
+
+We describe the general approach of the package and then sections with
+details at different levels of ecological organisation. Each section
+provides a short description of the required methods and then example 
+applications. Detailed technical
+documents are provided as supplementary information (see Appendices),
+and updated versions of technical documents are also available within
+the {\plant} package itself.
 
@@
 
-\texttt{TREE} is a mechanistic model, meaning the dynamics of the system arise
-from rules specifying how individuals grow and interact. In ecology,
-there is a long history of using simple deterministic models
-to understand biological phenomena, such as Lotka-Volterra population dynamics
-\citep{MacArthur-1967, Leimar-2013} .
-\texttt{TREE} was developed with this style of analysis in mind, while also
-allowing for a richer set of ecological dynamics than is possible in
-population models lacking size structure. The package implements methods
-for physiological, population, and adaptive dynamics (Fig.
-\ref{fig:schematic}) following methods described by \citet{Falster-2011}
-and \citet{Falster-2015}. The core rules in \texttt{TREE} are about the
-short-term physiological functioning of an individual plant and how
-these are influenced by its traits, size and light environment. Dynamics
-at higher levels of organisation then arise as emergent properties,
-driven by growth physiology, competition for light and disturbance (Fig.
-\ref{fig:schematic}). Demographic phenomena can be studied at the three
-levels: individual plants, stands of competing plants, and entire
-metapopulations.
+% NOTE: Keep the repetition with above under control.
+
+{\plant} is a mechanistic model, meaning the dynamics of the system
+arise from rules specifying how individuals grow and interact.  The
+core rules in {\plant} are about the short-term physiological
+functioning of an individual plant and how these are influenced by its
+traits, size and light environment.
+%
+On top of this physiological model, the package implements methods for
+population and adaptive dynamics (Fig.  \ref{fig:schematic}) following
+methods described by \citet{Falster-2011} and
+\citet{Falster-2015}. 
+%
+The dynamics at higher levels of organisation arise as emergent
+properties, driven by growth physiology, competition for light and
+disturbance (Fig.  \ref{fig:schematic}). Demographic phenomena can be
+studied at the three levels: individual plants, stands of competing
+plants and entire metapopulations.
+%
+We developed {\plant} as an extended thought experiment for helping to
+understand competition and succession by having these driven by
+low-level processes rather than being explicitly implemented in the
+model.
+% NOTE: I think that foreshaddowing some of the competition kernels
+% idea here is sensible.
+%
+% TODO: Somewhere, perhaps above, perhaps highlight the issue of doing
+% coexistance modelling with explicit modelling - the models are so
+% well understood it's easy to drive the behaviour.  But we want to
+% have something where we just get coexistence to study how tradeoffs
+% generate it.
+%
+% TODO: perhaps salavageable?
+%
+%   {\plant} was developed with this style of analysis in mind, while
+%   also allowing for a richer set of ecological dynamics than is
+%   possible in population models lacking size structure.
 
@@
 
-The core of \texttt{TREE} is a model for a plant's physiological strategy as
-specified by its traits (Fig. \ref{fig:schematic}a). This sub-model
-estimates rate of biomass production for a plant, given its size, light
-environment, and the supplied physiological constants. Gross photosynthetic 
-income is
-estimated from total leaf area and the light distribution across the
-plant's canopy. Costs of tissue respiration and turnover are then
-subtracted. The remaining biomass is then allocated between growth and
-reproduction. The strategy model thus includes a list of physiological
-rules and associated parameters.
+The core of {\plant} is a model for an individual plant's
+physiological strategy as specified by its traits
+(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
+production for a plant, given its size and a light environment. Gross
+photosynthetic income is estimated from total leaf area and the light
+distribution across the plant's canopy. Costs of tissue respiration
+and turnover are then subtracted. The remaining biomass is then
+allocated between growth and reproduction.  
+%
+The key outputs needed by higher levels (see below) is growth rate,
+mortality rate and rate of seed production.  However additional
+quantities are also computed, such as total assimilation and the
+respiration, turnover and allocation rates for different tissues (see
+Appendix \ref{sec:plant})
 
-The core job of the physiological model is to take size, light
-environment and physiological parameters as inputs, and to return growth rate,
-mortality and fecundity as outputs, in other words to move from
-individual growth to demography. However additional quantities are also
-available as output, such as total assimilation and the respiration,
-turnover, and allocation rates for different tissues.
+The default physiological model used in {\plant} largely reflects that
+presented in \citet{Falster-2011}, but with extensions allowing for
+diameter growth also to be estimated. To properly model trait
+variation, different parameters (such as rates of leaf turnover,
+respiration, photosynthesis per leaf area) in the model are linked via
+assumed tradeoffs. For example, the trait \textsc{lma} is used to estimate the
+rate of leaf turnover, based on observed scaling relationship spanning
+across diverse vegetation and plant functional global
+\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
 
-The default physiological model used in \texttt{TREE} largely reflects that
-presented in \citet{Falster-2011}, but with extensions allowing for diameter
-growth also to be estimated. To properly model trait variation,
-different parameters (such as rates of leaf turnover, respiration, photosynthesis 
-per leaf area) in the model are linked via assumed tradeoffs. For
-example, the trait LMA is used to estimate the rate of leaf turnover,
-based on widely observed scaling relationship from \citet{Wright-2004}.
+The current parametrisation of the physiological model is based on
+\citet{Falster-2011}. Parameter values were drawn from available
+literature (see Appendix \ref{sec:FFW16}, table 3); however for some
+parameters limited or no data were readily available. Providing
+updated parameter sets is trivial; the challenge is to infer suitable
+values by constraining outputs from {\plant} against diverse data
+sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
+are still actively working and one of the major use cases we envisage
+for {\plant}.
 
-\subsection{Design features}
+While we have implemented one physiological model (\texttt{FFW16}),
+the package is designed to allow arbitrary additional physiologies to
+be introduced.  That is, rather than varying the parameters of the
+equations, the equations themselves can be changed, additional
+resources could be competed for and the higher level functionality of
+the package can still be used.  As an intermediate level, we allow for
+parameters to imply relationships among arbitrary other parameters,
+allowing straightforward modelling of trade-offs.  For example,
+setting leaf mass per unit area (\textsc{lma}) affects leaf turnover rate,
+because it is known that leaves which are cheaper to build have shorter 
+lifespan \citep{Wright-2004}.
 
-\texttt{TREE} includes a default physiological model (fully described and derived
-in the Appendix \ref{sec:FFW16}). However a feature of the package is that alternative
-physiologies can be used, and the parameters of each model can be
-altered. Trait differences are accounted for by altering relevant
-parameters (see Appendix \ref{sec:parameters}).
+\subsection{Applications}
 
-\subsection{Use cases and examples}
-
-With above functionality, \texttt{TREE} can be used to estimate essential
-physiological rates for individual plants (Fig. \ref{fig:plant}). The
-function \texttt{grow\_plant\_to\_size} takes a given strategy and light
-environment and grows the plant, producing a trajectory of size over
-time (Fig. \ref{fig:plant}a). This is achieved by integrating an
-Ordinary Differential Equation (ODE) of size-dependent growth rate (see
-Appendix \ref{sec:demography} for more
-details). As the plant grows, allocation to different tissue types varies, and
-the composition of the plant changes to include more stem and less
-leaf (Fig. \ref{fig:plant}b). This in turn affects maintenance and
-turnover costs and the growth rate of plants (Fig. \ref{fig:plant}c).
-This shift in allocation alone generates a distinctive hump-shaped pattern of absolute
-growth with size \citep{King-2011}, and equally the widely regarded
-decline in relative growth rate with size from birth onwards
+With the basic physiological model, {\plant} can be used to estimate
+essential physiological rates for individual plants
+(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
+takes a given strategy and light environment and grows the plant,
+producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
+is achieved by integrating an Ordinary Differential Equation of
+size-dependent growth rate (see Appendices \ref{sec:demography} and
+\ref{sec:plant} for more details). As the plant grows, allocation to
+different tissue types varies and the composition of the plant
+changes to include more stem and less leaf
+(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
+costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
+shift in allocation alone generates a distinctive hump-shaped pattern
+of absolute growth with size \citep{King-2011} and equally the widely
+regarded decline in relative growth rate with size from birth onwards
 \citep{Enquist-2007}.
@@
 By varying the light environment and measuring growth rate, the whole
-plant light compensation point (WPLCP) can be computed (Fig.
-\ref{fig:plant}d). The WPLCP is the light level at which the plant 
-stops growing, and is increasingly regarded as the most useful
+plant light compensation point (\textsc{wplcp}) can be computed (Fig.
+\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant 
+stops growing and is increasingly regarded as the most useful
 measure of a strategy's shade tolerance
-\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, WPLCP
+\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
 increases with plant size, due to increased costs of building and
-maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, WPLCP
-decreases with LMA, because high LMA species have lower leaf turnover
-\citep{Baltzer-2007, Lusk-2013}.
+maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
+decreases with \textsc{lma}, because high \textsc{lma} species have lower leaf turnover
+\citep{Baltzer-2007, Lusk-2013} (Fig.
+\ref{fig:plant}d).
 
@@
 
-Modelling the development of a competitive hierarchy within a patch
-requires that both the initial size distribution and the inflow of new
-recruits be specified. We have primarily been interested in dynamics of
-a patch recovering from disturbance, and have therefore started with an
-empty patch and constant flow of seeds from a global seed rain (Fig.
-\ref{fig:schematic}b).
+We assume that patches are vertically but not horizontally
+structured. In other words, we account for size differences but not spatial 
+layout within the patch. Similar assumptions are found in many models
+simulating size-structured dynamics
+\citep{Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
+we assume that patches are very large -- this means the population dynamics 
+within the patch behave deterministically \citep{Deroos-1997} and can therefore 
+be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
+\ref{sec:demography}). The same \textsc{pde} captures average behaviour
+across a large number of small patches \citep{Moorcroft-2001}. The model may 
+therefore capture average demographic behaviour across a range of patch sizes.
 
-\subsection{Design features}
-
-When modelling a patch, we are interested in modelling changes in the
-size-density-distribution \(n(h,a)\) over time, i.e.~the density of
-individuals with height \(h\) within a patch of age \(a\). We assume
-that patches are large, and are vertically but not horizontally
-structured. Thus, shading in \texttt{TREE} is modelled as a mean field with
-respect to spatial layout. Similar assumptions are found in many models
-simulating size-structured dynamics
-\citep{Moorcroft-2001, Huston-1987, Smith-2014}. Under these
-assumptions, the dynamics of \(n\) behave deterministically and can be
-approximated via a Partial Differential Equation (PDE) (see Appendix
-\ref{sec:demography}). The same PDE has been shown to capture average behaviour
-across a large number of small patches \citep{Moorcroft-2001}. Ideally
-one would also consider spatial interactions within patches,
-however such models are very computationally demanding, and also
-impossible to solve in a deterministic fashion \citep{Pacala-1996}.
-Moreover, it remains unclear whether resolving spatial details within a
-patch would provide further ecological detail, in part because the
+Ideally one would also consider spatial interactions within patches,
+however such models are very computationally demanding and also
+impossible to solve in a deterministic fashion \citep{Huston-1987, 
+Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
+patch would provide further useful ecological detail, in part because the
 process of competitive thinning tends to break down spatial clusters
@@
 
+When modelling a patch, we are interested in modelling changes in the size-
+density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
+individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
+Density is measured as individuals per unit height and per unit ground area.
+Modelling the development of  \(n(x,h,a)\) requires that both the initial size
+distribution and the inflow of new recruits be specified. We have primarily
+been interested in dynamics of a patch recovering from disturbance and have
+therefore started with an empty patch and constant flow of seeds from a global
+seed rain (Fig. \ref{fig:schematic}b).
+
 Our approach for modelling size-structured population dynamics
-is based on the Escalator Boxcar Train technique (EBT)
-\citep{Deroos-1997, Deroos-1992, Deroos-1988}. The EBT solves the PDE
+is based on the Escalator Boxcar Train technique (\textsc{EBT})
+\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{EBT} solves the \textsc{pde}
 describing development of \(n\) by approximating the density function
@@
 cohorts are then transported along the characteristic equations of the
-PDE. Biologically these are the growth trajectories of individuals,
+\textsc{pde}. Biologically these are the growth trajectories of individuals,
 conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
@@
 
-We introduce two extensions to the EBT: i) a new method for estimating
-the size-density distribution, and ii) a technique for handling strong
+We introduce two extensions to the \textsc{EBT}: i) a new method for estimating
+the size-density distribution and ii) a technique for handling strong
 size-asymmetric competitive feedbacks, such as occur under strong
-competition for light. The original EBT
-\citep{Deroos-1997, Deroos-1992, Deroos-1988} proceeds by approximating
-the first and second moments of the density distribution \(n\) within
-each cohort. In \texttt{TREE} we use a different -- but from a theoretical
-perspective equally valid --  approach. Instead of tracking first and second
-moments of the size-distribution within cohorts, we track a point mass
-estimate of \(n\) along the trajectory for the boundary
-of each cohort. We found this approach preferable because it does not
-require us to maintain a separate set of equations for the smallest
-cohort \citep{Deroos-1997}.
+competition for light (see Appendix \ref{sec:demography} for
+details). The original \textsc{EBT} \citep{Deroos-1988,
+  Deroos-1992, Deroos-1997} proceeds by approximating the first and
+second moments of the density distribution \(n\) within each
+cohort. In {\plant} we use a different approach. Instead of tracking
+first and second moments of the size-distribution within cohorts, we
+track a point mass estimate of \(n\) along the trajectory for the
+boundary of each cohort. We found this approach preferable because it
+does not require us to maintain a separate set of equations for the
+smallest cohort \citep{Deroos-1997}.
 
@@
 adaptive refining of time points at which new cohorts are introduced
-into the population. In the original EBT, cohorts were introduced at a
+into the population. In the original \textsc{EBT}, cohorts were introduced at a
 fixed time interval. Under competition, the growth trajectories of
 individuals born at similar times can diverge substantially over time
-(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography}).
-This is can become a problem later during patch
-development because much of the population can be located between
-widely-spaced cohorts, delivering low numerical precision. To maintain
-numerical precision we therefore use an iterative algorithm to adaptively refine
-the cohort spacing until the trajectories of adjacent cohorts remain
-adequately resolved throughout development of the patch (see
-Appendix \ref{sec:demography} for details). This refinement causes an uneven
-distribution of cohort introduction times across patch age (Fig.
-\ref{fig:patch}a), generally with tighter spacing of cohorts in 
-younger patches.
+(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
+\ref{sec:patch} and \ref{sec:cohort-spacing}). This can become a
+problem later during patch development because much of the population
+can be located between widely-spaced cohorts, delivering low numerical
+precision. To maintain numerical precision we therefore use an
+iterative algorithm to adaptively refine the cohort spacing until the
+trajectories of adjacent cohorts remain adequately resolved throughout
+development of the patch (see Appendix \ref{sec:demography} for
+details). This refinement causes an uneven distribution of cohort
+introduction times across patch age (Fig.  \ref{fig:patch}a), with
+considerably tighter spacing of cohorts in younger patches.
 
-\subsection{Use cases and examples}
+\subsection{Applications}
 
-Modelling of size-structured dynamics via solving of a deterministic PDE
+Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
 enables multiple complex non-linear demographic phenomena to be
-investigated.
-
-First, \texttt{TREE} can simulate patterns of patch development after
+investigated. First, {\plant} can simulate patterns of patch development after
 disturbance, including self-thinning behaviours and successional
@@
 were limited to a single species
-\citep[e.g.][]{Barnes-2004, Coomes-2007}, \texttt{TREE} accommodates multiple
+\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
  species with different traits (Fig. \ref{fig:schematic}b; Fig.
@@
 
-Second, \texttt{TREE} allows for the effects of competition and succession on
+Second, {\plant} allows for the effects of competition and succession on
 productivity and biomass accumulation within a patch to be investigated
@@
 combination of physiological and structural changes
-\citep{Binkley-2002, Smith-2001, Ogawa-2010, Coomes-2007}. \texttt{TREE} enables
+\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
 the effects of competitive feedbacks to be incorporated, in addition to
-any intrinsic physiological factors. Moreover, \texttt{TREE} allows for the
+any intrinsic physiological factors. Moreover, {\plant} allows for the
 contributions of different species to be mapped (Fig. \ref{fig:patch}c).
-
-While many vegetation models still lack size structure, the importance of
-size-structure for modelling phenomena such as allocation and production is
-being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
 
@@
 disease outbreak
-\citep{Connell-1978, White-1979, Chambers-2013, Bormann-1979, Clark-1991, Coomes-2007}.
+\citep{White-1979, Bormann-1979, Chambers-2013}.
 The vegetation thus comprises a collection of patches differing in time
@@
 
-\subsection{Design features}
-
-Having simulated a single patch, it is straightforward to scale up from
-a single patch to an entire metapopulation, at least for the situation
-of metapopulations at an equilibrium mix of patch ages and where
-disturbances remove all established vegetation. The frequency-density
+In {\plant} we assume an infinite number of patches all experiencing
+the same disturbance regime; we also assume that the disturbances remove
+all established vegetation.  With this approach we can scale up from
+simulating a single patch to simulate an entire metapopulation.
+The frequency-density
 \(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
 are a large number of patches means the dynamics of \(p\) behave
-deterministically and can be approximated via a second PDE
-\citep{Vonfoerster-1959, Mckendrick-1926} (see Appendix
+deterministically and can be approximated via a second \textsc{pde}
+\citep{ Vonfoerster-1959} (see Appendix
 \ref{sec:demography} for
@@
 
-The main numerical challenge in stepping from single patch
-to metapopulation is to identify the equilibrium seed rain of the
-metapopulation. For a given seed rain input, the model produces a seed rain output. 
-The equilibrium seed rain is the vector of seed rain
-where input seed rain equals output seed rain. Finding this equilibrium
-requires finding a root of a multidimensional function, subject to the
+The main numerical challenge in stepping from single patch to
+metapopulation is to identify the equilibrium seed rain of the
+metapopulation. An "input" seed rain is required to simulate 
+the demography of the metapopulation, this in turn produces an "output" 
+number of seeds. The equilibrium seed rain is the vector of seed 
+rain where input seed
+rain equals output seed rain. Finding this equilibrium requires
+finding a root of a multidimensional function, subject to the
 constraints that this root is stable (e.g., the trivial equilibrium of
-zero seed rain always exists, but is often not stable) (Appendix \ref{sec:fitness});
-
-\subsection{Use cases and examples}
+zero seed rain always exists, but is often not stable) (Appendix
+\ref{sec:equilibrium}).
 
@@
 feature of the size-structured metapopulation concept implemented here
-is that it reconciles equilibrium and non-equilibrium approaches to
-modelling ecological dynamics
-\citep{Levin-1974, Bormann-1979, Connell-1978, Coomes-2007}. An
+is that it balances equilibrium and non-equilibrium approaches to
+modelling ecological dynamics. An
 equilibrium may be approached at the level of metapopulation, meaning
@@
 
-Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
-us to close the demographic feedback loop and thereby create a
+\subsection{Applications}
+% TODO[RGF]:  But we need
+% something that indicates how this can be used.  I suggest a complete
+% overhaul of this paragraph as that gets closest to a way of
+% using the model.
+
+Beyond its conceptual appeal, this idea of a dynamic equilibrium
+allows us to close the demographic feedback loop and thereby create a
 self-regulating system. The equilibrium seed rain for multi-species
 communities is approached demographically, incorporating the combined
-effects of the model's physiological rules, species traits, competitive
-interactions and the prevailing disturbance regime on the
-size-distributions of plants within a metapopulation of patches. Unlike
-some vegetation models, there is no need to set the density of seeds or
-seedlings, these arise as outputs from the model. Characteristics of the
-vegetation such as total leaf area, biomass, productivity and allocation
-to different plant parts also arise as emergent properties (Fig.
-\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
-growth rate with respect to size emerge at both patch scale and scale of
-metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
-physiological model produces a variety of relationships, dependent on patch
-age. These emergent phenomena might then be compared to the observed
-distributions in large forest tracts \citep{Muller-2006}.
+effects of the model's physiological rules, species traits,
+competitive interactions and the prevailing disturbance regime on the
+size-distributions of plants within a metapopulation of
+patches. Unlike some vegetation models, there is no need to set the
+density of seeds or seedlings, these arise as outputs from the
+model. Characteristics of the vegetation such as total leaf area,
+biomass, productivity and allocation to different plant parts also
+arise as emergent properties (Fig.  \ref{fig:patch}c). Similarly,
+distributions of abundance, leaf area and growth rate with respect to
+size emerge at both patch scale and scale of metapopulation
+(Fig. \ref{fig:emergent}). Note in particular how a single
+physiological model produces a variety of height-density
+relationships, dependent on patch age, as well as an emergent average
+across all patches (Fig. \ref{fig:emergent}, see more detail in
+appendix \ref{sec:emergent}). This emergent relationship might then be
+compared to the observed distributions in large forest plots
+\citep{Muller-2006}.
 
@@
 reproduction ratio \(R\) is simply the total seed output of the new
-type, averaged across the metapopulation. The new type can invade when
+type, averaged across the metapopulation. A new type can invade when
 \(R > 1\).
 
-\subsection{Use cases and examples}
+\subsection{Applications}
 
 With fitness defined, it becomes possible to model adaptive dynamics
-and community assembly \citep{Dieckmann-2007, Geritz-1998}. This is best
+and community assembly \citep{Geritz-1998,Dieckmann-2007}. This is best
 thought of as occurring on a fitness landscape, a surface with fitness
@@
 reproductive success of individuals possessing a particular trait value
-is measured in the size- and patch-structured metapopulation, and
+is measured in the size- and patch-structured metapopulation and
 depends on both the traits and densities of other species in the current
@@
 
-Fitness landscapes will be positive in regions that allow invasion, and
-zero for species that are at equilibrium. The slope of the fitness
-landscape at a resident indicates the direction of selection; a negative
-slope (dashed line in Fig. \ref{fig:fitness}a) indicates selection for
-smaller trait values. Fitness landscapes are constantly reshaped
-as species are introduced and evolve, reflecting the non-linear
-demographic feedbacks that occur under frequency-dependent evolution
- \citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient, can lead to an
-evolutionary branching point \citep{Geritz-1998,Dieckmann-2007};
-at this point there is no directional selection (solid line in
-figure \ref{fig:fitness}a) but invasion is possible both above and below
-it in trait space. It is often possible
-to find a configuration where fitness is negative everywhere (solid line
-in Fig. \ref{fig:fitness}b) except for the zero points where species are
-present. This community is immune to invasion from any other strategies
-\citep{Geritz-1998}.
+Fitness landscapes will be positive in regions that allow invasion,
+and zero for species that are at equilibrium. The slope of the fitness
+landscape at a resident indicates the direction of selection; a
+negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
+selection for smaller trait values (see appendix \ref{sec:fitness} for
+details). Fitness landscapes are constantly reshaped as species are
+introduced and evolve, reflecting the non-linear demographic feedbacks
+that occur under frequency-dependent evolution
+\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
+can lead to an evolutionary branching point
+\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
+directional selection (solid line in Fig. \ref{fig:fitness}a) but
+invasion is possible both above and below it in trait space. It is
+often possible to find a configuration where fitness is negative
+everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
+points where species are present. This community is immune to invasion
+from any other strategies \citep{Geritz-1998}.
 
 Identifying conditions promoting trait diversity is a valuable use-case
-for \texttt{TREE}. Coexisting plant species differ in a range of physiological
+for {\plant}. Coexisting plant species differ in a range of physiological
 traits, yet the conditions allowing for these different types to coexist
@@
 
-\texttt{TREE} is written in C++ and \texttt{R}, using the package \texttt{Rcpp}
+{\plant} is written in C++ and R, using the package \texttt{Rcpp}
 \citep{Eddelbuettel-2011, Eddelbuettel-2013} to bridge between the two
@@
 
-\texttt{TREE} makes use of much existing software, including the programming
-language C++; the \texttt{R} computational environment \citep{R-2015}, the \texttt{R}
+{\plant} makes use of much existing software, including the programming
+language C++; the R computational environment \citep{R-2015}, the R
 packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
-\texttt{R6}\citep{Chang-2014}, and \texttt{testhat}
+\texttt{R6} \citep{Chang-2014} and \texttt{testhat}
 \citep{Wickham-2011}; and the Boost Library for C++
-\citep{Schaling-2014} via the \texttt{R} package \texttt{BH}
+\citep{Schaling-2014} via the R package \texttt{BH}
 \citep{Eddelbuettel-2015}. Source code is hosted at
-\href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
-Installation instructions are available at this homepage, and binary
-releases will also be available there.
-
-The current parametrisation of the physiological model is based on
-\citet{Falster-2011}. Parameter values were drawn from available literature (see
-Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
-readily available. Providing updated parameter sets is trivial (see Appendix
-\ref{sec:parameters}). Ideally, parameters would be inferred through a combination of
-independent estimates and constraining outputs from \texttt{TREE} against diverse data sources
-\citep{Lebauer-2012, Keenan-2013}.
+\href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
+Installation instructions are available at this homepage, binary
+releases will also be available there.  (Note that {\plant} does not
+currently compile on Windows, but this will change soon as R core
+upgrade the Windows C++ compiler).
 
@@
 
-The \texttt{TREE} package makes it easier for researchers and forest practitioners
+The {\plant} package makes it easier for researchers and forest practitioners
 alike to investigate a variety of non-linear demographic phenomena.  Rather
-than provide a model with a few routines, we have developed an extensible
+than provide a model with a few entry points, we have developed an extensible
 framework that we hope can be used to answer a number of questions at a
@@
 interest in the assembly of trait mixtures in communities. However a variety
-of other uses would be possible, and also extensions to address other facets
+of other uses would be possible and also extensions to address other aspects
 of community assembly. For example,  the physiology of growth could be
@@
 structure, so remain unable to investigate demographic phenomena
-\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. At the same time,
-much attention is paid to detailed leaf-level physiology. \texttt{TREE} aims to bridge
+\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. At the same time,
+much attention is paid to detailed leaf-level physiology. {\plant} aims to bridge
 this gap that emerged in the model landscape, where detailed physiology and
@@
 (RP04-174). M Westoby was supported by a fellowship from Australian
-Research Council.
+Research Council. We thank {\AA} Brn{\"a}nnstr{\"o}m, J Camac, 
+A De Roos, U Dieckmann, R Duursma, and J Johansson for helpful discussions.
 
@@
 
-\caption{Processes modelled within \texttt{TREE} include physiological, population, and
+\caption{Processes modelled within {\plant} include physiological, population and
 evolutionary dynamics.
@@
 and light environment. (a) Growth trajectories for plants differing in light
-environment and trait (LMA, low = 0.1, solid line; high =  0.267, dashed
+environment and trait (\textsc{lma}, low = 0.1, solid line; high =  0.267, dashed
 line). Lower light environments slow trajectories and interact non-linearly
 with traits. (b) Over time, the fraction of living tissue switches from leaf
-towards sapwood, with high LMA species having relatively more mass in leaf
-than low LMA species. (c) Growth rates peak around 5m high, but this position
+towards sapwood, with high \textsc{lma} species having relatively more mass in leaf
+than low \textsc{lma} species. (c) Growth rates peak around 5m high, but this position
 varies with both trait and light level (note that this is the derivative of
@@
 growth rate is zero (x intercept).  The light level used in panels (a) and (c)
-is indicated with dots.} [Could you put delta symbols on the y axis rather than d?
-Also there is alot going in these figures with 4 lines each.
-You could make it easier for the reader by adding the dashed lines to the keys]
+is indicated with dots.  See Appendix \ref{sec:plant} for details and code.}
 \label{fig:plant}
@@
 introduction times generated by an adaptive refinement algorithm.
- Red lines are the low LMA
-species from figure \ref{fig:plant}, blue lines are the high LMA species.  The
-initial growth rate advantage of the low LMA species (see Figure
-\ref{fig:plant} c) means it quickly over tops the high LMA species,
-suppressing its growth.  Self-thinning leaves space for the high LMA species
+ Red lines are the low \textsc{lma}
+species from figure \ref{fig:plant}, blue lines are the high \textsc{lma} species.  The
+initial growth rate advantage of the low \textsc{lma} species (see Figure
+\ref{fig:plant} c) means it quickly over tops the high \textsc{lma} species,
+suppressing its growth.  Self-thinning leaves space for the high \textsc{lma} species
 to establish below the canopy of the fast species. Panel (b) shows the same
@@
 level for each species.  The dotted line is the total leaf area index (sum
-across both species).}
+across both species). See Appendix \ref{sec:patch} for details and code.}
 \label{fig:patch}
@@
 \caption{Emergent size-distributions across a metapopulation.
-Grey lines show patterns within individual patches differing in time since disturbance (age), with blue lines indicating patches of a specific age.
-Panels show (a) patterns of abundance, (b) leaf area and (c) height growth rate against size.
-Note that units of panels (a) and (b) are per unit height and per unit ground area, respectively.}
+Grey lines show patterns within individual patches differing in time
+since disturbance (age), with blue lines indicating patches of a
+specific age. Red lines show average relationship across the metapopulation, 
+obtained by weighting the grey lines by patch and individual abundance. 
+Panels show (a) patterns of abundance, (b) leaf area and
+(c) height growth rate against size. Note that units of panels (a) and
+(b) are per unit height and per unit ground area, respectively.  See Appendix \ref{sec:emergent} for details and code.}
 \label{fig:emergent}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
-\caption{Example fitness landscape for LMA, showing potential for stable
+\caption{Example fitness landscape for \textsc{lma}, showing potential for stable
 coexistence of multiple types.  Fitness is the log of the population growth
 rate - at zero species are at equilibrium, above zero they will increase in
-density.  (a) Landscape generated by a single species.  At LMA of 0.1 (the
-dashed line) there is directional selection for smaller LMA.  At LMA of 0.08
+density.  (a) Landscape generated by a single species.  At \textsc{lma} of 0.1 (the
+dashed line) there is directional selection for smaller \textsc{lma}.  At \textsc{lma} of 0.08
 (solid line) there is a branching point (see inset for more detail).  (b)
-Landscape generated by two species, holing the first at LMA of 0.08.
+Landscape generated by two species, holing the first at \textsc{lma} of 0.08.
 Introducing a species at the maximum fitness value from panel (a) reduces the
 region of positive fitness considerably (dashed line).  Dotted lines show
-subsequent invasions and replacements of the second species, and the solid
+subsequent invasions and replacements of the second species, the solid
 line shows a stable fitness landscape where both species exist on a peak in
-the fitness landscape.  At this point no further invasion is possible.}
+the fitness landscape.  At this point no further invasion is
+possible.  See Appendix \ref{sec:fitness} for details and code.}
 \label{fig:fitness}
@@
 \clearpage
+\setcounter{secnumdepth}{1}
 
@@
 
-\section{Default physiological model for \texttt{TREE}}\label{sec:FFW16}
+\section{Default physiological model for {\plant}}\label{sec:FFW16}
 
-See attached file \url{tree_physiology.pdf}
+See attached file \url{vignette_physiology.pdf}
 
-\section{Modelling demography in the \texttt{TREE} package}\label{sec:demography}
+\section{Modelling demography in the {\plant} package}\label{sec:demography}
 
-See attached file \url{tree_demography.pdf}
+See attached file \url{vignette_demography.pdf}
 
-\section{Solving for equilibrium seed rain}\label{sec:seed}
+\section{Plant level properties}\label{sec:plant}
+
+See attached file \url{vignette_plant.R}
+
+\section{Cohort spacing}\label{sec:cohort-spacing}
+
+See attached file \url{vignette_cohort_spacing.R}
+
+\section{Solving for equilibrium seed rain}\label{sec:equilibrium}
 
@@
 
-\section{Calculating fitness in the \texttt{TREE} package}\label{sec:fitness}
+\section{Patch level dynamics}\label{sec:patch}
+
+See attached file \url{vignette_patch.R}
+
+\section{Patch level emergent properties}\label{sec:emergent}
+
+See attached file \url{vignette_emergent.R}
+
+\section{Calculating fitness in the {\plant}  package}\label{sec:fitness}
 

### 2015-05-28 ms.tex
Reorganise schematic figures, and all figures generally
output -> figures
schematic is contained within figures

Fixes #22

--- 2015-05-27 ms.tex
@@
 \centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{output/patch.pdf}
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
 \caption{Population dynamics for two species competing within a patch. Lines

### 2015-05-28 ms.tex
Incorporate feedback from Sean
Clsoes #21

--- 2015-05-28 ms.tex
@@
 \runninghead{TREE}
-% TODO: need more keywords here
-\keywords{demography}
+\keywords{demography, emergent, fitness, growth,
+physiology, metapopulation, mortality, reproduction, size-structure,
+tradeoff}
 
@@
 
-\textbf{Key-words}: demography, emergent property, fitness, growth,
-physiology, metapopulation, mortality, reproduction, size-structure,
-tradeoff
-
 \section{Introduction}\label{introduction}
@@
 photosynthate also rises. Yet as individuals get larger, an increasing
-fraction of photosynthate must be diverted towards
-building and maintaining support tissues rather than leaves
-\citep{Givnish-1988, Enquist-2007}, and reproduction \citep{Thomas-2011}. Consequently 
+fraction of that photosynthate is diverted towards
+building and maintaining support tissues
+\citep{Givnish-1988, Enquist-2007} and reproduction \citep{Thomas-2011}. Consequently 
 rates of growth and mortality change with individual size
@@
 maturation determine start and end points of ontogenetic trajectories
-\citep{Westoby-2002}. Over longer time-frames, natural selection favours
-traits that provide a competitive advantage, such as faster growth, larger
-seeds or taller height \citep{Falster-2003}.
-Consequently, selective forces may amplify size-related effects on demography.
+\citep{Westoby-2002}. Over longer time-frames, natural selection generates
+traits that amplify size-related effects on demography,
+such as faster growth, larger seeds or taller height \citep{Falster-2003}. 
 Accounting for the effects of size and traits in models is therefore fundamental to developing theory on the structure and dynamics of the biosphere
@@
 \citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
-parameterised in terms of growth and demography, and consequently did a
-poor job of simulating carbon fluxes \todo{ref}. The next generation of models
-focussed more on physiology and, presumably for computational reasons,
+parameterised in terms of growth using statistical functions; this meant 
+that data had to be collected for each species, limiting widespread application. 
+The next generation of models focussed on establishing core physiology for 
+a limited number of functional types, and, presumably for computational reasons,
 sacrificed details of demography in order to achieve global coverage
-\citep{Haxeltine-1996}. This history has brought us to the
+\citep{Haxeltine-1996,Woodward-1995}. This history has brought us to the
 current perplexing situation where all but a couple
-\citep{Moorcroft-2001, Smith-2014} of the most popular vegetation
+\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
 models lack size-structure. Instead, most models
@@
 been difficult to connect mechanisms described in these simplistic
-models with field data and real traits, which are almost always strongly
-size-structured.
+models with type of data often collected in the field, such as physiological
+leaf traits, and demographic data on growth and mortality in long-term monitoring 
+plots \citep{Adler-2013}.
 
@@
 previous work investigating the effect of traits on vegetation
-properties \citep{Falster-2011} and the mechanisms facilitating
-coexistence of trait mixtures \citep{Falster-2015}. These studies
-provided the foundation methods needed \todo{needed for what.. its hinted above... maybe rearrange the sentences} here we provide robust and
-extensible implementations. We describe the general approach of the
-package and then a series of use cases, focussing at different levels of
+properties and the mechanisms facilitating
+coexistence of trait mixtures \citep{Falster-2011,Falster-2015}. Building 
+on earlier studies \citep{Deroos-1997,Kohyama-1993,Moorcroft-2001}, these
+papers outlined methods for driving trait-, size- and patch-structured 
+dynamics from a core physiological model. Here we provide 
+a robust and extensible implementation of those methods. 
+We describe the general approach of the
+package and then sections with details at different levels of
 ecological organisation. Each section provides a short description of
-the required methods, design considerations, and then some worked
-examples illustrating potential use cases and design features. Detailed
-technical documents are provided as supplementary information, and updated
-versions of technical documents are also available within the \texttt{TREE}
-package itself. The use cases highlighted reflect our own interests in
-understanding trait-based demography and community assembly. Yet we
-expect the methods provided to be useful in other contexts. For example,
-prior to \texttt{TREE} there was no \texttt{R} package that allowed one to simply grow a
-plant over time, based on well-understood physiology. Surely this is one
-of the simplest demographic actions one might hope for, with diverse
-potential applications in ecology, evolution, forestry, agriculture, and
-vegetation modelling. We hope the mechanistic platform provided here
-will complement the increasingly sophisticated statistical methods now
-available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.
+the required methods, design considerations, and then 
+example applications. Detailed
+technical documents are provided as supplementary information (see Appendices),
+and updated versions of technical documents are also available within the \texttt{TREE}
+package itself.
+
+While the applications highlighted reflect our own interests in understanding
+trait-based demography and community assembly, we expect the methods provided
+to be useful in variety of contexts. For example, prior to \texttt{TREE} there
+was no \texttt{R} package that allowed one to simply grow a plant over time,
+based on well-understood physiology. Surely this is one of the simplest
+demographic actions one might hope for, with diverse potential applications in
+ecology, evolution, forestry, agriculture, and vegetation modelling. More
+specifically, \texttt{TREE} provides a  platform for investigating how
+morphological and physiological traits influence  growth rate and shade
+tolerance, thus potentially giving a theoretical foundation  for observed
+empirical patterns \citep{Wright-2010,Baltzer-2007}. Being driven by traits,
+the model can be extended to potentially very many species. Users can also 
+add their own physiological models. Individual plants
+can then be  situated in a patch with competitors and the demography of a
+competing population  investigated, thereby capturing phenomena such as multi-
+species self-thinning,  successional transitions and age-related changes in
+vegetation productivity. \texttt{TREE}  models the emergent behaviour of
+patches deterministically, based on a continual  size-density distribution. As
+such, it is not required to specify a particular patch size.  Finally,
+\texttt{TREE} allows for emergent phenomena across a metapopulation of patches
+to  be investigated, and thus provides a tractable mechanism to scale from
+individual plant to  vegetation, incorporating the non-linear effects of size
+and traits on vegetation outcomes. In  this sense \texttt{TREE} is similar to
+the Ecosystem Demography \citep{Moorcroft-2001}  and LPJ-GUESS
+\citep{Smith-2014} models. However, whereas these other models were designed
+to primarily capture large-scale regional biogeocehcmisty driven by climatic
+variables,  \texttt{TREE} was designed so that the rules driving the
+demography of individual trees, patches and metapopulations could be
+manipulated and studied in detail. The three models are therefore highly 
+complimentary, and also connect with the increasingly sophisticated statistical 
+methods now available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.
 
@@
 example, the trait LMA is used to estimate the rate of leaf turnover,
-based on widely observed scaling relationship from \citet{Wright-2004}.
+based on observed scaling relationship spanning across diverse vegetation and
+plant functional  global \citep{Wright-2004} (see 
+Appendix \ref{sec:FFW16} for details).
 
@@
 
-\subsection{Use cases and examples}
+\subsection{Applications}
 
@@
 When modelling a patch, we are interested in modelling changes in the
-size-density-distribution \(n(h,a)\) over time, i.e.~the density of
-individuals with height \(h\) within a patch of age \(a\). We assume
-that patches are large, and are vertically but not horizontally
-structured. Thus, shading in \texttt{TREE} is modelled as a mean field with
-respect to spatial layout. Similar assumptions are found in many models
+size-density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
+individuals with traits \(x\) and height \(h\) within a patch of age \(a\). Density
+is measured as individuals per unit height and per unit ground area.
+
+We assume that patches are large, and are vertically but not horizontally
+structured. In other words, we account for size differences but not spatial 
+layout within the patch. Similar assumptions are found in many models
 simulating size-structured dynamics
-\citep{Moorcroft-2001, Huston-1987, Smith-2014}. Under these
-assumptions, the dynamics of \(n\) behave deterministically and can be
+\citep{Moorcroft-2001, Huston-1987, Smith-2014,Kohyama-1993}. The assumption of a large
+patch means the population of plants within the patch -- and thus the 
+dynamics of \(n\) -- behave deterministically \citep{Deroos-1997}, and can therefore be
 approximated via a Partial Differential Equation (PDE) (see Appendix
 \ref{sec:demography}). The same PDE has been shown to capture average behaviour
-across a large number of small patches \citep{Moorcroft-2001}. Ideally
-one would also consider spatial interactions within patches,
+across a large number of small patches \citep{Moorcroft-2001}. The model may 
+therefore capture average demographic behaviour across a range of patch sizes.
+
+Ideally one would also consider spatial interactions within patches,
 however such models are very computationally demanding, and also
-impossible to solve in a deterministic fashion \citep{Pacala-1996}.
+impossible to solve in a deterministic fashion \citep{Huston-1987, 
+Pacala-1996}.
 Moreover, it remains unclear whether resolving spatial details within a
@@
 metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
-physiological model produces a variety of relationships, dependent on patch
-age. These emergent phenomena might then be compared to the observed
-distributions in large forest tracts \citep{Muller-2006}.
+physiological model produces a variety of height-density relationships, 
+dependent on patch age, as well as an emergent average across all patches 
+(Fig. \ref{fig:emergent}). This emergent relationship might then be 
+compared to the observed distributions in large forest plots \citep{Muller-2006}.
 
@@
 Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
-readily available. Providing updated parameter sets is trivial (see Appendix
-\ref{sec:parameters}). Ideally, parameters would be inferred through a combination of
-independent estimates and constraining outputs from \texttt{TREE} against diverse data sources
-\citep{Lebauer-2012, Keenan-2013}.
+readily available. Providing updated parameter sets is trivial; the challenge 
+is to infer suitable values by constraining  outputs from \texttt{TREE} against diverse data sources
+\citep{Lebauer-2012, Keenan-2013}. This is an area in which we are still actively working.
 

### 2015-05-28 R/tables.R
Automatically add parameters to physiology table
Fixes #17

@@ -1,14 +1,38 @@
 table_plant_parameters <- function(filename, dest){
+  tab <- read.csv(filename, stringsAsFactors=FALSE,
+                  check.names=FALSE, comment.char="#")
 
-  xx <- read.csv(filename, stringsAsFactors=FALSE)
+  to_bold <- tab$Symbol == ""
+  tab$Description[to_bold] <- sprintf("\\textbf{%s}", tab$Description[to_bold])
 
-  to_bold <- which(xx$Symbol =="" )
+  to_math <- tab$Symbol != ""
+  tab$Symbol[to_math] <- sprintf("$%s$", tab$Symbol[to_math])
 
-  xx$Description[to_bold] <- sprintf("\\textbf{%s}", xx$Description[to_bold])
+  to_math <- grepl("[_^]", tab$Units)
+  tab$Units[to_math] <-
+    sprintf("$%s$",
+            gsub("(kg|m|yr)", "\\\\mathrm{\\1}", tab$Units[to_math]))
 
-  x <- xtable(xx,
+  to_code <- tab[["\\plant"]] != ""
+  code <- tab[["\\plant"]][to_code]
+
+  s <- FFW16_Strategy()
+  val <- s[code]
+  err <- sapply(val, is.null)
+  if (any(err)) {
+    stop("Missing parameters: ", paste(code[err], collapse=", "))
+  }
+
+  oo <- options(scipen=999)
+  on.exit(options(oo))
+  tab$Value[to_code] <- prettyNum(unlist(val))
+
+  tab[["\\plant"]][to_code] <- sprintf("\\texttt{%s}",
+                                      gsub("_", "\\\\_", code))
+
+  x <- xtable(tab,
               hline.after=c(1),
-              align='lp{7cm}lll')
+              align='lp{7cm}llll')
   y <- print(x, sanitize.text.function=as.character,
              include.rownames=FALSE, floating=FALSE,
              print.results=FALSE)

### 2015-05-28 ms.tex
Changes up to p 12

@@
 
+\newcommand{\plant}{\textsc{plant}}
+
 \usepackage[colorinlistoftodos]{todonotes}
 
-\title{\texttt{TREE}: A package for modelling plant TRait Ecology and Evolution}
+\title{{\plant}: A package for modelling forest trait ecology \& evolution}
 \author{Daniel S. Falster, Richard G. FitzJohn, and Mark Westoby}
@@
 \mstype{Applications Note}
-\runninghead{TREE}
+\runninghead{plant}
 \keywords{demography, emergent, fitness, growth,
@@
 \item
-  Here we introduce the \texttt{TREE} package (TRait Ecology and Evolution), an
+  Here we introduce the {\plant} package (TRait Ecology and Evolution), an
   extensible framework for modelling plant demography across the entire
   life cycle and in structured metapopulations, via coupled differential equations.
-  At its core, \texttt{TREE} is an
+  At its core, {\plant} is an
   individual-based mechanistic model where plant physiology and demography is mediated by
@@
 
-Plant growth and demography are fundamentally size- and trait- structured,
-influencing dynamics over time-scales ranging from instantaneous physiological
-effects to long-term evolutionary outcomes \citep{Harper-1977, Westoby-2002}. As
-an individual increases its leaf area, its potential to generate
-photosynthate also rises. Yet as individuals get larger, an increasing
-fraction of that photosynthate is diverted towards
-building and maintaining support tissues
-\citep{Givnish-1988, Enquist-2007} and reproduction \citep{Thomas-2011}. Consequently 
-rates of growth and mortality change with individual size
-\citep{Muller-2006, Thomas-2011, Ruger-2011, Wright-2010}. The traits of
-a species also interact with size to shape its demography. Leaf
-construction costs and wood density affect growth rates
-\citep{Wright-2010}. Other traits such as seed size and height at
+Plant growth and demography are fundamentally size- and trait-
+structured, influencing dynamics over time-scales ranging from
+instantaneous physiological effects to long-term evolutionary outcomes
+\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
+area, it increases its potential to generate photosynthate.  In
+contrast, as indivuduals grow, they must use increasing fractions of
+this photosynthetic income to building and maintaining support tissues
+rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
+reproduction \citep{Thomas-2011}. Consequently rates of growth and
+mortality change with individual size \citep{Muller-2006, Thomas-2011,
+  Ruger-2011, Wright-2010}.
+%
+The patterns of changing growth rate vary by plant traits; leaf
+construction costs and wood density affect growth rates and rates of
+tissue replacement \citep{Wright-2010}, while seed size and height at
 maturation determine start and end points of ontogenetic trajectories
-\citep{Westoby-2002}. Over longer time-frames, natural selection generates
-traits that amplify size-related effects on demography,
-such as faster growth, larger seeds or taller height \citep{Falster-2003}. 
-Accounting for the effects of size and traits in models is therefore fundamental to developing theory on the structure and dynamics of the biosphere
-\citep{Purves-2008, Moorcroft-2001, Falster-2003}, in particular for notable
-phenomena such as successional dynamics, self thinning, trait evolution
-and species coexistence.
+\citep{Westoby-2002}
+%
+Strong feedbacks emerge between individuals within a forest due to
+competition for light and other resources, such that the growth rate
+of one individual depends on the size and traits of nearby
+individuals.  This makes it difficult to predict demographic processes
+without modelling both individual growth and the interactions between
+individuals.
+% TODO[RGF]: link back things about natural selection, successional
+% dynamics, self thinning, trait evolution and species coexistance.
 
+% NOTE[RGF]: I thought this was way overrefrenced and so just kept a
+% few oldies.  Please re-add any key ones that you want to see here.
 Although the importance of size-structure for vegetation function and
 demography has been long been recognised
-\citep{Harper-1977, Hara-1984, Shugart-1980, Huston-1987, Moorcroft-2001,
-Kohyama-1993, Enquist-2007, Pacala-1996, Coomes-2007},
+\citep[e.g.,][]{Harper-1977, Shugart-1980, Hara-1984},
 most large-scale mechanistic models of plant growth either omit it
@@
 reason is the computational
-challenge of modelling size-structured dynamics. During the 1980s and
+challenge of modelling size-structured dynamics. 
+% TODO[RGF]: I don't see the link with the above?  Can this be
+% shortened/tightened?  It's too long for a 3k word paper
+During the 1980s and
 early 1990s, size-structured ``gap'' models were widely used
@@
 The next generation of models focussed on establishing core physiology for 
-a limited number of functional types, and, presumably for computational reasons,
+a limited number of functional types, and
 sacrificed details of demography in order to achieve global coverage
 \citep{Haxeltine-1996,Woodward-1995}. This history has brought us to the
-current perplexing situation where all but a couple
+current situation where all but a couple
 \citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
@@
 
-Interestingly, most models used in theoretical ecology to explore
+Similarly, most models used in theoretical ecology to explore
 questions about species coexistence also lack size-structured dynamics
-\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013, Levin-1974,MacArthur-1967,Tilman-1985}.
-It is common to assume competitive interactions are influenced by
-size-related traits such as adult size and seed size, but detailed
-size-structured demography is rarely considered, at least for plant
-models \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has
-been difficult to connect mechanisms described in these simplistic
-models with type of data often collected in the field, such as physiological
-leaf traits, and demographic data on growth and mortality in long-term monitoring 
-plots \citep{Adler-2013}.
+\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013,
+  Levin-1974,MacArthur-1967,Tilman-1985}.  It is common to assume
+competitive interactions are influenced by size-related traits such as
+adult size and seed size, but detailed size-structured demography is
+rarely considered, at least for plant models \cite[for animal
+examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
+difficult to connect mechanisms described in these simplistic models
+with type of data often collected in the field, such as physiological
+leaf traits, and demographic data on growth and mortality in long-term
+monitoring plots \citep{Adler-2013}.
 
-In this note, we describe the \texttt{TREE} package for R
-\citep{R-2015}, a mechanistic framework for studying the effects of size
-structure and trait variation on the demography of individual plants, of
-patches of competing plants, and of meta-populations structured by a
-prevailing disturbance regime. The package came about as a way to extend
-previous work investigating the effect of traits on vegetation
-properties and the mechanisms facilitating
-coexistence of trait mixtures \citep{Falster-2011,Falster-2015}. Building 
-on earlier studies \citep{Deroos-1997,Kohyama-1993,Moorcroft-2001}, these
-papers outlined methods for driving trait-, size- and patch-structured 
-dynamics from a core physiological model. Here we provide 
-a robust and extensible implementation of those methods. 
-We describe the general approach of the
-package and then sections with details at different levels of
-ecological organisation. Each section provides a short description of
-the required methods, design considerations, and then 
-example applications. Detailed
-technical documents are provided as supplementary information (see Appendices),
-and updated versions of technical documents are also available within the \texttt{TREE}
-package itself.
+In this note, we introduce the {\plant} package for R \citep{R-2015};
+a mechanistic framework for studying the effects of size structure and
+trait variation on the demography of individual plants, of patches of
+competing plants, and of meta-populations structured by a prevailing
+disturbance regime.
+%
+The {\plant} package builds on much existing work investigating the
+effect of traits on vegetation properties and the mechanisms
+facilitating coexistence of trait mixtures
+\citep{Kohyama-1993,Deroos-1997,Moorcroft-2001,Falster-2011,Falster-2015}.
 
+% TODO[RGF]: restructure here - I'm getting lost
+{\plant} provides a platform for investigating how morphological and
+physiological traits influence growth rate and shade tolerance, thus
+potentially giving a theoretical foundation for observed empirical
+patterns \citep{Wright-2010,Baltzer-2007}.
+
+% NOTE[RGF]: I dropped the appeal to novelty (growing an individual
+% plant) because I don't find them very convincing.
 While the applications highlighted reflect our own interests in understanding
 trait-based demography and community assembly, we expect the methods provided
-to be useful in variety of contexts. For example, prior to \texttt{TREE} there
-was no \texttt{R} package that allowed one to simply grow a plant over time,
-based on well-understood physiology. Surely this is one of the simplest
-demographic actions one might hope for, with diverse potential applications in
-ecology, evolution, forestry, agriculture, and vegetation modelling. More
-specifically, \texttt{TREE} provides a  platform for investigating how
-morphological and physiological traits influence  growth rate and shade
-tolerance, thus potentially giving a theoretical foundation  for observed
-empirical patterns \citep{Wright-2010,Baltzer-2007}. Being driven by traits,
+to be useful in variety of contexts. . Being driven by traits,
 the model can be extended to potentially very many species. Users can also 
@@
 species self-thinning,  successional transitions and age-related changes in
-vegetation productivity. \texttt{TREE}  models the emergent behaviour of
+vegetation productivity. {\plant}  models the emergent behaviour of
 patches deterministically, based on a continual  size-density distribution. As
 such, it is not required to specify a particular patch size.  Finally,
-\texttt{TREE} allows for emergent phenomena across a metapopulation of patches
+{\plant} allows for emergent phenomena across a metapopulation of patches
 to  be investigated, and thus provides a tractable mechanism to scale from
 individual plant to  vegetation, incorporating the non-linear effects of size
-and traits on vegetation outcomes. In  this sense \texttt{TREE} is similar to
+and traits on vegetation outcomes. In  this sense {\plant} is similar to
 the Ecosystem Demography \citep{Moorcroft-2001}  and LPJ-GUESS
@@
 to primarily capture large-scale regional biogeocehcmisty driven by climatic
-variables,  \texttt{TREE} was designed so that the rules driving the
+variables,  {\plant} was designed so that the rules driving the
 demography of individual trees, patches and metapopulations could be
@@
 complimentary, and also connect with the increasingly sophisticated statistical 
-methods now available to analyse size-structured data \citep[e.g.][]{Metcalf-2013}.
+methods now available to analyse size-structured data
+\citep[e.g.][]{Metcalf-2013}.
+
+We describe the general approach of the package and then sections with
+details at different levels of ecological organisation. Each section
+provides a short description of the required methods, design
+considerations, and then example applications. Detailed technical
+documents are provided as supplementary information (see Appendices),
+and updated versions of technical documents are also available within
+the {\plant} package itself.
 
@@
 
-\texttt{TREE} is a mechanistic model, meaning the dynamics of the system arise
-from rules specifying how individuals grow and interact. In ecology,
-there is a long history of using simple deterministic models
-to understand biological phenomena, such as Lotka-Volterra population dynamics
-\citep{MacArthur-1967, Leimar-2013} .
-\texttt{TREE} was developed with this style of analysis in mind, while also
-allowing for a richer set of ecological dynamics than is possible in
-population models lacking size structure. The package implements methods
-for physiological, population, and adaptive dynamics (Fig.
-\ref{fig:schematic}) following methods described by \citet{Falster-2011}
-and \citet{Falster-2015}. The core rules in \texttt{TREE} are about the
-short-term physiological functioning of an individual plant and how
-these are influenced by its traits, size and light environment. Dynamics
-at higher levels of organisation then arise as emergent properties,
-driven by growth physiology, competition for light and disturbance (Fig.
-\ref{fig:schematic}). Demographic phenomena can be studied at the three
-levels: individual plants, stands of competing plants, and entire
-metapopulations.
+% NOTE: Keep the repetition with above under control.
+% NOTE: LV are not really mechanistic
+
+{\plant} is a mechanistic model, meaning the dynamics of the system
+arise from rules specifying how individuals grow and interact.  The
+core rules in {\plant} are about the short-term physiological
+functioning of an individual plant and how these are influenced by its
+traits, size and light environment.
+%
+On top of this physiolgical model, the package implements methods for
+population and adaptive dynamics (Fig.  \ref{fig:schematic}) following
+methods described by \citet{Falster-2011} and
+\citet{Falster-2015}. 
+%
+The dynamics at higher levels of organisation arise as emergent
+properties, driven by growth physiology, competition for light and
+disturbance (Fig.  \ref{fig:schematic}). Demographic phenomena can be
+studied at the three levels: individual plants, stands of competing
+plants, and entire metapopulations.
+%
+We developed {\plant} as an extended thought experiment for helping to
+understand competition and succession by having these driven by
+low-level processes rather than being explicitly implemented in the
+model.
+% NOTE: I think that foreshaddowing some of the competition kernels
+% idea here is sensible.
+%
+% TODO: Somewhere, perhaps above, perhaps highlight the issue of doing
+% coexistance modelling with explicit modelling - the models are so
+% well understood it's easy to drive the behaviour.  But we want to
+% have something where we just get coexistence to study how tradeoffs
+% generate it.
+%
+% TODO: perhaps salavageable?
+%
+%   {\plant} was developed with this style of analysis in mind, while
+%   also allowing for a richer set of ecological dynamics than is
+%   possible in population models lacking size structure.
 
@@
 
-The core of \texttt{TREE} is a model for a plant's physiological strategy as
-specified by its traits (Fig. \ref{fig:schematic}a). This sub-model
-estimates rate of biomass production for a plant, given its size, light
-environment, and the supplied physiological constants. Gross photosynthetic 
-income is
-estimated from total leaf area and the light distribution across the
-plant's canopy. Costs of tissue respiration and turnover are then
-subtracted. The remaining biomass is then allocated between growth and
-reproduction. The strategy model thus includes a list of physiological
-rules and associated parameters.
+The core of {\plant} is a model for an individual plant's
+physiological strategy as specified by its traits
+(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
+production for a plant, given its size and a light environment. Gross
+photosynthetic income is estimated from total leaf area and the light
+distribution across the plant's canopy. Costs of tissue respiration
+and turnover are then subtracted. The remaining biomass is then
+allocated between growth and reproduction.  
+%
+The key outputs needed by higher levels (see below) is growth rate,
+mortality rate and rate of seed production.  However additional
+quantities are also computed, such as total assimilation and the
+respiration, turnover, and allocation rates for different tissues (see
+Appendix \ref{sec:plant})
 
-The core job of the physiological model is to take size, light
-environment and physiological parameters as inputs, and to return growth rate,
-mortality and fecundity as outputs, in other words to move from
-individual growth to demography. However additional quantities are also
-available as output, such as total assimilation and the respiration,
-turnover, and allocation rates for different tissues.
+The default physiological model used in {\plant} largely reflects that
+presented in \citet{Falster-2011}, but with extensions allowing for
+diameter growth also to be estimated. To properly model trait
+variation, different parameters (such as rates of leaf turnover,
+respiration, photosynthesis per leaf area) in the model are linked via
+assumed tradeoffs. For example, the trait LMA is used to estimate the
+rate of leaf turnover, based on observed scaling relationship spanning
+across diverse vegetation and plant functional global
+\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
 
-The default physiological model used in \texttt{TREE} largely reflects that
-presented in \citet{Falster-2011}, but with extensions allowing for diameter
-growth also to be estimated. To properly model trait variation,
-different parameters (such as rates of leaf turnover, respiration, photosynthesis 
-per leaf area) in the model are linked via assumed tradeoffs. For
-example, the trait LMA is used to estimate the rate of leaf turnover,
-based on observed scaling relationship spanning across diverse vegetation and
-plant functional  global \citep{Wright-2004} (see 
-Appendix \ref{sec:FFW16} for details).
+The current parametrisation of the physiological model is based on
+\citet{Falster-2011}. Parameter values were drawn from available
+literature (see Appendix \ref{sec:FFW16}, table 3); however for some
+parameters limited or no data were readily available. Providing
+updated parameter sets is trivial; the challenge is to infer suitable
+values by constraining outputs from {\plant} against diverse data
+sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
+are still actively working, and one of the major use cases we envisage
+for {\plant}.
 
@@
 
-\texttt{TREE} includes a default physiological model (fully described and derived
-in the Appendix \ref{sec:FFW16}). However a feature of the package is that alternative
-physiologies can be used, and the parameters of each model can be
-altered. Trait differences are accounted for by altering relevant
-parameters (see Appendix \ref{sec:parameters}).
+While we have implemented one physiological model (\texttt{FFW16}),
+the package is designed to allow arbitrary additional physiologies to
+be introduced.  That is, rather than varying the parameters of the
+equations, the equations themselves can be changed, additional
+resources could be competed for, and the higher level functionality of
+the package can still be used.  As an intermediate level, we allow for
+parameters to imply relationships among arbitrary other parameters,
+allowing straightforward modelling of trade-offs.  For example,
+setting leaf mass per unit area (LMA) affects ....
+% TODO[RGF]: please complete (running out of time here)
 
@@
 
-With above functionality, \texttt{TREE} can be used to estimate essential
-physiological rates for individual plants (Fig. \ref{fig:plant}). The
-function \texttt{grow\_plant\_to\_size} takes a given strategy and light
-environment and grows the plant, producing a trajectory of size over
-time (Fig. \ref{fig:plant}a). This is achieved by integrating an
-Ordinary Differential Equation (ODE) of size-dependent growth rate (see
-Appendix \ref{sec:demography} for more
-details). As the plant grows, allocation to different tissue types varies, and
-the composition of the plant changes to include more stem and less
-leaf (Fig. \ref{fig:plant}b). This in turn affects maintenance and
-turnover costs and the growth rate of plants (Fig. \ref{fig:plant}c).
-This shift in allocation alone generates a distinctive hump-shaped pattern of absolute
-growth with size \citep{King-2011}, and equally the widely regarded
-decline in relative growth rate with size from birth onwards
+With the basic physiological model, {\plant} can be used to estimate
+essential physiological rates for individual plants
+(Fig. \ref{fig:plant}). The function \texttt{grow\_plant\_to\_size}
+takes a given strategy and light environment and grows the plant,
+producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
+is achieved by integrating an Ordinary Differential Equation (ODE) of
+size-dependent growth rate (see Appendices \ref{sec:demography} and
+\ref{sec:plant} for more details). As the plant grows, allocation to
+different tissue types varies, and the composition of the plant
+changes to include more stem and less leaf
+(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
+costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
+shift in allocation alone generates a distinctive hump-shaped pattern
+of absolute growth with size \citep{King-2011}, and equally the widely
+regarded decline in relative growth rate with size from birth onwards
 \citep{Enquist-2007}.
@@
 
+% TODO: missing something saying we are considering an infinitely
+% large population here.
 Modelling the development of a competitive hierarchy within a patch
@@
 
+% TODO: This is not really a design feature. That heading is almost
+% extinct in the plant section and missing in the last - perhaps just
+% merge with the above section?
 \subsection{Design features}
@@
 approximated via a Partial Differential Equation (PDE) (see Appendix
-\ref{sec:demography}). The same PDE has been shown to capture average behaviour
+\ref{sec:demography}). The same PDE captures average behaviour
 across a large number of small patches \citep{Moorcroft-2001}. The model may 
@@
 Pacala-1996}.
+% TODO: I added "useful" in here, but more changes perhaps?
 Moreover, it remains unclear whether resolving spatial details within a
-patch would provide further ecological detail, in part because the
+patch would provide further useful ecological detail, in part because the
 process of competitive thinning tends to break down spatial clusters
@@
 
+% TODO: early in this paragraph include a pointer to the appropriate
+% vignette that justifies it
 We introduce two extensions to the EBT: i) a new method for estimating
@@
 size-asymmetric competitive feedbacks, such as occur under strong
-competition for light. The original EBT
-\citep{Deroos-1997, Deroos-1992, Deroos-1988} proceeds by approximating
-the first and second moments of the density distribution \(n\) within
-each cohort. In \texttt{TREE} we use a different -- but from a theoretical
-perspective equally valid --  approach. Instead of tracking first and second
-moments of the size-distribution within cohorts, we track a point mass
-estimate of \(n\) along the trajectory for the boundary
-of each cohort. We found this approach preferable because it does not
-require us to maintain a separate set of equations for the smallest
-cohort \citep{Deroos-1997}.
+competition for light. The original EBT \citep{Deroos-1997,
+  Deroos-1992, Deroos-1988} proceeds by approximating the first and
+second moments of the density distribution \(n\) within each
+cohort. In {\plant} we use a different approach. Instead of tracking
+first and second moments of the size-distribution within cohorts, we
+track a point mass estimate of \(n\) along the trajectory for the
+boundary of each cohort. We found this approach preferable because it
+does not require us to maintain a separate set of equations for the
+smallest cohort \citep{Deroos-1997}.
 
@@
 individuals born at similar times can diverge substantially over time
-(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography}).
+(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography} and \ref{sec:cohort-spacing}).
 This is can become a problem later during patch
@@
 distribution of cohort introduction times across patch age (Fig.
-\ref{fig:patch}a), generally with tighter spacing of cohorts in 
+\ref{fig:patch}a), with considerably tighter spacing of cohorts in 
 younger patches.
@@
 
+% TODO: not a paragraph.  Delete?  Or merge together with one of the
+% others?
 Modelling of size-structured dynamics via solving of a deterministic PDE
@@
 
-First, \texttt{TREE} can simulate patterns of patch development after
+First, {\plant} can simulate patterns of patch development after
 disturbance, including self-thinning behaviours and successional
@@
 combination of physiological and structural changes
-\citep{Binkley-2002, Smith-2001, Ogawa-2010, Coomes-2007}. \texttt{TREE} enables
+\citep{Binkley-2002, Smith-2001, Ogawa-2010, Coomes-2007}. {\plant} enables
 the effects of competitive feedbacks to be incorporated, in addition to
-any intrinsic physiological factors. Moreover, \texttt{TREE} allows for the
+any intrinsic physiological factors. Moreover, {\plant} allows for the
 contributions of different species to be mapped (Fig. \ref{fig:patch}c).
 
-While many vegetation models still lack size structure, the importance of
-size-structure for modelling phenomena such as allocation and production is
-being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
+% TODO: this orphan sentence might belong somewhere else?
+% While many vegetation models still lack size structure, the importance of
+% size-structure for modelling phenomena such as allocation and production is
+% being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
 
@@
 Identifying conditions promoting trait diversity is a valuable use-case
-for \texttt{TREE}. Coexisting plant species differ in a range of physiological
+for {\plant}. Coexisting plant species differ in a range of physiological
 traits, yet the conditions allowing for these different types to coexist
@@
 
-\texttt{TREE} makes use of much existing software, including the programming
-language C++; the \texttt{R} computational environment \citep{R-2015}, the \texttt{R}
+{\plant} makes use of much existing software, including the programming
+language C++; the R computational environment \citep{R-2015}, the R
 packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
@@
 \citep{Wickham-2011}; and the Boost Library for C++
-\citep{Schaling-2014} via the \texttt{R} package \texttt{BH}
+\citep{Schaling-2014} via the R package \texttt{BH}
 \citep{Eddelbuettel-2015}. Source code is hosted at
-\href{https://github.com/traitecoevo/tree}{github.com/traitecoevo/tree}.
+\href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
 Installation instructions are available at this homepage, and binary
@@
 
-The current parametrisation of the physiological model is based on
-\citet{Falster-2011}. Parameter values were drawn from available literature (see
-Appendix \ref{sec:FFW16}); however for some parameters limited or no data were
-readily available. Providing updated parameter sets is trivial; the challenge 
-is to infer suitable values by constraining  outputs from \texttt{TREE} against diverse data sources
-\citep{Lebauer-2012, Keenan-2013}. This is an area in which we are still actively working.
-
 \section{Closing comments}
 
-The \texttt{TREE} package makes it easier for researchers and forest practitioners
+The {\plant} package makes it easier for researchers and forest practitioners
 alike to investigate a variety of non-linear demographic phenomena.  Rather
@@
 \citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. At the same time,
-much attention is paid to detailed leaf-level physiology. \texttt{TREE} aims to bridge
+much attention is paid to detailed leaf-level physiology. {\plant} aims to bridge
 this gap that emerged in the model landscape, where detailed physiology and
@@
 
-\caption{Processes modelled within \texttt{TREE} include physiological, population, and
+\caption{Processes modelled within {\plant} include physiological, population, and
 evolutionary dynamics.
@@
 
-\section{Default physiological model for \texttt{TREE}}\label{sec:FFW16}
+\section{Default physiological model for {\plant}}\label{sec:FFW16}
 
-See attached file \url{tree_physiology.pdf}
+See attached file \url{vignette_physiology.pdf}
 
-\section{Modelling demography in the \texttt{TREE} package}\label{sec:demography}
+\section{Modelling demography in the {\plant} package}\label{sec:demography}
 
-See attached file \url{tree_demography.pdf}
+See attached file \url{vignette_demography.pdf}
+
+\section{Plant level properties}\label{sec:plant}
+
+See attached file \url{vignette_plant.R}
+
+\section{Cohort spacing}\label{sec:cohort-spacing}
+
+See attached file \url{vignette_cohort_spacing.R}
 
@@
 
-\section{Calculating fitness in the \texttt{TREE} package}\label{sec:fitness}
+\section{Calculating fitness in the {\plant}  package}\label{sec:fitness}
 

### 2015-05-28 R/figures_patch.R
Compute average leaf area over metapopulation

@@ -60,9 +60,18 @@ figure_patch <- function(data) {
   axis(1, labels=FALSE)
 
   ## Then, the total leaf area.
+  cols <- c("black", cols)
+  lty <- c(3, 1, 1)
   y <- cbind(total=rowSums(data$area_leaf), data$area_leaf)
-  matplot(t, y, lty=c(3, 1, 1), col=c("black", cols), type="l",
+  matplot(t, y, lty=lty, col=cols, type="l",
           xlab="Time (years)", ylab="Leaf area index", las=1)
   label_panel(3)
   mtext("Time (years)", 1, 2.2, cex=.7)
+
+  # find average value over metapopulation
+  y_av <- apply(y, 2, function(x) tree:::trapezium(t, x*data$patch_density))
+  for(i in seq_len(3)) {
+    axis(4, at=y_av[i], tck=0.1, col.ticks=cols[i], lty = lty[i])
+  }
+  axis(1, at=108, label="Av")
 }

### 2015-05-28 ms.tex
More comments and changes

@@
 individuals born at similar times can diverge substantially over time
-(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography} and \ref{sec:cohort-spacing}).
-This is can become a problem later during patch
-development because much of the population can be located between
-widely-spaced cohorts, delivering low numerical precision. To maintain
-numerical precision we therefore use an iterative algorithm to adaptively refine
-the cohort spacing until the trajectories of adjacent cohorts remain
-adequately resolved throughout development of the patch (see
-Appendix \ref{sec:demography} for details). This refinement causes an uneven
-distribution of cohort introduction times across patch age (Fig.
-\ref{fig:patch}a), with considerably tighter spacing of cohorts in 
-younger patches.
+(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
+\ref{sec:patch}, and \ref{sec:cohort-spacing}). This is can become a
+problem later during patch development because much of the population
+can be located between widely-spaced cohorts, delivering low numerical
+precision. To maintain numerical precision we therefore use an
+iterative algorithm to adaptively refine the cohort spacing until the
+trajectories of adjacent cohorts remain adequately resolved throughout
+development of the patch (see Appendix \ref{sec:demography} for
+details). This refinement causes an uneven distribution of cohort
+introduction times across patch age (Fig.  \ref{fig:patch}a), with
+considerably tighter spacing of cohorts in younger patches.
 
@@
 disease outbreak
+% TODO: Too many references
 \citep{Connell-1978, White-1979, Chambers-2013, Bormann-1979, Clark-1991, Coomes-2007}.
@@
 
+% TODO[RGF]: Again, I'd drop the "design features" thing here - it's
+% not helpful I think.
 \subsection{Design features}
 
-Having simulated a single patch, it is straightforward to scale up from
-a single patch to an entire metapopulation, at least for the situation
-of metapopulations at an equilibrium mix of patch ages and where
-disturbances remove all established vegetation. The frequency-density
+In {\plant} we assume an infinite number of patches all experiencing
+the same disturbance regime, and asusme that the disturbances remove
+all established vegetation.  With this approach we can scale up from
+simulating a single patch to simulate an entire metapopulation.
+% TODO: Is the 1926 paper *really* needed?
+The frequency-density
 \(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
@@
 
-The main numerical challenge in stepping from single patch
-to metapopulation is to identify the equilibrium seed rain of the
-metapopulation. For a given seed rain input, the model produces a seed rain output. 
-The equilibrium seed rain is the vector of seed rain
-where input seed rain equals output seed rain. Finding this equilibrium
-requires finding a root of a multidimensional function, subject to the
+The main numerical challenge in stepping from single patch to
+metapopulation is to identify the equilibrium seed rain of the
+metapopulation. For a given seed rain input, the model produces a seed
+rain output.
+% TODO: need to be clearer about what seed rain output means in this
+% context?
+The equilibrium seed rain is the vector of seed rain where input seed
+rain equals output seed rain. Finding this equilibrium requires
+finding a root of a multidimensional function, subject to the
 constraints that this root is stable (e.g., the trivial equilibrium of
-zero seed rain always exists, but is often not stable) (Appendix \ref{sec:fitness});
+zero seed rain always exists, but is often not stable) (Appendix
+\ref{sec:equilibrium}).
 
@@
 
+% TODO[RGF]: This section is closer to the design features than to the
+% use cases, I think.  Not sure what to suggest here.  But we need
+% something that indicates how this can be used.  I suggest a complete
+% overhaul of the second paragraph as that gets closest to a way of
+% using the model.
+
 As previously noted \citep{Kohyama-1993, Moorcroft-2001, Falster-2011} a
 feature of the size-structured metapopulation concept implemented here
-is that it reconciles equilibrium and non-equilibrium approaches to
+is that it balances equilibrium and non-equilibrium approaches to
 modelling ecological dynamics
@@
 
-Beyond its conceptual appeal, this idea of a dynamic equilibrium allows
-us to close the demographic feedback loop and thereby create a
+Beyond its conceptual appeal, this idea of a dynamic equilibrium
+allows us to close the demographic feedback loop and thereby create a
 self-regulating system. The equilibrium seed rain for multi-species
 communities is approached demographically, incorporating the combined
-effects of the model's physiological rules, species traits, competitive
-interactions and the prevailing disturbance regime on the
-size-distributions of plants within a metapopulation of patches. Unlike
-some vegetation models, there is no need to set the density of seeds or
-seedlings, these arise as outputs from the model. Characteristics of the
-vegetation such as total leaf area, biomass, productivity and allocation
-to different plant parts also arise as emergent properties (Fig.
-\ref{fig:patch}c). Similarly, distributions of abundance, leaf area and
-growth rate with respect to size emerge at both patch scale and scale of
-metapopulation (Fig. \ref{fig:emergent}). Note in particular how a single
-physiological model produces a variety of height-density relationships, 
-dependent on patch age, as well as an emergent average across all patches 
-(Fig. \ref{fig:emergent}). This emergent relationship might then be 
-compared to the observed distributions in large forest plots \citep{Muller-2006}.
+effects of the model's physiological rules, species traits,
+competitive interactions and the prevailing disturbance regime on the
+size-distributions of plants within a metapopulation of
+patches. Unlike some vegetation models, there is no need to set the
+density of seeds or seedlings, these arise as outputs from the
+model. Characteristics of the vegetation such as total leaf area,
+biomass, productivity and allocation to different plant parts also
+arise as emergent properties (Fig.  \ref{fig:patch}c). Similarly,
+distributions of abundance, leaf area and growth rate with respect to
+size emerge at both patch scale and scale of metapopulation
+(Fig. \ref{fig:emergent}). Note in particular how a single
+physiological model produces a variety of height-density
+relationships, dependent on patch age, as well as an emergent average
+across all patches (Fig. \ref{fig:emergent}, see more detail in
+appendix \ref{sec:emergent}). This emergent relationship might then be
+compared to the observed distributions in large forest plots
+\citep{Muller-2006}.
 
@@
 reproduction ratio \(R\) is simply the total seed output of the new
-type, averaged across the metapopulation. The new type can invade when
+type, averaged across the metapopulation. A new type can invade when
 \(R > 1\).
@@
 
-Fitness landscapes will be positive in regions that allow invasion, and
-zero for species that are at equilibrium. The slope of the fitness
-landscape at a resident indicates the direction of selection; a negative
-slope (dashed line in Fig. \ref{fig:fitness}a) indicates selection for
-smaller trait values. Fitness landscapes are constantly reshaped
-as species are introduced and evolve, reflecting the non-linear
-demographic feedbacks that occur under frequency-dependent evolution
- \citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient, can lead to an
-evolutionary branching point \citep{Geritz-1998,Dieckmann-2007};
-at this point there is no directional selection (solid line in
-figure \ref{fig:fitness}a) but invasion is possible both above and below
-it in trait space. It is often possible
-to find a configuration where fitness is negative everywhere (solid line
-in Fig. \ref{fig:fitness}b) except for the zero points where species are
-present. This community is immune to invasion from any other strategies
-\citep{Geritz-1998}.
+Fitness landscapes will be positive in regions that allow invasion,
+and zero for species that are at equilibrium. The slope of the fitness
+landscape at a resident indicates the direction of selection; a
+negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
+selection for smaller trait values (see appendix \ref{sec:fitness} for
+details). Fitness landscapes are constantly reshaped as species are
+introduced and evolve, reflecting the non-linear demographic feedbacks
+that occur under frequency-dependent evolution
+\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
+can lead to an evolutionary branching point
+\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
+directional selection (solid line in figure \ref{fig:fitness}a) but
+invasion is possible both above and below it in trait space. It is
+often possible to find a configuration where fitness is negative
+everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
+points where species are present. This community is immune to invasion
+from any other strategies \citep{Geritz-1998}.
 
@@
 packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
-\texttt{R6}\citep{Chang-2014}, and \texttt{testhat}
+\texttt{R6} \citep{Chang-2014}, and \texttt{testhat}
 \citep{Wickham-2011}; and the Boost Library for C++
@@
 Installation instructions are available at this homepage, and binary
-releases will also be available there.
+releases will also be available there.  (Note that {\plant} does not
+currently compile on Windows, but this will change soon as R core
+upgrade the Windows C++ compiler).
 
@@
 alike to investigate a variety of non-linear demographic phenomena.  Rather
-than provide a model with a few routines, we have developed an extensible
+than provide a model with a few entry points, we have developed an extensible
 framework that we hope can be used to answer a number of questions at a
@@
 interest in the assembly of trait mixtures in communities. However a variety
-of other uses would be possible, and also extensions to address other facets
+of other uses would be possible, and also extensions to address other aspects
 of community assembly. For example,  the physiology of growth could be
@@
 growth rate is zero (x intercept).  The light level used in panels (a) and (c)
-is indicated with dots.} [Could you put delta symbols on the y axis rather than d?
+is indicated with dots.  See Appendix \ref{sec:plant} for details and code.}
+% TODO[Note from James]: [Could you put delta symbols on the y axis rather than d?
 Also there is alot going in these figures with 4 lines each.
@@
 \caption{Emergent size-distributions across a metapopulation.
-Grey lines show patterns within individual patches differing in time since disturbance (age), with blue lines indicating patches of a specific age.
-Panels show (a) patterns of abundance, (b) leaf area and (c) height growth rate against size.
-Note that units of panels (a) and (b) are per unit height and per unit ground area, respectively.}
+Grey lines show patterns within individual patches differing in time
+since disturbance (age), with blue lines indicating patches of a
+specific age. Panels show (a) patterns of abundance, (b) leaf area and
+(c) height growth rate against size. Note that units of panels (a) and
+(b) are per unit height and per unit ground area, respectively.  See Appendix \ref{sec:emergent} for details and code.}
 \label{fig:emergent}
@@
 line shows a stable fitness landscape where both species exist on a peak in
-the fitness landscape.  At this point no further invasion is possible.}
+the fitness landscape.  At this point no further invasion is
+possible.  See Appendix \ref{sec:fitness} for details and code.}
 \label{fig:fitness}
@@
 
-\section{Solving for equilibrium seed rain}\label{sec:seed}
+\section{Solving for equilibrium seed rain}\label{sec:equilibrium}
 
 See attached file \url{vignette_equilibrium.R}
+
+\section{Patch level dynamics}\label{sec:patch}
+
+See attached file \url{vignette_patch.R}
+
+\section{Patch level emergent properties}\label{sec:emergent}
+
+See attached file \url{vignette_emergent.R}
 

### 2015-05-28 R/figures_patch.R
Fix partial match

@@ -73,5 +73,5 @@ figure_patch <- function(data) {
   for(i in seq_len(3)) {
     axis(4, at=y_av[i], tck=0.1, col.ticks=cols[i], lty = lty[i])
   }
-  axis(1, at=108, label="Av")
+  axis(1, at=108, labels="Av")
 }

### 2015-05-28 R/figures_emergent.R
Plot averages with respect to height across metapopulation.
For #13

Squashed commit of the following:

commit 1d0640e721e018abc5ba7dfc459a45810ebe6172
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 20:34:40 2015 +1000

    Increase resolution

commit d8df1369e081494d722500912f395b5e02079c65
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 18:56:47 2015 +1000

    interpolate using spline_loglog

commit 85cec05505f81e0f17f692dd60d5fc98b659b9e2
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 17:09:50 2015 +1000

    First cut at get

@@ -1,5 +1,10 @@
-make_emergent_data <- function() {
+make_emergent_data <- function(hires=TRUE) {
   p0 <- ebt_base_parameters()
+  if(hires) {
+    p0$control$schedule_eps <- 2e-6
+    p0$control$ode_tol_rel <- 1e-6
+    p0$control$ode_tol_abs <- 1e-6
+  }
   p0$disturbance_mean_interval <- 30.0
   p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
   p1 <- build_schedule(p1)
@@ -13,12 +18,22 @@ figure_emergent <- function(data) {
   last <- function(x) {
     x[[length(x)]]
   }
+  splinefun_loglog2 <- function(x,y,xx){
+    ret <- splinefun_loglog(x,y)(xx)
+    ii <- xx <= max(x, na.rm=TRUE) & xx >= min(x, na.rm=TRUE) 
+    ret[!ii] <-0 
+    ret
+  }
 
   species <- data$species[[1]]
   height <- t(species["height", , ])
   log_density <- t(species["log_density", , ])
   density <- exp(log_density)
 
+  # times for background grey lines
+  j <- seq(1, length(data$time), by=10)
+
+  # time to highlight in blue
   times <- c(5, 10, 20, 40, last(data$time))
   i <- vapply(times, closest, integer(1), data$time)
   blues <- c("#DEEBF7", "#C6DBEF", "#9ECAE1", "#6BAED6",
@@ -34,11 +49,37 @@ figure_emergent <- function(data) {
   xxx <- lapply(tmp, function(x) matrix(x$species[[1]]$ode_rates, 4)[1, ])
   growth_rate <- do.call("cbind", plant:::pad_matrix(xxx))
 
-  op <- par(mfrow=c(3, 1), mar=c(2.1, 4.1, .5, .5), oma=c(2, 0, 0, 0))
+  # interpolate density in each patch to common set of heights
+  hh <- seq_log_range(range(height, na.rm=TRUE), 500)
+  xxx <- lapply(seq_along(data$time), function(i) 
+              splinefun_loglog2(height[,i],density[,i], hh))
+  n_hh <- do.call("cbind", plant:::pad_matrix(xxx))
+
+  # interpolate leaf area in each patch to common set of heights
+  xxx <- lapply(seq_along(data$time), function(i) 
+              splinefun_loglog2(height[,i], leaf_area[,i], hh))
[…]

### 2015-05-29 release.sh
Script for making release
For #27

@@ -1,0 +1,8 @@
+#!/bin/sh
+set -e
+remake all
+rm -rf plant_paper
+mkdir plant_paper
+cp -rp vignettes figures ms.pdf plant_paper
+zip -r plant_paper.zip plant_paper
+rm -rf plant_paper

### 2015-05-29 ms.tex
Plot averages with respect to height across metapopulation.
For #13

Squashed commit of the following:

commit 1d0640e721e018abc5ba7dfc459a45810ebe6172
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 20:34:40 2015 +1000

    Increase resolution

commit d8df1369e081494d722500912f395b5e02079c65
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 18:56:47 2015 +1000

    interpolate using spline_loglog

commit 85cec05505f81e0f17f692dd60d5fc98b659b9e2
Author: Daniel Falster <daniel.falster@mq.edu.au>
Date:   Thu May 28 17:09:50 2015 +1000

    First cut at get

@@
 \usepackage{ms}
-\usepackage[sort&compress]{natbib}
+\usepackage[]{natbib}
 \raggedright
@@
 \item
-  Here we introduce the {\plant} package (TRait Ecology and Evolution), an
-  extensible framework for modelling plant demography across the entire
-  life cycle and in structured metapopulations, via coupled differential equations.
+  Here we introduce the {\plant} package, an
+  extensible framework for modelling trait-based demography, ecology and evolution 
+  in vegetation, via coupled differential equations.
   At its core, {\plant} is an
@@
 area, it increases its potential to generate photosynthate.  In
-contrast, as indivuduals grow, they must use increasing fractions of
+contrast, as individuals grow, they must use increasing fractions of
 this photosynthetic income to building and maintaining support tissues
@@
 reproduction \citep{Thomas-2011}. Consequently rates of growth and
-mortality change with individual size \citep{Muller-2006, Thomas-2011,
-  Ruger-2011, Wright-2010}.
+mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
 %
@@
 
-% NOTE[RGF]: I thought this was way overrefrenced and so just kept a
-% few oldies.  Please re-add any key ones that you want to see here.
 Although the importance of size-structure for vegetation function and
 demography has been long been recognised
-\citep[e.g.,][]{Harper-1977, Shugart-1980, Hara-1984},
+\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987},
 most large-scale mechanistic models of plant growth either omit it
 entirely, or only include coarse approximations
-\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. The likely
+\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. The likely
 reason is the computational
@@
 early 1990s, size-structured ``gap'' models were widely used
-\citep[e.g.][]{Huston-1987, Shugart-1980}. These gap models were
+\citep[e.g.][]{Shugart-1980,Huston-1987}. These gap models were
 parameterised in terms of growth using statistical functions; this meant 
@@
 The next generation of models focussed on establishing core physiology for 
-a limited number of functional types, and
+a limited number of functional types and
 sacrificed details of demography in order to achieve global coverage
-\citep{Haxeltine-1996,Woodward-1995}. This history has brought us to the
+\citep{Woodward-1995,Haxeltine-1996}. This history has brought us to the
 current situation where all but a couple
@@
 are focused on modelling fluxes carbon and water, but say little about
-demographic behaviour, and consequently are also unable to account for
+demographic behaviour; consequently they are also unable to account for
 any non-linear demographic feedbacks that may arise.
+% While many vegetation models still lack size structure, the importance of
+% size-structure for modelling phenomena such as allocation and production is
+% being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
 
@@
 questions about species coexistence also lack size-structured dynamics
-\citep[e.g.][]{Calcagno-2006, Geritz-1995, Leimar-2013,
-  Levin-1974,MacArthur-1967,Tilman-1985}.  It is common to assume
+\citep[e.g.][]{MacArthur-1967,Levin-1974,Tilman-1985,Geritz-1995, 
+Calcagno-2006, Leimar-2013}.  It is common to assume
 competitive interactions are influenced by size-related traits such as
@@
 with type of data often collected in the field, such as physiological
-leaf traits, and demographic data on growth and mortality in long-term
+leaf traits and demographic data on growth and mortality in long-term
 monitoring plots \citep{Adler-2013}.
@@
 trait variation on the demography of individual plants, of patches of
-competing plants, and of meta-populations structured by a prevailing
+competing plants and of meta-populations structured by a prevailing
 disturbance regime.
@@
 potentially giving a theoretical foundation for observed empirical
-patterns \citep{Wright-2010,Baltzer-2007}.
+patterns \citep{Baltzer-2007,Wright-2010}.
 
@@
 {\plant} allows for emergent phenomena across a metapopulation of patches
-to  be investigated, and thus provides a tractable mechanism to scale from
+to  be investigated, thereby providing a workable mechanism to scale from
 individual plant to  vegetation, incorporating the non-linear effects of size
-and traits on vegetation outcomes. In  this sense {\plant} is similar to
-the Ecosystem Demography \citep{Moorcroft-2001}  and LPJ-GUESS
+and traits on vegetation outcomes. 
+% To strigul?
+In  this sense {\plant} is similar to
+the \textsc{Ecosystem Demography} \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
 \citep{Smith-2014} models. However, whereas these other models were designed
-to primarily capture large-scale regional biogeocehcmisty driven by climatic
+to primarily capture large-scale regional biogeochemistry driven by climatic
 variables,  {\plant} was designed so that the rules driving the
@@
 manipulated and studied in detail. The three models are therefore highly 
-complimentary, and also connect with the increasingly sophisticated statistical 
+complimentary, they also connect with the increasingly sophisticated statistical 
 methods now available to analyse size-structured data
@@
 details at different levels of ecological organisation. Each section
-provides a short description of the required methods, design
-considerations, and then example applications. Detailed technical
+provides a short description of the required methods and then example 
+applications. Detailed technical
 documents are provided as supplementary information (see Appendices),
@@
 % NOTE: Keep the repetition with above under control.
-% NOTE: LV are not really mechanistic
 
@@
 %
-On top of this physiolgical model, the package implements methods for
+On top of this physiological model, the package implements methods for
 population and adaptive dynamics (Fig.  \ref{fig:schematic}) following
@@
 studied at the three levels: individual plants, stands of competing
-plants, and entire metapopulations.
+plants and entire metapopulations.
 %
@@
 quantities are also computed, such as total assimilation and the
-respiration, turnover, and allocation rates for different tissues (see
+respiration, turnover and allocation rates for different tissues (see
 Appendix \ref{sec:plant})
@@
 respiration, photosynthesis per leaf area) in the model are linked via
-assumed tradeoffs. For example, the trait LMA is used to estimate the
+assumed tradeoffs. For example, the trait \textsc{lma} is used to estimate the
 rate of leaf turnover, based on observed scaling relationship spanning
@@
 sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
-are still actively working, and one of the major use cases we envisage
+are still actively working and one of the major use cases we envisage
 for {\plant}.
-
-\subsection{Design features}
 
@@
 equations, the equations themselves can be changed, additional
-resources could be competed for, and the higher level functionality of
+resources could be competed for and the higher level functionality of
 the package can still be used.  As an intermediate level, we allow for
@@
 allowing straightforward modelling of trade-offs.  For example,
-setting leaf mass per unit area (LMA) affects ....
-% TODO[RGF]: please complete (running out of time here)
+setting leaf mass per unit area (\textsc{lma}) affects leaf turnover rate,
+because it is known that leaves which are cheaper to build have shorter 
+lifespan \citep{Wright-2004}.
 
@@
 producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
-is achieved by integrating an Ordinary Differential Equation (ODE) of
+is achieved by integrating an Ordinary Differential Equation of
 size-dependent growth rate (see Appendices \ref{sec:demography} and
 \ref{sec:plant} for more details). As the plant grows, allocation to
-different tissue types varies, and the composition of the plant
+different tissue types varies and the composition of the plant
 changes to include more stem and less leaf
@@
 shift in allocation alone generates a distinctive hump-shaped pattern
-of absolute growth with size \citep{King-2011}, and equally the widely
+of absolute growth with size \citep{King-2011} and equally the widely
 regarded decline in relative growth rate with size from birth onwards
@@
 
-% TODO: missing something saying we are considering an infinitely
-% large population here.
-Modelling the development of a competitive hierarchy within a patch
-requires that both the initial size distribution and the inflow of new
-recruits be specified. We have primarily been interested in dynamics of
-a patch recovering from disturbance, and have therefore started with an
-empty patch and constant flow of seeds from a global seed rain (Fig.
-\ref{fig:schematic}b).
-
-% TODO: This is not really a design feature. That heading is almost
-% extinct in the plant section and missing in the last - perhaps just
-% merge with the above section?
-\subsection{Design features}
-
-When modelling a patch, we are interested in modelling changes in the
-size-density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
-individuals with traits \(x\) and height \(h\) within a patch of age \(a\). Density
-is measured as individuals per unit height and per unit ground area.
-
-We assume that patches are large, and are vertically but not horizontally
+We assume that patches are vertically but not horizontally
 structured. In other words, we account for size differences but not spatial 
@@
 simulating size-structured dynamics
-\citep{Moorcroft-2001, Huston-1987, Smith-2014,Kohyama-1993}. The assumption of a large
-patch means the population of plants within the patch -- and thus the 
-dynamics of \(n\) -- behave deterministically \citep{Deroos-1997}, and can therefore be
-approximated via a Partial Differential Equation (PDE) (see Appendix
-\ref{sec:demography}). The same PDE captures average behaviour
+\citep{Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
+we assume that patches are very large -- this means the population dynamics 
+within the patch behave deterministically \citep{Deroos-1997} and can therefore 
+be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
+\ref{sec:demography}). The same \textsc{pde} captures average behaviour
 across a large number of small patches \citep{Moorcroft-2001}. The model may 
@@
 Ideally one would also consider spatial interactions within patches,
-however such models are very computationally demanding, and also
+however such models are very computationally demanding and also
 impossible to solve in a deterministic fashion \citep{Huston-1987, 
-Pacala-1996}.
-% TODO: I added "useful" in here, but more changes perhaps?
-Moreover, it remains unclear whether resolving spatial details within a
+Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
 patch would provide further useful ecological detail, in part because the
@@
 
-% TODO: early in this paragraph include a pointer to the appropriate
-% vignette that justifies it
-We introduce two extensions to the EBT: i) a new method for estimating
-the size-density distribution, and ii) a technique for handling strong
+We introduce two extensions to the \textsc{EBT}: i) a new method for estimating
+the size-density distribution and ii) a technique for handling strong
 size-asymmetric competitive feedbacks, such as occur under strong
-competition for light. The original EBT \citep{Deroos-1997,
-  Deroos-1992, Deroos-1988} proceeds by approximating the first and
+competition for light (see Appendix \ref{sec:demography} for
+details). The original \textsc{EBT} \citep{Deroos-1988,
+  Deroos-1992, Deroos-1997} proceeds by approximating the first and
 second moments of the density distribution \(n\) within each
@@
 adaptive refining of time points at which new cohorts are introduced
-into the population. In the original EBT, cohorts were introduced at a
+into the population. In the original \textsc{EBT}, cohorts were introduced at a
 fixed time interval. Under competition, the growth trajectories of
@@
 (Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
-\ref{sec:patch}, and \ref{sec:cohort-spacing}). This is can become a
+\ref{sec:patch} and \ref{sec:cohort-spacing}). This can become a
 problem later during patch development because much of the population
@@
 
-% TODO: not a paragraph.  Delete?  Or merge together with one of the
-% others?
-Modelling of size-structured dynamics via solving of a deterministic PDE
+Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
 enables multiple complex non-linear demographic phenomena to be
-investigated.
-
-First, {\plant} can simulate patterns of patch development after
+investigated. First, {\plant} can simulate patterns of patch development after
 disturbance, including self-thinning behaviours and successional
@@
 combination of physiological and structural changes
-\citep{Binkley-2002, Smith-2001, Ogawa-2010, Coomes-2007}. {\plant} enables
+\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
 the effects of competitive feedbacks to be incorporated, in addition to
@@
 contributions of different species to be mapped (Fig. \ref{fig:patch}c).
-
-% TODO: this orphan sentence might belong somewhere else?
-% While many vegetation models still lack size structure, the importance of
-% size-structure for modelling phenomena such as allocation and production is
-% being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
 
@@
 disease outbreak
-% TODO: Too many references
-\citep{Connell-1978, White-1979, Chambers-2013, Bormann-1979, Clark-1991, Coomes-2007}.
+\citep{White-1979, Bormann-1979, Chambers-2013}.
 The vegetation thus comprises a collection of patches differing in time
@@
 
-% TODO[RGF]: Again, I'd drop the "design features" thing here - it's
-% not helpful I think.
-\subsection{Design features}
-
 In {\plant} we assume an infinite number of patches all experiencing
-the same disturbance regime, and asusme that the disturbances remove
+the same disturbance regime; we also assume that the disturbances remove
 all established vegetation.  With this approach we can scale up from
 simulating a single patch to simulate an entire metapopulation.
-% TODO: Is the 1926 paper *really* needed?
 The frequency-density
@@
 are a large number of patches means the dynamics of \(p\) behave
-deterministically and can be approximated via a second PDE
-\citep{Vonfoerster-1959, Mckendrick-1926} (see Appendix
+deterministically and can be approximated via a second \textsc{pde}
+\citep{ Vonfoerster-1959} (see Appendix
 \ref{sec:demography} for
@@
 metapopulation is to identify the equilibrium seed rain of the
-metapopulation. For a given seed rain input, the model produces a seed
-rain output.
-% TODO: need to be clearer about what seed rain output means in this
-% context?
-The equilibrium seed rain is the vector of seed rain where input seed
+metapopulation. An "input" seed rain is required to simulate 
+the demography of the metapopulation, this in turn produces an "output" 
+number of seeds. The equilibrium seed rain is the vector of seed 
+rain where input seed
 rain equals output seed rain. Finding this equilibrium requires
@@
 
-\subsection{Applications}
-
-% TODO[RGF]: This section is closer to the design features than to the
-% use cases, I think.  Not sure what to suggest here.  But we need
-% something that indicates how this can be used.  I suggest a complete
-% overhaul of the second paragraph as that gets closest to a way of
-% using the model.
-
 As previously noted \citep{Kohyama-1993, Moorcroft-2001, Falster-2011} a
@@
 is that it balances equilibrium and non-equilibrium approaches to
-modelling ecological dynamics
-\citep{Levin-1974, Bormann-1979, Connell-1978, Coomes-2007}. An
+modelling ecological dynamics. An
 equilibrium may be approached at the level of metapopulation, meaning
@@
 residents continue to grow until they are disturbed.
+
+\subsection{Applications}
+% TODO[RGF]:  But we need
+% something that indicates how this can be used.  I suggest a complete
+% overhaul of this paragraph as that gets closest to a way of
+% using the model.
 
@@
 With fitness defined, it becomes possible to model adaptive dynamics
-and community assembly \citep{Dieckmann-2007, Geritz-1998}. This is best
+and community assembly \citep{Geritz-1998,Dieckmann-2007}. This is best
 thought of as occurring on a fitness landscape, a surface with fitness
@@
 \citep{Geritz-1998,Dieckmann-2007}; at this point there is no
-directional selection (solid line in figure \ref{fig:fitness}a) but
+directional selection (solid line in Fig. \ref{fig:fitness}a) but
 invasion is possible both above and below it in trait space. It is
@@
 packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
-\texttt{R6} \citep{Chang-2014}, and \texttt{testhat}
+\texttt{R6} \citep{Chang-2014} and \texttt{testhat}
 \citep{Wickham-2011}; and the Boost Library for C++
@@
 \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
-Installation instructions are available at this homepage, and binary
+Installation instructions are available at this homepage, binary
 releases will also be available there.  (Note that {\plant} does not
@@
 interest in the assembly of trait mixtures in communities. However a variety
-of other uses would be possible, and also extensions to address other aspects
+of other uses would be possible and also extensions to address other aspects
 of community assembly. For example,  the physiology of growth could be
@@
 structure, so remain unable to investigate demographic phenomena
-\citep{Cramer-2001, Dekauwe-2014, Kelley-2013, Sitch-2003}. At the same time,
+\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. At the same time,
 much attention is paid to detailed leaf-level physiology. {\plant} aims to bridge
@@
 
-\caption{Processes modelled within {\plant} include physiological, population, and
+\caption{Processes modelled within {\plant} include physiological, population and
 evolutionary dynamics.
@@
 is indicated with dots.  See Appendix \ref{sec:plant} for details and code.}
-% TODO[Note from James]: [Could you put delta symbols on the y axis rather than d?
-Also there is alot going in these figures with 4 lines each.
-You could make it easier for the reader by adding the dashed lines to the keys]
 \label{fig:plant}
@@
 since disturbance (age), with blue lines indicating patches of a
-specific age. Panels show (a) patterns of abundance, (b) leaf area and
+specific age. Red lines show average relationship across the metapopulation, 
+obtained by weighting the grey lines by patch and individual abundance. 
+Panels show (a) patterns of abundance, (b) leaf area and
 (c) height growth rate against size. Note that units of panels (a) and
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
-\caption{Example fitness landscape for LMA, showing potential for stable
+\caption{Example fitness landscape for \textsc{lma}, showing potential for stable
 coexistence of multiple types.  Fitness is the log of the population growth
 rate - at zero species are at equilibrium, above zero they will increase in
-density.  (a) Landscape generated by a single species.  At LMA of 0.1 (the
-dashed line) there is directional selection for smaller LMA.  At LMA of 0.08
+density.  (a) Landscape generated by a single species.  At \textsc{lma} of 0.1 (the
+dashed line) there is directional selection for smaller \textsc{lma}.  At \textsc{lma} of 0.08
 (solid line) there is a branching point (see inset for more detail).  (b)
-Landscape generated by two species, holing the first at LMA of 0.08.
+Landscape generated by two species, holing the first at \textsc{lma} of 0.08.
 Introducing a species at the maximum fitness value from panel (a) reduces the
 region of positive fitness considerably (dashed line).  Dotted lines show
-subsequent invasions and replacements of the second species, and the solid
+subsequent invasions and replacements of the second species, the solid
 line shows a stable fitness landscape where both species exist on a peak in

### 2015-05-29 update_wiki.sh
Scripts for updating the wiki

@@ -1,0 +1,15 @@
+#!/bin/sh
+remake vignettes_md
+
+if [ -d wiki/ ]; then
+    git -C wiki pull
+else
+    git clone https://github.com/traitecoevo/plant_paper.wiki.git wiki
+fi
+
+cp vignettes/*.md wiki
+cp -r vignettes/figure wiki
+git -C wiki add .
+SHA=$(git rev-parse --short HEAD)
+git -C wiki commit -m "Updated wiki to ${SHA}"
+git -C wiki push

### 2015-05-29 vignettes/src/fitness.R

@@ -1,5 +1,5 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _fitness_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _calculating fitness_"
 ## ---
 
 library(plant)

### 2015-05-29 ms.tex
Update SM links to match file to be uploaded

@@
 
-See attached file \url{vignette_physiology.pdf}
+See attached file \url{vignettes/physiology.pdf}
 
@@
 
-See attached file \url{vignette_demography.pdf}
+See attached file \url{vignettes/demography.pdf}
 
@@
 
-See attached file \url{vignette_plant.R}
+See attached file \url{vignettes/plant.pdf}, knitr source
+\url{vignettes/plant.Rmd}, plain source, \url{vignettes/R/plant.R}
 
-\section{Cohort spacing}\label{sec:cohort-spacing}
+\section{Cohort spacing algorithm}\label{sec:cohort-spacing}
 
-See attached file \url{vignette_cohort_spacing.R}
+See attached file \url{vignettes/cohort_spacing.pdf}, knitr source
+\url{vignettes/cohort_spacing.Rmd}, plain source, \url{vignettes/R/cohort_spacing.R}
 
-\section{Solving for equilibrium seed rain}\label{sec:equilibrium}
+\section{Finding demographic equilibrium}\label{sec:equilibrium}
 
-See attached file \url{vignette_equilibrium.R}
+See attached file \url{vignettes/equilibrium.pdf}, knitr source
+\url{vignettes/equilibrium.Rmd}, plain source, \url{vignettes/R/equilibrium.R}
 
@@
 
-See attached file \url{vignette_patch.R}
+See attached file \url{vignettes/patch.pdf}, knitr source
+\url{vignettes/patch.Rmd}, plain source, \url{vignettes/R/patch.R}
 
@@
 
-See attached file \url{vignette_emergent.R}
+See attached file \url{vignettes/emergent.pdf}, knitr source
+\url{vignettes/emergent.Rmd}, plain source, \url{vignettes/R/emergent.R}
 
-\section{Calculating fitness in the {\plant}  package}\label{sec:fitness}
+\section{Calculating fitness}\label{sec:fitness}
 
-See attached file \url{vignette_fitness.R}
+See attached file \url{vignettes/fitness.pdf}, knitr source
+\url{vignettes/fitness.Rmd}, plain source, \url{vignettes/R/fitness.R}
 

### 2015-05-29 ms.tex
Reworking introduction

@@
   Here we introduce the {\plant} package, an
-  extensible framework for modelling trait-based demography, ecology and evolution 
-  in vegetation, via coupled differential equations.
+  extensible framework for modelling size- and trait-based demography, 
+  ecology and evolution in vegetations.
   At its core, {\plant} is an
@@
   technique. Combined effects of trait-, size- and patch-structured
-  dynamics are integrated into population level estimates of
-  reproductive fitness.
+  dynamics are integrated into metapopulation level estimates of
+  reproductive fitness and vegetation structure.
 \item
@@
 \item
+% todo
   We provide worked examples illustrating {\plant}'s use to study the
@@
 mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
-%
 The patterns of changing growth rate vary by plant traits; leaf
@@
 \citep{Westoby-2002}
-%
 Strong feedbacks emerge between individuals within a forest due to
@@
 without modelling both individual growth and the interactions between
-individuals.
-% TODO[RGF]: link back things about natural selection, successional
-% dynamics, self thinning, trait evolution and species coexistance.
+individuals. The non-linear feedbacks arising from competitive 
+interactions are thought to be crucial for understanding phenomena such 
+as self thinning, successional transitions, trait evolution and 
+species coexistence.
 
-Although the importance of size-structure for vegetation function and
-demography has been long been recognised
-\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987},
-most large-scale mechanistic models of plant growth either omit it
-entirely, or only include coarse approximations
-\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. The likely
-reason is the computational
-challenge of modelling size-structured dynamics. 
-% TODO[RGF]: I don't see the link with the above?  Can this be
-% shortened/tightened?  It's too long for a 3k word paper
-During the 1980s and
-early 1990s, size-structured ``gap'' models were widely used
-\citep[e.g.][]{Shugart-1980,Huston-1987}. These gap models were
-parameterised in terms of growth using statistical functions; this meant 
-that data had to be collected for each species, limiting widespread application. 
-The next generation of models focussed on establishing core physiology for 
-a limited number of functional types and
-sacrificed details of demography in order to achieve global coverage
-\citep{Woodward-1995,Haxeltine-1996}. This history has brought us to the
-current situation where all but a couple
-\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular vegetation
-models lack size-structure. Instead, most models
-are focused on modelling fluxes carbon and water, but say little about
-demographic behaviour; consequently they are also unable to account for
-any non-linear demographic feedbacks that may arise.
-% While many vegetation models still lack size structure, the importance of
-% size-structure for modelling phenomena such as allocation and production is
-% being increasingly noted \citep{Moorcroft-2001, Dekauwe-2014, Falster-2011}.
-
-Similarly, most models used in theoretical ecology to explore
-questions about species coexistence also lack size-structured dynamics
-\citep[e.g.][]{MacArthur-1967,Levin-1974,Tilman-1985,Geritz-1995, 
+Although the importance of size-structure for vegetation dynamics and diversity, 
+has been long been recognised
+\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987}, current research in these
+areas is dominated by models and theory that either omit size-structure 
+entirely, or only include coarse approximations. Theoretical investigations 
+exploring questions about niche-based species coexistence and ecological drift 
+(neutrality) both primarily rely  on unstructured population models 
+\citep[e.g.][]{MacArthur-1967, Tilman-1985, Geritz-1995, Hubbell-2001,
 Calcagno-2006, Leimar-2013}. It is common to assume
@@
 examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
-difficult to connect mechanisms described in these simplistic models
-with type of data often collected in the field, such as physiological
-leaf traits and demographic data on growth and mortality in long-term
-monitoring plots \citep{Adler-2013}.
+difficult to connect and test the mechanisms described in these simplistic models
+with types of data collected in the field, such as physiological
+leaf traits or demographic data from long-term monitoring plots \citep{Adler-2013}.
+Similarly, most large-scale models of vegetation dynamics do not include
+size-structured demography \citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. 
+During the  1980s, size-structured ``gap'' models were widely used
+\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation of models 
+focussed more heavily on biogeochemistry at the expense of demography 
+\citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the importance
+of size-structure is starting again to be recognised \citep{Moorcroft-2001, 
+ Falster-2011, Dekauwe-2014}, all but a couple \citep[e.g.][]{Moorcroft-2001, Smith-2014} 
+of the most popular, large-scale vegetation models lack any real demography or 
+size-structure.
 
@@
 competing plants and of meta-populations structured by a prevailing
-disturbance regime.
-%
-The {\plant} package builds on much existing work investigating the
-effect of traits on vegetation properties and the mechanisms
-facilitating coexistence of trait mixtures
-\citep{Kohyama-1993,Deroos-1997,Moorcroft-2001,Falster-2011,Falster-2015}.
+disturbance regime. In the theoretical ecology literature it is common to
+analyse the full behaviour of a model under its equilibrium conditions.
+{\plant} was developed with this style of analysis in mind, while also 
+allowing for a richer set of ecological dynamics than is possible in 
+population models lacking size structure. {\plant} is thus part
+of an extended thought experiment investigating how species differing
+in traits are able to coexist with one another \citep{Falster-2011,Falster-2015}.
+Understanding and incorporating the demographic outcomes of trait variation 
+is ultimately needed if we are to properly understand how trade-offs in plant function 
+enable species coexistence.
 
-% TODO[RGF]: restructure here - I'm getting lost
-{\plant} provides a platform for investigating how morphological and
-physiological traits influence growth rate and shade tolerance, thus
-potentially giving a theoretical foundation for observed empirical
-patterns \citep{Baltzer-2007,Wright-2010}.
+At the same time, we expect the {\plant} model to be useful for modelling
+other  phenomena, beyond questions about species coexistence. The internal
+structure  of the {\plant} model is superficially similar to the
+\textsc{Ecosystem Demography}  \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
+\citep{Smith-2014} models. Like  {\plant}, these models include size-
+structured competition within a metapopulation of  "patches". However,
+whereas these models were designed to capture large-scale regional
+biogeochemistry driven by climatic variables and rely on simulations,
+{\plant} was designed so that the rules  driving the dynamics of the system
+could be manipulated and emergent demographic  behaviours studied in detail.
+In {\plant} it is possible to grow plants in isolation, in patches of
+competing plants and in disturbance-driven metapopulations. The package is
+thus designed to sit in the middle ground between a large-scale coupled
+simulation model and simple abstract theoretical device. As examples, we
+expect the package will prove  useful for investigating how morphological and
+physiological traits influence growth rate and shade tolerance, and emergent
+phenomena such as multi-species self-thinning,  successional transitions and
+age-related changes in vegetation productivity.
 
-% NOTE[RGF]: I dropped the appeal to novelty (growing an individual
-% plant) because I don't find them very convincing.
-While the applications highlighted reflect our own interests in understanding
-trait-based demography and community assembly, we expect the methods provided
-to be useful in variety of contexts. . Being driven by traits,
-the model can be extended to potentially very many species. Users can also 
-add their own physiological models. Individual plants
-can then be  situated in a patch with competitors and the demography of a
-competing population  investigated, thereby capturing phenomena such as multi-
-species self-thinning,  successional transitions and age-related changes in
-vegetation productivity. {\plant}  models the emergent behaviour of
-patches deterministically, based on a continual  size-density distribution. As
-such, it is not required to specify a particular patch size.  Finally,
-{\plant} allows for emergent phenomena across a metapopulation of patches
-to  be investigated, thereby providing a workable mechanism to scale from
-individual plant to  vegetation, incorporating the non-linear effects of size
-and traits on vegetation outcomes. 
-% To strigul?
-In  this sense {\plant} is similar to
-the \textsc{Ecosystem Demography} \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
-\citep{Smith-2014} models. However, whereas these other models were designed
-to primarily capture large-scale regional biogeochemistry driven by climatic
-variables,  {\plant} was designed so that the rules driving the
-demography of individual trees, patches and metapopulations could be
-manipulated and studied in detail. The three models are therefore highly 
-complimentary, they also connect with the increasingly sophisticated statistical 
-methods now available to analyse size-structured data
-\citep[e.g.][]{Metcalf-2013}.
-
-We describe the general approach of the package and then sections with
-details at different levels of ecological organisation. Each section
-provides a short description of the required methods and then example 
-applications. Detailed technical
-documents are provided as supplementary information (see Appendices),
-and updated versions of technical documents are also available within
-the {\plant} package itself.
+Below we describe the general approach of the package, then provide  sections
+with details at different levels of ecological organisation.  Detailed
+technical documents are provided as supplementary information  (see
+Appendices), and updated versions of technical documents are also available
+within the {\plant} package itself.
 
 \section{The approach}
-
-% NOTE: Keep the repetition with above under control.
 
@@
 functioning of an individual plant and how these are influenced by its
-traits, size and light environment.
-%
+traits, size and light environment. Being driven by traits, the model can be extended to potentially 
+very many species. Users can also  add their own physiological models.
+
 On top of this physiological model, the package implements methods for
-population and adaptive dynamics (Fig.  \ref{fig:schematic}) following
-methods described by \citet{Falster-2011} and
-\citet{Falster-2015}. 
-%
-The dynamics at higher levels of organisation arise as emergent
-properties, driven by growth physiology, competition for light and
-disturbance (Fig.  \ref{fig:schematic}). Demographic phenomena can be
-studied at the three levels: individual plants, stands of competing
-plants and entire metapopulations.
-%
-We developed {\plant} as an extended thought experiment for helping to
-understand competition and succession by having these driven by
-low-level processes rather than being explicitly implemented in the
-model.
-% NOTE: I think that foreshaddowing some of the competition kernels
-% idea here is sensible.
-%
-% TODO: Somewhere, perhaps above, perhaps highlight the issue of doing
-% coexistance modelling with explicit modelling - the models are so
-% well understood it's easy to drive the behaviour.  But we want to
-% have something where we just get coexistence to study how tradeoffs
-% generate it.
-%
-% TODO: perhaps salavageable?
-%
-%   {\plant} was developed with this style of analysis in mind, while
-%   also allowing for a richer set of ecological dynamics than is
-%   possible in population models lacking size structure.
+population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
+described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
+phenomena can be studied at the three levels: individual plants, stands of
+competing plants and entire metapopulations. The dynamics at these higher
+levels of organisation arise as emergent properties, driven by growth
+physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
+We model the emergent behaviour of patches deterministically, based on a
+continual  size-density distribution. As such, it is not required to specify a
+particular patch size.
 
@@
 allocated between growth and reproduction.  
-%
 The key outputs needed by higher levels (see below) is growth rate,
@@
 simulating size-structured dynamics
-\citep{Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
+\citep{Shugart-1980,Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
 we assume that patches are very large -- this means the population dynamics 

### 2015-05-29 vignettes/src/parameters.R
Basic parameters vignette

@@ -1,0 +1,113 @@
+## ---
+## title: "plant: A package for modelling forest trait ecology & evolution: _modifying parameters of the physiological model_"
+## ---
+
+library(plant)
+
+## There are a large number of parameters to the physiological model,
+## but all are changable.  The default strategy is detailed in the
+## "physiology" vignette:
+s <- FFW16_Strategy()
+names(s)
+
+## The strategy object here is a special object of class `r class(s)`.
+## In contrast with most of the "reference" objects used by `plant`,
+## this is simply a list with a class attribute.  Some validation will
+## be done on the parameters every time it is passed into the C++ bits
+## of the model.
+
+## All parameters except for `control` are floating point (decimal)
+## numbers:
+unlist(s[names(s) != "control"])
+
+## All of these parameters can be directly changed.  For example, we
+## can double the leaf mass per unit area (lma) value:
+s$lma <- s$lma * 2
+
+## and then from this construct a plant:
+pl <- FFW16_Plant(s)
+pl$strategy$lma
+
+## `lma` affects a few places in the model; see [the source
+## code](https://github.com/traitecoevo/plant/blob/master/src/ffw16_strategy.cpp),
+## but only really for converting from leaf area to leaf mass (leaf
+## mass being leaf area multiplied by leaf mass per leaf area).
+## However, as a component of the leaf economic spectrum we imagine
+## LMA as affecting a number of other components of the model.
+
+## We capture this through what we call a "hyperparameterisation";
+## additional parameters and functions that mean that changing one
+## parameter might affect a number of other lower-level parameters.
+## Our default hyperparameterisation is via the
+## `FFW16_hyperpar` function:
+FFW16_hyperpar
+
+## You will rarely need to call this function directly (see below) but
+## setting lma affects parameters `k_l`, `c_p1` and `c_Rl`
+FFW16_hyperpar(trait_matrix(0.1, "lma"), s)
+
+## These are:
+##   * `k_l`: Turnover rate for leaves
+##   * `c_p1`: Leaf photosynthesis per area
+##   * `c_Rl`: Leaf respiration per mass
+
+## This means that it is possible to implement different trade-offs
+## between parameters relatively easily by modifying the
+## hyperparameterisation function in R, rather than having to modify
+## the underlying physiological model in C++.
+
+## `trait_matrix` function here is a simple wrapper that just makes
[…]

### 2015-05-29 ms.tex
Basic parameters vignette
Comments and edits to the first two sections

@@
 
-\newcommand{\plant}{\textsc{plant}}
+\newcommand{\plant}{\texttt{plant}}
 
@@
   extensible framework for modelling size- and trait-based demography, 
-  ecology and evolution in vegetations.
+  ecology and evolution in simulated forests.
   At its core, {\plant} is an
@@
 instantaneous physiological effects to long-term evolutionary outcomes
-\citep{Harper-1977, Westoby-2002}. As an individual increases its leaf
-area, it increases its potential to generate photosynthate.  In
-contrast, as individuals grow, they must use increasing fractions of
-this photosynthetic income to building and maintaining support tissues
-rather than to new leaves \citep{Givnish-1988, Enquist-2007} and
-reproduction \citep{Thomas-2011}. Consequently rates of growth and
-mortality change with individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
-The patterns of changing growth rate vary by plant traits; leaf
-construction costs and wood density affect growth rates and rates of
-tissue replacement \citep{Wright-2010}, while seed size and height at
-maturation determine start and end points of ontogenetic trajectories
-\citep{Westoby-2002}
+\citep{Harper-1977, Westoby-2002}. 
+%
+As an individual increases its leaf area, it increases its potential
+to generate photosynthate.
+%
+In opposition to this, as individuals grow they must use increasing
+fractions of their photosynthetic income to building tissues other
+than new leaves; to maintaining support tissues \citep{Givnish-1988,
+  Enquist-2007} and to reproduction \citep{Thomas-2011}.
+%
+Consequently rates of growth, mortality and reproduction change with
+individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  The
+patterns of changing growth rate vary with plant traits; for example,
+leaf construction costs and wood density affect growth rates and rates
+of tissue replacement \citep{Wright-2010}, while seed size and height
+at maturation determine start and end points of ontogenetic
+trajectories \citep{Westoby-2002}.
+%
 Strong feedbacks emerge between individuals within a forest due to
@@
 of one individual depends on the size and traits of nearby
-individuals.  This makes it difficult to predict demographic processes
-without modelling both individual growth and the interactions between
-individuals. The non-linear feedbacks arising from competitive 
-interactions are thought to be crucial for understanding phenomena such 
-as self thinning, successional transitions, trait evolution and 
-species coexistence.
+individuals.
+%
+This makes it difficult to predict demographic processes without
+modelling both individual growth and the interactions between
+individuals.
+%
+It is therefore difficult to link observable species traits to
+phenomena such as self-thinning, successional transitions, trait
+evolution and species coexistence.
 
-Although the importance of size-structure for vegetation dynamics and diversity, 
-has been long been recognised
-\citep[e.g.,][]{Harper-1977, Shugart-1980, Huston-1987}, current research in these
-areas is dominated by models and theory that either omit size-structure 
-entirely, or only include coarse approximations. Theoretical investigations 
-exploring questions about niche-based species coexistence and ecological drift 
-(neutrality) both primarily rely  on unstructured population models 
-\citep[e.g.][]{MacArthur-1967, Tilman-1985, Geritz-1995, Hubbell-2001,
-Calcagno-2006, Leimar-2013}. It is common to assume
-competitive interactions are influenced by size-related traits such as
-adult size and seed size, but detailed size-structured demography is
-rarely considered, at least for plant models \cite[for animal
-examples, see][]{Deroos-1988, Deroos-1992}. Consequently, it has been
-difficult to connect and test the mechanisms described in these simplistic models
-with types of data collected in the field, such as physiological
-leaf traits or demographic data from long-term monitoring plots \citep{Adler-2013}.
-Similarly, most large-scale models of vegetation dynamics do not include
-size-structured demography \citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. 
-During the  1980s, size-structured ``gap'' models were widely used
-\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation of models 
-focussed more heavily on biogeochemistry at the expense of demography 
-\citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the importance
-of size-structure is starting again to be recognised \citep{Moorcroft-2001, 
- Falster-2011, Dekauwe-2014}, all but a couple \citep[e.g.][]{Moorcroft-2001, Smith-2014} 
-of the most popular, large-scale vegetation models lack any real demography or 
+Although the importance of size-structure for vegetation dynamics and
+diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
+  Shugart-1980, Huston-1987}, current research in this area is
+dominated by models and theory that either omit size-structure
+entirely, or only include coarse approximations. 
+%
+Theoretical investigations exploring questions about niche-based
+species coexistence and ecological drift (neutrality) both primarily
+rely on unstructured population models \citep[e.g.][]{MacArthur-1967,
+  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
+%
+It is common to assume competitive interactions are influenced by
+size-related traits such as adult size and seed size, but detailed
+size-structure is rarely considered, at least for plant models
+\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
+% TODO: I don't have the background to write it, but (at the risk of
+% making a long paragraph even longer) is it worth adding something
+% saying something about how we'd expect size structure to be
+% particularly important for plants because they grow up through a
+% canopy?  Perhaps not useful?
+Consequently, it has been difficult to connect and test the mechanisms
+described in these simplistic models with types of data collected in
+the field, such as physiological leaf traits or demographic data from
+long-term monitoring plots \citep{Adler-2013}.
+%
+Similarly, most large-scale models of vegetation dynamics do not
+include size-structured demography \citep{Cramer-2001, Sitch-2003,
+  Kelley-2013,Dekauwe-2014}.  During the 1980s, size-structured
+``gap'' models were widely used
+\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
+of models focused more heavily on biogeochemistry at the expense of
+demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the
+importance of size-structure is starting again to be recognised
+\citep{Moorcroft-2001, Falster-2011, Dekauwe-2014}, all but a couple
+\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular,
+large-scale vegetation models lack any real demography or
 size-structure.
@@
 competing plants and of meta-populations structured by a prevailing
-disturbance regime. In the theoretical ecology literature it is common to
-analyse the full behaviour of a model under its equilibrium conditions.
-{\plant} was developed with this style of analysis in mind, while also 
-allowing for a richer set of ecological dynamics than is possible in 
-population models lacking size structure. {\plant} is thus part
-of an extended thought experiment investigating how species differing
-in traits are able to coexist with one another \citep{Falster-2011,Falster-2015}.
-Understanding and incorporating the demographic outcomes of trait variation 
-is ultimately needed if we are to properly understand how trade-offs in plant function 
-enable species coexistence.
+disturbance regime.
+% TODO: the first sentence reads weirdly to me: the paragraph now says
+% "we introduce our model / other models are like X / our model is
+% also like X" I'm not sure what the key point that is being driven at
+% here, either.
+%
+% TODO: I've reordered it and changed the emphasis here.  Can you
+% check that you agree and if so add a reference.
+We developed {\plant} primarily to investigate if and how species
+differing in traits are able to coexist with one another
+\citep{Falster-2011,Falster-2015}.  In order to allow interactions to
+be driven by species traits and arise through competition for shared
+resources, we needed to model a community with a full range of sizes,
+as species may not be able to recruit under the canopy that they
+generate (REF).
 
-At the same time, we expect the {\plant} model to be useful for modelling
-other  phenomena, beyond questions about species coexistence. The internal
-structure  of the {\plant} model is superficially similar to the
-\textsc{Ecosystem Demography}  \citep{Moorcroft-2001}  and \textsc{LPJ-GUESS}
-\citep{Smith-2014} models. Like  {\plant}, these models include size-
-structured competition within a metapopulation of  "patches". However,
-whereas these models were designed to capture large-scale regional
+At the same time, we expect the {\plant} model to be useful for
+modelling other phenomena. The internal structure of the {\plant}
+model is similar in approach to the ``Ecosystem Demography''
+\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
+models. Like {\plant}, these models include size- structured
+competition within a metapopulation of ``patches''. However, whereas
+these models were designed to capture large-scale regional
 biogeochemistry driven by climatic variables and rely on simulations,
-{\plant} was designed so that the rules  driving the dynamics of the system
-could be manipulated and emergent demographic  behaviours studied in detail.
-In {\plant} it is possible to grow plants in isolation, in patches of
-competing plants and in disturbance-driven metapopulations. The package is
-thus designed to sit in the middle ground between a large-scale coupled
-simulation model and simple abstract theoretical device. As examples, we
-expect the package will prove  useful for investigating how morphological and
-physiological traits influence growth rate and shade tolerance, and emergent
-phenomena such as multi-species self-thinning,  successional transitions and
-age-related changes in vegetation productivity.
+{\plant} was designed so that the rules driving the dynamics of the
+system could be manipulated and emergent demographic behaviours
+studied in detail.  In {\plant} it is possible to grow plants in
+isolation, in patches of competing plants and in disturbance-driven
+metapopulations. The package is thus designed to sit in the middle
+ground between a large-scale coupled simulation model and simple
+abstract theoretical device. 
+% TODO: Consider moving to the conclusion?
+As examples, we expect the package will prove useful for investigating
+how morphological and physiological traits influence growth rate and
+shade tolerance, and emergent phenomena such as multi-species
+self-thinning, successional transitions and age-related changes in
+vegetation productivity.
 
-Below we describe the general approach of the package, then provide  sections
-with details at different levels of ecological organisation.  Detailed
-technical documents are provided as supplementary information  (see
-Appendices), and updated versions of technical documents are also available
-within the {\plant} package itself.
+% TODO: they aren't actually available within plant yet, and
+% realistically won't be by the updated.  I've added a placeholder
+% that contains appropriate links.
+Below we describe the general approach of the package, then provide
+sections with details at nested levels of ecological organisation.
+Detailed technical documents are provided as supplementary information
+(see Appendices), and updated versions of technical documents are also
+available within the {\plant} package itself.
 
@@
 functioning of an individual plant and how these are influenced by its
-traits, size and light environment. Being driven by traits, the model can be extended to potentially 
-very many species. Users can also  add their own physiological models.
+traits, size and light environment. Being driven by traits, the model
+can be extended to potentially very many species. Users can also add
+their own physiological models, though this is a work in progress.
 
@@
 respiration, photosynthesis per leaf area) in the model are linked via
-assumed tradeoffs. For example, the trait \textsc{lma} is used to estimate the
+assumed trade-offs. For example, the trait \textsc{lma} is used to estimate the
 rate of leaf turnover, based on observed scaling relationship spanning
@@
 Our approach for modelling size-structured population dynamics
-is based on the Escalator Boxcar Train technique (\textsc{EBT})
-\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{EBT} solves the \textsc{pde}
+is based on the Escalator Boxcar Train technique (\textsc{ebt})
+\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves the \textsc{pde}
 describing development of \(n\) by approximating the density function
@@
 
-We introduce two extensions to the \textsc{EBT}: i) a new method for estimating
+We introduce two extensions to the \textsc{ebt}: i) a new method for estimating
 the size-density distribution and ii) a technique for handling strong
@@
 competition for light (see Appendix \ref{sec:demography} for
-details). The original \textsc{EBT} \citep{Deroos-1988,
+details). The original \textsc{ebt} \citep{Deroos-1988,
   Deroos-1992, Deroos-1997} proceeds by approximating the first and
@@
 adaptive refining of time points at which new cohorts are introduced
-into the population. In the original \textsc{EBT}, cohorts were introduced at a
+into the population. In the original \textsc{ebt}, cohorts were introduced at a
 fixed time interval. Under competition, the growth trajectories of
@@
 metapopulation is to identify the equilibrium seed rain of the
-metapopulation. An "input" seed rain is required to simulate 
-the demography of the metapopulation, this in turn produces an "output" 
+metapopulation. An ``input'' seed rain is required to simulate 
+the demography of the metapopulation, this in turn produces an ``output'' 
 number of seeds. The equilibrium seed rain is the vector of seed 
@@
 across all patches (Fig. \ref{fig:emergent}, see more detail in
-appendix \ref{sec:emergent}). This emergent relationship might then be
+Appendix \ref{sec:emergent}). This emergent relationship might then be
 compared to the observed distributions in large forest plots
@@
 negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
-selection for smaller trait values (see appendix \ref{sec:fitness} for
+selection for smaller trait values (see Appendix \ref{sec:fitness} for
 details). Fitness landscapes are constantly reshaped as species are
@@
 unknown. The simple example here can be extended to multiple dimensions
-and can be used to investigate how tradeoffs facilitate coexistence
+and can be used to investigate how trade-offs facilitate coexistence
 \citep{Falster-2015}
@@
 packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
-\texttt{R6} \citep{Chang-2014} and \texttt{testhat}
+\texttt{R6} \citep{Chang-2014} and \texttt{testthat}
 \citep{Wickham-2011}; and the Boost Library for C++
@@
 (b) Competitive hierarchies  are modelled by tracking the height distribution of
-individuals across within a "patch" after a disturbance. The intensity
+individuals across within a ``patch'' after a disturbance. The intensity
 of shading indicates the density of individuals at a given height for
@@
 in panels (a) same growth trajectories for individuals on the
-boundary of successive "cohorts" for each species, using a schedule of 
+boundary of successive ``cohorts'' for each species, using a schedule of 
 introduction times generated by an adaptive refinement algorithm.
  Red lines are the low \textsc{lma}
-species from figure \ref{fig:plant}, blue lines are the high \textsc{lma} species.  The
+species from Fig. \ref{fig:plant}, blue lines are the high \textsc{lma} species.  The
 initial growth rate advantage of the low \textsc{lma} species (see Figure
@@
 
-\section{Modifying parameters of physiological model}\label{sec:parameters}
+\section{Modifying parameters of the physiological model}\label{sec:parameters}
 
-See attached file \url{vignette_parameters.R}
+See attached file \url{vignettes/parameters.pdf}, knitr source
+\url{vignettes/parameters.Rmd}, plain source, \url{vignettes/R/parameters.R}
 
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% TeX-PDF-mode: t
+%%% End:
+
+%  LocalWords:  Westoby photosynthate Givnish Enquist Ruger feedbacks
+%  LocalWords:  ontogenetic Shugart Tilman Geritz Hubbell Calcagno BH
+%  LocalWords:  Cramer Sitch Dekauwe focussed successional Haxeltine
+%  LocalWords:  Moorcroft lpj FFW lma wplcp Baltzer Lusk Kohyama pde
+%  LocalWords:  Deroos Pacala Strigul ebt adaptively Coomes Binkley
+%  LocalWords:  Ogawa Bormann Gyllenberg Vonfoerster Metz Dieckmann
+%  LocalWords:  Rcpp Eddelbuettel templated RcppR testhat Wickham DS
+%  LocalWords:  Schaling RG Brn nnstr Camac De Roos Duursma Johansson
+%  LocalWords:  knitr

### 2015-05-29 ms.tex
Reworking of metapopulation section

@@
  species with different traits (Fig. \ref{fig:schematic}b; Fig.
-\ref{fig:patch}).
-
-Second, {\plant} allows for the effects of competition and succession on
+\ref{fig:patch}). Second, {\plant} allows for the effects of competition and succession on
 productivity and biomass accumulation within a patch to be investigated
@@
 disease outbreak
-\citep{White-1979, Bormann-1979, Chambers-2013}.
+\citep{White-1979, Bormann-1979, Coomes-2007, Chambers-2013}.
 The vegetation thus comprises a collection of patches differing in time
@@
 
-In {\plant} we assume an infinite number of patches all experiencing
-the same disturbance regime; we also assume that the disturbances remove
+In {\plant} we assume an infinite number of patches, all experiencing
+the same disturbance regime and linked via seed dispersal (Fig.
+\ref{fig:schematic}b). We also assume that the disturbances remove
 all established vegetation.  With this approach we can scale up from
@@
 deterministically and can be approximated via a second \textsc{pde}
-\citep{ Vonfoerster-1959} (see Appendix
-\ref{sec:demography} for
+(see Appendix \ref{sec:demography} for
 details). Scaling from patch to metapopulation is then achieved by
@@
 
-As previously noted \citep{Kohyama-1993, Moorcroft-2001, Falster-2011} a
-feature of the size-structured metapopulation concept implemented here
+A feature of the size-structured metapopulation concept implemented here
 is that it balances equilibrium and non-equilibrium approaches to
-modelling ecological dynamics. An
+modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001, 
+Falster-2011}. An
 equilibrium may be approached at the level of metapopulation, meaning
@@
 \subsection{Applications}
-% TODO[RGF]:  But we need
-% something that indicates how this can be used.  I suggest a complete
-% overhaul of this paragraph as that gets closest to a way of
-% using the model.
 
-Beyond its conceptual appeal, this idea of a dynamic equilibrium
-allows us to close the demographic feedback loop and thereby create a
-self-regulating system. The equilibrium seed rain for multi-species
-communities is approached demographically, incorporating the combined
-effects of the model's physiological rules, species traits,
-competitive interactions and the prevailing disturbance regime on the
-size-distributions of plants within a metapopulation of
-patches. Unlike some vegetation models, there is no need to set the
-density of seeds or seedlings, these arise as outputs from the
-model. Characteristics of the vegetation such as total leaf area,
-biomass, productivity and allocation to different plant parts also
-arise as emergent properties (Fig.  \ref{fig:patch}c). Similarly,
-distributions of abundance, leaf area and growth rate with respect to
-size emerge at both patch scale and scale of metapopulation
-(Fig. \ref{fig:emergent}). Note in particular how a single
-physiological model produces a variety of height-density
-relationships, dependent on patch age, as well as an emergent average
-across all patches (Fig. \ref{fig:emergent}, see more detail in
-Appendix \ref{sec:emergent}). This emergent relationship might then be
-compared to the observed distributions in large forest plots
-\citep{Muller-2006}.
+Beyond its conceptual appeal, the idea of a dynamic equilibrium captured by
+the trait-, size- and patch structured metapopulation model potentially
+resolves several challenges faced by unstructured models. First, by scaling up
+from patch to metapopulation, we close the demographic feedback loop to create
+a self-regulating system. Unlike some vegetation models, there is no need to
+set the density of seeds or seedlings, these arise as outputs from the model,
+based on the combined effects of the model's physiological rules, species
+traits, competitive interactions and the prevailing disturbance regime.
 
-\section{Invasion fitness}
+Second, the metapopulation concept allows for the effects of disturbance
+regimes on vegetation structure and dynamics to be properly incorporated. In
+{\plant} characteristics for the entire vegetation (metapopulation) -- such
+as total leaf area, biomass and productivity -- are obtained by averaging
+patch-level outcomes over the frequency-distribution of patch ages. Because
+young patches are more common in the landscape, metapopulation averages are
+strongly influenced by the structure of younger patches (e.g. 
+Fig. \ref{fig:patch}c). Moreover, predictions are for a frequency distribution 
+of states across the landscape. Similarly, the distributions of  abundance and growth rate
+observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} may
+can be modelled as  emergent phenomena, where variety of patch-level
+relationships combine to  produce  an emergent average across all patches
+(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
+size-and patch- structured models may be more suitable for comparing with 
+satellite and survey data, than unstructured models representing the world as 
+homogeneous green slime \citep{Moorcroft-2001,Moorcroft-2001}.
 
-With the extension to metapopulation, we are modelling demography across
+Finally, with the extension to metapopulation, we are modelling demography across
 the entire life cycle and are thus in a position to estimate the fitness
@@
 \(R > 1\).
-
-\subsection{Applications}
 

### 2015-05-29 ms.tex
Further tweaks to main text

@@
 \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
-% TODO: I don't have the background to write it, but (at the risk of
-% making a long paragraph even longer) is it worth adding something
-% saying something about how we'd expect size structure to be
-% particularly important for plants because they grow up through a
-% canopy?  Perhaps not useful?
+%
 Consequently, it has been difficult to connect and test the mechanisms
@@
 include size-structured demography \citep{Cramer-2001, Sitch-2003,
-  Kelley-2013,Dekauwe-2014}.  During the 1980s, size-structured
+  Purves-2008,Dekauwe-2014}.  During the 1980s, size-structured
 ``gap'' models were widely used
@@
 of models focused more heavily on biogeochemistry at the expense of
-demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996}. While the
+demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996,Cramer-2001, 
+Sitch-2003}. While the
 importance of size-structure is starting again to be recognised
-\citep{Moorcroft-2001, Falster-2011, Dekauwe-2014}, all but a couple
-\citep[e.g.][]{Moorcroft-2001, Smith-2014} of the most popular,
+\citep{Moorcroft-2001, Purves-2008, Falster-2011, Dekauwe-2014}, all but a couple
+of the most popular,
 large-scale vegetation models lack any real demography or
-size-structure.
+size-structure \citep[e.g.][]{Moorcroft-2001, Smith-2014}.
 
@@
 disturbance regime.
-% TODO: the first sentence reads weirdly to me: the paragraph now says
-% "we introduce our model / other models are like X / our model is
-% also like X" I'm not sure what the key point that is being driven at
-% here, either.
 % 
-% TODO: I've reordered it and changed the emphasis here.  Can you
-% check that you agree and if so add a reference.
 We developed {\plant} primarily to investigate if and how species
@@
 as species may not be able to recruit under the canopy that they
-generate (REF).
+generate \citep{Moorcroft-2001,Huston-1987}.
 
@@
 system could be manipulated and emergent demographic behaviours
-studied in detail.  In {\plant} it is possible to grow plants in
+studied deterministically and in detail. In {\plant} it is possible to grow plants in
 isolation, in patches of competing plants and in disturbance-driven
-metapopulations. The package is thus designed to sit in the middle
+metapopulations. Moreover, we put considerable effort into resolving
+competitive hierarchies. 
+The package is thus designed to sit in the middle
 ground between a large-scale coupled simulation model and simple
 abstract theoretical device.
-% TODO: Consider moving to the conclusion?
-As examples, we expect the package will prove useful for investigating
-how morphological and physiological traits influence growth rate and
-shade tolerance, and emergent phenomena such as multi-species
-self-thinning, successional transitions and age-related changes in
-vegetation productivity.
 
-% TODO: they aren't actually available within plant yet, and
-% realistically won't be by the updated.  I've added a placeholder
-% that contains appropriate links.
 Below we describe the general approach of the package, then provide
@@
 Detailed technical documents are provided as supplementary information
-(see Appendices), and updated versions of technical documents are also
-available within the {\plant} package itself.
+(see Appendices), updated versions of these technical documents will also
+be made available within the {\plant} package itself.
 
@@
 strongly influenced by the structure of younger patches (e.g. 
-Fig. \ref{fig:patch}c). Moreover, predictions are for a frequency distribution 
-of states across the landscape. Similarly, the distributions of  abundance and growth rate
-observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} may
-can be modelled as  emergent phenomena, where variety of patch-level
-relationships combine to  produce  an emergent average across all patches
+Fig. \ref{fig:patch}c).  Similarly, the distributions of  abundance and growth rate
+observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} arise
+as  emergent phenomena. Moreover, the predictions from {\plant} 
+are for a frequency distribution of patch states across the landscape 
 (Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
-size-and patch- structured models may be more suitable for comparing with 
-satellite and survey data, than unstructured models representing the world as 
-homogeneous green slime \citep{Moorcroft-2001,Moorcroft-2001}.
+size-and patch- structured models may be more suitable for linking with 
+satellite and forest survey data, than unstructured models representing the 
+world as homogeneous green slime \citep{Moorcroft-2001, Purves-2008}.
 
 Finally, with the extension to metapopulation, we are modelling demography across
-the entire life cycle and are thus in a position to estimate the fitness
-of different types. Invasion fitness quantifies the ability of a rare
+the entire life cycle and are thus in a position to estimate invasion fitness -- 
+the ability of a rare
 individual with particular traits to establish in a community of
@@
 reproductive success of individuals possessing a particular trait value
-is measured in the size- and patch-structured metapopulation and
 depends on both the traits and densities of other species in the current
-resident community. The reason this is interesting is that under
-frequency- and density-dependence, evolution tends to favour different
-trait mixtures compared to those predicted by models that optimise some
-feature of the community, such as total biomass or carbon production or
-seed production in the absence of competition
-\citep{Falster-2003, Dieckmann-2007}.
-
-Fitness landscapes will be positive in regions that allow invasion,
+resident community. Fitness landscapes will be positive in regions that allow invasion,
 and zero for species that are at equilibrium. The slope of the fitness
@@
 points where species are present. This community is immune to invasion
-from any other strategies \citep{Geritz-1998}.
+from any other strategies and represents a prediction for the 
+trait mixture that might might be favoured by natural selection \citep{Geritz-1998}.
 
@@
 and can be used to investigate how trade-offs facilitate coexistence
-\citep{Falster-2015}
+\citep{Falster-2015}.
 
@@
 of other uses would be possible and also extensions to address other aspects
-of community assembly. For example,  the physiology of growth could be
-modified to account for rising atmospheric CO2 or the effects of competition
+of community assembly. As examples, we expect the package will prove useful
+for investigating how morphological and physiological traits influence growth
+rate and shade tolerance, and emergent phenomena such as multi-species self-
+thinning, successional transitions and age-related changes in vegetation
+productivity. The effects of competition
 for multiple resources could likewise be investigated.
@@
 structure, so remain unable to investigate demographic phenomena
-\citep{Cramer-2001, Sitch-2003, Kelley-2013,Dekauwe-2014}. At the same time,
+\citep{Cramer-2001, Sitch-2003, Dekauwe-2014}. At the same time,
 much attention is paid to detailed leaf-level physiology. {\plant} aims to bridge
@@
 Research Council. We thank {\AA} Brn{\"a}nnstr{\"o}m, J Camac, 
-A De Roos, U Dieckmann, R Duursma, and J Johansson for helpful discussions.
+A De Roos, U Dieckmann, R Duursma, J Johansson and C Prentice 
+for helpful discussions.
 

### 2015-05-29 ms.tex
Equal first authors

@@
 \title{{\plant}: A package for modelling forest trait ecology \& evolution}
-\author{Daniel S. Falster, Richard G. FitzJohn, and Mark Westoby}
+\author{Daniel S. Falster\textdagger\textasteriskcentered, Richard G. FitzJohn\textdagger, and Mark Westoby}
 \affiliation{
 Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
-Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
+\textdagger These authors contributed equally.\\ 
+\textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
 A manuscript in consideration as a research paper for

### 2015-05-29 ms.tex
Minor tweaks

@@
 isolation, in patches of competing plants and in disturbance-driven
-metapopulations. Moreover, we put considerable effort into resolving
+metapopulations. 
+% TODO: can you change this?  the amount of effort we put in is
+% decoupled from how useful other people may find it.
+Moreover, we put considerable effort into resolving
 competitive hierarchies. 
@@
 Detailed technical documents are provided as supplementary information
-(see Appendices), updated versions of these technical documents will also
-be made available within the {\plant} package itself.
+(see Appendices), updated versions of these technical documents are
+available within the {\plant} package itself.
 
@@
 
-\section{Effects of size, trait and environment on demography of individual plants}
+\section{Effects of size, trait and environment on demography of individual plants}Q
 
@@
 
-Beyond its conceptual appeal, the idea of a dynamic equilibrium captured by
-the trait-, size- and patch structured metapopulation model potentially
-resolves several challenges faced by unstructured models. First, by scaling up
-from patch to metapopulation, we close the demographic feedback loop to create
-a self-regulating system. Unlike some vegetation models, there is no need to
-set the density of seeds or seedlings, these arise as outputs from the model,
-based on the combined effects of the model's physiological rules, species
-traits, competitive interactions and the prevailing disturbance regime.
+The dynamic equilibrium captured by the trait-, size- and patch
+structured metapopulation model potentially resolves several
+challenges faced by unstructured models. First, by scaling up from
+patch to metapopulation, we close the demographic feedback loop to
+create a self-regulating system. Unlike some vegetation models, there
+is no need to set the density of seeds or seedlings, these arise as
+outputs from the model, based on the combined effects of the model's
+physiological rules, species traits, competitive interactions and the
+prevailing disturbance regime.
 
@@
 thinning, successional transitions and age-related changes in vegetation
-productivity. The effects of competition
-for multiple resources could likewise be investigated.
+productivity.
 
-It is hoped that the approach here can be used to inform both future empirical
-research and model development in neighbouring areas. As noted above, most
-large-scale mechanistic models of plant growth do not explicitly model size
-structure, so remain unable to investigate demographic phenomena
-\citep{Cramer-2001, Sitch-2003, Dekauwe-2014}. At the same time,
-much attention is paid to detailed leaf-level physiology. {\plant} aims to bridge
-this gap that emerged in the model landscape, where detailed physiology and
-size-structured dynamics combine to drive forest structure and function via
-traits.
+% TODO: In favour of dropping this whole paragraph?
+It is hoped that the approach here can be used to inform both future
+empirical research and model development in neighbouring areas.
+{\plant} aims to bridge the gap between detailed physiological models
+and large-scale mechanistic models of plant growth \citep{Cramer-2001,
+  Sitch-2003, Dekauwe-2014}.  As noted above, most large-scale
+mechanistic models of plant growth do not explicitly model size
+structure, so remain unable to investigate demographic phenomena }. At
+the same time, much attention is paid to detailed leaf-level
+physiology. {\plant} aims to bridge this gap that emerged in the model
+landscape, where detailed physiology and size-structured dynamics
+combine to drive forest structure and function via traits.
 

### 2015-05-29 ms.tex
More vignette name fixes

@@
 mechanistic models of plant growth do not explicitly model size
-structure, so remain unable to investigate demographic phenomena }. At
+structure, so remain unable to investigate demographic phenomena. At
 the same time, much attention is paid to detailed leaf-level
@@
 
-\section{Default physiological model for {\plant}}\label{sec:FFW16}
+\section{Plant physiological model}\label{sec:FFW16}
 
@@
 
-\section{Modelling demography in the {\plant} package}\label{sec:demography}
+\section{Modelling demography of plants, patches and metapopulations}\label{sec:demography}
 

### 2015-05-29 update_wiki.sh

@@ -7,7 +7,7 @@ else
     git clone https://github.com/traitecoevo/plant_paper.wiki.git wiki
 fi
 
-cp vignettes/*.md wiki
+cp vignettes/*.md vignettes/*.pdf wiki
 cp -r vignettes/figure wiki
 git -C wiki add .
 SHA=$(git rev-parse --short HEAD)

### 2015-05-29 R/figures_emergent.R
Oust relict calls to tree

@@ -68,7 +68,7 @@ figure_emergent <- function(data) {
   # find average across patches for each height, 
   # i.e. across rows, weighting by patch abundance 
   metapopulation_average <- function(x) {
-    tree:::trapezium(data$time, x*data$patch_density)
+    plant:::trapezium(data$time, x*data$patch_density)
   }  
   n_av <- apply(n_hh,      1, metapopulation_average) 
   l_av <- apply(l_hh*n_hh, 1, metapopulation_average) / n_av

### 2015-05-29 R/figures_patch.R

@@ -69,7 +69,7 @@ figure_patch <- function(data) {
   mtext("Time (years)", 1, 2.2, cex=.7)
 
   # find average value over metapopulation
-  y_av <- apply(y, 2, function(x) tree:::trapezium(t, x*data$patch_density))
+  y_av <- apply(y, 2, function(x) plant:::trapezium(t, x*data$patch_density))
   for(i in seq_len(3)) {
     axis(4, at=y_av[i], tck=0.1, col.ticks=cols[i], lty = lty[i])
   }

### 2015-05-29 vignettes/src/cohort_spacing.R
Minor editting of text in vignettes
Closes #23

Also added new metapopulation calculations for patch and emergent

@@ -3,9 +3,11 @@
 ## ---
 
 ## This vignette shows some details of cohort splitting.  It's
-## probably not very interesting to most people.  It also uses a lot
+## probably not very interesting to most people, only those
+## interested in knowing how the EBT technique works in detail.  It also uses a lot
 ## of non-exported, non-documented functions from plant so you'll see
 ## a lot of `plant:::` prefixes.
+
 library(plant)
 library(parallel)
 
@@ -48,9 +50,9 @@ sr_1 <- run_with_times(p, t1)
 sr_2 <- run_with_times(p, t2)
 sr_3 <- run_with_times(p, t3)
 
-## Seed rain increases as time is cohorts are introduced more finely,
+## Seed rain increases as cohorts are introduced more finely,
 ## though at a potentially saturating rate.  We're doing lots more
-## work at the more refined end though!
+## work at the more refined end!
 sr <- c(sr_1, sr_2, sr_3)
 sr
 
@@ -73,13 +75,13 @@ run_with_insert <- function(i, p, t) {
 }
 
 ## The internal function `plant:::run_ebt_error` runs the EBT and
-## computes errors as the integration proceeds and helps shed some
+## computes errors as the integration proceeds; this helps shed some
 ## light.
 i <- seq_len(length(t1) - 1)
 res <- unlist(mclapply(i, run_with_insert, p, t1))
 tm <- (t1[-1L] + t1[-length(t1)]) / 2
 
-## The biggest deviations in output seed rain cpome about half way
+## The biggest deviations in output seed rain come about half way
 ## through the schedule:
 plot(i, res - sr_1, xlab="Index", ylab="Seed rain", las=1)
 abline(h=0, col="grey")
@@ -92,7 +94,7 @@ abline(h=0, col="grey")
 ebt <- plant:::FFW16_EBT(p)
 ebt$run()
 
-## This is the contribution of different cohorts to see rain (x axis
+## Now look at the contribution of different cohorts to see rain (x axis
 ## log scaled for clarity).  In this case almost all the contribution
 ## comes from early cohorts (this is essentially a single-age stand of
 ## pioneers).  Overlaid on this are the five cohorts with the largest

### 2015-05-29 vignettes/src/emergent.R

@@ -2,11 +2,10 @@
 ## title: "plant: A package for modelling forest trait ecology & evolution: _patch level emergent properties_"
 ## ---
 
-## The aim here is to use plant to investigate dynamics of plants
-## within a single patch, but to focus on patch level properties,
-## rather than properties of plants within the patch (though I guess
-## that the leaf area index calculations in vignette:patch are also
-## patch level properties).
+## The aim here is to use the plant package to investigate dynamics 
+## within a patch of competing plants, focussing on emergent patch-level 
+## properties, rather than properties of plants within the patch.
+
 library(plant)
 
 p0 <- ebt_base_parameters()
@@ -14,7 +13,7 @@ p0$control$equilibrium_nsteps <- 30
 p0$control$equilibrium_solver_name <- "hybrid"
 p0$disturbance_mean_interval <- 30.0
 
-## Working with a single species at equilibrium
+## We'll work with a single species at equilibrium
 p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
 p1_eq <- equilibrium_seed_rain(p1)
 data1 <- run_ebt_collect(p1_eq)
@@ -50,10 +49,10 @@ matplot(height, density, type="l", lty=1,
         log="y")
 
 ## Note that the densities here can be *extremely* low, and yet
-## cohorts continue to grow in size; this is the "atto-fox problem",
+## individuals within the cohorts continue to grow in size; this is the "[atto-fox problem](http://en.wikipedia.org/wiki/Lotka-Volterra_equation)",
 ## thou here we drop down as low as `r formatC(signif(min(density, na.rm=TRUE) / 1e-24, 1))`
 ## yocto plants / m / metre squared (a yocto-x being 1 millionth of an
-## atto-x).  *Anyway*:
+## atto-x).  *Anyway*....
 
 ## The trajectories are easier to understand if a few are
 ## highlighted.
@@ -112,3 +111,50 @@ points(height[1, i], growth_rate[1, i], pch=19, col=cols)
 text(height[1, i] + strwidth("x"), growth_rate[1, i],
      paste0(round(times), c(" years", rep("", length(times) - 1))),
      adj=c(0, 0))
+
+
+## The above plots show relationships with patches of a given age
+## What about the average relationship across the entire metapopulation? 
+## To get that, we  average (integrate) over the distribution over patch 
[…]

### 2015-05-29 vignettes/src/equilibrium.R

@@ -106,14 +106,14 @@ abline(v=p2_eq$seed_rain, col=1:2, lty=3)
 
 ## Note that the first guess position of the red species is higher
 ## than the black species, but in the end the output seed rain is
-## lower.  This is the difficulty in computing multispecies equilibria
+## lower.  This is the difficulty in computing multi species equilibria
 ## - the different solutions affect each other.  In general
-## multidimenensional root finding is difficult; even knowning that
+## multi-dimensional root finding is difficult; even knowing that
 ## there are roots is not straightforward, much less proving that
 ## we've converged on the "correct" root (for example [0,0] is a root
 ## in this case but that root is not stable).  plant uses some
 ## heuristics to try to ensure that the root returned is an attracting
-## point but sequentially applying rounds of iteration and nonlinear
+## point but sequentially applying rounds of iteration and non-linear
 ## root finding algorithms, as well as rescaling seed rains to repel
 ## from known unstable roots.
 

### 2015-05-29 vignettes/src/fitness.R

@@ -69,7 +69,7 @@ abline(h=0, col="grey")
 points(lma_b, 0, pch=19)
 
 ## Holding the first species at 0.08 we can introduce additional
-## species (it's close enough to the optimium here, though in general
+## species (it's close enough to the optimum here, though in general
 ## this point might move substantially as new species are introduced).
 
 ## Consider introducing a new species at the point of maximum fitness:

### 2015-05-29 vignettes/src/parameters.R

@@ -5,7 +5,7 @@
 library(plant)
 
 ## There are a large number of parameters to the physiological model,
-## but all are changable.  The default strategy is detailed in the
+## but all are changeable.  The default strategy is detailed in the
 ## "physiology" vignette:
 s <- FFW16_Strategy()
 names(s)
@@ -35,15 +35,15 @@ pl$strategy$lma
 ## However, as a component of the leaf economic spectrum we imagine
 ## LMA as affecting a number of other components of the model.
 
-## We capture this through what we call a "hyperparameterisation";
+## We capture this through what we call a "hyper-parameterisation";
 ## additional parameters and functions that mean that changing one
 ## parameter might affect a number of other lower-level parameters.
-## Our default hyperparameterisation is via the
+## Our default hyper-parameterisation is via the
 ## `FFW16_hyperpar` function:
 FFW16_hyperpar
 
 ## You will rarely need to call this function directly (see below) but
-## setting lma affects parameters `k_l`, `c_p1` and `c_Rl`
+## note how setting of lma affects parameters `k_l`, `c_p1` and `c_Rl`
 FFW16_hyperpar(trait_matrix(0.1, "lma"), s)
 
 ## These are:
@@ -53,15 +53,15 @@ FFW16_hyperpar(trait_matrix(0.1, "lma"), s)
 
 ## This means that it is possible to implement different trade-offs
 ## between parameters relatively easily by modifying the
-## hyperparameterisation function in R, rather than having to modify
+## hyper-parameterisation function in R, rather than having to modify
 ## the underlying physiological model in C++.
 
-## `trait_matrix` function here is a simple wrapper that just makes
-## sure that a trait matrix is in the right format:
+## The `trait_matrix` function is a simple wrapper that just makes
+## sure the trait matrix has the right format:
 trait_matrix(0.1, "lma")
 trait_matrix(c(0.1, 500), "rho")
 
-## To make use of the hyperparameterisation, the preferred way of
+## To make use of the hyper-parameterisation, the preferred way of
 ## setting parameters is through the utility functions `strategy` and
 ## `strategy_list`.  These take a `Parameters` object:
 p <- FFW16_Parameters()
@@ -69,7 +69,7 @@ p <- FFW16_Parameters()
 ## The Parameters object mostly contains information about the patch:
 names(p)
 
[…]

### 2015-05-29 vignettes/src/patch.R

@@ -61,7 +61,7 @@ h[, 60]
 matplot(t, h, lty=1, col=make_transparent("black", 0.25), type="l",
         las=1, xlab="Time (years)", ylab="Height (m)")
 
-## The light environment is stored over each timestep:
+## The light environment is stored over each time step:
 xlim <- c(0, 1.1)
 ylim <- range(data1$light_env[[length(data1$light_env)]][, "height"])
 plot(NA, xlim=xlim, ylim=ylim, las=1,
@@ -84,7 +84,7 @@ for (i in seq_along(times)) {
 }
 
 ## The amount of light at the ground level is perhaps the most
-## relevent metric:
+## relevant metric:
 y <- sapply(data1$light_env, function(x) x[1, "canopy_openness"])
 plot(data1$time, y, type="l", las=1,
      ylim=c(0, 1), xlab="Time (years)", ylab="Canopy openness")
@@ -176,3 +176,22 @@ plot(t2, lai2, type="l", las=1, lty=2,
      xlab="Time (years)", ylab="Leaf area index")
 lines(t2, lai2_1, col=cols[[1]])
 lines(t2, lai2_2, col=cols[[2]])
+
+## To find the average value over the metapopulation we weight by patch abundance:
+metapopulation <- function(x){
+  plant:::trapezium(t2, x*data2$patch_density)
+}
+
+lai2_av <- metapopulation(lai2)
+lai2_1_av <- metapopulation(lai2_1)
+lai2_2_av <- metapopulation(lai2_2)
+
+plot(t2, lai2, type="l", las=1, lty=2,
+     xlab="Time (years)", ylab="Leaf area index")
+lines(t2, lai2_1, col=cols[[1]])
+lines(t2, lai2_2, col=cols[[2]])
+
+axis(4, at=lai2_av,   tck=0.1, lty = 2, labels=NA)
+axis(4, at=lai2_1_av, tck=0.1, col.ticks=cols[[1]], labels=NA)
+axis(4, at=lai2_2_av, tck=0.1, col.ticks=cols[[1]], labels=NA)
+axis(1, at=108, labels="Av")

### 2015-05-29 vignettes/src/plant.R

@@ -2,15 +2,18 @@
 ## title: "plant: A package for modelling forest trait ecology & evolution: _plant level properties_"
 ## ---
 
-## This vignette covers the sort of analysis in the plant figure in
-## the manuscript.  It also shows some of the features of working
+## This vignette illustrates the sort of analysis used to generate 
+## the plant figure (Fig 2) in the manuscript, i.e. modelling the
+## dynamics of individual plants. It also shows some of the features of working
 ## with plant components of the model.
 library(plant)
 
 ## Plants are constructed with the `FFW16_Plant` function.  That
 ## function takes as its only argument a "strategy" object; the
 ## default is `FFW16_Strategy`, but alternative strategies can be
-## provided (see below)
+## provided (see below). The "strategy" object contains all 
+## the physiological underpinning the dynamics of individual plants
+## and entire metapopulations.
 pl <- FFW16_Plant()
 
 ## Plants are an [R6](https://github.com/wch/R6) class, and have a
@@ -27,17 +30,17 @@ pl$fecundity
 pl$mortality
 
 ## Height, fecundity and mortality are the three key variables
-## propagated by the system of differential equations:
+## propagated by the internal system of differential equations:
 pl$ode_state
 
 ## To compute rates of change for these variables we need a light
 ## environment.  The function `fixed_environment` creates an
-## environment that has the same light level (here 100%) at all
+## environment that has the same canopy openness (here 100%) at all
 ## heights.  The plant *does not affect* this light environment.
 env <- fixed_environment(1.0)
 
-## The `compute_vars_phys` method computes whole-plant assmilation for
-## the plant, and from that growth rates:
+## The `compute_vars_phys` method computes net mass production for
+## the plant, and from that demographic rates:
 pl$ode_rates
 pl$compute_vars_phys(env)
 pl$ode_rates
@@ -58,19 +61,21 @@ pl$internals$area_leaf
 ## most of the internal calculations.
 ##
 ## Because of this, we have a "PlantPlus" object that exposes more of
-## a strategy (NOTE: a little more than this because it also includes
-## stem diameter growth)
+## a strategy, and in addition stem diameter growth:
 pp <- FFW16_PlantPlus(pl$strategy)
 pp$compute_vars_phys(env)
 names(pp$internals)
 
-## Some of the internals require `compute_vars_internals` to be run:
[…]

### 2015-05-29 ms.tex
Tidying up many loose ends
closes #19, close #18, closes #24

Check main text for spelling and grammar, further tighten text.
Scientific notation for units in figures.
Running head
Word count
links to other vignettes in demography and physiology documents.
Add details for stem diameter to physiology.

@@
 publication in MEE as part of the Special Feature \emph{Demography
-  beyond the Population}.}
+  beyond the Population}.\\
+Word count: ~3900 words}
 \date{}
@@
 \mstype{Applications Note}
-\runninghead{plant}
+\runninghead{The {\plant} package}
 \keywords{demography, emergent, fitness, growth,
@@
 \item
-% todo
-  We provide worked examples illustrating {\plant}'s use to study the
-  influence of functional traits on growth of individual plants, of
-  whole patches and on assembly of ecological communities.
+  {\plant} provides a transparent platform for investigating how  
+  physiological rules and functional trade-offs interact with competition and 
+  disturbance regimes to influence vegetation demography, structure and 
+  diversity.
 \end{enumerate}
@@
 
-Plant growth and demography are fundamentally size- and trait-
-structured, influencing dynamics over time-scales ranging from
+Plant growth and demography are fundamentally size- and 
+trait-structured, influencing dynamics over time-scales ranging from
 instantaneous physiological effects to long-term evolutionary outcomes
@@
 Consequently rates of growth, mortality and reproduction change with
-individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  The
-patterns of changing growth rate vary with plant traits; for example,
-leaf construction costs and wood density affect growth rates and rates
-of tissue replacement \citep{Wright-2010}, while seed size and height
-at maturation determine start and end points of ontogenetic
+individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.  
+Ontogenetic 
+patterns of growth also vary with traits; for example,
+leaf and wood construction costs influence growth rates 
+\citep{Wright-2010}, while seed size and height
+at maturation determine the start and end points of ontogenetic
 trajectories \citep{Westoby-2002}.
 %
-Strong feedbacks emerge between individuals within a forest due to
+Strong feedbacks emerge between individuals within a forest via
 competition for light and other resources, such that the growth rate
 of one individual depends on the size and traits of nearby
+individuals \cite{Pacala-1996}.
+%
+Such feedbacks make it difficult to link observable species traits to
+phenomena such as self-thinning, successional transitions, trait
+evolution and species coexistence, without
+modelling both individual growth and the competitive interactions between
 individuals.
-%
-This makes it difficult to predict demographic processes without
-modelling both individual growth and the interactions between
-individuals.
-%
-It is therefore difficult to link observable species traits to
-phenomena such as self-thinning, successional transitions, trait
-evolution and species coexistence.
 
@@
 Theoretical investigations exploring questions about niche-based
-species coexistence and ecological drift (neutrality) both primarily
-rely on unstructured population models \citep[e.g.][]{MacArthur-1967,
+species coexistence and ecological drift (neutrality) both rely
+primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
   Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
 %
-It is common to assume competitive interactions are influenced by
+It is common to assume competitive outcomes are influenced by
 size-related traits such as adult size and seed size, but detailed
-size-structure is rarely considered, at least for plant models
+size-structure is rarely considered, at least in plant models
 \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
-%
-Consequently, it has been difficult to connect and test the mechanisms
-described in these simplistic models with types of data collected in
-the field, such as physiological leaf traits or demographic data from
-long-term monitoring plots \citep{Adler-2013}.
 %
@@
 large-scale vegetation models lack any real demography or
-size-structure \citep[e.g.][]{Moorcroft-2001, Smith-2014}.
+size-structure \citep[see][]{Moorcroft-2001, Smith-2014}.
 
@@
 % 
-We developed {\plant} primarily to investigate if and how species
+We developed {\plant} primarily to further investigate if and how species
 differing in traits are able to coexist with one another
-\citep{Falster-2011,Falster-2015}.  In order to allow interactions to
-be driven by species traits and arise through competition for shared
-resources, we needed to model a community with a full range of sizes,
-as species may not be able to recruit under the canopy that they
-generate \citep{Moorcroft-2001,Huston-1987}.
+\citep[following][]{Falster-2011,Falster-2015}.  Size-structured models
+provide unique opportunities for coexistence via differentiation in
+successional strategy \citep{Moorcroft-2001,Huston-1987}.
 
 At the same time, we expect the {\plant} model to be useful for
-modelling other phenomena. The internal structure of the {\plant}
+modelling other demographic phenomena. The internal structure of the {\plant}
 model is similar in approach to the ``Ecosystem Demography''
 \citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
-models. Like {\plant}, these models include size- structured
+models. Like {\plant}, these models include size-structured
 competition within a metapopulation of ``patches''. However, whereas
-these models were designed to capture large-scale regional
-biogeochemistry driven by climatic variables and rely on simulations,
+these models rely on simulations and were designed to capture large-scale regional
+biogeochemistry driven by climatic variables,
 {\plant} was designed so that the rules driving the dynamics of the
@@
 isolation, in patches of competing plants and in disturbance-driven
-metapopulations. 
-% TODO: can you change this?  the amount of effort we put in is
-% decoupled from how useful other people may find it.
-Moreover, we put considerable effort into resolving
-competitive hierarchies. 
-The package is thus designed to sit in the middle
-ground between a large-scale coupled simulation model and simple
+metapopulations. The package is thus designed to bridge the gap between 
+a large-scale coupled simulation model and simple
 abstract theoretical device.
@@
 
-\section{Effects of size, trait and environment on demography of individual plants}Q
+\section{Effects of size, trait and environment on demography of individual plants}
 
@@
 production for a plant, given its size and a light environment. Gross
-photosynthetic income is estimated from total leaf area and the light
+photosynthetic income is calculated from total leaf area and the light
 distribution across the plant's canopy. Costs of tissue respiration
@@
 allocated between growth and reproduction.  
-The key outputs needed by higher levels (see below) is growth rate,
+The key outputs needed by higher levels (see below) are height growth rate,
 mortality rate and rate of seed production.  However additional
@@
 The default physiological model used in {\plant} largely reflects that
-presented in \citet{Falster-2011}, but with extensions allowing for
+presented in \citet{Falster-2011,Falster-2015}, but with extensions allowing for
 diameter growth also to be estimated. To properly model trait
@@
 respiration, photosynthesis per leaf area) in the model are linked via
-assumed trade-offs. For example, the trait \textsc{lma} is used to estimate the
-rate of leaf turnover, based on observed scaling relationship spanning
-across diverse vegetation and plant functional global
-\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
+assumed trade-offs.
 
@@
 \citet{Falster-2011}. Parameter values were drawn from available
-literature (see Appendix \ref{sec:FFW16}, table 3); however for some
+literature (see Table 3 in Appendix \ref{sec:FFW16}); however for some
 parameters limited or no data were readily available. Providing
-updated parameter sets is trivial; the challenge is to infer suitable
+updated parameter sets is trivial (see  Appendix \ref{sec:parameters}); 
+the challenge is to infer suitable
 values by constraining outputs from {\plant} against diverse data
@@
 be introduced.  That is, rather than varying the parameters of the
-equations, the equations themselves can be changed, additional
-resources could be competed for and the higher level functionality of
-the package can still be used.  As an intermediate level, we allow for
-parameters to imply relationships among arbitrary other parameters,
-allowing straightforward modelling of trade-offs.  For example,
-setting leaf mass per unit area (\textsc{lma}) affects leaf turnover rate,
-because it is known that leaves which are cheaper to build have shorter 
-lifespan \citep{Wright-2004}.
+equations, the equations themselves can be changed and the higher level 
+functionality of the package can still be used.  As an intermediate level, 
+we allow for changes in key parameters to trigger changes 
+in other parameters (i.e. hyper-parametrisation),
+allowing straightforward modelling of trade-offs.  
+For example, the trait \textsc{lma} is used to estimate the
+rate of leaf turnover, based on an observed scaling relationship spanning
+across diverse vegetation and plant functional types
+\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
 
@@
 producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
-is achieved by integrating an Ordinary Differential Equation of
+is achieved by integrating an ordinary differential equation of
 size-dependent growth rate (see Appendices \ref{sec:demography} and
@@
 stops growing and is increasingly regarded as the most useful
-measure of a strategy's shade tolerance
+measure of a species's shade tolerance
 \citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
@@
 shading on rates of biomass production, although competition for other
-resources such as nitrogen or water could also be considered with
-suitable extensions.
+resources such as nitrogen or water could also be considered by extending the 
+physiological model.
 
@@
 \ref{sec:demography}). The same \textsc{pde} captures average behaviour
-across a large number of small patches \citep{Moorcroft-2001}. The model may 
+across a large number of small patches \citep{Moorcroft-2001}. {\plant} may 
 therefore capture average demographic behaviour across a range of patch sizes.
@@
 however such models are very computationally demanding and also
-impossible to solve in a deterministic fashion \citep{Huston-1987, 
+impossible to solve in a deterministic fashion \citep{Shugart-1980, 
 Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
@@
 
-When modelling a patch, we are interested in modelling changes in the size-
+When modelling a patch, the focus is on changes to the size-
 density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
@@
 \citep{Falster-2011}. Leaf area cover and rates of biomass production
-tend to vary with patch age. Such changes might come about via a
-combination of physiological and structural changes
+tend to vary with patch age 
 \citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
 the effects of competitive feedbacks to be incorporated, in addition to
-any intrinsic physiological factors. Moreover, {\plant} allows for the
-contributions of different species to be mapped (Fig. \ref{fig:patch}c).
+any intrinsic physiological factors, yielding emergent non-linear outcomes
+ (Fig. \ref{fig:patch}c).
 
@@
 weighting the temporal dynamics of a single patch by \(p(a)\), the
-relative abundance of patches aged \(a\) in the metapopulation.
+relative abundance of patches aged \(a\) in the metapopulation 
+ (Fig. \ref{fig:patch}c).
 
@@
 the demography of the metapopulation, this in turn produces an ``output'' 
-number of seeds. The equilibrium seed rain is the vector of seed 
-rain where input seed
-rain equals output seed rain. Finding this equilibrium requires
-finding a root of a multidimensional function, subject to the
+number of seeds. The equilibrium occurs when a given
+input seed
+rain produces an equivalent output seed rain. Finding this equilibrium requires
+finding the root of a multi-dimensional function, subject to the
 constraints that this root is stable (e.g., the trivial equilibrium of
@@
 
-A feature of the size-structured metapopulation concept implemented here
-is that it balances equilibrium and non-equilibrium approaches to
+An attractive  feature of the metapopulation concept implemented here
+is that integrates equilibrium and non-equilibrium approaches to
 modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001, 
@@
 
-The dynamic equilibrium captured by the trait-, size- and patch
-structured metapopulation model potentially resolves several
+The dynamic equilibrium captured by the trait-, size- and 
+patch-structured metapopulation model potentially resolves several
 challenges faced by unstructured models. First, by scaling up from
@@
 (Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
-size-and patch- structured models may be more suitable for linking with 
-satellite and forest survey data, than unstructured models representing the 
-world as homogeneous green slime \citep{Moorcroft-2001, Purves-2008}.
+size- and patch-structured models provide unique opportunities for linking with 
+satellite and forest survey data \citep{Moorcroft-2001, Purves-2008}.
 
@@
 individual with particular traits to establish in a community of
-established residents at their equilibrium densities \citep{Metz-1992}.
-Invasion fitness
-is ideally calculated as the long-term per capita growth rate of the new
-type; however, in structured metapopulations, a more convenient
-indicator of per-capita growth rate is given by the logarithm of the
-basic reproduction ratio \citep{Gyllenberg-2001, Metz-2001}. The basic
-reproduction ratio \(R\) is simply the total seed output of the new
-type, averaged across the metapopulation. A new type can invade when
-\(R > 1\).
+established residents at their equilibrium densities \citep{Geritz-1998,Dieckmann-2007}.
+While ideally calculated as the long-term per capita growth rate, 
+a more convenient
+indicator of invasion fitness in structured metapopulations is given by the logarithm of the
+basic reproduction ratio, \(R\) \citep{Gyllenberg-2001, Metz-2001}. 
+\(R\) is simply the total seed output of the new
+type, averaged across the metapopulation. So a new type can invade when
+ \(\log R > 0\).
 
-With fitness defined, it becomes possible to model adaptive dynamics
-and community assembly \citep{Geritz-1998,Dieckmann-2007}. This is best
+With fitness defined, it becomes possible to model trait-based adaptive dynamics
+and community assembly \citep{Falster-2015}. This is best
 thought of as occurring on a fitness landscape, a surface with fitness
@@
 introduced and evolve, reflecting the non-linear demographic feedbacks
-that occur under frequency-dependent evolution
-\citep{Geritz-1998,Dieckmann-2007}. Following the fitness gradient,
+that occur under frequency-dependent evolution. Stale coexistence is possible when 
+species with different traits can mutually invade \citep{Geritz-1998}
+Following the fitness gradient,
 can lead to an evolutionary branching point
@@
 from any other strategies and represents a prediction for the 
-trait mixture that might might be favoured by natural selection \citep{Geritz-1998}.
+trait mixture that might might be favoured by natural selection 
+\citep{Geritz-1998,Falster-2015}.
 
 Identifying conditions promoting trait diversity is a valuable use-case
-for {\plant}. Coexisting plant species differ in a range of physiological
-traits, yet the conditions allowing for these different types to coexist
-in face of competition for a common set of resources remain largely
-unknown. The simple example here can be extended to multiple dimensions
+for {\plant}. The simple example here can be extended to multiple dimensions
 and can be used to investigate how trade-offs facilitate coexistence
-\citep{Falster-2015}.
+\citep{Falster-2015}. Co-occurring plant species differ in a range of physiological
+traits, yet the conditions allowing for stable coexistence remain largely
+unknown \citep{Adler-2013}.
 
@@
 of community assembly. As examples, we expect the package will prove useful
-for investigating how morphological and physiological traits influence growth
-rate and shade tolerance, and emergent phenomena such as multi-species self-
+for i) investigating how morphological and physiological traits influence growth
+rate and shade tolerance, ii) emergent phenomena such as multi-species self-
 thinning, successional transitions and age-related changes in vegetation
-productivity.
-
-% TODO: In favour of dropping this whole paragraph?
-It is hoped that the approach here can be used to inform both future
-empirical research and model development in neighbouring areas.
-{\plant} aims to bridge the gap between detailed physiological models
-and large-scale mechanistic models of plant growth \citep{Cramer-2001,
-  Sitch-2003, Dekauwe-2014}.  As noted above, most large-scale
-mechanistic models of plant growth do not explicitly model size
-structure, so remain unable to investigate demographic phenomena. At
-the same time, much attention is paid to detailed leaf-level
-physiology. {\plant} aims to bridge this gap that emerged in the model
-landscape, where detailed physiology and size-structured dynamics
-combine to drive forest structure and function via traits.
+productivity, and iii) as a bridge between empirical data, large-scale 
+simulators of global biogeochemical cycles and simple abstract theoretical models.
 
@@
 (b) Competitive hierarchies  are modelled by tracking the height distribution of
-individuals across within a ``patch'' after a disturbance. The intensity
+individuals within a ``patch''. The intensity
 of shading indicates the density of individuals at a given height for
@@
 across the metapopulation, which in turn determines the fitness landscape.
-Resident traits adjust through directional selection up the fitness landscape
+Resident traits adjust through directional selection
 and through the introduction of new species. Adapted from
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
-\caption{Physiological dynamics for individual plants vary with size, trait
-and light environment. (a) Growth trajectories for plants differing in light
-environment and trait (\textsc{lma}, low = 0.1, solid line; high =  0.267, dashed
-line). Lower light environments slow trajectories and interact non-linearly
-with traits. (b) Over time, the fraction of living tissue switches from leaf
+\caption{Physiological dynamics for individual plants varying in size, trait
+and light environment. Solid lines are for high \textsc{lma} strategy, dashed
+lines are for low \textsc{lma} strategy. 
+(a) Growth trajectories are influenced by light environment and trait . 
+(b) Over time, the fraction of living tissue switches from leaf
 towards sapwood, with high \textsc{lma} species having relatively more mass in leaf
-than low \textsc{lma} species. (c) Growth rates peak around 5m high, but this position
-varies with both trait and light level (note that this is the derivative of
-panel (a)). (d) Growth rates vary with plant size (seedling and mature shown)
-and with traits; growth rate differences are more pronounced in high light.
-Whole plant light compensation points emerge from the model as the point where
+than low \textsc{lma} species. (c) Size-dependent growth rates (derivative of
+panel (a)) peak at around 5m high, but this position varies 
+with both trait and light level. 
+(d) Declines in growth rate with light vary with plant size 
+and traits.
+Whole plant light compensation points (indicated by filled dots) emerge as the point where
 growth rate is zero (x intercept).  The light level used in panels (a) and (c)
-is indicated with dots.  See Appendix \ref{sec:plant} for details and code.}
+is indicated with coloured ticks.  See Appendix \ref{sec:plant} for further details and code.}
 \label{fig:plant}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
-\caption{Population dynamics for two species competing within a patch. Lines
-in panels (a) same growth trajectories for individuals on the
+\caption{Population dynamics for two species competing within a patch. 
+Red and blue lines are low and high \textsc{lma} species from Fig. \ref{fig:plant}. 
+Lines in panels (a) show growth trajectories for individuals on the
 boundary of successive ``cohorts'' for each species, using a schedule of 
-introduction times generated by an adaptive refinement algorithm.
- Red lines are the low \textsc{lma}
-species from Fig. \ref{fig:plant}, blue lines are the high \textsc{lma} species.  The
-initial growth rate advantage of the low \textsc{lma} species (see Figure
+introduction times generated by {\plant}'s adaptive algorithm.
+The initial growth rate advantage of the low \textsc{lma} species (see Figure
 \ref{fig:plant} c) means it quickly over tops the high \textsc{lma} species,
-suppressing its growth.  Self-thinning leaves space for the high \textsc{lma} species
-to establish below the canopy of the fast species. Panel (b) shows the same
+suppressing its growth.  Self-thinning creates space for the high \textsc{lma} species
+to establish below the canopy of the faster species. Panel (b) shows the same
 trajectories as in panel (a), but with shading indicating the density $n$ of
@@
 level for each species.  The dotted line is the total leaf area index (sum
-across both species). See Appendix \ref{sec:patch} for details and code.}
+across both species). See Appendix \ref{sec:patch} for further details and code.}
 \label{fig:patch}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/emergent.pdf}
-\caption{Emergent size-distributions across a metapopulation.
+\caption{Emergent size-distributions and demography across a metapopulation.
 Grey lines show patterns within individual patches differing in time
@@
 Panels show (a) patterns of abundance, (b) leaf area and
-(c) height growth rate against size. Note that units of panels (a) and
-(b) are per unit height and per unit ground area, respectively.  See Appendix \ref{sec:emergent} for details and code.}
+(c) height growth rate against size. Units of panels (a) and
+(b) are amount per unit height per unit ground area.  
+See Appendix \ref{sec:emergent} for further details and code.}
 \label{fig:emergent}
@@
 coexistence of multiple types.  Fitness is the log of the population growth
-rate - at zero species are at equilibrium, above zero they will increase in
+rate -- at zero species are at equilibrium, above zero they will increase in
 density.  (a) Landscape generated by a single species.  At \textsc{lma} of 0.1 (the
@@
 (solid line) there is a branching point (see inset for more detail).  (b)
-Landscape generated by two species, holing the first at \textsc{lma} of 0.08.
+Landscape generated by two species, holding the first at \textsc{lma} of 0.08.
 Introducing a species at the maximum fitness value from panel (a) reduces the
@@
 the fitness landscape.  At this point no further invasion is
-possible.  See Appendix \ref{sec:fitness} for details and code.}
+possible.  See Appendix \ref{sec:fitness} for further details and code.}
 \label{fig:fitness}

### 2015-05-29 R/figures_emergent.R

@@ -74,15 +74,18 @@ figure_emergent <- function(data) {
   l_av <- apply(l_hh*n_hh, 1, metapopulation_average) / n_av
   g_av <- apply(g_hh*n_hh, 1, metapopulation_average) / n_av
  
-  op <- par(mfrow=c(3, 1), mar=c(2.1, 4.1, .5, .5), oma=c(2, 0.5, 0, 0))
+  op <- par(mfrow=c(3, 1), mar=c(2.1, 6.1, .5, .5), oma=c(2, 0.5, 0, 0))
   on.exit(par(op))
 
   ## First panel: density vs height:
   xlim <- c(0, max(height, na.rm=TRUE) * 1.05)
   matplot(height[, j], density[, j], type="l", lty=1, col="lightgrey",
-          xlim=xlim, xlab="Height (m)", ylab="Density (1 / m / m2)", las=1,
-          xaxt="n", log="y")
+          xlim=xlim, xlab="Height (m)", ylab=expression("Density"~~("#"~m^-1~m^-2)), las=1,
+          xaxt="n", log="y", yaxt="n")
   axis(1, labels=FALSE)
+  aty <- zapsmall(log10(axTicks(2)))
+  labels <- sapply(aty,function(i) as.expression(bquote(10^ .(i))))
+  axis(2,at=10^aty,labels=labels, las=1)
   matlines(height[, i], density[, i], col=cols, lty=1, type="l")
   points(height[1, i], density[1, i], pch=19, col=cols)
   text(height[1, i] + strwidth("x"), density[1, i],
@@ -93,7 +96,7 @@ figure_emergent <- function(data) {
 
   matplot(height[, j], leaf_area[, j], type="l", lty=1, col="lightgrey",
           xlim=xlim, xlab="Height (m)",
-          ylab="Leaf area density (m2 / m2 / m)", las=1,
+          ylab=expression("Leaf area density"~~(m^2~m^-1~m^-2)), las=1,
           xaxt="n")
   axis(1, labels=FALSE)
   matlines(height[, i], leaf_area[, i], col=cols, lty=1, type="l")
@@ -106,7 +109,7 @@ figure_emergent <- function(data) {
 
   matplot(height[, j], growth_rate[, j], type="l", lty=1, col="lightgrey",
           xlim=xlim, xlab="Height (m)",
-          ylab="Height growth rate (m / year)", las=1)
+          ylab=expression("Height growth rate"~~(m~yr^-1)), las=1)
[…]

### 2015-05-29 R/figures_fitness.R

@@ -73,7 +73,7 @@ figure_fitness_landscape <- function(data) {
   plot(lma[i], w$second1[i], type="l", lty=2,
        log="x", xlim=xlim, ylim=ylim, ylab="Fitness", las=1)
   label_panel(2)
-  mtext("Leaf mass per unit area (LMA; kg / m2)", 1, 3, cex=1)
+  mtext(expression("Leaf mass per unit area"~("LMA;"~kg~m^-2)), 1, 3, cex=1)
   lines(lma[i], w$second2[i], lty=3)
   lines(lma[i], w$second3[i], lty=3)
   lines(lma, w$attractor2)

### 2015-05-29 R/figures_plant.R

@@ -128,7 +128,7 @@ figure_plant <- function() {
   ylim_dhdt <- c(-0.05 * ymax_dhdt, ymax_dhdt)
 
   plot(dhdt_h[["1_1"]], ylim=ylim_dhdt, lty=1, col=cols_light[[1]],
-       type="l", las=1, xlab="Height (m)", ylab="Height growth rate (m/yr)")
+       type="l", las=1, xlab="Height (m)", ylab=expression("Height growth rate"~~(m~yr^-1)))
   lines(dhdt_h[["2_1"]], type="l", lty=2, col=cols_light[[1]])
   lines(dhdt_h[["1_2"]], type="l", lty=1, col=cols_light[[2]])
   lines(dhdt_h[["2_2"]], type="l", lty=2, col=cols_light[[2]])
@@ -139,7 +139,7 @@ figure_plant <- function() {
 
   ## Panel d: wplcp
   plot(NA, xlim=c(0, 1), ylim=ylim_dhdt, las=1,
-       xlab="Canopy openness (%)", ylab="Height growth rate (m/yr)")
+       xlab="Canopy openness (%)", ylab=expression("Height growth rate"~~(m~yr^-1)))
   abline(h=0)
   label_panel(4)
   for (i in seq_along(d1)) {

### 2015-06-01 ms.tex
Add graphicx in
Somehow this is not added?  I thought it was and it doesn't compile
here without.  But then I don't see a change in the log.  Issue
is possibly due to a minimal tex installation here.
Spelling how does it work
Typo-squashing edits throughout.
Mostly from Andrea's read-through

@@
 \newcommand{\plant}{\texttt{plant}}
-
-\usepackage[colorinlistoftodos]{todonotes}
+\usepackage{graphicx}
 
@@
   At its core, {\plant} is an
-  individual-based mechanistic model where plant physiology and demography is mediated by
+  individual-based mechanistic model where plant physiology and demography are mediated by
   traits. Individuals from multiple species can be grown in isolation,
@@
 fractions of their photosynthetic income to building tissues other
-than new leaves; to maintaining support tissues \citep{Givnish-1988,
-  Enquist-2007} and to reproduction \citep{Thomas-2011}.
+than new leaves; for example, to maintaining support tissues
+\citep{Givnish-1988, Enquist-2007} or to reproduction
+\citep{Thomas-2011}.
 %
@@
 of one individual depends on the size and traits of nearby
-individuals \cite{Pacala-1996}.
+individuals \citep{Pacala-1996}.
 %
@@
 Although the importance of size-structure for vegetation dynamics and
-diversity, has been long been recognised \citep[e.g.,][]{Harper-1977,
+diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
   Shugart-1980, Huston-1987}, current research in this area is
@@
 It is common to assume competitive outcomes are influenced by
-size-related traits such as adult size and seed size, but detailed
+size-related traits such as adult and seed size, but detailed
 size-structure is rarely considered, at least in plant models
 \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
-%
+% TODO: Andrea notes that this section does not flow well.  I agree,
+% but let's leave it until after submission.
 Similarly, most large-scale models of vegetation dynamics do not
@@
 models. Like {\plant}, these models include size-structured
-competition within a metapopulation of ``patches''. However, whereas
+competition within a metapopulation of ``patches''. However, while
 these models rely on simulations and were designed to capture large-scale regional
-biogeochemistry driven by climatic variables,
-{\plant} was designed so that the rules driving the dynamics of the
+biogeochemistry driven by climatic variables, we designed
+{\plant} so that the rules driving the dynamics of the
 system could be manipulated and emergent demographic behaviours
-studied deterministically and in detail. In {\plant} it is possible to grow plants in
+studied deterministically and in detail. In {\plant}, it is possible to grow plants in
 isolation, in patches of competing plants and in disturbance-driven
 metapopulations. The package is thus designed to bridge the gap between 
-a large-scale coupled simulation model and simple
+a large-scale coupled simulation model and a simple
 abstract theoretical device.
@@
 functioning of an individual plant and how these are influenced by its
-traits, size and light environment. Being driven by traits, the model
-can be extended to potentially very many species. Users can also add
-their own physiological models, though this is a work in progress.
+traits, size and light environment. By being driven by traits, the
+model can be extended to potentially very many species. Users will
+also be able to add their own physiological models, though this is a
+work in progress.
 
@@
 We model the emergent behaviour of patches deterministically, based on a
-continual  size-density distribution. As such, it is not required to specify a
-particular patch size.
+continual  size-density distribution. As such, specifying a patch size
+is not required.
 
@@
 
-The core of {\plant} is a model for an individual plant's
+% NOTE: Add something saying: we define a species as a group of
+% individuals with identical traits?
+The core of {\plant} is a model for an individual species'
 physiological strategy as specified by its traits
 (Fig. \ref{fig:schematic}a). This model estimates rate of biomass
-production for a plant, given its size and a light environment. Gross
-photosynthetic income is calculated from total leaf area and the light
-distribution across the plant's canopy. Costs of tissue respiration
-and turnover are then subtracted. The remaining biomass is then
-allocated between growth and reproduction.  
-The key outputs needed by higher levels (see below) are height growth rate,
-mortality rate and rate of seed production.  However additional
-quantities are also computed, such as total assimilation and the
-respiration, turnover and allocation rates for different tissues (see
-Appendix \ref{sec:plant})
+production for a plant, given its size and the current light
+environment. Gross photosynthetic income is calculated from the total
+leaf area and the light distribution across the plant's canopy. Costs
+of tissue respiration and turnover are then subtracted. The remaining
+biomass is then allocated between growth and reproduction.  The key
+outputs needed by higher levels of the model (see below) are height
+growth rate, mortality rate and rate of seed production.  However
+additional quantities are also computed, such as total assimilation
+and the respiration, turnover and allocation rates for different
+tissues (see Appendix \ref{sec:plant})
 
+% TODO: Might need example/citation for trade-offs here?
 The default physiological model used in {\plant} largely reflects that
-presented in \citet{Falster-2011,Falster-2015}, but with extensions allowing for
-diameter growth also to be estimated. To properly model trait
-variation, different parameters (such as rates of leaf turnover,
+presented in \citet{Falster-2011,Falster-2015}, but with extensions
+allowing for diameter growth also to be estimated. To properly model
+trait variation, different parameters (such as rates of leaf turnover,
 respiration, photosynthesis per leaf area) in the model are linked via
-assumed trade-offs.
+assumed trade-offs (see Appendix \ref{sec:parameters}).  We refer to
+this default model as \texttt{FFW16}, reflecting our initials in this
+paper.
 
@@
 \citet{Falster-2011}. Parameter values were drawn from available
-literature (see Table 3 in Appendix \ref{sec:FFW16}); however for some
-parameters limited or no data were readily available. Providing
-updated parameter sets is trivial (see  Appendix \ref{sec:parameters}); 
-the challenge is to infer suitable
-values by constraining outputs from {\plant} against diverse data
-sources \citep{Lebauer-2012, Keenan-2013}. This is an area in which we
-are still actively working and one of the major use cases we envisage
-for {\plant}.
+literature (see Table 3 in Appendix \ref{sec:FFW16}); however data for
+some parameters were limited. Providing updated parameter sets is
+trivial (see Appendix \ref{sec:parameters}); the challenge is to infer
+suitable values by constraining outputs from {\plant} against diverse
+data sources \citep{Lebauer-2012, Keenan-2013}. This is an area in
+which we are still actively working and one of the major use cases we
+envisage for {\plant}.
 
@@
 be introduced.  That is, rather than varying the parameters of the
-equations, the equations themselves can be changed and the higher level 
-functionality of the package can still be used.  As an intermediate level, 
-we allow for changes in key parameters to trigger changes 
-in other parameters (i.e. hyper-parametrisation),
-allowing straightforward modelling of trade-offs.  
-For example, the trait \textsc{lma} is used to estimate the
-rate of leaf turnover, based on an observed scaling relationship spanning
-across diverse vegetation and plant functional types
+equations, the equations themselves can be changed and the higher
+level functionality of the package can still be used.  As an
+intermediate level, we allow for changes in key parameters to trigger
+changes in other parameters (i.e. hyper-parametrisation), allowing
+straightforward modelling of trade-offs.  For example, the trait
+\textsc{lma} (leaf mass per unit leaf area) is used to estimate the
+rate of leaf turnover, based on an observed scaling relationship
+spanning across diverse vegetation and plant functional types
 \citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
@@
 
-Ideally one would also consider spatial interactions within patches,
+Ideally, one would also consider spatial interactions within patches,
 however such models are very computationally demanding and also
@@
 
-Our approach for modelling size-structured population dynamics
-is based on the Escalator Boxcar Train technique (\textsc{ebt})
-\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves the \textsc{pde}
-describing development of \(n\) by approximating the density function
-with a collection of cohorts spanning the size spectrum. Following a
-disturbance, a series of cohorts are introduced into each patch. These
-cohorts are then transported along the characteristic equations of the
-\textsc{pde}. Biologically these are the growth trajectories of individuals,
-conditioned on them surviving (e.g. Fig. \ref{fig:patch}a). Meanwhile,
-growth and mortality combine to alter the density of individuals along
-the trajectories (Fig. \ref{fig:patch}b).
+Our approach for modelling size-structured population dynamics is
+based on the Escalator Boxcar Train (\textsc{ebt}) technique
+\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves
+the \textsc{pde} describing development of \(n\) by approximating the
+density function with a collection of cohorts spanning the size
+spectrum. Following a disturbance, a series of cohorts are introduced
+into each patch. These cohorts are then transported along the
+characteristic equations of the \textsc{pde}. Biologically these are
+the growth trajectories of individuals, conditioned on them surviving
+(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
+to alter the density of individuals along the trajectories
+(Fig. \ref{fig:patch}b).
 
-We introduce two extensions to the \textsc{ebt}: i) a new method for estimating
-the size-density distribution and ii) a technique for handling strong
-size-asymmetric competitive feedbacks, such as occur under strong
-competition for light (see Appendix \ref{sec:demography} for
-details). The original \textsc{ebt} \citep{Deroos-1988,
-  Deroos-1992, Deroos-1997} proceeds by approximating the first and
-second moments of the density distribution \(n\) within each
-cohort. In {\plant} we use a different approach. Instead of tracking
-first and second moments of the size-distribution within cohorts, we
-track a point mass estimate of \(n\) along the trajectory for the
-boundary of each cohort. We found this approach preferable because it
-does not require us to maintain a separate set of equations for the
-smallest cohort \citep{Deroos-1997}.
+We introduce two extensions to the \textsc{ebt}: i) a new method for
+estimating the size-density distribution and ii) a technique for
+handling strong size-asymmetric competitive feedbacks, such as occur
+under strong competition for light (see Appendix \ref{sec:demography}
+for details). The original \textsc{ebt} proceeds by approximating the
+first and second moments of the density distribution \(n\) within each
+cohort \citep{Deroos-1997}.
+%
+In {\plant} we use a different approach. Instead of tracking first and
+second moments of the size-distribution within cohorts, we track a
+point mass estimate of \(n\) along the trajectory for the boundary of
+each cohort. We found this approach preferable because it does not
+require us to maintain a separate set of equations for the smallest
+cohort.
 
@@
 adaptive refining of time points at which new cohorts are introduced
-into the population. In the original \textsc{ebt}, cohorts were introduced at a
-fixed time interval. Under competition, the growth trajectories of
-individuals born at similar times can diverge substantially over time
-(Fig. \ref{fig:patch}a; Appendix \ref{sec:demography},
-\ref{sec:patch} and \ref{sec:cohort-spacing}). This can become a
-problem later during patch development because much of the population
-can be located between widely-spaced cohorts, delivering low numerical
-precision. To maintain numerical precision we therefore use an
-iterative algorithm to adaptively refine the cohort spacing until the
-trajectories of adjacent cohorts remain adequately resolved throughout
-development of the patch (see Appendix \ref{sec:demography} for
-details). This refinement causes an uneven distribution of cohort
-introduction times across patch age (Fig.  \ref{fig:patch}a), with
-considerably tighter spacing of cohorts in younger patches.
+into the population. In the original \textsc{ebt}, cohorts were
+introduced at a fixed time interval. Under competition, the growth
+trajectories of individuals born at similar times can diverge
+substantially over time (Fig. \ref{fig:patch}a; Appendix
+\ref{sec:demography}, \ref{sec:cohort-spacing} and
+\ref{sec:patch}). This can become a problem later during patch
+development because much of the population can be located between
+widely-spaced cohorts, causing low numerical accuracy. To maintain
+numerical accuracy, we use an iterative algorithm to adaptively refine
+the cohort spacing until the trajectories of adjacent cohorts remain
+adequately resolved throughout development of the patch (see Appendix
+\ref{sec:demography} for details). This refinement causes an uneven
+distribution of cohort introduction times across patch age (Fig.
+\ref{fig:patch}a), with considerably tighter spacing of cohorts in
+younger patches.
 
@@
 
-Modelling of size-structured dynamics via solving of a deterministic \textsc{pde}
-enables multiple complex non-linear demographic phenomena to be
-investigated. First, {\plant} can simulate patterns of patch development after
-disturbance, including self-thinning behaviours and successional
-replacement. Whereas previous approaches for modelling self-thinning
-were limited to a single species
-\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates multiple
- species with different traits (Fig. \ref{fig:schematic}b; Fig.
-\ref{fig:patch}). Second, {\plant} allows for the effects of competition and succession on
-productivity and biomass accumulation within a patch to be investigated
-\citep{Falster-2011}. Leaf area cover and rates of biomass production
-tend to vary with patch age 
-\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. {\plant} enables
-the effects of competitive feedbacks to be incorporated, in addition to
-any intrinsic physiological factors, yielding emergent non-linear outcomes
- (Fig. \ref{fig:patch}c).
+Modelling of size-structured dynamics via solving of a deterministic
+\textsc{pde} enables multiple demographic phenomena to be
+investigated. First, {\plant} can simulate patterns of patch
+development after a disturbance, including self-thinning and
+successional replacement. Whereas previous approaches for modelling
+self-thinning were limited to a single species
+\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates
+multiple species with different traits (Fig. \ref{fig:schematic}b;
+Fig.  \ref{fig:patch}). Second, {\plant} allows for the effects of
+competition and succession on productivity and biomass accumulation
+within a patch to be investigated \citep{Falster-2011}. Leaf area
+cover and rates of biomass production tend to vary with patch age
+\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. 
+% TODO: I really don't know what the last bit of this sentence is
+% meant to be saying.  Obviously it's something to do with the
+% "emergent" properties of the patch, but not clear to me.
+{\plant} enables the effects of competitive feedbacks to be
+incorporated, in addition to any intrinsic physiological factors,
+yielding emergent non-linear outcomes (Fig. \ref{fig:patch}c).
 
@@
 
-Most vegetation is subject to a disturbance regime, such that patches of
-the landscape are cleared at some rate, by fire, cyclone, landslide, or
-disease outbreak
-\citep{White-1979, Bormann-1979, Coomes-2007, Chambers-2013}.
-The vegetation thus comprises a collection of patches differing in time
-since last disturbance and linked via seed dispersal (Fig.
-\ref{fig:schematic}b). Such a system is often refereed to as a
+% TODO: I removed the four(!) references here as I don't think this is
+% controversial.
+All vegetation is subject to a disturbance regime, such that patches
+of the landscape are cleared at some rate, by fire, cyclone od
+landslide.  The vegetation thus comprises a series of patches
+differing in time since last disturbance and linked via seed dispersal
+(Fig.  \ref{fig:schematic}b). Such a system is often referred to as a
 structured metapopulation \citep{Gyllenberg-2001}.
@@
 In {\plant} we assume an infinite number of patches, all experiencing
-the same disturbance regime and linked via seed dispersal (Fig.
-\ref{fig:schematic}b). We also assume that the disturbances remove
-all established vegetation.  With this approach we can scale up from
-simulating a single patch to simulate an entire metapopulation.
-The frequency-density
-\(p(a)\) of patches aged \(a\) in the landscape is needed. Assuming there
-are a large number of patches means the dynamics of \(p\) behave
-deterministically and can be approximated via a second \textsc{pde}
-(see Appendix \ref{sec:demography} for
-details). Scaling from patch to metapopulation is then achieved by
-weighting the temporal dynamics of a single patch by \(p(a)\), the
-relative abundance of patches aged \(a\) in the metapopulation 
- (Fig. \ref{fig:patch}c).
+the same disturbance regime and sharing a common seed dispersal pool
+(i.e., the island model; Fig. \ref{fig:schematic}b). We also assume
+that the disturbances remove all established vegetation.  With this
+approach we can scale up from simulating a single patch to simulate an
+entire metapopulation.  The frequency-density \(p(a)\) of patches aged
+\(a\) in the landscape is needed. Assuming there are a large number of
+patches means the dynamics of \(p\) behave deterministically and can
+be approximated via a second \textsc{pde} (see Appendix
+\ref{sec:demography} for details). Scaling from patch to
+metapopulation is then achieved by weighting the temporal dynamics of
+a single patch by \(p(a)\), the relative abundance of patches aged
+\(a\) in the metapopulation (Fig. \ref{fig:patch}c).
 
-The main numerical challenge in stepping from single patch to
+The main numerical challenge in stepping from a single patch to a
 metapopulation is to identify the equilibrium seed rain of the
-metapopulation. An ``input'' seed rain is required to simulate 
-the demography of the metapopulation, this in turn produces an ``output'' 
-number of seeds. The equilibrium occurs when a given
-input seed
-rain produces an equivalent output seed rain. Finding this equilibrium requires
-finding the root of a multi-dimensional function, subject to the
-constraints that this root is stable (e.g., the trivial equilibrium of
-zero seed rain always exists, but is often not stable) (Appendix
+metapopulation. An ``input'' seed rain is required to simulate the
+demography of the metapopulation, this in turn produces an ``output''
+number of seeds. The equilibrium occurs when a given input seed rain
+produces an equivalent output seed rain. Finding this equilibrium
+requires finding the root of a multi-dimensional function, subject to
+the constraint that this root is stable (e.g., the trivial equilibrium
+of zero seed rain always exists, but is often not stable; Appendix
 \ref{sec:equilibrium}).
@@
 An attractive feature of the metapopulation concept implemented here
-is that integrates equilibrium and non-equilibrium approaches to
+is that integrates both equilibrium and non-equilibrium approaches to
 modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001,
-Falster-2011}. An
-equilibrium may be approached at the level of metapopulation, meaning
-the seed rain and size structure across the entire metapopulation is
-approximately stable. Yet the structure of vegetation within individual
-patches is constantly in flux. Patches continue to age and their
-residents continue to grow until they are disturbed.
+  Falster-2011}. An equilibrium may be approached at the level of
+metapopulation, meaning the seed rain and size structure across the
+entire metapopulation is approximately stable. Yet the structure of
+vegetation within individual patches is constantly in flux. Patches
+continue to age and their residents continue to grow until they are
+disturbed.
 
@@
 create a self-regulating system. Unlike some vegetation models, there
-is no need to set the density of seeds or seedlings, these arise as
-outputs from the model, based on the combined effects of the model's
-physiological rules, species traits, competitive interactions and the
-prevailing disturbance regime.
+is no need to set the density of seeds or seedlings.  These densities
+arise as outputs from the model, based on the combined effects of the
+model's physiological rules, species traits, competitive interactions
+and the prevailing disturbance regime.
 
+% NOTE: Andrea points out that young patches may not be the most
+% common patch type in the metapopulation - it will depend on the
+% regime.  And on the definition of young.
 Second, the metapopulation concept allows for the effects of disturbance
@@
 patch-level outcomes over the frequency-distribution of patch ages. Because
-young patches are more common in the landscape, metapopulation averages are
+young patches tend to be more common in the landscape, metapopulation averages are
 strongly influenced by the structure of younger patches (e.g. 
@@
 
-Finally, with the extension to metapopulation, we are modelling demography across
-the entire life cycle and are thus in a position to estimate invasion fitness -- 
-the ability of a rare
+Finally, with the extension to the metapopulation level, we are
+modelling demography across the entire life cycle and are thus in a
+position to estimate invasion fitness -- the ability of a rare
 individual with particular traits to establish in a community of
-established residents at their equilibrium densities \citep{Geritz-1998,Dieckmann-2007}.
-While ideally calculated as the long-term per capita growth rate, 
-a more convenient
-indicator of invasion fitness in structured metapopulations is given by the logarithm of the
-basic reproduction ratio, \(R\) \citep{Gyllenberg-2001, Metz-2001}. 
-\(R\) is simply the total seed output of the new
-type, averaged across the metapopulation. So a new type can invade when
- \(\log R > 0\).
+established residents at their equilibrium densities
+\citep{Geritz-1998,Dieckmann-2007}.  While ideally calculated as the
+long-term per capita growth rate, a more convenient indicator of
+fitness in structured metapopulations is given by the
+logarithm of the basic reproduction ratio, \(R\)
+\citep{Gyllenberg-2001, Metz-2001}.
+% TODO: Ideally, don't start a sentence with an abbreviation
+\(R\) is simply the total seed output of the new type, averaged across
+the metapopulation. So a new type can invade when \(\log R > 0\).
 
-With fitness defined, it becomes possible to model trait-based adaptive dynamics
-and community assembly \citep{Falster-2015}. This is best
+With this measure of fitness, we can model trait-based adaptive
+dynamics and community assembly \citep{Falster-2015}. This is best
 thought of as occurring on a fitness landscape, a surface with fitness
-plotted against one or more species traits (Fig. \ref{fig:fitness}). The
-reproductive success of individuals possessing a particular trait value
-depends on both the traits and densities of other species in the current
-resident community. Fitness landscapes will be positive in regions that allow invasion,
-and zero for species that are at equilibrium. The slope of the fitness
-landscape at a resident indicates the direction of selection; a
-negative slope (dashed line in Fig. \ref{fig:fitness}a) indicates
-selection for smaller trait values (see Appendix \ref{sec:fitness} for
+plotted against one or more species traits
+(Fig. \ref{fig:fitness}). The reproductive success of individuals
+possessing a particular trait value depends on both the traits and
+densities of other species in the current resident community. Fitness
+landscapes will be positive in regions that allow invasion, and zero
+for species that are at equilibrium. The slope of the fitness
+landscape with respect to traits indicates the direction of selection
+(dashed line in Fig. \ref{fig:fitness}a) indicates selection for
+smaller trait values (see Appendix \ref{sec:fitness} for
 details). Fitness landscapes are constantly reshaped as species are
 introduced and evolve, reflecting the non-linear demographic feedbacks
-that occur under frequency-dependent evolution. Stale coexistence is possible when 
-species with different traits can mutually invade \citep{Geritz-1998}
-Following the fitness gradient,
-can lead to an evolutionary branching point
-\citep{Geritz-1998,Dieckmann-2007}; at this point there is no
-directional selection (solid line in Fig. \ref{fig:fitness}a) but
-invasion is possible both above and below it in trait space. It is
-often possible to find a configuration where fitness is negative
-everywhere (solid line in Fig. \ref{fig:fitness}b) except for the zero
-points where species are present. This community is immune to invasion
-from any other strategies and represents a prediction for the 
-trait mixture that might might be favoured by natural selection 
-\citep{Geritz-1998,Falster-2015}.
+that occur under frequency-dependent evolution. Stable coexistence is
+possible when species with different traits can mutually invade
+\citep{Geritz-1998} Following the fitness gradient may lead to an
+evolutionary branching point \citep{Geritz-1998,Dieckmann-2007}; at
+this point there is no directional selection (solid line in
+Fig. \ref{fig:fitness}a) but invasion is possible both above and below
+it in trait space. It is often possible to find a configuration where
+fitness is negative everywhere (solid line in Fig. \ref{fig:fitness}b)
+except for the zero points where species are present. This community
+is immune to invasion from any other strategies and represents a
+prediction for the trait mixture that might might be favoured by
+natural selection \citep{Geritz-1998,Falster-2015}.
 
@@
 \citep{Eddelbuettel-2011, Eddelbuettel-2013} to bridge between the two
-languages. Due to the computationally intensive nature of the model, the
-core physiological and demographic components are in C++ to maximise
-speed. We used templated types that allow the demographic and evolutionary
-components to be driven by alternative physiological models while
-retaining the same interface at higher levels. Despite mostly being
-programmed in C++, all components of the model, down to individual
-plants, can be extracted and interacted with dynamically in R, using the
-\texttt{RcppR6} package \citep{RcppR6}.
+languages. Due to the computationally intensive nature of the model,
+the core physiological and demographic components are in C++ to
+maximise speed. We used templated types that allow the demographic and
+evolutionary components to be driven by alternative physiological
+models while retaining the same interface at higher levels. Despite
+mostly being written in C++, all components of the model, down to
+individual plants, can be extracted and interacted with dynamically in
+R, using the \texttt{RcppR6} package \citep{RcppR6}.
 
-{\plant} makes use of much existing software, including the programming
-language C++; the R computational environment \citep{R-2015}, the R
-packages \texttt{Rcpp} \citep{Eddelbuettel-2011, Eddelbuettel-2013},
-\texttt{R6} \citep{Chang-2014} and \texttt{testthat}
-\citep{Wickham-2011}; and the Boost Library for C++
+{\plant} makes use of much existing software, including the R
+computational environment \citep{R-2015}, the R packages \texttt{Rcpp}
+\citep{Eddelbuettel-2011, Eddelbuettel-2013}, \texttt{R6}
+\citep{Chang-2014}; and the Boost Library for C++
 \citep{Schaling-2014} via the R package \texttt{BH}
@@
 \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
-Installation instructions are available at this homepage, binary
+Installation instructions are available at this webpage, binary
 releases will also be available there.  (Note that {\plant} does not
@@
 
-The {\plant} package makes it easier for researchers and forest practitioners
-alike to investigate a variety of non-linear demographic phenomena.  Rather
-than provide a model with a few entry points, we have developed an extensible
-framework that we hope can be used to answer a number of questions at a
-variety of scales. The specific use-cases highlighted here arise from our
-interest in the assembly of trait mixtures in communities. However a variety
-of other uses would be possible and also extensions to address other aspects
-of community assembly. As examples, we expect the package will prove useful
-for i) investigating how morphological and physiological traits influence growth
-rate and shade tolerance, ii) emergent phenomena such as multi-species self-
-thinning, successional transitions and age-related changes in vegetation
-productivity, and iii) as a bridge between empirical data, large-scale 
-simulators of global biogeochemical cycles and simple abstract theoretical models.
+The {\plant} package makes it easier for researchers and forest
+practitioners alike to investigate a variety of non-linear demographic
+phenomena.  Rather than provide a model with a few entry points, we
+have developed an extensible framework that we hope can be used to
+answer a number of questions at a variety of scales. The specific
+use-cases highlighted here arise from our interest in the assembly of
+trait mixtures in communities. However a variety of other uses would
+be possible to address other aspects of community assembly. For
+example, we expect the package will prove useful for i) investigating
+how morphological and physiological traits influence growth rate and
+shade tolerance, ii) emergent phenomena such as multi-species self-
+thinning, successional transitions and age-related changes in
+vegetation productivity, and iii) as a bridge between empirical data,
+large-scale simulators of global biogeochemical cycles and simple
+abstract theoretical models.
 
@@
 
-DS Falster was supported by an ARC discovery grant (DP110102086). RG
-FitzJohn was supported by the Science and Industry Endowment Fund
-(RP04-174). M Westoby was supported by a fellowship from Australian
-Research Council. We thank {\AA} Brn{\"a}nnstr{\"o}m, J Camac, 
-A De Roos, U Dieckmann, R Duursma, J Johansson and C Prentice 
-for helpful discussions.
+DS Falster was supported by an Australian Research Council (ARC)
+discovery grant (DP110102086). RG FitzJohn was supported by the
+Science and Industry Endowment Fund (RP04-174). M Westoby was
+supported by a fellowship from the ARC. We thank {\AA}
+Br{\"a}nnstr{\"o}m, J Camac, A De Roos, U Dieckmann, R Duursma, J
+Johansson and C Prentice for helpful discussions.
 
@@
 Dynamics across all three levels are driven by the
-physiological sub model determining an individual's physiological and
+physiological sub-model determining an individual's physiological and
 demographic rates on the basis of its traits, size and light environment (a).
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
-\caption{Physiological dynamics for individual plants varying in size, trait
-and light environment. Solid lines are for high \textsc{lma} strategy, dashed
-lines are for low \textsc{lma} strategy. 
-(a) Growth trajectories are influenced by light environment and trait . 
-(b) Over time, the fraction of living tissue switches from leaf
-towards sapwood, with high \textsc{lma} species having relatively more mass in leaf
-than low \textsc{lma} species. (c) Size-dependent growth rates (derivative of
-panel (a)) peak at around 5m high, but this position varies 
-with both trait and light level. 
-(d) Declines in growth rate with light vary with plant size 
-and traits.
-Whole plant light compensation points (indicated by filled dots) emerge as the point where
-growth rate is zero (x intercept).  The light level used in panels (a) and (c)
-is indicated with coloured ticks.  See Appendix \ref{sec:plant} for further details and code.}
+\caption{Physiological dynamics for individual plants varying in size,
+  trait and light environment. Solid lines are for the high
+  \textsc{lma} (leaf mass per unit area) strategy, dashed lines are
+  for the low \textsc{lma} strategy.  (a) Growth trajectories are
+  influenced by both light environment and traits.  (b) Over time, the
+  fraction of living tissue switches from leaf towards sapwood, with
+  high \textsc{lma} species having relatively more mass in leaf than
+  low \textsc{lma} species. (c) Size-dependent growth rates
+  (derivative of panel (a)) peak at around 5m high, but this position
+  varies with both trait and light level.  (d) Declines in growth rate
+  with light vary with plant size and traits.  Whole plant light
+  compensation points (indicated by dots) emerge as the point where
+  growth rate is zero (x intercept).  The light level used in panels
+  (a) and (c) is indicated with coloured ticks.  See Appendix
+  \ref{sec:plant} for further details and code.}
 \label{fig:plant}
@@
 \caption{Population dynamics for two species competing within a patch.
-Red and blue lines are low and high \textsc{lma} species from Fig. \ref{fig:plant}. 
-Lines in panels (a) show growth trajectories for individuals on the
-boundary of successive ``cohorts'' for each species, using a schedule of 
-introduction times generated by {\plant}'s adaptive algorithm.
-The initial growth rate advantage of the low \textsc{lma} species (see Figure
-\ref{fig:plant} c) means it quickly over tops the high \textsc{lma} species,
-suppressing its growth.  Self-thinning creates space for the high \textsc{lma} species
-to establish below the canopy of the faster species. Panel (b) shows the same
-trajectories as in panel (a), but with shading indicating the density $n$ of
-individuals at given size and patch age (light = low density, dark = high density).  
-(c) The leaf area index at ground
-level for each species.  The dotted line is the total leaf area index (sum
-across both species). See Appendix \ref{sec:patch} for further details and code.}
+  Red and blue lines are low and high \textsc{lma} species from
+  Fig. \ref{fig:plant}.  Lines in panel (a) show growth trajectories
+  for individuals on the boundary of successive ``cohorts'' for each
+  species, using a schedule of introduction times generated by
+  {\plant}'s adaptive algorithm.  The initial growth rate advantage of
+  the low \textsc{lma} species (see Figure \ref{fig:plant}c) means it
+  quickly over tops the high \textsc{lma} species, suppressing its
+  growth.  Self-thinning creates space for the high \textsc{lma}
+  species to establish below the canopy of the faster species. Panel
+  (b) shows the same trajectories as in panel (a), but with shading
+  indicating the density $n$ of individuals at given size and patch
+  age (light = low density, dark = high density).  (c) The leaf area
+  index at ground level for each species.  The dotted line is the
+  total leaf area index (summed across both species), and the lines on
+  the right side are average when integrated over the whole
+  metapopulation structure.  See Appendix \ref{sec:patch} for further
+  details and code.}
 \label{fig:patch}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
-\caption{Example fitness landscape for \textsc{lma}, showing potential for stable
-coexistence of multiple types.  Fitness is the log of the population growth
-rate -- at zero species are at equilibrium, above zero they will increase in
-density.  (a) Landscape generated by a single species.  At \textsc{lma} of 0.1 (the
-dashed line) there is directional selection for smaller \textsc{lma}.  At \textsc{lma} of 0.08
-(solid line) there is a branching point (see inset for more detail).  (b)
-Landscape generated by two species, holding the first at \textsc{lma} of 0.08.
-Introducing a species at the maximum fitness value from panel (a) reduces the
-region of positive fitness considerably (dashed line).  Dotted lines show
-subsequent invasions and replacements of the second species, the solid
-line shows a stable fitness landscape where both species exist on a peak in
-the fitness landscape.  At this point no further invasion is
-possible.  See Appendix \ref{sec:fitness} for further details and code.}
+\caption{Example fitness landscape for \textsc{lma} (leaf mass per
+  unit leaf area), showing potential for stable coexistence of
+  multiple types.  Fitness is the log of the population growth rate --
+  at zero species are at equilibrium, above zero they will increase in
+  density.  (a) Landscape generated by a single species (indicated by
+  dots).  At \textsc{lma} of 0.1 (the dashed line, open dot) there is
+  directional selection for smaller \textsc{lma}.  At \textsc{lma} of
+  0.08 (solid line, filled dot) there is a branching point (see inset
+  for more detail).  (b) Landscape generated by two species, holding
+  the first at \textsc{lma} of 0.08.  Introducing a species at the
+  maximum fitness value from panel (a) reduces the region of positive
+  fitness considerably (dashed line, open dot).  Dotted lines show
+  subsequent invasions and replacements of the second species, the
+  solid line shows a stable fitness landscape where both species exist
+  on a peak in the fitness landscape.  At this point no further
+  invasion is possible.  See Appendix \ref{sec:fitness} for further
+  details and code.}
 \label{fig:fitness}

### 2015-06-01 ms.tex
Few small edits

@@
 
-% TODO: Might need example/citation for trade-offs here?
 The default physiological model used in {\plant} largely reflects that
@@
 this default model as \texttt{FFW16}, reflecting our initials in this
-paper.
+paper. For example, the trait
+\textsc{lma} (leaf mass per unit leaf area) is used to estimate the
+rate of leaf turnover, based on an observed scaling relationship
+spanning across diverse vegetation and plant functional types
+\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
 
@@
 changes in other parameters (i.e. hyper-parametrisation), allowing
-straightforward modelling of trade-offs.  For example, the trait
-\textsc{lma} (leaf mass per unit leaf area) is used to estimate the
-rate of leaf turnover, based on an observed scaling relationship
-spanning across diverse vegetation and plant functional types
-\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
+straightforward modelling of trade-offs.
 
@@
 \textsc{pde} enables multiple demographic phenomena to be
-investigated. First, {\plant} can simulate patterns of patch
-development after a disturbance, including self-thinning and
+investigated. First, {\plant} can simulate demographic 
+phenomena typically observed in patches
+recovering from a disturbance, including self-thinning and
 successional replacement. Whereas previous approaches for modelling
@@
 multiple species with different traits (Fig. \ref{fig:schematic}b;
-Fig.  \ref{fig:patch}). Second, {\plant} allows for the effects of
-competition and succession on productivity and biomass accumulation
-within a patch to be investigated \citep{Falster-2011}. Leaf area
-cover and rates of biomass production tend to vary with patch age
-\citep{Binkley-2002, Smith-2001, Coomes-2007, Ogawa-2010}. 
-% TODO: I really don't know what the last bit of this sentence is
-% meant to be saying.  Obviously it's something to do with the
-% "emergent" properties of the patch, but not clear to me.
-{\plant} enables the effects of competitive feedbacks to be
-incorporated, in addition to any intrinsic physiological factors,
-yielding emergent non-linear outcomes (Fig. \ref{fig:patch}c).
+Fig.  \ref{fig:patch}). Moreover, these features emerge from the 
+combined effects of physiology and competition for resources, rather 
+than being prescribed in the model. Second, {\plant} allows for
+the combined effects of physiology and succession on emergent 
+characteristics of the entire patch to be investigated
+\citep{Falster-2011}. Leaf area cover and rates of biomass 
+production tend to vary  non-linearly with patch 
+age \citep{Binkley-2002, Smith-2001, Ogawa-2010};
+in {\plant} such outcomes arise naturally as emergent properties, 
+driven by demography (Fig. \ref{fig:patch}c).
 
@@
 
-% TODO: I removed the four(!) references here as I don't think this is
-% controversial.
 All vegetation is subject to a disturbance regime, such that patches
@@
 \citep{Gyllenberg-2001, Metz-2001}.
-% TODO: Ideally, don't start a sentence with an abbreviation
-\(R\) is simply the total seed output of the new type, averaged across
-the metapopulation. So a new type can invade when \(\log R > 0\).
+The basic reproduction ratio is simply the total seed output of 
+the new type, averaged across the metapopulation. So a new type can 
+invade when \(\log R > 0\).
 

### 2015-06-01 ms.tex
No conflicts of interest
Work around manuscript central :fire::fire:

@@
 \newcommand{\plant}{\texttt{plant}}
+\newcommand{\smurl}[1]{{\footnotesize\url{#1}}}
 \usepackage{graphicx}
@@
 Br{\"a}nnstr{\"o}m, J Camac, A De Roos, U Dieckmann, R Duursma, J
-Johansson and C Prentice for helpful discussions.
+Johansson and C Prentice for helpful discussions. The authors have no 
+conflicts of interest to declare.
 
@@
 
-See attached file \url{vignettes/plant.pdf}, knitr source
-\url{vignettes/plant.Rmd}, plain source, \url{vignettes/R/plant.R}
+\emph{Note}: due to Manuscript Central not accepting \textsc{html}
+files, we have had to convert this (and the following appendices) to
+\textsc{pdf}s.  \textsc{html} versions are available on the wiki:
+https://github.com/traitecoevo/plant\_paper/wiki --- the \textsc{html}
+versions have less pixelated graphics and are easier to copy and
+paste out of.
+
+See attached file \url{vignettes/plant.pdf}, 
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/plant.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/plant}
 
@@
 
-See attached file \url{vignettes/cohort_spacing.pdf}, knitr source
-\url{vignettes/cohort_spacing.Rmd}, plain source, \url{vignettes/R/cohort_spacing.R}
+See attached file \url{vignettes/cohort_spacing.pdf},
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/cohort_spacing.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/cohort_spacing}
 
@@
 
-See attached file \url{vignettes/equilibrium.pdf}, knitr source
-\url{vignettes/equilibrium.Rmd}, plain source, \url{vignettes/R/equilibrium.R}
+See attached file \url{vignettes/equilibrium.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/equilibrium.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/equilibrium}
 
@@
 
-See attached file \url{vignettes/patch.pdf}, knitr source
-\url{vignettes/patch.Rmd}, plain source, \url{vignettes/R/patch.R}
+See attached file \url{vignettes/patch.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/patch.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/patch}
 
@@
 
-See attached file \url{vignettes/emergent.pdf}, knitr source
-\url{vignettes/emergent.Rmd}, plain source, \url{vignettes/R/emergent.R}
+See attached file \url{vignettes/emergent.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/emergent.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/emergent}
 
@@
 
-See attached file \url{vignettes/fitness.pdf}, knitr source
-\url{vignettes/fitness.Rmd}, plain source, \url{vignettes/R/fitness.R}
+See attached file \url{vignettes/fitness.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/fitness.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/fitness}
 
 \section{Modifying parameters of the physiological model}\label{sec:parameters}
-
-See attached file \url{vignettes/parameters.pdf}, knitr source
-\url{vignettes/parameters.Rmd}, plain source, \url{vignettes/R/parameters.R}
+See attached file \url{vignettes/parameters.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/parameters.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/parameters}
 

### 2015-06-01 release.sh

@@ -6,7 +6,10 @@ mkdir release
 cp ms.pdf release
 
 mkdir supporting_material
-cp vignettes/*.html vignettes/*.pdf supporting_material
+for file in $(find vignettes -name '*.html' | sed -e 's|.*/||'); do
+    echo "Converting ${file}"
+    (cd vignettes; pandoc "$file" -o ../supporting_material/"${file%.html}.pdf" --template template.tex)
+done
 zip -r release/supporting_material.zip supporting_material
 rm -rf supporting_material
 

### 2015-08-21 ms.tex

@@
 When modelling a patch, the focus is on changes to the size-
-density-distribution \(n(x,h,a)\) with patch age \(a\), i.e. the density of
+density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density of
 individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
 Density is measured as individuals per unit height and per unit ground area.
-Modelling the development of  \(n(x,h,a)\) requires that both the initial size
+Modelling the development of  \(N(x, H,a)\) requires that both the initial size
 distribution and the inflow of new recruits be specified. We have primarily
@@
 \citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves
-the \textsc{pde} describing development of \(n\) by approximating the
+the \textsc{pde} describing development of \(N\) by approximating the
 density function with a collection of cohorts spanning the size
@@
 for details). The original \textsc{ebt} proceeds by approximating the
-first and second moments of the density distribution \(n\) within each
+first and second moments of the density distribution \(N\) within each
 cohort \citep{Deroos-1997}.
@@
 second moments of the size-distribution within cohorts, we track a
-point mass estimate of \(n\) along the trajectory for the boundary of
+point mass estimate of \(N\) along the trajectory for the boundary of
 each cohort. We found this approach preferable because it does not
@@
 approach we can scale up from simulating a single patch to simulate an
-entire metapopulation.  The frequency-density \(p(a)\) of patches aged
+entire metapopulation.  The frequency-density \(P(a)\) of patches aged
 \(a\) in the landscape is needed. Assuming there are a large number of
-patches means the dynamics of \(p\) behave deterministically and can
+patches means the dynamics of \(P\) behave deterministically and can
 be approximated via a second \textsc{pde} (see Appendix
@@
 metapopulation is then achieved by weighting the temporal dynamics of
-a single patch by \(p(a)\), the relative abundance of patches aged
+a single patch by \(P(a)\), the relative abundance of patches aged
 \(a\) in the metapopulation (Fig. \ref{fig:patch}c).
@@
   (b) shows the same trajectories as in panel (a), but with shading
-  indicating the density $n$ of individuals at given size and patch
+  indicating the density $N$ of individuals at given size and patch
   age (light = low density, dark = high density).  (c) The leaf area

### 2015-08-21 vignettes/src/parameters.R
Further parameter renaming

@@ -43,13 +43,13 @@ pl$strategy$lma
 FFW16_hyperpar
 
 ## You will rarely need to call this function directly (see below) but
-## note how setting of lma affects parameters `k_l`, `c_p1` and `c_Rl`
+## note how setting of lma affects parameters `k_l`, `a_p1` and `r_l`
 FFW16_hyperpar(trait_matrix(0.1, "lma"), s)
 
 ## These are:
 ##   * `k_l`: Turnover rate for leaves
-##   * `c_p1`: Leaf photosynthesis per area
-##   * `c_Rl`: Leaf respiration per mass
+##   * `a_p1`: Leaf photosynthesis per area
+##   * `r_l`: Leaf respiration per mass
 
 ## This means that it is possible to implement different trade-offs
 ## between parameters relatively easily by modifying the
@@ -95,8 +95,8 @@ length(ss)
 ## We can then use standard r command to extract variable from this list
 sapply(ss, function(x) x$lma)
 sapply(ss, function(x) x$k_l)
-sapply(ss, function(x) x$c_p1)
-sapply(ss, function(x) x$c_Rl)
+sapply(ss, function(x) x$a_p1)
+sapply(ss, function(x) x$r_l)
 
 ## There's a convenience function `plant_list` that returns a set of
 ## *plants* based on a vector of traits:

### 2015-08-24 ms.tex
Minor comments from reviewers

@@
 of one individual depends on the size and traits of nearby
-individuals \citep{Pacala-1996}.
+individuals \citep{Shugart-1980, Pacala-1996}.
 %
@@
 competing plants and of meta-populations structured by a prevailing
-disturbance regime.
+disturbance regime.ƒ
 %
@@
 \citep[following][]{Falster-2011,Falster-2015}.  Size-structured models
-provide unique opportunities for coexistence via differentiation in
+provide unique opportunities for studying coexistence via differentiation in
 successional strategy \citep{Moorcroft-2001,Huston-1987}.
@@
 Detailed technical documents are provided as supplementary information
-(see Appendices), updated versions of these technical documents are
+(see Appendices); updated versions of these technical documents are
 available within the {\plant} package itself.
@@
 
-{\plant} is a mechanistic model, meaning the dynamics of the system
-arise from rules specifying how individuals grow and interact.  The
+{\plant} is an individual-based model, meaning the dynamics of the population arise from rules specifying how individuals grow and interact. The
 core rules in {\plant} are about the short-term physiological
@@
 presented in \citet{Falster-2011,Falster-2015}, but with extensions
-allowing for diameter growth also to be estimated. To properly model
-trait variation, different parameters (such as rates of leaf turnover,
+allowing for diameter growth also to be estimated. We refer to
+this default model as \texttt{FFW16}, reflecting our initials in this
+paper.  To model
+trait variation properly, different parameters (such as rates of leaf turnover,
 respiration, photosynthesis per leaf area) in the model are linked via
-assumed trade-offs (see Appendix \ref{sec:parameters}).  We refer to
-this default model as \texttt{FFW16}, reflecting our initials in this
-paper. For example, the trait
+assumed trade-offs (see Appendix \ref{sec:parameters}).  For example, the trait
 \textsc{lma} (leaf mass per unit leaf area) is used to estimate the
@@
 literature (see Table 3 in Appendix \ref{sec:FFW16}); however data for
-some parameters were limited. Providing updated parameter sets is
+some parameters were limited. Providing updated parameter sets to the
+algorithm is
 trivial (see Appendix \ref{sec:parameters}); the challenge is to infer
@@
 of absolute growth with size \citep{King-2011} and equally the widely
-regarded decline in relative growth rate with size from birth onwards
+observed decline in relative growth rate with size from birth onwards
 \citep{Enquist-2007}.
@@
 
-In {\plant} we assume an infinite number of patches, all experiencing
+In {\plant} and following \citep{Kohyama-1993, Moorcroft-2001,
+  Falster-2011} we assume an infinite number of patches, all experiencing
 the same disturbance regime and sharing a common seed dispersal pool
@@
 demography of the metapopulation, this in turn produces an ``output''
-number of seeds. The equilibrium occurs when a given input seed rain
+number of seeds. The equilibrium occurs when, for each species,
+a given input seed rain
 produces an equivalent output seed rain. Finding this equilibrium
@@
 established residents at their equilibrium densities
-\citep{Geritz-1998,Dieckmann-2007}.  While ideally calculated as the
+\citep{Metz-1992, Geritz-1998}.  While ideally calculated as the
 long-term per capita growth rate, a more convenient indicator of
@@
 With this measure of fitness, we can model trait-based adaptive
-dynamics and community assembly \citep{Falster-2015}. This is best
+dynamics and community assembly of forest communities, following
+ \citep{Falster-2015}. This is best
 thought of as occurring on a fitness landscape, a surface with fitness
@@
 possible when species with different traits can mutually invade
-\citep{Geritz-1998} Following the fitness gradient may lead to an
-evolutionary branching point \citep{Geritz-1998,Dieckmann-2007}; at
+\citep{Geritz-1998, Chesson-2000}. Following the fitness gradient may lead to an
+evolutionary branching point \citep{Geritz-1998}; at
 this point there is no directional selection (solid line in
@@
 prediction for the trait mixture that might might be favoured by
-natural selection \citep{Geritz-1998,Falster-2015}.
+natural selection \citep{Geritz-1998}.
 

### 2015-08-24 R/tables.R
Allow function to work with missing values
(needed because can't yet retrieve default vals for hyper parameters)

@@ -15,22 +15,28 @@ table_plant_parameters <- function(filename, dest){
   tab$Units[to_math] <-
             gsub(" ", "\\,", tab$Units[to_math], fixed =TRUE)
 
-  to_code <- tab[["\\plant"]] != ""
+  s <- FFW16_Strategy()
+  tab$Value <- NA_real_
+
+  # Now extract actual values from plant
+  to_code <- tab[["\\plant"]] != "" & tab[["\\plant"]] %in% names(s)
   code <- tab[["\\plant"]][to_code]
 
-  s <- FFW16_Strategy()
   val <- s[code]
   err <- sapply(val, is.null)
   if (any(err)) {
-    stop("Missing parameters: ", paste(code[err], collapse=", "))
-  }
+    message("Missing parameters: ", paste(code[err], collapse=", "))
+  } 
 
   oo <- options(scipen=999)
   on.exit(options(oo))
-  tab$Value[to_code] <- prettyNum(unlist(val))
+  tab$Value[to_code] <- unlist(lapply(val, prettyNum))
+
+  # Display names in latex
+  to_code <- tab[["\\plant"]] != ""
 
   tab[["\\plant"]][to_code] <- sprintf("\\texttt{%s}",
-                                      gsub("_", "\\\\_", code))
+                                      gsub("_", "\\\\_", tab[["\\plant"]][to_code]))
 
   x <- xtable(tab,
               hline.after=c(1),

### 2015-08-27 ms.tex
Start adreessing reviewers comments

@@
 \item
+% TODO: Rework to downgrade physiological model. Stochastic plants
+
   Here we introduce the {\plant} package, an
@@
 
-Although the importance of size-structure for vegetation dynamics and
-diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
+Although the importance of size-structure for trait evolution, vegetation
+dynamics and diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
   Shugart-1980, Huston-1987}, current research in this area is
@@
 \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
-% TODO: Andrea notes that this section does not flow well.  I agree,
-% but let's leave it until after submission.
-Similarly, most large-scale models of vegetation dynamics do not
-include size-structured demography \citep{Cramer-2001, Sitch-2003,
-  Purves-2008,Dekauwe-2014}.  During the 1980s, size-structured
-``gap'' models were widely used
-\citep[e.g.][]{Shugart-1980,Huston-1987}. However, the next generation
-of models focused more heavily on biogeochemistry at the expense of
-demography \citep[e.g.][]{Woodward-1995,Haxeltine-1996,Cramer-2001, 
-Sitch-2003}. While the
-importance of size-structure is starting again to be recognised
-\citep{Moorcroft-2001, Purves-2008, Falster-2011, Dekauwe-2014}, all but a couple
-of the most popular,
-large-scale vegetation models lack any real demography or
-size-structure \citep[see][]{Moorcroft-2001, Smith-2014}.
+%
+Similarly, many of the models used to study global vegetation dynamics across
+the last 20 yrs have not included size-structured demography
+\citep[for comparisons of some major models see][]{Sitch-2008,
+Dekauwe-2014}.
+%
+While the importance of explicitly including size-structured dynamics is
+increasingly recognised in both areas of evolutionary ecology
+\citep[e.g.][]{Falster-2015} and vegetation dynamics
+\citep[e.g.][]{Moorcroft-2001, Purves-2008, Smith-2014,
+Weng-2015,Sakschewski-2015}, our understanding on how these features
+impact on the ecology and evolution of vegetation communities remains
+relatively limited.
 
@@
 competing plants and of meta-populations structured by a prevailing
-disturbance regime.ƒ
+disturbance regime.
 %
@@
 differing in traits are able to coexist with one another
-\citep[following][]{Falster-2011,Falster-2015}.  Size-structured models
+\citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
 provide unique opportunities for studying coexistence via differentiation in
-successional strategy \citep{Moorcroft-2001,Huston-1987}.
+successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
+At the same time, we expect {\plant} to be useful for
+modelling other demographic phenomena.
 
-At the same time, we expect the {\plant} model to be useful for
-modelling other demographic phenomena. The internal structure of the {\plant}
-model is similar in approach to the ``Ecosystem Demography''
-\citep{Moorcroft-2001} and \textsc{lpj-guess} \citep{Smith-2014}
-models. Like {\plant}, these models include size-structured
-competition within a metapopulation of ``patches''. However, while
-these models rely on simulations and were designed to capture large-scale regional
-biogeochemistry driven by climatic variables, we designed
-{\plant} so that the rules driving the dynamics of the
-system could be manipulated and emergent demographic behaviours
-studied deterministically and in detail. In {\plant}, it is possible to grow plants in
-isolation, in patches of competing plants and in disturbance-driven
-metapopulations. The package is thus designed to bridge the gap between
-a large-scale coupled simulation model and a simple
-abstract theoretical device.
+Broadly, the {\plant} model falls into a class we refer to Trait- Size- and
+Patch- structured Models (\textsc{TSPMs}). \textsc{TSPMs} are direct
+descendants of the ``gap'' models developed in 1980s
+\citep[e.g.][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
+include \textsc{ED} \citep{Moorcroft-2001}, \textsc{LPJ- GUESS}
+\citep{Smith-2014} and \textsc{LPJml-FIT} \citep{Sakschewski-2015}. Common to
+\textsc{TSPMs} are that they explicitly model size-structured competition
+among individual plants within a metapopulation of ``patches'', where
+individuals can also differ in their traits and thus demographic behaviour.
+\textsc {TSPMs} also typically ignore the spatial structure of plants within
+patches and the spatial structure of patches relative to one another; instead
+focussing computational effort on resolving the dynamics of size-structured
+competition for light and successional turnover. Differences among
+\textsc{TSPMs} arise from both the core physiological models used (describing
+how the demography of individual plants responds to resource availability),
+and the numerical techniques applied to scaling the physiological model up to
+estimate emergent behaviours of individuals, patches and vegetation.
 
 Below we describe the general approach of the package, then provide
-sections with details at nested levels of ecological organisation.
+sections outlining potential applications at nested levels of ecological organisation.
 Detailed technical documents are provided as supplementary information
@@
 
-{\plant} is an individual-based model, meaning the dynamics of the population arise from rules specifying how individuals grow and interact. The
+{\plant} is an individual-based model, meaning the dynamics of the
+population arise from rules specifying how individuals grow and interact. 
+Being driven by traits, the model can be extended to potentially very many species. The
 core rules in {\plant} are about the short-term physiological
 functioning of an individual plant and how these are influenced by its
-traits, size and light environment. By being driven by traits, the
-model can be extended to potentially very many species. Users will
-also be able to add their own physiological models, though this is a
-work in progress.
+traits, size and light environment.  While {\plant}
+includes a core physiological model, we use it here only as an example to
+demonstrate the different ways in which any given physiological model
+could be analysed. To be clear, it is not the purpose
+of this paper to justify the particular physiological model used, to provide detailed
+comparisons to data, or necessarily promote this physiological model over
+those in other \textsc{TSPMs}. (These are all topics worthy of
+substantially more attention than can be afforded here.) Moreover, an important
+feature of {\plant} is that it allows for users to add their own
+physiological models.
 
-On top of this physiological model, the package implements methods for
+On top of the physiological model, {\plant} implements methods for
 population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
@@
 physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
-We model the emergent behaviour of patches deterministically, based on a
-continual  size-density distribution. As such, specifying a patch size
-is not required.
+We offer the ability to model the emergent behaviour of patches with
+specified size, as well as a deterministic size-independent approximation
+in which specifying a patch size is not required.
 
@@
 
-% NOTE: Add something saying: we define a species as a group of
-% individuals with identical traits?
-The core of {\plant} is a model for an individual species'
-physiological strategy as specified by its traits
-(Fig. \ref{fig:schematic}a). This model estimates rate of biomass
+The core of {\plant} is a model for an individual species' physiological
+strategy as specified by its traits (Fig. \ref{fig:schematic}a). For
+present purposes, a species is defined as a group of individuals with identical
+traits. The effects of trait variation is to modify parameters of the core
+physiological model, which in turn translate into different demographic outcomes.
+
+The default physiological model used in {\plant} largely reflects
+that presented in \citet{Falster-2011,Falster-2015}, but with extensions
+allowing for diameter growth also to be estimated. We refer to this
+default model as \texttt{FFW16}, reflecting the author initials and
+publication year of this paper. The \texttt{FFW16} model estimates rate
+of biomass
 production for a plant, given its size and the current light
@@
 biomass is then allocated between growth and reproduction. The key
-outputs needed by higher levels of the model (see below) are height
+outputs needed by higher levels of the {\plant} are height
 growth rate, mortality rate and rate of seed production.  However
@@
 and the respiration, turnover and allocation rates for different
-tissues (see Appendix \ref{sec:plant})
+tissues and these intermediates can also be accessed (see Appendix
+\ref{sec:plant}).
 
-The default physiological model used in {\plant} largely reflects that
-presented in \citet{Falster-2011,Falster-2015}, but with extensions
-allowing for diameter growth also to be estimated. We refer to
-this default model as \texttt{FFW16}, reflecting our initials in this
-paper.  To model
-trait variation properly, different parameters (such as rates of leaf turnover,
-respiration, photosynthesis per leaf area) in the model are linked via
-assumed trade-offs (see Appendix \ref{sec:parameters}).  For example, the trait
-\textsc{lma} (leaf mass per unit leaf area) is used to estimate the
-rate of leaf turnover, based on an observed scaling relationship
-spanning across diverse vegetation and plant functional types
-\citep{Wright-2004} (see Appendix \ref{sec:FFW16} for details).
-
-The current parametrisation of the physiological model is based on
-\citet{Falster-2011}. Parameter values were drawn from available
-literature (see Table 3 in Appendix \ref{sec:FFW16}); however data for
-some parameters were limited. Providing updated parameter sets to the
-algorithm is
-trivial (see Appendix \ref{sec:parameters}); the challenge is to infer
-suitable values by constraining outputs from {\plant} against diverse
-data sources \citep{Lebauer-2012, Keenan-2013}. This is an area in
-which we are still actively working and one of the major use cases we
-envisage for {\plant}.
-
-While we have implemented one physiological model (\texttt{FFW16}),
-the package is designed to allow arbitrary additional physiologies to
-be introduced.  That is, rather than varying the parameters of the
-equations, the equations themselves can be changed and the higher
-level functionality of the package can still be used.  As an
-intermediate level, we allow for changes in key parameters to trigger
-changes in other parameters (i.e. hyper-parametrisation), allowing
-straightforward modelling of trade-offs.
-
-\subsection{Applications}
+While we have implemented one physiological model (\texttt{FFW16}), the
+package is designed to allow arbitrary additional physiologies to be
+introduced. Users are able to modify the physiological model in three
+different ways. First, one can vary the parameters of the \texttt{FFW16} model
+directly (see Table XXX for list, for examples see Appendix
+\ref{sec:parameters}). Second, we allow for changes in key parameters to
+trigger changes in other parameters (i.e. a hyper-parametrisation), enabling
+straightforward modelling of trade-offs. For example, the trait \textsc{lma}
+(leaf mass per unit leaf area) is used to estimate the rate of leaf turnover
+($k_l$), based on an observed scaling relationship spanning across diverse
+vegetation and plant functional types \citep{Wright-2004}:
+$$k_l = \beta_{kl1} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{kl2}},$$ 
+where $k_l$ and $\phi$ (lma) are both parameters of the core physiological
+model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
+linkages are defined within a user-supplied hyper-parameterisation function
+(see Appendix \ref{sec:FFW16} for details). Finally, rather than varying the
+parameters of the existing equations in the \texttt{FFW16} model, the
+equations themselves can be changed and the higher level functionality of the
+package can still be used.  The latter corresponds to defining a new
+physiological model. As an example, in the \texttt{FFW16b} physiology, we have modified
+the function determining allocation to reproduction. The specifications
+for writing a new physiological model are outlined further in Appendix 
+XXX.
 
@@
 stops growing and is increasingly regarded as the most useful
-measure of a species's shade tolerance
+measure of a species' shade tolerance
 \citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
@@
 resources such as nitrogen or water could also be considered by extending the
-physiological model.
-
-We assume that patches are vertically but not horizontally
-structured. In other words, we account for size differences but not spatial 
-layout within the patch. Similar assumptions are found in many models
-simulating size-structured dynamics
-\citep{Shugart-1980,Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}. In addition, 
-we assume that patches are very large -- this means the population dynamics 
-within the patch behave deterministically \citep{Deroos-1997} and can therefore 
-be approximated via a Partial Differential Equation (\textsc{pde}) (see Appendix
-\ref{sec:demography}). The same \textsc{pde} captures average behaviour
-across a large number of small patches \citep{Moorcroft-2001}. {\plant} may 
-therefore capture average demographic behaviour across a range of patch sizes.
-
-Ideally, one would also consider spatial interactions within patches,
-however such models are very computationally demanding and also
+physiological model. We assume that patches are vertically but not horizontally
+structured; in other words, we account for size differences but not spatial
+layout within the patch. Similar assumptions are found in other \textsc{TSPMs}
+\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
+While ideally, one would also consider spatial interactions within patches,
+such models are very computationally demanding and also
 impossible to solve in a deterministic fashion \citep{Shugart-1980,
@@
 
-When modelling a patch, the focus is on changes to the size-
+When modelling a patch of competing, the focus is on changes to the size-
 density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density of
-individuals with traits \(x\) and height \(h\) within a patch of age \(a\).
+individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
 Density is measured as individuals per unit height and per unit ground area.
-Modelling the development of  \(N(x, H,a)\) requires that both the initial size
+Modelling the development of  \(N(x, H, a)\) requires that both the initial size
 distribution and the inflow of new recruits be specified. We have primarily
@@
 
-Our approach for modelling size-structured population dynamics is
+The {\plant} model offers two methods for modelling the dynamics of \(N(x,
+H, a)\) in a competing population. In the first ``stochastic'' mode, users
+specify an average seed rain and patch size, {\plant} then generates a
+vector of seed arrival events and simulates the stochastic
+development of the resulting finite-sized population (see Appendices XXX for an
+example).
+
+For most applications, however, a preferable alternative to modelling
+dynamics within a finite-sized stochastic patch, is to use {\plant}'s 
+``determinstic'' approximation. The ``determinstic'' mode assumes that
+patches are very large; at this point the population dynamics within the
+patch approach their deterministic limit and can therefore
+be approximated via a Partial Differential Equation (\textsc{pde})
+\citep{Deroos-1997, Kohyama-1993}.  The same \textsc{pde} is also thought to
+capture the average behaviour across a large number of small patches
+\citep{Moorcroft-2001}. Importantly, the ``determinstic'' mode is
+also much faster than the stochastic model and without noise (more
+details in Appendix \ref{sec:demography}).
+
+Our approach for modelling deterministic size-structured population dynamics is
 based on the Escalator Boxcar Train (\textsc{ebt}) technique
@@
 adaptive refining of time points at which new cohorts are introduced
-into the population. In the original \textsc{ebt}, cohorts were
-introduced at a fixed time interval. Under competition, the growth
+into the population, an idea first applied in \citet{Falster-2011} and
+described further in \citet{Falster-2015}.
+In the original \textsc{ebt}, cohorts were introduced at a fixed time
+interval. Under competition, however, the growth
 trajectories of individuals born at similar times can diverge
@@
 
-\subsection{Applications}
-
 Modelling of size-structured dynamics via solving of a deterministic
@@
 recovering from a disturbance, including self-thinning and
-successional replacement. Whereas previous approaches for modelling
-self-thinning were limited to a single species
-\citep[e.g.][]{Barnes-2004, Coomes-2007}, {\plant} accommodates
+successional replacement. Whereas other approaches for modelling
+self-thinning are limited to a single species
+\citep[e.g.][]{Barnes-2004, Coomes-2007}, \textsc{TSPMs} can accommodate
 multiple species with different traits (Fig. \ref{fig:schematic}b;
@@
 combined effects of physiology and competition for resources, rather
-than being prescribed in the model. Second, {\plant} allows for
+than being prescribed in the model. Second, \textsc{TSPMs} allows for
 the combined effects of physiology and succession on emergent
 characteristics of the entire patch to be investigated
-\citep{Falster-2011}. Leaf area cover and rates of biomass 
+\citep{Moorcroft-2001, Falster-2011}. Leaf area cover and rates of biomass
 production tend to vary  non-linearly with patch
-age \citep{Binkley-2002, Smith-2001, Ogawa-2010};
+age \citep{Smith-2001, Binkley-2002, Ogawa-2010};
 in {\plant} such outcomes arise naturally as emergent properties,
@@
 All vegetation is subject to a disturbance regime, such that patches
-of the landscape are cleared at some rate, by fire, cyclone od
+of the landscape are cleared at some rate, by fire, cyclone or
 landslide.  The vegetation thus comprises a series of patches
@@
 that the disturbances remove all established vegetation.  With this
-approach we can scale up from simulating a single patch to simulate an
-entire metapopulation.  The frequency-density \(P(a)\) of patches aged
+approach we can scale up from deterministically solving the dynamics of
+a single patch to solving the dynamics of an entire metapopulation.
+The frequency-density \(P(a)\) of patches aged
 \(a\) in the landscape is needed. Assuming there are a large number of
@@
 
-\subsection{Applications}
-
 The dynamic equilibrium captured by the trait-, size- and
@@
 patch to metapopulation, we close the demographic feedback loop to
-create a self-regulating system. Unlike some vegetation models, there
-is no need to set the density of seeds or seedlings.  These densities
+create a self-regulating system. These densities
 arise as outputs from the model, based on the combined effects of the
@@
 
-% NOTE: Andrea points out that young patches may not be the most
-% common patch type in the metapopulation - it will depend on the
-% regime.  And on the definition of young.
 Second, the metapopulation concept allows for the effects of disturbance
@@
 as total leaf area, biomass and productivity -- are obtained by averaging
-patch-level outcomes over the frequency-distribution of patch ages. Because
-young patches tend to be more common in the landscape, metapopulation averages are
-strongly influenced by the structure of younger patches (e.g.
+patch-level outcomes over the frequency-distribution of patch ages (e.g.
 Fig. \ref{fig:patch}c).  Similarly, the distributions of  abundance and growth rate
@@
 established residents at their equilibrium densities
-\citep{Metz-1992, Geritz-1998}.  While ideally calculated as the
+\citep{Metz-1992, Chesson-2000}.  While ideally calculated as the
 long-term per capita growth rate, a more convenient indicator of
@@
 
+Something hear about adaptive dynamics and fitness
+
 Identifying conditions promoting trait diversity is a valuable use-case
@@
 
+\section{Comparison to existing software}
+
+Increased interest in capturing the behaviour of size-structured  populations
+has prompted development of several tools, each with its own distinct
+strengths.
+
+The \textsc{EBT} software (see De Roos XXXX in this special feature) provides 
+a flexible framework that has been used to study a range of marine organisms.
+Compared to the \textsc{EBT}, {\plant} is more focussed on plants, and 
+to that end offers refinements needed to handle modelling of the strong 
+size-asymmetric feedbacks that occur in plant communities, as well as the 
+extension to structured metapopulation.
+
+Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
+\citep[ver 1 and 2][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
+\citep{Smith-2014} and a\textsc{LPJml-FIT}  \citep{Sakschewski-2015}. These
+models were designed for simulation of bio-geochemical cycles and vegetation
+structure across broad geographic areas; as such their architectures are
+optimised for these use-cases. By contrast, {\plant} was  designed so that the
+physiological rules driving the dynamics of the system could be easily
+manipulated and emergent demographic behaviours (including fitness) studied in
+detail. These different use-cases underpin important design differences among
+the models. Whereas the primary purpose of these other models is to scale a
+given physiology up to predict  broad-scale changes in the terrestrial
+biosphere, {\plant} allows users to  interact with the model at multiple
+levels of ecological organisation and  study the dynamics in detail. With any
+given physiology, it is possible to  grow plants in isolation, in patches of
+competing plants and in disturbance- driven metapopulations. {\plant} also
+allows for greater focus on population dynamics, reporting not only vegetation
+properties but also individual level densities, growth rates and fitness.
+{\plant} thus aims to bridge the gap between existing large-scale coupled
+simulation models like \textsc{ED} and the extremely simple models that are
+still widely  used in theoretical ecology.
+
+As the only other mdoel with open source code, we also note some more specific 
+differences between \textsc{ED} and {\plant}.
+\begin{itemize}
+  \item More effort is devoted in {\plant} to resolving competitive
+  hierarchies within developing patches. To that end we use a finer-scale 
+  adaptive time-step; allow for vertical overlap in the canopies of different 
+  plants and expend more effort on resolving the temporal spacing of cohorts
+  within a single patch. These details are essential to retrieving stable
+  estimates of fitness but possibly make little different to patch-scale
+  outcomes like total biomass or leaf area cover.
+
+  \item  \textsc{ED} includes sub-models for handling water, nutrient and 
+  temperature effects, and is thus better at capturing water, carbon and 
+  nutrient fluxes.
+
+  \item In order to approximate a single archetypal patch and thus study 
+  metapopulations at equilibrium, {\plant} requires a function giving the 
+  average photosynthesis per year as a function of canopy depth. While we
+  derive this function from a more-detailed sub-model, it is currently not
+  possible to explicitly simulate intra-annual cycles in environmental
+  drivers.
+
+  \item The physiological model in  \textsc{ED} has undergone more rigorous
+  comparisons with data \citep[see][]{Medvigy-2009}.  Providing updated 
+  parameter sets to the \texttt{FFW16} model is trivial (see Appendix 
+  \ref{sec:parameters}); the challenge is to infer
+  suitable values by constraining outputs from {\plant} against diverse
+  data sources \citep[e.g.][]{Medvigy-2009, Lebauer-2012}.
+
+  \item {\plant} is distributed as an R package an as such offers greater
+  possibilities for interactive use. 
+\end{itemize}
+
+Finally, like {\plant} the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
+to scale from individual-level behaviours to emergent properties of
+vegetation, including estimating invasion fitness. Yet the approaches taken
+for solving the dynamics are very different. The \textsc{LM3-PPA} relies on
+complex analytical solutions to the size- structured interactions, whose
+derivation required simplifying elements of the physiology and ecology in the
+model. By contrast in {\plant} we have retained the full range of
+physiological and ecological feedbacks, and instead relied on using robust
+numerical methods to solve the system dynamics (see Appendix 
+\ref{sec:demography} for further details).
+
 \section{Closing comments}
@@
 Science and Industry Endowment Fund (RP04-174). M Westoby was
-supported by a fellowship from the ARC. We thank {\AA}
-Br{\"a}nnstr{\"o}m, J Camac, A De Roos, U Dieckmann, R Duursma, J
-Johansson and C Prentice for helpful discussions. The authors have no 
-conflicts of interest to declare.
-
-\clearpage
-\bibliography{refs}
-
-\clearpage
-
-\section{Figures}\label{figures}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/schematic}
-
-\caption{Processes modelled within {\plant} include physiological, population and
-evolutionary dynamics.
-Dynamics across all three levels are driven by the
-physiological sub-model determining an individual's physiological and
-demographic rates on the basis of its traits, size and light environment (a).
-(b) Competitive hierarchies  are modelled by tracking the height distribution of
-individuals within a ``patch''. The intensity
-of shading indicates the density of individuals at a given height for
-different species (distinguished by colours). A size- and patch-structured
-metapopulation consists of a distribution of patches linked by seed dispersal.
-Disturbances occasionally remove all vegetation within a patch, resetting the
-patch. (c) The traits of the resident species determine the shading environment
-across the metapopulation, which in turn determines the fitness landscape.
-Resident traits adjust through directional selection
-and through the introduction of new species. Adapted from
-\citet{Falster-2011}, \citet{Falster-2015}. }
-
-\label{fig:schematic}
-\end{figure}
-
-\newpage
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
-\caption{Physiological dynamics for individual plants varying in size,
-  trait and light environment. Solid lines are for the high
-  \textsc{lma} (leaf mass per unit area) strategy, dashed lines are
-  for the low \textsc{lma} strategy.  (a) Growth trajectories are
-  influenced by both light environment and traits.  (b) Over time, the
-  fraction of living tissue switches from leaf towards sapwood, with
-  high \textsc{lma} species having relatively more mass in leaf than
-  low \textsc{lma} species. (c) Size-dependent growth rates
-  (derivative of panel (a)) peak at around 5m high, but this position
-  varies with both trait and light level.  (d) Declines in growth rate
-  with light vary with plant size and traits.  Whole plant light
-  compensation points (indicated by dots) emerge as the point where
-  growth rate is zero (x intercept).  The light level used in panels
-  (a) and (c) is indicated with coloured ticks.  See Appendix
-  \ref{sec:plant} for further details and code.}
-\label{fig:plant}
-\end{figure}
-
-\newpage
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
-\caption{Population dynamics for two species competing within a patch.
-  Red and blue lines are low and high \textsc{lma} species from
-  Fig. \ref{fig:plant}.  Lines in panel (a) show growth trajectories
-  for individuals on the boundary of successive ``cohorts'' for each
-  species, using a schedule of introduction times generated by
-  {\plant}'s adaptive algorithm.  The initial growth rate advantage of
-  the low \textsc{lma} species (see Figure \ref{fig:plant}c) means it
-  quickly over tops the high \textsc{lma} species, suppressing its
-  growth.  Self-thinning creates space for the high \textsc{lma}
-  species to establish below the canopy of the faster species. Panel
-  (b) shows the same trajectories as in panel (a), but with shading
-  indicating the density $N$ of individuals at given size and patch
-  age (light = low density, dark = high density).  (c) The leaf area
-  index at ground level for each species.  The dotted line is the
-  total leaf area index (summed across both species), and the lines on
-  the right side are average when integrated over the whole
-  metapopulation structure.  See Appendix \ref{sec:patch} for further
-  details and code.}
-\label{fig:patch}
-\end{figure}
-
-\newpage
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/emergent.pdf}
-\caption{Emergent size-distributions and demography across a metapopulation.
-Grey lines show patterns within individual patches differing in time
-since disturbance (age), with blue lines indicating patches of a
-specific age. Red lines show average relationship across the metapopulation, 
-obtained by weighting the grey lines by patch and individual abundance. 
-Panels show (a) patterns of abundance, (b) leaf area and
-(c) height growth rate against size. Units of panels (a) and
-(b) are amount per unit height per unit ground area.  
-See Appendix \ref{sec:emergent} for further details and code.}
-\label{fig:emergent}
-\end{figure}
-
-\newpage
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
-\caption{Example fitness landscape for \textsc{lma} (leaf mass per
-  unit leaf area), showing potential for stable coexistence of
-  multiple types.  Fitness is the log of the population growth rate --
-  at zero species are at equilibrium, above zero they will increase in
-  density.  (a) Landscape generated by a single species (indicated by
-  dots).  At \textsc{lma} of 0.1 (the dashed line, open dot) there is
-  directional selection for smaller \textsc{lma}.  At \textsc{lma} of
-  0.08 (solid line, filled dot) there is a branching point (see inset
-  for more detail).  (b) Landscape generated by two species, holding
-  the first at \textsc{lma} of 0.08.  Introducing a species at the
-  maximum fitness value from panel (a) reduces the region of positive
-  fitness considerably (dashed line, open dot).  Dotted lines show
-  subsequent invasions and replacements of the second species, the
-  solid line shows a stable fitness landscape where both species exist
-  on a peak in the fitness landscape.  At this point no further
-  invasion is possible.  See Appendix \ref{sec:fitness} for further
-  details and code.}
-\label{fig:fitness}
-\end{figure}
-
-\clearpage
-\setcounter{secnumdepth}{1}
-
-\begin{appendices}\label{sec:appendices}
-
-\section{Plant physiological model}\label{sec:FFW16}
-
-See attached file \url{vignettes/physiology.pdf}
-
-\section{Modelling demography of plants, patches and metapopulations}\label{sec:demography}
-
-See attached file \url{vignettes/demography.pdf}
-
-\section{Plant level properties}\label{sec:plant}
-
-\emph{Note}: due to Manuscript Central not accepting \textsc{html}
-files, we have had to convert this (and the following appendices) to
-\textsc{pdf}s.  \textsc{html} versions are available on the wiki:
-https://github.com/traitecoevo/plant\_paper/wiki --- the \textsc{html}
-versions have less pixelated graphics and are easier to copy and
-paste out of.
-
-See attached file \url{vignettes/plant.pdf}, 
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/plant.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/plant}
-
-\section{Cohort spacing algorithm}\label{sec:cohort-spacing}
-
-See attached file \url{vignettes/cohort_spacing.pdf},
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/cohort_spacing.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/cohort_spacing}
-
-\section{Finding demographic equilibrium}\label{sec:equilibrium}
-
-See attached file \url{vignettes/equilibrium.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/equilibrium.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/equilibrium}
-
-\section{Patch level dynamics}\label{sec:patch}
-
-See attached file \url{vignettes/patch.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/patch.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/patch}
-
-\section{Patch level emergent properties}\label{sec:emergent}
-
-See attached file \url{vignettes/emergent.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/emergent.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/emergent}
-
-\section{Calculating fitness}\label{sec:fitness}
-
-See attached file \url{vignettes/fitness.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/fitness.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/fitness}
-
-\section{Modifying parameters of the physiological model}\label{sec:parameters}
-See attached file \url{vignettes/parameters.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/parameters.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/parameters}
-
-\end{appendices}
+supported by a fellowship from the ARC. DSF thanks {\AA}
+Br{\"a}nnstr{\"o}m and U Dieckmann for introducing him to the \textsc{ebt}
+framework and the eco-evolutionary framework implemented in {\plant}, and
+for discussions around prior software. We also thank J Camac, A De Roos,
+R Duursma, J Johansson and C Prentice for helpful discussions. The authors 
+have no conflicts of interest to declare.
 \end{document}

### 2015-08-27 ms.tex
Minor comments in a few places on changes

@@
 
+% TODO[RGF]: Is this new acronym needed?  It's not really the place of
+% this contribution to define a new class of models, and given that
+% this all comes from a response to a reviewer, IMO it bends the paper
+% to be too much like a response to the reviewer.
 Broadly, the {\plant} model falls into a class we refer to Trait- Size- and
@@
 traits, size and light environment.  While {\plant}
-includes a core physiological model, we use it here only as an example to
+includes a core physiological model, we use it here primarily as an example to
 demonstrate the different ways in which any given physiological model
@@
 
+%% TODO[RGF]: I think that this needs softening.  Users can't really
+%% add new models; that has to be done by *developers*, really
+%% (because of the design the models are compiled into the package.
 While we have implemented one physiological model (\texttt{FFW16}), the
@@
 physiological model. As an example, in the \texttt{FFW16b} physiology, we have modified
-the function determining allocation to reproduction. The specifications
-for writing a new physiological model are outlined further in Appendix 
-XXX.
+the function determining allocation to reproduction.
 
@@
 
-When modelling a patch of competing, the focus is on changes to the size-
+When modelling a patch of competing plants, the focus is on changes to the size-
 density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density of
@@
 
+% TODO[RGF]: I think that this section should be reduced considerably,
+% ideally into a single paragraph.  The first three "paragraphs"
+% should be largely retained, shrinking the third.
+
 Increased interest in capturing the behaviour of size-structured  populations
@@
 manipulated and emergent demographic behaviours (including fitness) studied in
-detail. These different use-cases underpin important design differences among
+detail.
+% From here down can be largely dropped I think
+These different use-cases underpin important design differences among
 the models. Whereas the primary purpose of these other models is to scale a

### 2015-08-27 ms.tex
Replace deleted material, revert from a3b9a6d
Accidentally removed in a3b9a6d0f10d1b673c0d06679a9c7117169bd577

@@
 have no conflicts of interest to declare.
+
+\clearpage
+\bibliography{refs}
+
+\clearpage
+
+\section{Figures}\label{figures}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/schematic}
+
+\caption{Processes modelled within {\plant} include physiological, population and
+evolutionary dynamics.
+Dynamics across all three levels are driven by the
+physiological sub-model determining an individual's physiological and
+demographic rates on the basis of its traits, size and light environment (a).
+(b) Competitive hierarchies  are modelled by tracking the height distribution of
+individuals within a ``patch''. The intensity
+of shading indicates the density of individuals at a given height for
+different species (distinguished by colours). A size- and patch-structured
+metapopulation consists of a distribution of patches linked by seed dispersal.
+Disturbances occasionally remove all vegetation within a patch, resetting the
+patch. (c) The traits of the resident species determine the shading environment
+across the metapopulation, which in turn determines the fitness landscape.
+Resident traits adjust through directional selection
+and through the introduction of new species. Adapted from
+\citet{Falster-2011}, \citet{Falster-2015}. }
+
+\label{fig:schematic}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/plant.pdf}
+\caption{Physiological dynamics for individual plants varying in size,
+  trait and light environment. Solid lines are for the high
+  \textsc{lma} (leaf mass per unit area) strategy, dashed lines are
+  for the low \textsc{lma} strategy.  (a) Growth trajectories are
+  influenced by both light environment and traits.  (b) Over time, the
+  fraction of living tissue switches from leaf towards sapwood, with
+  high \textsc{lma} species having relatively more mass in leaf than
+  low \textsc{lma} species. (c) Size-dependent growth rates
+  (derivative of panel (a)) peak at around 5m high, but this position
+  varies with both trait and light level.  (d) Declines in growth rate
+  with light vary with plant size and traits.  Whole plant light
+  compensation points (indicated by dots) emerge as the point where
+  growth rate is zero (x intercept).  The light level used in panels
+  (a) and (c) is indicated with coloured ticks.  See Appendix
+  \ref{sec:plant} for further details and code.}
+\label{fig:plant}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/patch.pdf}
+\caption{Population dynamics for two species competing within a patch.
+  Red and blue lines are low and high \textsc{lma} species from
+  Fig. \ref{fig:plant}.  Lines in panel (a) show growth trajectories
+  for individuals on the boundary of successive ``cohorts'' for each
+  species, using a schedule of introduction times generated by
+  {\plant}'s adaptive algorithm.  The initial growth rate advantage of
+  the low \textsc{lma} species (see Figure \ref{fig:plant}c) means it
+  quickly over tops the high \textsc{lma} species, suppressing its
+  growth.  Self-thinning creates space for the high \textsc{lma}
+  species to establish below the canopy of the faster species. Panel
+  (b) shows the same trajectories as in panel (a), but with shading
+  indicating the density $N$ of individuals at given size and patch
+  age (light = low density, dark = high density).  (c) The leaf area
+  index at ground level for each species.  The dotted line is the
+  total leaf area index (summed across both species), and the lines on
+  the right side are average when integrated over the whole
+  metapopulation structure.  See Appendix \ref{sec:patch} for further
+  details and code.}
+\label{fig:patch}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/emergent.pdf}
+\caption{Emergent size-distributions and demography across a metapopulation.
+Grey lines show patterns within individual patches differing in time
+since disturbance (age), with blue lines indicating patches of a
+specific age. Red lines show average relationship across the metapopulation, 
+obtained by weighting the grey lines by patch and individual abundance. 
+Panels show (a) patterns of abundance, (b) leaf area and
+(c) height growth rate against size. Units of panels (a) and
+(b) are amount per unit height per unit ground area.  
+See Appendix \ref{sec:emergent} for further details and code.}
+\label{fig:emergent}
+\end{figure}
+
+\newpage
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
+\caption{Example fitness landscape for \textsc{lma} (leaf mass per
+  unit leaf area), showing potential for stable coexistence of
+  multiple types.  Fitness is the log of the population growth rate --
+  at zero species are at equilibrium, above zero they will increase in
+  density.  (a) Landscape generated by a single species (indicated by
+  dots).  At \textsc{lma} of 0.1 (the dashed line, open dot) there is
+  directional selection for smaller \textsc{lma}.  At \textsc{lma} of
+  0.08 (solid line, filled dot) there is a branching point (see inset
+  for more detail).  (b) Landscape generated by two species, holding
+  the first at \textsc{lma} of 0.08.  Introducing a species at the
+  maximum fitness value from panel (a) reduces the region of positive
+  fitness considerably (dashed line, open dot).  Dotted lines show
+  subsequent invasions and replacements of the second species, the
+  solid line shows a stable fitness landscape where both species exist
+  on a peak in the fitness landscape.  At this point no further
+  invasion is possible.  See Appendix \ref{sec:fitness} for further
+  details and code.}
+\label{fig:fitness}
+\end{figure}
+
+\clearpage
+\setcounter{secnumdepth}{1}
+
+\begin{appendices}\label{sec:appendices}
+
+\section{Plant physiological model}\label{sec:FFW16}
+
+See attached file \url{vignettes/physiology.pdf}
+
+\section{Modelling demography of plants, patches and metapopulations}\label{sec:demography}
+
+See attached file \url{vignettes/demography.pdf}
+
+\section{Plant level properties}\label{sec:plant}
+
+\emph{Note}: due to Manuscript Central not accepting \textsc{html}
+files, we have had to convert this (and the following appendices) to
+\textsc{pdf}s.  \textsc{html} versions are available on the wiki:
+https://github.com/traitecoevo/plant\_paper/wiki --- the \textsc{html}
+versions have less pixelated graphics and are easier to copy and
+paste out of.
+
+See attached file \url{vignettes/plant.pdf}, 
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/plant.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/plant}
+
+\section{Cohort spacing algorithm}\label{sec:cohort-spacing}
+
+See attached file \url{vignettes/cohort_spacing.pdf},
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/cohort_spacing.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/cohort_spacing}
+
+\section{Finding demographic equilibrium}\label{sec:equilibrium}
+
+See attached file \url{vignettes/equilibrium.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/equilibrium.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/equilibrium}
+
+\section{Patch level dynamics}\label{sec:patch}
+
+See attached file \url{vignettes/patch.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/patch.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/patch}
+
+\section{Patch level emergent properties}\label{sec:emergent}
+
+See attached file \url{vignettes/emergent.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/emergent.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/emergent}
+
+\section{Calculating fitness}\label{sec:fitness}
+
+See attached file \url{vignettes/fitness.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/fitness.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/fitness}
+
+\section{Modifying parameters of the physiological model}\label{sec:parameters}
+See attached file \url{vignettes/parameters.pdf}
+source:
+\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/parameters.R}
+html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/parameters}
+
+\end{appendices}
 \end{document}

### 2015-08-28 ms.tex
Work throughout to address reviewers comments
Included input from D Falster and M Westoby

@@
 \textasteriskcentered Email for correspondence: \texttt{daniel.falster@mq.edu.au}\\
-A manuscript in consideration as a research paper for
+A manuscript in consideration as an Applications Note for
 publication in MEE as part of the Special Feature \emph{Demography
@@
 \item
-% TODO: Rework to downgrade physiological model. Stochastic plants
-
   Here we introduce the {\plant} package, an
@@
   in patches of competing plants, or in metapopulations under a
-  disturbance regime. Dynamics within patches of competing plants are
-  resolved using novel extensions of the ``Escalator Boxcar Train''
+  disturbance regime. Dynamics within patches of competing plants can be
+  resolved either in a stochastic, finite-sized patch; or deterministically, using
+  novel extensions of the ``Escalator Boxcar Train''
   technique. Combined effects of trait-, size- and patch-structured
   dynamics are integrated into metapopulation level estimates of
-  reproductive fitness and vegetation structure.
+  invasion fitness and vegetation structure. Since fitness emerges as a
+  function of traits and is integrated over a full metapopulation, plant
+  provides a novel arena for exploring eco-evolutionary dynamics.
 \item
@@
 %
-In opposition to this, as individuals grow they must use increasing
-fractions of their photosynthetic income to building tissues other
-than new leaves; for example, to maintaining support tissues
+On the other hand as individuals grow larger they must allocate increasing
+fractions of their photosynthetic income to activities other
+than building new leaves; for example, to maintaining support tissues
 \citep{Givnish-1988, Enquist-2007} or to reproduction
@@
 primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
-  Tilman-1985, Geritz-1995, Hubbell-2001, Calcagno-2006}.
+  Tilman-1985, Geritz-1998, Hubbell-2001, Calcagno-2006}.
 %
@@
 Similarly, many of the models used to study global vegetation dynamics across
-the last 20 yrs have not included size-structured demography
+the last 20 years have not included size-structured demography
 \citep[for comparisons of some major models see][]{Sitch-2008,
@@
 \citep[e.g.][]{Moorcroft-2001, Purves-2008, Smith-2014,
-Weng-2015,Sakschewski-2015}, our understanding on how these features
+Weng-2015, Sakschewski-2015}, our understanding on how these features
 impact on the ecology and evolution of vegetation communities remains
@@
 %
-We developed {\plant} primarily to further investigate if and how species
-differing in traits are able to coexist with one another
+Our own purpose in developing {\plant} has been mainly to investigate how species
+differing in traits may be able to coexist with one another
 \citep[following ][]{Falster-2011, Falster-2015}; as size-structured models
@@
 successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
-At the same time, we expect {\plant} to be useful for
-modelling other demographic phenomena.
+At the same time, we expect {\plant} to be useful for modelling other demographic phenomena.
 
-% TODO[RGF]: Is this new acronym needed?  It's not really the place of
-% this contribution to define a new class of models, and given that
-% this all comes from a response to a reviewer, IMO it bends the paper
-% to be too much like a response to the reviewer.
-Broadly, the {\plant} model falls into a class we refer to Trait- Size- and
-Patch- structured Models (\textsc{TSPMs}). \textsc{TSPMs} are direct
-descendants of the ``gap'' models developed in 1980s
+Broadly, the {\plant} model falls into a class we refer to as Trait- Size- and
+Patch- structured Models \citep[\textsc{TSPMs}, following][]{Falster-2011}. \textsc{TSPMs} are direct
+descendants from the ``gap'' models developed in 1980s
 \citep[e.g.][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
@@
 \citep{Smith-2014} and \textsc{LPJml-FIT} \citep{Sakschewski-2015}. Common to
-\textsc{TSPMs} are that they explicitly model size-structured competition
+\textsc{TSPMs} is that they explicitly model size-structured competition
 among individual plants within a metapopulation of ``patches'', where
 individuals can also differ in their traits and thus demographic behaviour.
-\textsc {TSPMs} also typically ignore the spatial structure of plants within
-patches and the spatial structure of patches relative to one another; instead
-focussing computational effort on resolving the dynamics of size-structured
+\textsc {TSPMs} typically ignore the spatial structure of plants within
+patches and the spatial structure of patches relative to one another. Instead
+they focus computational effort on resolving the dynamics of size-structured
 competition for light and successional turnover. Differences among
-\textsc{TSPMs} arise from both the core physiological models used (describing
+\textsc{TSPMs} arise from the core physiological models used (describing
 how the demography of individual plants responds to resource availability),
-and the numerical techniques applied to scaling the physiological model up to
+and also from the numerical techniques applied to scaling the physiological model up to
 estimate emergent behaviours of individuals, patches and vegetation.
@@
 
-\section{The approach}
+\section{Overview of approach}
 
@@
 core rules in {\plant} are about the short-term physiological
-functioning of an individual plant and how these are influenced by its
-traits, size and light environment.  While {\plant}
-includes a core physiological model, we use it here primarily as an example to
-demonstrate the different ways in which any given physiological model
-could be analysed. To be clear, it is not the purpose
-of this paper to justify the particular physiological model used, to provide detailed
-comparisons to data, or necessarily promote this physiological model over
-those in other \textsc{TSPMs}. (These are all topics worthy of
-substantially more attention than can be afforded here.) Moreover, an important
-feature of {\plant} is that it allows for users to add their own
-physiological models.
+functioning of an individual plant and how this is influenced by its
+traits, size and light environment.  While {\plant} includes a core physiological model, the present paper does not aim to justify the particular physiological model used, to provide detailed
+comparisons to data, or to promote this physiological model over those in other \textsc{TSPMs}.
+(These are all topics deserving substantially more attention than can be afforded here.) An important feature of {\plant} considered as a software package is that it allows for users to substitute their own physiological models.
 
@@
 physiology, competition for light and disturbance (Fig.  \ref{fig:schematic}).
-We offer the ability to model the emergent behaviour of patches with
-specified size, as well as a deterministic size-independent approximation
-in which specifying a patch size is not required.
+{\plant} offers the capacity to model emergent behaviour of patches with
+specified size, and also a deterministic size-independent approximation
+where specifying a patch size is not required. Trait evolution and community
+assembly can then be modelled using the estimates of invasion fitness
+provided by {\plant}.
 
@@
 allowing for diameter growth also to be estimated. We refer to this
-default model as \texttt{FFW16}, reflecting the author initials and
+default model as \texttt{FFW16}, reflecting the author's initials and
 publication year of this paper. The \texttt{FFW16} model estimates rate
@@
 of tissue respiration and turnover are then subtracted. The remaining
-biomass is then allocated between growth and reproduction. The key
-outputs needed by higher levels of the {\plant} are height
+biomass production is then allocated between growth and reproduction. The key
+outputs needed by higher levels of {\plant} are height
 growth rate, mortality rate and rate of seed production.  However
@@
 and the respiration, turnover and allocation rates for different
-tissues and these intermediates can also be accessed (see Appendix
+tissues, and these intermediates can also be accessed (for further details on the \texttt{FFW16} model see Appendix
 \ref{sec:plant}).
 
-%% TODO[RGF]: I think that this needs softening.  Users can't really
-%% add new models; that has to be done by *developers*, really
-%% (because of the design the models are compiled into the package.
-While we have implemented one physiological model (\texttt{FFW16}), the
+While we have implemented a particular physiological model (\texttt{FFW16}), the
 package is designed to allow arbitrary additional physiologies to be
-introduced. Users are able to modify the physiological model in three
+introduced. Users are able to modify the physiological model in two
 different ways. First, one can vary the parameters of the \texttt{FFW16} model
@@
 \ref{sec:parameters}). Second, we allow for changes in key parameters to
-trigger changes in other parameters (i.e. a hyper-parametrisation), enabling
+bring about changes in other parameters (a hyper-parameterisation), enabling
 straightforward modelling of trade-offs. For example, the trait \textsc{lma}
@@
 equations themselves can be changed and the higher level functionality of the
-package can still be used.  The latter corresponds to defining a new
-physiological model. As an example, in the \texttt{FFW16b} physiology, we have modified
-the function determining allocation to reproduction.
+package can still be used.  Changed equations would amount to defining a new
+physiological model in a new C++ file. As an example, in
+the \texttt{FFW16b} physiology, we have modified the function determining
+allocation to reproduction.
 
-With the basic physiological model, {\plant} can be used to estimate
+With a basic physiological model in place, {\plant} can be used to estimate
 essential physiological rates for individual plants
@@
 (Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
-costs and the growth rate of plants (Fig. \ref{fig:plant}c).  This
-shift in allocation alone generates a distinctive hump-shaped pattern
-of absolute growth with size \citep{King-2011} and equally the widely
-observed decline in relative growth rate with size from birth onwards
-\citep{Enquist-2007}.
+costs and the growth rate of the plant (Fig. \ref{fig:plant}c).  This
+shift in allocation by itself can generate the widely-observed hump-shaped pattern
+of absolute height growth with size \citep{King-2011} and the decline in relative mass growth rate with size from birth onwards \citep{Enquist-2007}.
 
-By varying the light environment and measuring growth rate, the whole
-plant light compensation point (\textsc{wplcp}) can be computed (Fig.
+By varying the light environment and measuring growth rate, the whole-plant light compensation point (\textsc{wplcp}) can be computed (Fig.
 \ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant
@@
 maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
-decreases with \textsc{lma}, because high \textsc{lma} species have lower leaf turnover
+decreases with \textsc{lma}, because high \textsc{lma} species have slower leaf turnover
 \citep{Baltzer-2007, Lusk-2013} (Fig.
@@
 \citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
-While ideally, one would also consider spatial interactions within patches,
+While ideally one would also consider spatial interactions within patches,
 such models are very computationally demanding and also
@@
 therefore started with an empty patch and constant flow of seeds from a global
-seed rain (Fig. \ref{fig:schematic}b).
+seed rain (Fig. \ref{fig:schematic}b). A non-trivial extension could be added to allow some plants to survive disturbance \citep[following][]{Kohyama-1993}.
 
@@
 H, a)\) in a competing population. In the first ``stochastic'' mode, users
-specify an average seed rain and patch size, {\plant} then generates a
-vector of seed arrival events and simulates the stochastic
+specify an average seed rain and patch size, and {\plant} then generates a
+vector of seed arrival events and simulates stochastic
 development of the resulting finite-sized population (see Appendices XXX for an
@@
 
-For most applications, however, a preferable alternative to modelling
-dynamics within a finite-sized stochastic patch, is to use {\plant}'s 
-``determinstic'' approximation. The ``determinstic'' mode assumes that
-patches are very large; at this point the population dynamics within the
-patch approach their deterministic limit and can therefore
+For most applications, however, rather than modelling
+dynamics within a finite-sized stochastic patch, it will be preferable to use {\plant}'s
+``deterministic'' approximation. The deterministic mode assumes that
+patches are very large, so that the population dynamics within the
+patch approach their deterministic limit and can
 be approximated via a Partial Differential Equation (\textsc{pde})
@@
 capture the average behaviour across a large number of small patches
-\citep{Moorcroft-2001}. Importantly, the ``determinstic'' mode is
-also much faster than the stochastic model and without noise (more
+\citep{Moorcroft-2001}. Importantly, the ``deterministic'' mode is
+much faster than the stochastic model and without noise (more
 details in Appendix \ref{sec:demography}).
@@
 based on the Escalator Boxcar Train (\textsc{ebt}) technique
-\citep{Deroos-1988, Deroos-1992, Deroos-1997}. The \textsc{ebt} solves
+\citep{Deroos-1988, Deroos-1992, Deroos-1997, Brannstrom-2013}. The \textsc{ebt} solves
 the \textsc{pde} describing development of \(N\) by approximating the
@@
 adequately resolved throughout development of the patch (see Appendix
-\ref{sec:demography} for details). This refinement causes an uneven
+\ref{sec:demography} for details). This refinement gives rise to uneven
 distribution of cohort introduction times across patch age (Fig.
@@
 
-Modelling of size-structured dynamics via solving of a deterministic
+Modelling of size-structured dynamics via solving a deterministic
 \textsc{pde} enables multiple demographic phenomena to be
@@
 multiple species with different traits (Fig. \ref{fig:schematic}b;
-Fig.  \ref{fig:patch}). Moreover, these features emerge from the
+Fig.  \ref{fig:patch}). Second, the self-thinning and
+successional replacement emerge from the
 combined effects of physiology and competition for resources, rather
-than being prescribed in the model. Second, \textsc{TSPMs} allows for
-the combined effects of physiology and succession on emergent
+than being prescribed in the model.
+As a consequence, it is possible to ask how alternative trait values would translate into demographic fitness (see further below).
+Third,  \textsc{TSPMs} allows for the combined effects of physiology and succession on emergent
 characteristics of the entire patch to be investigated
@@
 
-All vegetation is subject to a disturbance regime, such that patches
+All vegetation in {\plant} is subject to a disturbance regime, such that patches
 of the landscape are cleared at some rate, by fire, cyclone or
@@
 metapopulation. An ``input'' seed rain is required to simulate the
-demography of the metapopulation, this in turn produces an ``output''
-number of seeds. The equilibrium occurs when, for each species,
-a given input seed rain
-produces an equivalent output seed rain. Finding this equilibrium
+demography of the metapopulation, and this in turn produces an ``output''
+number of seeds. Equilibrium occurs when, for each species,
+a given input seed rain produces an equivalent output seed rain. Finding this equilibrium
 requires finding the root of a multi-dimensional function, subject to
-the constraint that this root is stable (e.g., the trivial equilibrium
+the constraint that this root is stable (the trivial equilibrium
 of zero seed rain always exists, but is often not stable; Appendix
-\ref{sec:equilibrium}).
+\ref{sec:equilibrium}).  Real seed rains are of course quite noisy,
+but this approach still allows for the mean tendencies to be estimated.
 
-An attractive feature of the metapopulation concept implemented here
+An attractive feature of the metapopulation concept implemented in \textsc{TSPMs}
 is that integrates both equilibrium and non-equilibrium approaches to
@@
 patch to metapopulation, we close the demographic feedback loop to
-create a self-regulating system. These densities
-arise as outputs from the model, based on the combined effects of the
-model's physiological rules, species traits, competitive interactions
+create a self-regulating system.
+The equilibrium seed rains and size-density distributions  arise for each species present as outputs from the model, based on the combined effects of the model's physiological rules, species traits, competitive interactions
 and the prevailing disturbance regime.
@@
 fitness in structured metapopulations is given by the
-logarithm of the basic reproduction ratio, \(R\)
-\citep{Gyllenberg-2001, Metz-2001}.
-The basic reproduction ratio is simply the total seed output of
-the new type, averaged across the metapopulation. So a new type can
+logarithm of the basic reproduction ratio, \(R\) \citep{Gyllenberg-2001, Metz-2001}.
+The basic reproduction ratio is simply the average number of seeds produced from one seed of a new type over its lifetime, averaged across the metapopulation.
+So a new type can
 invade when \(\log R > 0\).
@@
 dynamics and community assembly of forest communities, following
- \citep{Falster-2015}. This is best
-thought of as occurring on a fitness landscape, a surface with fitness
-plotted against one or more species traits
+ \citep{Falster-2015}. This is best thought of as occurring on a fitness landscape, a surface with fitness plotted against one or more species traits
 (Fig. \ref{fig:fitness}). The reproductive success of individuals
-possessing a particular trait value depends on both the traits and
-densities of other species in the current resident community. Fitness
-landscapes will be positive in regions that allow invasion, and zero
+depends both on their own traits and on densities of other occupants in the current resident community. Fitness landscapes will be positive in regions that allow invasion, and zero
 for species that are at equilibrium. The slope of the fitness
@@
 smaller trait values (see Appendix \ref{sec:fitness} for
-details). Fitness landscapes are constantly reshaped as species are
-introduced and evolve, reflecting the non-linear demographic feedbacks
+details)). Fitness landscapes are continually reshaped as species are
+introduced and their trait values evolve, reflecting the non-linear demographic feedbacks
 that occur under frequency-dependent evolution. Stable coexistence is
@@
 evolutionary branching point \citep{Geritz-1998}; at
-this point there is no directional selection (solid line in
+such a point there is no directional selection (solid line in
 Fig. \ref{fig:fitness}a) but invasion is possible both above and below
@@
 is immune to invasion from any other strategies and represents a
-prediction for the trait mixture that might might be favoured by
+prediction for the trait mixture that might be favoured by
 natural selection \citep{Geritz-1998}.
 
-Something hear about adaptive dynamics and fitness
+In general,  a variety of algorithms from the area of adaptive dynamics could be applied for modelling community assembly \citep[e.g.][]{Geritz-1998, Falster-2015}, using the estimate of invasion fitness provided by {\plant}.
 
-Identifying conditions promoting trait diversity is a valuable use-case
+Identifying conditions promoting trait diversity is a valuable role
 for {\plant}. The simple example here can be extended to multiple dimensions
@@
 Installation instructions are available at this webpage, binary
-releases will also be available there.  (Note that {\plant} does not
-currently compile on Windows, but this will change soon as R core
+releases will also be available there.  {\plant} does not
+currently compile on Windows, but this will change as soon as R core
 upgrade the Windows C++ compiler).
@@
 
-% TODO[RGF]: I think that this section should be reduced considerably,
-% ideally into a single paragraph.  The first three "paragraphs"
-% should be largely retained, shrinking the third.
-
 Increased interest in capturing the behaviour of size-structured  populations
 has prompted development of several tools, each with its own distinct
-strengths.
-
-The \textsc{EBT} software (see De Roos XXXX in this special feature) provides 
+strengths. The \textsc{EBT} software (see De Roos XXXX in this special feature) provides
 a flexible framework that has been used to study a range of marine organisms.
@@
 size-asymmetric feedbacks that occur in plant communities, as well as the
-extension to structured metapopulation.
-
-Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
+extension to structured metapopulation. Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
 \citep[ver 1 and 2][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
@@
 models were designed for simulation of bio-geochemical cycles and vegetation
-structure across broad geographic areas; as such their architectures are
-optimised for these use-cases. By contrast, {\plant} was  designed so that the
+structure across broad geographic areas, and their architectures are
+optimised for such use. By contrast, {\plant} was  designed so that the
 physiological rules driving the dynamics of the system could be easily
-manipulated and emergent demographic behaviours (including fitness) studied in
-detail.
-% From here down can be largely dropped I think
-These different use-cases underpin important design differences among
-the models. Whereas the primary purpose of these other models is to scale a
-given physiology up to predict  broad-scale changes in the terrestrial
-biosphere, {\plant} allows users to  interact with the model at multiple
-levels of ecological organisation and  study the dynamics in detail. With any
-given physiology, it is possible to  grow plants in isolation, in patches of
-competing plants and in disturbance- driven metapopulations. {\plant} also
-allows for greater focus on population dynamics, reporting not only vegetation
-properties but also individual level densities, growth rates and fitness.
-{\plant} thus aims to bridge the gap between existing large-scale coupled
-simulation models like \textsc{ED} and the extremely simple models that are
-still widely  used in theoretical ecology.
-
-As the only other mdoel with open source code, we also note some more specific 
-differences between \textsc{ED} and {\plant}.
-\begin{itemize}
-  \item More effort is devoted in {\plant} to resolving competitive
-  hierarchies within developing patches. To that end we use a finer-scale 
-  adaptive time-step; allow for vertical overlap in the canopies of different 
-  plants and expend more effort on resolving the temporal spacing of cohorts
-  within a single patch. These details are essential to retrieving stable
-  estimates of fitness but possibly make little different to patch-scale
-  outcomes like total biomass or leaf area cover.
-
-  \item  \textsc{ED} includes sub-models for handling water, nutrient and 
-  temperature effects, and is thus better at capturing water, carbon and 
-  nutrient fluxes.
-
-  \item In order to approximate a single archetypal patch and thus study 
-  metapopulations at equilibrium, {\plant} requires a function giving the 
-  average photosynthesis per year as a function of canopy depth. While we
-  derive this function from a more-detailed sub-model, it is currently not
-  possible to explicitly simulate intra-annual cycles in environmental
-  drivers.
-
-  \item The physiological model in  \textsc{ED} has undergone more rigorous
-  comparisons with data \citep[see][]{Medvigy-2009}.  Providing updated 
-  parameter sets to the \texttt{FFW16} model is trivial (see Appendix 
-  \ref{sec:parameters}); the challenge is to infer
-  suitable values by constraining outputs from {\plant} against diverse
-  data sources \citep[e.g.][]{Medvigy-2009, Lebauer-2012}.
-
-  \item {\plant} is distributed as an R package an as such offers greater
-  possibilities for interactive use. 
-\end{itemize}
-
-Finally, like {\plant} the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
+manipulated and emergent demographic behaviours (including properties of individual plants, patches , metapopulations of patches and evolutionary fitness) studied in
+detail. Finally, like {\plant} the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
 to scale from individual-level behaviours to emergent properties of
@@
 
-The {\plant} package makes it easier for researchers and forest
+The {\plant} package can make it easier for researchers and forest
 practitioners alike to investigate a variety of non-linear demographic
@@
 
-DS Falster was supported by an Australian Research Council (ARC)
-discovery grant (DP110102086). RG FitzJohn was supported by the
-Science and Industry Endowment Fund (RP04-174). M Westoby was
+Falster was supported by an Australian Research Council (ARC)
+discovery grant (DP110102086). FitzJohn was supported by the
+Science and Industry Endowment Fund (RP04-174). Westoby was
 supported by a fellowship from the ARC. DSF thanks {\AA}
@@
   high \textsc{lma} species having relatively more mass in leaf than
-  low \textsc{lma} species. (c) Size-dependent growth rates
-  (derivative of panel (a)) peak at around 5m high, but this position
+  low \textsc{lma} species, due to their longer leaf lifespan. (c) Size-dependent growth rates
+  (derivative of panel (a)) peak at around 5m high, but location of the peak
   varies with both trait and light level.  (d) Declines in growth rate
@@
   the low \textsc{lma} species (see Figure \ref{fig:plant}c) means it
-  quickly over tops the high \textsc{lma} species, suppressing its
+  quickly overtops the high \textsc{lma} species, suppressing its
   growth.  Self-thinning creates space for the high \textsc{lma}
@@
   total leaf area index (summed across both species), and the lines on
-  the right side are average when integrated over the whole
+  the right side are averages when integrated over the whole
   metapopulation structure.  See Appendix \ref{sec:patch} for further

### 2015-08-28 ms.tex
Reorganise links to appendices in main text

@@
 sections outlining potential applications at nested levels of ecological organisation.
-Detailed technical documents are provided as supplementary information
-(see Appendices); updated versions of these technical documents are
-available within the {\plant} package itself.
+More detailed technical documents are provided as supplementary information
+(see Appendices; updated versions of these technical documents are
+available within the {\plant} package itself). These outline \ref{sec:demography}) the system of
+equations being solved when modelling demography of plants, patches and metapopulations; \ref{sec:FFW16}) a description of the \texttt{FFW16}
+physiological model, and \ref{sec:code}) a compendium of worked code examples.
 
@@
 tissues, and these intermediates can also be accessed (for further details on the \texttt{FFW16} model see Appendix
-\ref{sec:plant}).
+\ref{sec:code}).
 
@@
 directly (see Table XXX for list, for examples see Appendix
-\ref{sec:parameters}). Second, we allow for changes in key parameters to
+\ref{sec:code}). Second, we allow for changes in key parameters to
 bring about changes in other parameters (a hyper-parameterisation), enabling
@@
 size-dependent growth rate (see Appendices \ref{sec:demography} and
-\ref{sec:plant} for more details). As the plant grows, allocation to
+\ref{sec:code} for more details). As the plant grows, allocation to
 different tissue types varies and the composition of the plant
@@
 substantially over time (Fig. \ref{fig:patch}a; Appendix
-\ref{sec:demography}, \ref{sec:cohort-spacing} and
-\ref{sec:patch}). This can become a problem later during patch
+\ref{sec:demography}, and \ref{sec:code}). This can become a problem later during patch
 development because much of the population can be located between
@@
 of zero seed rain always exists, but is often not stable; Appendix
-\ref{sec:equilibrium}).  Real seed rains are of course quite noisy,
+\ref{sec:code}).  Real seed rains are of course quite noisy,
 but this approach still allows for the mean tendencies to be estimated.
@@
 are for a frequency distribution of patch states across the landscape
-(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:emergent}). As such,
+(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:code}). As such,
 size- and patch-structured models provide unique opportunities for linking with
@@
 (dashed line in Fig. \ref{fig:fitness}a) indicates selection for
-smaller trait values (see Appendix \ref{sec:fitness} for
+smaller trait values (see Appendix \ref{sec:code} for
 details)). Fitness landscapes are continually reshaped as species are
@@
   (a) and (c) is indicated with coloured ticks.  See Appendix
-  \ref{sec:plant} for further details and code.}
+  \ref{sec:code} for further details and code.}
 \label{fig:plant}
@@
   the right side are averages when integrated over the whole
-  metapopulation structure.  See Appendix \ref{sec:patch} for further
+  metapopulation structure.  See Appendix \ref{sec:code} for further
   details and code.}
@@
 (b) are amount per unit height per unit ground area.
-See Appendix \ref{sec:emergent} for further details and code.}
+See Appendix \ref{sec:code} for further details and code.}
 \label{fig:emergent}
@@
   on a peak in the fitness landscape.  At this point no further
-  invasion is possible.  See Appendix \ref{sec:fitness} for further
+  invasion is possible.  See Appendix \ref{sec:code} for further
   details and code.}
@@
 
-\section{Plant physiological model}\label{sec:FFW16}
-
-See attached file \url{vignettes/physiology.pdf}
-
 \section{Modelling demography of plants, patches and metapopulations}\label{sec:demography}
 
-See attached file \url{vignettes/demography.pdf}
+Describes the system of equations being solved in {\plant} when modelling demography of plants, patches and metapopulations. See attached file \url{vignettes/demography.pdf}
 
-\section{Plant level properties}\label{sec:plant}
+\section{Description of the \texttt{FFW16} physiological model}\label{sec:FFW16}
 
-\emph{Note}: due to Manuscript Central not accepting \textsc{html}
-files, we have had to convert this (and the following appendices) to
-\textsc{pdf}s.  \textsc{html} versions are available on the wiki:
-https://github.com/traitecoevo/plant\_paper/wiki --- the \textsc{html}
-versions have less pixelated graphics and are easier to copy and
-paste out of.
+Describes the core physiological model embedded in plant (\texttt{FFW16}). See attached file \url{vignettes/physiology.pdf}
 
-See attached file \url{vignettes/plant.pdf},
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/plant.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/plant}
+\section{Code examples}\label{sec:code}
 
-\section{Cohort spacing algorithm}\label{sec:cohort-spacing}
-
-See attached file \url{vignettes/cohort_spacing.pdf},
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/cohort_spacing.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/cohort_spacing}
-
-\section{Finding demographic equilibrium}\label{sec:equilibrium}
-
-See attached file \url{vignettes/equilibrium.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/equilibrium.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/equilibrium}
-
-\section{Patch level dynamics}\label{sec:patch}
-
-See attached file \url{vignettes/patch.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/patch.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/patch}
-
-\section{Patch level emergent properties}\label{sec:emergent}
-
-See attached file \url{vignettes/emergent.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/emergent.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/emergent}
-
-\section{Calculating fitness}\label{sec:fitness}
-
-See attached file \url{vignettes/fitness.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/fitness.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/fitness}
-
-\section{Modifying parameters of the physiological model}\label{sec:parameters}
-See attached file \url{vignettes/parameters.pdf}
-source:
-\smurl{https://github.com/traitecoevo/plant\_paper/blob/master/vignettes/src/parameters.R}
-html: \smurl{https://github.com/traitecoevo/plant\_paper/wiki/parameters}
+A compendium of worked code examples showing how to intearct with {\plant} from R
+See attached file \url{vignettes/combined.pdf}.
 

### 2015-08-28 R/utils.R
Generate the combined (code) vignette via remake

@@ -107,3 +107,14 @@ combine_md <- function(..., output) {
   ret <- c(header, unlist(dat))
   writeLines(ret, output)
 }
+
+## I'm 99% sure we've done this before.
+md_to_pdf <- function(filename) {
+  output <- sub("\\.md", ".pdf", basename(filename))
+  template <- c("--template", "template.tex")
+  opts <- "--toc"
+  owd <- setwd(dirname(filename))
+  on.exit(setwd(owd))
+  call_system(Sys_which("pandoc"),
+              c(basename(filename), "-o", output, template, opts))
+}

### 2015-08-28 vignettes/src/patch.R

@@ -96,8 +96,8 @@ plot(data1$time, y, type="l", las=1,
 ## Leaf area index is the driver that controls the canopy openness
 ## (via the light extinction coefficient `p1$c_ext`, following
 ## exponential extinction).  This is not returned by
-## \code{run_ebt_collect} so instead we need to rebuild patches using
-## \code{ebt_patch}.
+## `run_ebt_collect` so instead we need to rebuild patches using
+## `ebt_patch`.
 patches1 <- lapply(seq_along(data1$time), ebt_patch, data1)
 
 ## Each element of the resulting list is a Patch object, the same as

### 2015-08-28 ms.tex
Add lma numbers to the figure caption

@@
 
+% TODO: Table XXX?
 While we have implemented a particular physiological model (\texttt{FFW16}), the
@@
 
+% TODO: Appendix XXX
 The {\plant} model offers two methods for modelling the dynamics of \(N(x,
@@
 has prompted development of several tools, each with its own distinct
-strengths. The \textsc{EBT} software (see De Roos XXXX in this special feature) provides
+strengths. The \textsc{EBT} software (see De Roos 2016 in this special feature) provides
 a flexible framework that has been used to study a range of marine organisms.
@@
   trait and light environment. Solid lines are for the high
-  \textsc{lma} (leaf mass per unit area) strategy, dashed lines are
-  for the low \textsc{lma} strategy.  (a) Growth trajectories are
-  influenced by both light environment and traits.  (b) Over time, the
-  fraction of living tissue switches from leaf towards sapwood, with
-  high \textsc{lma} species having relatively more mass in leaf than
-  low \textsc{lma} species, due to their longer leaf lifespan. (c) Size-dependent growth rates
-  (derivative of panel (a)) peak at around 5m high, but location of the peak
-  varies with both trait and light level.  (d) Declines in growth rate
-  with light vary with plant size and traits.  Whole plant light
-  compensation points (indicated by dots) emerge as the point where
-  growth rate is zero (x intercept).  The light level used in panels
-  (a) and (c) is indicated with coloured ticks.  See Appendix
-  \ref{sec:code} for further details and code.}
+  \textsc{lma} (leaf mass per unit area) strategy (\textsc{lma} =
+  0.267), dashed lines are for the low \textsc{lma} strategy
+  (\textsc{lma} = 0.08).  (a) Growth trajectories are influenced by
+  both light environment and traits.  (b) Over time, the fraction of
+  living tissue switches from leaf towards sapwood, with high
+  \textsc{lma} species having relatively more mass in leaf than low
+  \textsc{lma} species, due to their longer leaf lifespan. (c)
+  Size-dependent growth rates (derivative of panel (a)) peak at around
+  5m high, but location of the peak varies with both trait and light
+  level.  (d) Declines in growth rate with light vary with plant size
+  and traits.  Whole plant light compensation points (indicated by
+  dots) emerge as the point where growth rate is zero (x intercept).
+  The light level used in panels (a) and (c) is indicated with
+  coloured ticks.  See Appendix \ref{sec:code} for further details and
+  code.}
 \label{fig:plant}

### 2015-08-28 ms.tex
Minor changes through ms

@@
   At its core, {\plant} is an
-  individual-based mechanistic model where plant physiology and demography are mediated by
+  individual-based model where plant physiology and demography are mediated by
   traits. Individuals from multiple species can be grown in isolation,
@@
 In this note, we introduce the {\plant} package for R \citep{R-2015};
-a mechanistic framework for studying the effects of size structure and
+a framework for studying the effects of size structure and
 trait variation on the demography of individual plants, of patches of
-competing plants and of meta-populations structured by a prevailing
+competing plants and of metapopulations structured by a prevailing
 disturbance regime.
@@
 outputs needed by higher levels of {\plant} are height
-growth rate, mortality rate and rate of seed production.  However
-additional quantities are also computed, such as total assimilation
+growth rate, mortality rate and rate of seed production (Fig. \ref{fig:schematic}).  Additional quantities are also computed, such as total assimilation
 and the respiration, turnover and allocation rates for different
 tissues, and these intermediates can also be accessed (for further details on the \texttt{FFW16} model see Appendix
-\ref{sec:code}).
+\ref{sec:FFW16} and \ref{sec:code}).
 
 % TODO: Table XXX?
+% NOTE: I've moved FFW16b to FFdev for now so that it matches the implementation.
 While we have implemented a particular physiological model (\texttt{FFW16}), the
@@
 equations themselves can be changed and the higher level functionality of the
-package can still be used.  Changed equations would amount to defining a new
-physiological model in a new C++ file. As an example, in
-the \texttt{FFW16b} physiology, we have modified the function determining
+package can still be used.  Changed equations would require defining a new
+physiological model in a new C++ file, and compiling that into the package.
+As an example, in the \texttt{FFWdev} physiology, we have modified the function determining
 allocation to reproduction.
@@
 \ref{sec:code}).  Real seed rains are of course quite noisy,
-but this approach still allows for the mean tendencies to be estimated.
+but this approach still allows for the mean tendencies to be estimated (which is required for computing fitness; see below).
 
@@
 smaller trait values (see Appendix \ref{sec:code} for
-details)). Fitness landscapes are continually reshaped as species are
+details). Fitness landscapes are continually reshaped as species are
 introduced and their trait values evolve, reflecting the non-linear demographic feedbacks
@@
 
-In general,  a variety of algorithms from the area of adaptive dynamics could be applied for modelling community assembly \citep[e.g.][]{Geritz-1998, Falster-2015}, using the estimate of invasion fitness provided by {\plant}.
-
 Identifying conditions promoting trait diversity is a valuable role
@@
 unknown \citep{Adler-2013}.
+% NOTE: is this OK as a merge (was an orphan sentence in previous version).
+A variety of algorithms from the area of adaptive dynamics could be applied for modelling community assembly \citep[e.g.][]{Geritz-1998, Falster-2015}, using the estimate of invasion fitness provided by {\plant}.
 
@@
 answer a number of questions at a variety of scales. The specific
-use-cases highlighted here arise from our interest in the assembly of
+applications highlighted here arise from our interest in the assembly of
 trait mixtures in communities. However a variety of other uses would

### 2015-09-01 ms.tex
Many changes throughout araising from thorough read through
Main text

- [x] Lots of small fixes, from thorough read-through
- [x] Figure 1: indicate bottom left section is different (routines not in plant)
- [x] Figure 3: rename axes (individuals not cohorts)

Physiology

- [x] Lots of small fixes, from thorough read-through
- [x] expand introduction and be clearer about what is new
- [x] name all hyper pars
- [x] expand list of pars in table
- [x] details on leaf photosynthesis calculations

Demography

- [x] Lots of small fixes, from thorough read-through
- [x] expand introduction and be cl

@@
   technique. Combined effects of trait-, size- and patch-structured
-  dynamics are integrated into metapopulation level estimates of
+  dynamics are integrated into metapopulation-level estimates of
   invasion fitness and vegetation structure. Since fitness emerges as a
@@
 
-Broadly, the {\plant} model falls into a class we refer to as Trait- Size- and
+Broadly, the {\plant} package falls into a class we refer to as Trait- Size- and
 Patch- structured Models \citep[\textsc{TSPMs}, following][]{Falster-2011}. \textsc{TSPMs} are direct
@@
 include \textsc{ED} \citep{Moorcroft-2001}, \textsc{LPJ- GUESS}
-\citep{Smith-2014} and \textsc{LPJml-FIT} \citep{Sakschewski-2015}. Common to
+\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. Common to
 \textsc{TSPMs} is that they explicitly model size-structured competition
@@
 
-Below we describe the general approach of the package, then provide
+Below we describe the general approach of the {\plant} package, then provide
 sections outlining potential applications at nested levels of ecological organisation.
@@
 available within the {\plant} package itself). These outline \ref{sec:demography}) the system of
-equations being solved when modelling demography of plants, patches and metapopulations; \ref{sec:FFW16}) a description of the \texttt{FFW16}
-physiological model, and \ref{sec:code}) a compendium of worked code examples.
+equations being solved when modelling demography of plants, patches and metapopulations; \ref{sec:FFW16}) a description of the core
+physiological model used {\plant}, and \ref{sec:code}) a compendium of worked code examples showing how to interact with the {\plant} package from R.
 
@@
 
-{\plant} is an individual-based model, meaning the dynamics of the
+The {\plant} package implements an individual-based model, meaning the dynamics of the
 population arise from rules specifying how individuals grow and interact.
@@
 functioning of an individual plant and how this is influenced by its
-traits, size and light environment.  While {\plant} includes a core physiological model, the present paper does not aim to justify the particular physiological model used, to provide detailed
+traits, size and light environment.
+While {\plant} includes a core physiological model, the present paper uses it only as an example, to illustrate how one can take a given physiological model and infer simulate ecological and evolutionary outcomes.
+Thus, we do not aim to justify the particular physiological model used, to provide detailed
 comparisons to data, or to promote this physiological model over those in other \textsc{TSPMs}.
@@
 
-% TODO: Table XXX?
-% NOTE: I've moved FFW16b to FFdev for now so that it matches the implementation.
 While we have implemented a particular physiological model (\texttt{FFW16}), the
@@
 different ways. First, one can vary the parameters of the \texttt{FFW16} model
-directly (see Table XXX for list, for examples see Appendix
-\ref{sec:code}). Second, we allow for changes in key parameters to
+directly (see Appendix \ref{sec:FFW16} for list, or Appendix
+\ref{sec:code} for code examples). Second, we allow for changes in key parameters to
 bring about changes in other parameters (a hyper-parameterisation), enabling
@@
 linkages are defined within a user-supplied hyper-parameterisation function
-(see Appendix \ref{sec:FFW16} for details). Finally, rather than varying the
+(see Appendix \ref{sec:FFW16} for details, or Appendix
+\ref{sec:code} for code examples). Finally, rather than varying the
 parameters of the existing equations in the \texttt{FFW16} model, the
@@
 (Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
-costs and the growth rate of the plant (Fig. \ref{fig:plant}c).  This
+costs and the growth rate of the plant.  This
 shift in allocation by itself can generate the widely-observed hump-shaped pattern
-of absolute height growth with size \citep{King-2011} and the decline in relative mass growth rate with size from birth onwards \citep{Enquist-2007}.
+of absolute height growth with size \citep[Fig. \ref{fig:plant}c;][]{King-2011} and the decline in relative mass growth rate with size from birth onwards \citep{Enquist-2007}.
 
@@
 \citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
-increases with plant size, due to increased costs of building and
+increases with plant size (Fig. \ref{fig:plant}d), due to increased costs of building and
 maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
-decreases with \textsc{lma}, because high \textsc{lma} species have slower leaf turnover
-\citep{Baltzer-2007, Lusk-2013} (Fig.
-\ref{fig:plant}d).
+decreases with \textsc{lma} (Fig. \ref{fig:plant}d), because high \textsc{lma} species have slower leaf turnover \citep{Baltzer-2007, Lusk-2013}.
 
@@
 When modelling a patch of competing plants, the focus is on changes to the size-
-density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density of
+density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density \(N\) of
 individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
@@
 \citep{Moorcroft-2001}. Importantly, the ``deterministic'' mode is
-much faster than the stochastic model and without noise (more
-details in Appendix \ref{sec:demography}).
+much faster than the stochastic model and without the demographic noise
+that arises in small populations (more details in Appendix \ref{sec:demography}).
 
@@
 
-We introduce two extensions to the \textsc{ebt}: i) a new method for
+We implement two extensions to the \textsc{ebt}: i) a new method for
 estimating the size-density distribution and ii) a technique for
@@
 In {\plant} we use a different approach. Instead of tracking first and
-second moments of the size-distribution within cohorts, we track a
-point mass estimate of \(N\) along the trajectory for the boundary of
-each cohort. We found this approach preferable because it does not
-require us to maintain a separate set of equations for the smallest
+second moments of the size-distribution \(N\) within cohorts, we track a
+point estimate of \(N\) along the characteristic for the boundary of
+each cohort. We found this approach preferable to the approach of
+approximating the first and second moments of \(N\) for each cohort
+because it does not require us to maintain a separate set of equations for the smallest
 cohort.
@@
 As a consequence, it is possible to ask how alternative trait values would translate into demographic fitness (see further below).
-Third,  \textsc{TSPMs} allows for the combined effects of physiology and succession on emergent
+Third,  \textsc{TSPMs} allow for the combined effects of physiology and succession on emergent
 characteristics of the entire patch to be investigated
@@
 
-In {\plant} and following \citep{Kohyama-1993, Moorcroft-2001,
-  Falster-2011} we assume an infinite number of patches, all experiencing
+In {\plant}, and following an approach established in previous TSPMs
+\citep{Kohyama-1993, Moorcroft-2001, Falster-2011}),
+we assume an infinite number of patches, all experiencing
 the same disturbance regime and sharing a common seed dispersal pool
@@
 An attractive feature of the metapopulation concept implemented in \textsc{TSPMs}
-is that integrates both equilibrium and non-equilibrium approaches to
+is that it integrates both equilibrium and non-equilibrium approaches to
 modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001,
@@
 
-The dynamic equilibrium captured by the trait-, size- and
-patch-structured metapopulation model potentially resolves several
+This dynamic equilibrium of the \textsc{TSPM} potentially resolves several
 challenges faced by unstructured models. First, by scaling up from
@@
 So a new type can
-invade when \(\log R > 0\).
+invade when \(\log R > 0\). The calculation of \(R\) available in
+{\plant} follows equations outlined by \citet{Falster-2015} (
+see Appendix \ref{sec:demography} for details).
 
@@
 unknown \citep{Adler-2013}.
-% NOTE: is this OK as a merge (was an orphan sentence in previous version).
-A variety of algorithms from the area of adaptive dynamics could be applied for modelling community assembly \citep[e.g.][]{Geritz-1998, Falster-2015}, using the estimate of invasion fitness provided by {\plant}.
+A variety of algorithms from the area of adaptive dynamics could be applied for modelling community assembly \citep[e.g.][]{Geritz-1998, Falster-2015}(Fig.
+\ref{fig:schematic}), using the estimate of invasion fitness provided by {\plant}.
 
@@
 a flexible framework that has been used to study a range of marine organisms.
-Compared to the \textsc{EBT}, {\plant} is more focussed on plants, and
+Compared to the \textsc{EBT}, {\plant} is customised for plants, and
 to that end offers refinements needed to handle modelling of the strong
@@
 \citep[ver 1 and 2][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
-\citep{Smith-2014} and a\textsc{LPJml-FIT}  \citep{Sakschewski-2015}. These
+\citep{Smith-2014} and a\textsc{LPJmL-FIT}  \citep{Sakschewski-2015}. These
 models were designed for simulation of bio-geochemical cycles and vegetation
@@
 
-The {\plant} package can make it easier for researchers and forest
+The {\plant} package is designed to make it easier for researchers and forest
 practitioners alike to investigate a variety of non-linear demographic
@@
 Science and Industry Endowment Fund (RP04-174). Westoby was
-supported by a fellowship from the ARC. DSF thanks {\AA}
+supported by a fellowship from the ARC. Falster thanks {\AA}
 Br{\"a}nnstr{\"o}m and U Dieckmann for introducing him to the \textsc{ebt}
@@
 patch. (c) The traits of the resident species determine the shading environment
-across the metapopulation, which in turn determines the fitness landscape.
-Resident traits adjust through directional selection
-and through the introduction of new species. Adapted from
-\citet{Falster-2011}, \citet{Falster-2015}. }
+across the metapopulation, which in turn determines the invasion fitness.
+The dashed line indicates that users must supply their own algorithms for
+modelling trait evolution and community assembly, using the estimate of
+invasion fitness provided by {\plant}. Figure adapted from
+\citet{Falster-2011}, \citet{Falster-2015}.
+}
 
@@
   \textsc{lma} species having relatively more mass in leaf than low
-  \textsc{lma} species, due to their longer leaf lifespan. (c)
+  \textsc{lma} species. (c)
   Size-dependent growth rates (derivative of panel (a)) peak at around
@@
   The light level used in panels (a) and (c) is indicated with
-  coloured ticks.  See Appendix \ref{sec:code} for further details and
-  code.}
+  coloured ticks.  See Appendix \ref{sec:code} for code
+  reproducing this figure.}
 \label{fig:plant}
@@
   the right side are averages when integrated over the whole
-  metapopulation structure.  See Appendix \ref{sec:code} for further
-  details and code.}
+  metapopulation structure.  See Appendix \ref{sec:code} for code
+  reproducing this figure.}
 \label{fig:patch}
@@
 obtained by weighting the grey lines by patch and individual abundance.
-Panels show (a) patterns of abundance, (b) leaf area and
-(c) height growth rate against size. Units of panels (a) and
+Panels show distributions of (a) individual density, (b) leaf area density
+and (c) height growth rate against size. Units of panels (a) and
 (b) are amount per unit height per unit ground area.
-See Appendix \ref{sec:code} for further details and code.}
+See Appendix \ref{sec:code} for code reproducing this figure.}
 \label{fig:emergent}
@@
   unit leaf area), showing potential for stable coexistence of
-  multiple types.  Fitness is the log of the population growth rate --
+  multiple types \citep[adapted from][]{Falster-2015}.  Fitness is the log of the population growth rate --
   at zero species are at equilibrium, above zero they will increase in
@@
   on a peak in the fitness landscape.  At this point no further
-  invasion is possible.  See Appendix \ref{sec:code} for further
-  details and code.}
+  invasion is possible.  See Appendix \ref{sec:code} for code
+  reproducing this figure.}
 \label{fig:fitness}
@@
 
-A compendium of worked code examples showing how to intearct with {\plant} from R
-See attached file \url{vignettes/combined.pdf}.
+A compendium of worked code examples showing how to interact with {\plant} from R
+See attached file \url{vignettes/code.pdf}.
 

### 2015-09-01 R/figures_patch.R

@@ -42,7 +42,7 @@ figure_patch <- function(data) {
             mgp=c(2.3, 1, 0))
   on.exit(par(op))
   matplot(t, h1, type="l", col=make_transparent(cols[[1]], 0.5), lty=1,
-          las=1, xlab="Time (years)", ylab="Cohort height (m)", xaxt="n")
+          las=1, xlab="Time (years)", ylab="Plant height (m)", xaxt="n")
   matlines(t, h2, col=make_transparent(cols[[2]], 0.5), lty=1)
   label_panel(1)
   axis(1, labels=FALSE)
@@ -53,7 +53,7 @@ figure_patch <- function(data) {
   n <- length(t)
   x <- matrix(rep(t, ncol(h1)), nrow(h1))
   plot(NA, xlim=range(t), ylim=range(h1, na.rm=TRUE),
-       las=1, xlab="Time (years)", ylab="Cohort height (m)", xaxt="n")
+       las=1, xlab="Time (years)", ylab="Plant height (m)", xaxt="n")
   segments(x[-1, ], h2[-1, ], x[-n, ], h2[-n, ], col=col2[-n, ], lend="butt")
   segments(x[-1, ], h1[-1, ], x[-n, ], h1[-n, ], col=col1[-n, ], lend="butt")
   label_panel(2)

### 2015-09-01 R/tables.R

@@ -11,7 +11,7 @@ table_plant_parameters <- function(filename, dest){
   to_math <- grepl("[_^]", tab$Units)
   tab$Units[to_math] <-
     sprintf("$%s$",
-            gsub("(kg|m|yr)", "\\\\mathrm{\\1}", tab$Units[to_math]))
+            gsub("(mol|kg|m|yr|d)", "\\\\mathrm{\\1}", tab$Units[to_math]))
   tab$Units[to_math] <-
             gsub(" ", "\\,", tab$Units[to_math], fixed =TRUE)
 
@@ -26,7 +26,7 @@ table_plant_parameters <- function(filename, dest){
   err <- sapply(val, is.null)
   if (any(err)) {
     message("Missing parameters: ", paste(code[err], collapse=", "))
-  } 
+  }
 
   oo <- options(scipen=999)
   on.exit(options(oo))

### 2015-09-01 R/utils.R

@@ -102,7 +102,7 @@ combine_md <- function(..., output) {
 
   header <-
     c("---",
-      'title: "plant: A package for modelling forest trait ecology & evolution"',
+      'title: "Worked examples showing how to interact with plant from R"',
       "---")
   ret <- c(header, unlist(dat))
   writeLines(ret, output)

### 2015-09-01 vignettes/src/cohort_spacing.R

@@ -1,5 +1,5 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _cohort spacing algorithm_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _Cohort spacing algorithm_"
 ## ---
 
 ## This vignette shows some details of cohort splitting.  It's

### 2015-09-01 vignettes/src/emergent.R

@@ -1,9 +1,9 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _patch level emergent properties_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _Patch-level emergent properties_"
 ## ---
 
-## The aim here is to use the plant package to investigate dynamics 
-## within a patch of competing plants, focussing on emergent patch-level 
+## The aim here is to use the plant package to investigate dynamics
+## within a patch of competing plants, focussing on emergent patch-level
 ## properties, rather than properties of plants within the patch.
 
 library(plant)
@@ -114,9 +114,9 @@ text(height[1, i] + strwidth("x"), growth_rate[1, i],
      adj=c(0, 0))
 
 ## The above plots show relationships with patches of a given age
-## What about the average relationship across the entire metapopulation? 
-## To get that, we  average (integrate) over the distribution over patch 
-## (for formula see demography vignette). To achieve this we first need 
+## What about the average relationship across the entire metapopulation?
+## To get that, we  average (integrate) over the distribution over patch
+## (for formula see demography vignette). To achieve this we first need
 ## the patch-level relationships to a series of fixed heights
 hh <- seq_log_range(range(height, na.rm=TRUE), 500)
 
@@ -127,7 +127,7 @@ f <- function(height, density, hout) {
   clamp_domain(splinefun_loglog(height, density), r, 0)(hout)
 }
 
-## Now interpolate the height-density relationship in each patch to 
+## Now interpolate the height-density relationship in each patch to
 ## the series of  specified heights
 xx <- lapply(seq_along(data1$time),
              function(i) f(height[, i], density[, i], hh))

### 2015-09-01 vignettes/src/equilibrium.R

@@ -1,5 +1,5 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _finding demographic equilibrium_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _Finding demographic equilibrium_"
 ## ---
 
 library(plant)

### 2015-09-01 vignettes/src/fitness.R

@@ -1,5 +1,5 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _calculating fitness_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _Calculating fitness_"
 ## ---
 
 library(plant)

### 2015-09-01 vignettes/src/parameters.R

@@ -1,5 +1,5 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _modifying parameters of the physiological model_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _Modifying parameters_"
 ## ---
 
 library(plant)
@@ -72,7 +72,7 @@ names(p)
 ## and it comes pre-set with the hyper-parameterisation function:
 identical(p$hyperpar, FFW16_hyperpar)
 
-## `c_ext` is the light extinction coefficient,
+## `k_I` is the light extinction coefficient,
 ## `disturbance_mean_interval` is the mean disturbance interval.  The
 ## `strategy_default` member is a Strategy:
 class(p$strategy_default)
@@ -92,7 +92,7 @@ FFW16_hyperpar(trait_matrix(lma, "lma"), s)
 ss <- strategy_list(lma, p)
 length(ss)
 
-## We can then use standard r command to extract variable from this list
+## We can then use standard R commands to extract variable from this list
 sapply(ss, function(x) x$lma)
 sapply(ss, function(x) x$k_l)
 sapply(ss, function(x) x$a_p1)

### 2015-09-01 vignettes/src/patch.R

@@ -1,5 +1,5 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _patch level dynamics_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _Patch-level dynamics_"
 ## ---
 
 ## The aim here is to use plant to investigate dynamics of plants
@@ -94,7 +94,7 @@ plot(data1$time, y, type="l", las=1,
 ## that generates a canopy that they cannot survive under.
 
 ## Leaf area index is the driver that controls the canopy openness
-## (via the light extinction coefficient `p1$c_ext`, following
+## (via the light extinction coefficient `p1$k_I`, following
 ## exponential extinction).  This is not returned by
 ## `run_ebt_collect` so instead we need to rebuild patches using
 ## `ebt_patch`.

### 2015-09-01 vignettes/src/plant.R

@@ -1,8 +1,8 @@
 ## ---
-## title: "plant: A package for modelling forest trait ecology & evolution: _plant level properties_"
+## title: "plant: A package for modelling forest trait ecology & evolution: _Plant-level properties_"
 ## ---
 
-## This vignette illustrates the sort of analysis used to generate 
+## This vignette illustrates the sort of analysis used to generate
 ## the plant figure (Fig 2) in the manuscript, i.e. modelling the
 ## dynamics of individual plants. It also shows some of the features of working
 ## with plant components of the model.
@@ -11,7 +11,7 @@ library(plant)
 ## Plants are constructed with the `FFW16_Plant` function.  That
 ## function takes as its only argument a "strategy" object; the
 ## default is `FFW16_Strategy`, but alternative strategies can be
-## provided (see below). The "strategy" object contains all 
+## provided (see below). The "strategy" object contains all
 ## the physiological underpinning the dynamics of individual plants
 ## and entire metapopulations.
 pl <- FFW16_Plant()
@@ -66,7 +66,7 @@ pp <- FFW16_PlantPlus(pl$strategy)
 pp$compute_vars_phys(env)
 names(pp$internals)
 
-## Some of the internals require `compute_vars_internals` to be run 
+## Some of the internals require `compute_vars_internals` to be run
 ## (the zapsmall function rounds numbers close to zero to zero):
 pp$compute_vars_growth()
 zapsmall(unlist(pp$internals))
@@ -75,7 +75,7 @@ zapsmall(unlist(pp$internals))
 ## This PlantPlus object also includes heartwood area and mass as two
 ## more variables for the ODE system (this might move into Plant
 ## soon -- see [this issue](https://github.com/traitecoevo/plant/issues/139)).
-## Tracking of these variables is needed to estimate stem diameter growth 
+## Tracking of these variables is needed to estimate stem diameter growth
 pp$ode_names
 
 ## Plants are a type of *reference object*.  They are different to
@@ -128,7 +128,7 @@ pl2$height
 ## And at this height, here is the total seed production:
 pl2$fecundity
 
-## To make this type of operation easier, we provide the function 
+## To make this type of operation easier, we provide the function
 ## `grow_plant_to_time`
 
 res <- grow_plant_to_time(FFW16_PlantPlus(FFW16_Strategy()), tt, env)

### 2015-09-01 ms.tex
More minor changes

@@
   Although the importance of size-
-  structure for demography is widely recognised, many mechanistic models
+  structure for demography is widely recognised, many models
   either omit it entirely, or include only coarse approximations.
@@
   Here we introduce the {\plant} package, an
-  extensible framework for modelling size- and trait-based demography,
+  extensible framework for modelling size- and trait--based demography,
   ecology and evolution in simulated forests.
@@
   novel extensions of the ``Escalator Boxcar Train''
-  technique. Combined effects of trait-, size- and patch-structured
-  dynamics are integrated into metapopulation-level estimates of
-  invasion fitness and vegetation structure. Since fitness emerges as a
-  function of traits and is integrated over a full metapopulation, plant
+  technique. These dynamics can be integrated into metapopulation--level estimates of
+  invasion fitness and vegetation structure. Because fitness emerges as a
+  function of traits, plant
   provides a novel arena for exploring eco-evolutionary dynamics.
@@
 Thus, we do not aim to justify the particular physiological model used, to provide detailed
-comparisons to data, or to promote this physiological model over those in other \textsc{TSPMs}.
-(These are all topics deserving substantially more attention than can be afforded here.) An important feature of {\plant} considered as a software package is that it allows for users to substitute their own physiological models.
+comparisons to data, or to promote this physiological model over those
+in other \textsc{TSPMs}. (These are all topics deserving substantially
+more attention than can be afforded here.)
+% TODO: I'm a bit concerned that while we know that this is possible,
+% it's not that well developed and does require compiling a new
+% version of plant.  So it's not that _usefully_ extendable.  On the
+% other hand, the population of users who will do this is not very
+% large.  It might be possible to allow runtime linkage though if I
+% push a little more of the cpp code into headers and do some tweaking
+% in RcppR6...
+An important feature of {\plant} considered as a software package is that it allows for users to substitute their own physiological models.
 
@@
 
-While we have implemented a particular physiological model (\texttt{FFW16}), the
-package is designed to allow arbitrary additional physiologies to be
-introduced. Users are able to modify the physiological model in two
-different ways. First, one can vary the parameters of the \texttt{FFW16} model
-directly (see Appendix \ref{sec:FFW16} for list, or Appendix
-\ref{sec:code} for code examples). Second, we allow for changes in key parameters to
-bring about changes in other parameters (a hyper-parameterisation), enabling
-straightforward modelling of trade-offs. For example, the trait \textsc{lma}
-(leaf mass per unit leaf area) is used to estimate the rate of leaf turnover
-($k_l$), based on an observed scaling relationship spanning across diverse
-vegetation and plant functional types \citep{Wright-2004}:
+While we have implemented a particular physiological model
+(\texttt{FFW16}), the package is designed to allow arbitrary
+additional physiologies to be introduced. Users are able to modify the
+physiological model in two different ways. First, one can vary the
+parameters of the \texttt{FFW16} model directly (see Appendix
+\ref{sec:FFW16} for list, or Appendix \ref{sec:code} for code
+examples). Second, we allow for changes in key parameters to bring
+about changes in other parameters (a hyper-parameterisation), enabling
+straightforward modelling of trade-offs. For example, by default the
+trait \textsc{lma} (leaf mass per unit leaf area) is used to estimate
+the rate of leaf turnover ($k_l$), based on an observed scaling
+relationship spanning across diverse vegetation and plant functional
+types \citep{Wright-2004}:
 $$k_l = \beta_{kl1} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{kl2}},$$
@@
 (see Appendix \ref{sec:FFW16} for details, or Appendix
-\ref{sec:code} for code examples). Finally, rather than varying the
-parameters of the existing equations in the \texttt{FFW16} model, the
-equations themselves can be changed and the higher level functionality of the
-package can still be used.  Changed equations would require defining a new
-physiological model in a new C++ file, and compiling that into the package.
-As an example, in the \texttt{FFWdev} physiology, we have modified the function determining
-allocation to reproduction.
+\ref{sec:code} for code examples).
+%
+Finally, rather than varying the parameters of the existing equations
+in the \texttt{FFW16} model, the equations themselves can be
+fundamenally changed.  This requires defining a new physiological
+model in a new C++ file, and compiling that into the package.  As an
+example, in the \texttt{FFWdev} physiology, we have modified the
+function determining allocation to reproduction.
 
-With a basic physiological model in place, {\plant} can be used to estimate
+With a physiological model in place, {\plant} can be used to estimate
 essential physiological rates for individual plants
@@
 impossible to solve in a deterministic fashion \citep{Shugart-1980,
-Pacala-1996}. Moreover, it remains unclear whether resolving spatial details within a
-patch would provide further useful ecological detail, in part because the
-process of competitive thinning tends to break down spatial clusters
+Pacala-1996}. This assumption can be motivated from the observation
+that competitive thinning tends to break down spatial clusters
 \citep{Strigul-2008}.
@@
 
-% TODO: Appendix XXX
 The {\plant} model offers two methods for modelling the dynamics of \(N(x,
@@
 vector of seed arrival events and simulates stochastic
-development of the resulting finite-sized population (see Appendices XXX for an
-example).
+development of the resulting finite-sized population.
 
@@
 substantially over time (Fig. \ref{fig:patch}a; Appendix
-\ref{sec:demography}, and \ref{sec:code}). This can become a problem later during patch
+\ref{sec:demography} and \ref{sec:code}). This can become a problem later during patch
 development because much of the population can be located between
@@
 In {\plant}, and following an approach established in previous TSPMs
-\citep{Kohyama-1993, Moorcroft-2001, Falster-2011}),
+\citep{Kohyama-1993, Moorcroft-2001, Falster-2011},
 we assume an infinite number of patches, all experiencing
@@
 unknown \citep{Adler-2013}.
-A variety of algorithms from the area of adaptive dynamics could be applied for modelling community assembly \citep[e.g.][]{Geritz-1998, Falster-2015}(Fig.
-\ref{fig:schematic}), using the estimate of invasion fitness provided by {\plant}.
+A variety of algorithms from the area of adaptive dynamics could be
+applied for modelling community assembly \citep[e.g.][Fig. \ref{fig:schematic}]{Geritz-1998, Falster-2015}, using the estimate of invasion fitness provided by {\plant}.
 
@@
 
-{\plant} is written in C++ and R, using the package \texttt{Rcpp}
-\citep{Eddelbuettel-2011, Eddelbuettel-2013} to bridge between the two
-languages. Due to the computationally intensive nature of the model,
-the core physiological and demographic components are in C++ to
-maximise speed. We used templated types that allow the demographic and
-evolutionary components to be driven by alternative physiological
-models while retaining the same interface at higher levels. Despite
-mostly being written in C++, all components of the model, down to
-individual plants, can be extracted and interacted with dynamically in
-R, using the \texttt{RcppR6} package \citep{RcppR6}.
+{\plant} is written in C++ and R, using the packages \texttt{Rcpp}
+\citep{Eddelbuettel-2011, Eddelbuettel-2013} and \texttt{RcppR6}
+\citep{RcppR6} to bridge between the two languages. Due to the
+computationally intensive nature of the model, the core physiological
+and demographic components are in C++ to maximise speed. We used
+templated types that allow the demographic and evolutionary components
+to be driven by alternative physiological models while retaining the
+same interface at higher levels. Despite mostly being written in C++,
+all components of the model, down to individual plants, can be
+extracted and interacted with dynamically in R.
 
@@
 currently compile on Windows, but this will change as soon as R core
-upgrade the Windows C++ compiler).
+upgrade the Windows C++ compiler.
 

### 2015-09-01 ms.tex
Update release script and word count for resubmission

@@
   beyond the Population}.\\
-Word count: ~3900 words}
+Word count: ~4200 words}
 \date{}
@@
 traits, size and light environment.
-While {\plant} includes a core physiological model, the present paper uses it only as an example, to illustrate how one can take a given physiological model and infer simulate ecological and evolutionary outcomes.
+While {\plant} includes a core physiological model, the present paper uses it only as an example, to illustrate how one can take a given physiological model and simulate ecological and evolutionary outcomes.
 Thus, we do not aim to justify the particular physiological model used, to provide detailed

### 2015-09-01 release.sh

@@ -13,13 +13,6 @@ mkdir release
 cp ms.pdf release
 
 mkdir supporting_material
-(cd vignettes; pandoc combined.html -o ../supporting_material/combined.pdf --template template.tex --toc)
-
-
-for file in $(find vignettes -name '*.html' | sed -e 's|.*/||'); do
-    echo "Converting ${file}"
-    (cd vignettes; pandoc "$file" -o ../supporting_material/"${file%.html}.pdf" --template template.tex)
-done
 cp vignettes/*.pdf supporting_material
 zip -r release/supporting_material.zip supporting_material
 rm -rf supporting_material

### 2015-11-09 ms.tex
Incorporate feedback from Ulf and Ake
Comments provided via email, have implemented some but not all of suggested changes

Changes reviwed with Mark Westoby. He suggested also removing quotes on some
items.

@@
 \title{{\plant}: A package for modelling forest trait ecology \& evolution}
-\author{Daniel S. Falster\textdagger\textasteriskcentered, Richard G. FitzJohn\textdagger, and Mark Westoby}
+\author{Daniel S. Falster\textdagger\textasteriskcentered$^1$, Richard G. FitzJohn\textdagger$^1$, \\{\AA}ke Br{\"a}nnstr{\"o}m$^{2,3}$, Ulf Dieckmann$^3$, Mark Westoby$^1$}
 \affiliation{
-Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
+$^1$ Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia\\
+$^2$ Department of Mathematics and Mathematical Statistics, Ume{\aa} University, 901 87 Ume{\aa}, Sweden \\
+$^3$ Evolution and Ecology Program, International Institute for Applied Systems Analysis, Schlossplatz 1, A-2361 Laxenburg, Austria \\
 \textdagger These authors contributed equally.\\
@@
  proportionately more energy on building and maintaining woody stems.
-  Although the importance of size-
+ Although the importance of size
  structure for demography is widely recognised, many models
@@
  Here we introduce the {\plant} package, an
-  extensible framework for modelling size- and trait--based demography,
+ extensible framework for modelling size- and trait-structured demography,
  ecology and evolution in simulated forests.
@@
  individual-based model where plant physiology and demography are mediated by
-  traits. Individuals from multiple species can be grown in isolation,
+ traits. Individual plants from multiple species can be grown in isolation,
  in patches of competing plants, or in metapopulations under a
-  disturbance regime. Dynamics within patches of competing plants can be
-  resolved either in a stochastic, finite-sized patch; or deterministically, using
-  novel extensions of the ``Escalator Boxcar Train''
-  technique. These dynamics can be integrated into metapopulation--level estimates of
+ disturbance regime. Patches can be modelled either as a stochastic
+ realisation of a finite-sized population, or as their deterministic limit.
+ These dynamics can be integrated into metapopulation--level
+ estimates of
  invasion fitness and vegetation structure. Because fitness emerges as a
-  function of traits, plant
+ function of traits, {\plant}
  provides a novel arena for exploring eco-evolutionary dynamics.
@@
  \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
-  While accessed from R, the core routines in {\plant} are written in C++.
+ Accessed from R, the core routines in {\plant} are written in C++.
  The package provides for alternative physiologies and for
-  hyper-parameterisation to capture trade-offs among parameters. A
+ capturing trade-offs among parameters. A
  detailed test suite is provided to ensure correct behaviour of the code.
@@
 %
-As an individual increases its leaf area, it increases its potential
+As an individual plant increases its leaf area, it increases its potential
 to generate photosynthate.
 %
-On the other hand as individuals grow larger they must allocate increasing
+On the other hand, as individuals grow larger, they must allocate increasing
 fractions of their photosynthetic income to activities other
@@
 %
-Consequently rates of growth, mortality and reproduction change with
+Consequently, rates of growth, mortality and reproduction change with
 individual size \citep{Muller-2006, Ruger-2011, Thomas-2011}.
@@
 \citep{Wright-2010}, while seed size and height
-at maturation determine the start and end points of ontogenetic
+at maturation determine start and end points of ontogenetic
 trajectories \citep{Westoby-2002}.
@@
 
-Although the importance of size-structure for trait evolution, vegetation
-dynamics and diversity has been long been recognised \citep[e.g.,][]{Harper-1977,
+Although the importance of size structure for trait evolution, vegetation
+dynamics and diversity has long been recognised \citep[e.g.,][]{Harper-1977,
  Shugart-1980, Huston-1987}, current research in this area is
-dominated by models and theory that either omit size-structure
+dominated by models and theory that either omit size structure
 entirely, or only include coarse approximations.
@@
 species coexistence and ecological drift (neutrality) both rely
-primarily on unstructured population models \citep[e.g.][]{MacArthur-1967,
+primarily on unstructured population models \citep[e.g.,][]{MacArthur-1967,
  Tilman-1985, Geritz-1998, Hubbell-2001, Calcagno-2006}.
 %
-It is common to assume competitive outcomes are influenced by
-size-related traits such as adult and seed size, but detailed
-size-structure is rarely considered, at least in plant models
+It is common to assume competitive outcomes to be influenced by
+size-related traits such as adult size and seed size, but detailed
+size structure is rarely considered, at least in plant models
 \cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
@@
 the last 20 years have not included size-structured demography
-\citep[for comparisons of some major models see][]{Sitch-2008,
+\citep[for comparisons of some major models, see][]{Sitch-2008,
 Dekauwe-2014}.
@@
 While the importance of explicitly including size-structured dynamics is
-increasingly recognised in both areas of evolutionary ecology
-\citep[e.g.][]{Falster-2015} and vegetation dynamics
-\citep[e.g.][]{Moorcroft-2001, Purves-2008, Smith-2014,
-Weng-2015, Sakschewski-2015}, our understanding on how these features
-impact on the ecology and evolution of vegetation communities remains
+increasingly recognised in both evolutionary ecology
+\citep[e.g.,][]{Falster-2015} and vegetation dynamics
+\citep[e.g.,][]{Moorcroft-2001, Purves-2008, Smith-2014,
+Weng-2015, Sakschewski-2015}, our understanding of how these features
+impact the ecology and evolution of vegetation communities remains
 relatively limited.
@@
 trait variation on the demography of individual plants, of patches of
-competing plants and of metapopulations structured by a prevailing
+competing plants and of metapopulations structured by a
 disturbance regime.
@@
 successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
-At the same time, we expect {\plant} to be useful for modelling other demographic phenomena.
+At the same time, we expect {\plant} to be useful for modelling other demographic and evolutionary phenomena.
 
-Broadly, the {\plant} package falls into a class we refer to as Trait- Size- and
-Patch- structured Models \citep[\textsc{TSPMs}, following][]{Falster-2011}. \textsc{TSPMs} are direct
-descendants from the ``gap'' models developed in 1980s
-\citep[e.g.][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
+Broadly, the {\plant} package falls into a class we refer to as trait-, size- and
+patch-structured models \citep[\textsc{TSPMs}, following][]{Falster-2011}. \textsc{TSPMs} are direct
+descendants from the ``gap'' models developed in the 1980s
+\citep[e.g.,][]{Shugart-1980, Huston-1987, Kohyama-1993}; other modern examples
 include \textsc{ED} \citep{Moorcroft-2001}, \textsc{LPJ- GUESS}
@@
 among individual plants within a metapopulation of ``patches'', where
-individuals can also differ in their traits and thus demographic behaviour.
+individuals can differ in their traits and thus demographic behaviour.
 \textsc {TSPMs} typically ignore the spatial structure of plants within
-patches and the spatial structure of patches relative to one another. Instead
+patches and the spatial structure of patches relative to one another. Instead,
 they focus computational effort on resolving the dynamics of size-structured
@@
 how the demography of individual plants responds to resource availability),
-and also from the numerical techniques applied to scaling the physiological model up to
-estimate emergent behaviours of individuals, patches and vegetation.
+and also from the numerical techniques applied to scaling the physiological model
+up to estimate emergent behaviours of individuals, patches and vegetation.
 
@@
 (see Appendices; updated versions of these technical documents are
-available within the {\plant} package itself). These outline \ref{sec:demography}) the system of
-equations being solved when modelling demography of plants, patches and metapopulations; \ref{sec:FFW16}) a description of the core
-physiological model used {\plant}, and \ref{sec:code}) a compendium of worked code examples showing how to interact with the {\plant} package from R.
+available within the {\plant} package itself). These documents outline \ref{sec:demography}) the system of
+equations being solved when modelling the demography of plants, patches and metapopulations; \ref{sec:FF16}) a description of the core
+physiological model used in {\plant}, and \ref{sec:code}) a compendium of worked code examples showing how to interact with the {\plant} package from R.
 
@@
 
-The {\plant} package implements an individual-based model, meaning the dynamics of the
+The {\plant} package implements an individual-based model, meaning that the dynamics of the
 population arise from rules specifying how individuals grow and interact.
@@
 more attention than can be afforded here.)
-% TODO: I'm a bit concerned that while we know that this is possible,
-% it's not that well developed and does require compiling a new
-% version of plant.  So it's not that _usefully_ extendable.  On the
-% other hand, the population of users who will do this is not very
-% large.  It might be possible to allow runtime linkage though if I
-% push a little more of the cpp code into headers and do some tweaking
-% in RcppR6...
 An important feature of {\plant} considered as a software package is that it allows for users to substitute their own physiological models.
@@
 On top of the physiological model, {\plant} implements methods for
-population and adaptive dynamics (Fig.  \ref{fig:schematic}) following methods
+population dynamics and evolutionary dynamics (Fig. \ref{fig:schematic}), following methods
 described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
-phenomena can be studied at the three levels: individual plants, stands of
-competing plants and entire metapopulations. The dynamics at these higher
+phenomena can be studied at three levels: individual plants, stands of
+competing plants and entire metapopulations. The dynamics at higher
 levels of organisation arise as emergent properties, driven by growth
 physiology, competition for light and disturbance (Fig. \ref{fig:schematic}).
-{\plant} offers the capacity to model emergent behaviour of patches with
-specified size, and also a deterministic size-independent approximation
+{\plant} offers the capacity to model emergent phenomena in patches with
+specified size, and also through a deterministic approximation
 where specifying a patch size is not required. Trait evolution and community
@@
 
-\section{Effects of size, trait and environment on demography of individual plants}
+\section{Effects of size, trait and light environment on the demography of individual plants}
 
@@
 The default physiological model used in {\plant} largely reflects
-that presented in \citet{Falster-2011,Falster-2015}, but with extensions
-allowing for diameter growth also to be estimated. We refer to this
-default model as \texttt{FFW16}, reflecting the author's initials and
-publication year of this paper. The \texttt{FFW16} model estimates rate
+that presented in \citet{Falster-2011, Falster-2015}, but with extensions
+allowing for growth in plant diameters also to be estimated. We refer to this
+default model as \texttt{FF16}, reflecting the initials of the first two authors
+and publication year of this paper. The \texttt{FF16} model estimates rate
 of biomass
@@
 leaf area and the light distribution across the plant's canopy. Costs
-of tissue respiration and turnover are then subtracted. The remaining
-biomass production is then allocated between growth and reproduction. The key
+of tissue respiration and turnover are subtracted. The remaining
+biomass production is allocated between growth and reproduction. The key
 outputs needed by higher levels of {\plant} are height
-growth rate, mortality rate and rate of seed production (Fig. \ref{fig:schematic}).  Additional quantities are also computed, such as total assimilation
-and the respiration, turnover and allocation rates for different
-tissues, and these intermediates can also be accessed (for further details on the \texttt{FFW16} model see Appendix
-\ref{sec:FFW16} and \ref{sec:code}).
+growth rate, mortality rate and rate of seed production (Fig. \ref{fig:schematic}). Additional quantities are also computed, such as total assimilation, as well as
+respiration, turnover and allocation rates for different
+tissues; these intermediates can also be accessed (for further details on the \texttt{FF16} model, see Appendices
+\ref{sec:FF16} and \ref{sec:code}).
 
 While we have implemented a particular physiological model
-(\texttt{FFW16}), the package is designed to allow arbitrary
+(\texttt{FF16}), the package is designed to allow arbitrary
 additional physiologies to be introduced. Users are able to modify the
 physiological model in two different ways. First, one can vary the
-parameters of the \texttt{FFW16} model directly (see Appendix
-\ref{sec:FFW16} for list, or Appendix \ref{sec:code} for code
+parameters of the \texttt{FF16} model directly (see Appendix
+\ref{sec:FF16} for the list of parameters and Appendix \ref{sec:code} for code
 examples). Second, we allow for changes in key parameters to bring
-about changes in other parameters (a hyper-parameterisation), enabling
-straightforward modelling of trade-offs. For example, by default the
-trait \textsc{lma} (leaf mass per unit leaf area) is used to estimate
+about changes in other parameters (a hyper-parameterisation), enabling the
+straightforward modelling of trade-offs. For example, the
+trait \textsc{lma} (leaf mass per unit leaf area) is used by default to estimate
 the rate of leaf turnover ($k_l$), based on an observed scaling
 relationship spanning across diverse vegetation and plant functional
-types \citep{Wright-2004}:
-$$k_l = \beta_{kl1} \, \left(\frac{\phi}{\phi_0}\right)^{-\beta_{kl2}},$$
-where $k_l$ and $\phi$ (lma) are both parameters of the core physiological
+types \citep{Wright-2004},
+$$k_l = \beta_{kl1} \, \left(\phi / \phi_0\right)^{-\beta_{kl2}},$$
+where $k_l$ and $\phi$ (\textsc{lma}) are both parameters of the core physiological
 model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
 linkages are defined within a user-supplied hyper-parameterisation function
-(see Appendix \ref{sec:FFW16} for details, or Appendix
+(see Appendix \ref{sec:FF16} for details, or Appendix
 \ref{sec:code} for code examples).
@@
 Finally, rather than varying the parameters of the existing equations
-in the \texttt{FFW16} model, the equations themselves can be
-fundamenally changed.  This requires defining a new physiological
-model in a new C++ file, and compiling that into the package.  As an
-example, in the \texttt{FFWdev} physiology, we have modified the
+in the \texttt{FF16} model, the equations themselves can be
+fundamentally changed. This requires defining a new physiological
+model in a new C++ file and compiling that into the package. As an
+example, in the \texttt{FFBDWdev} physiology, we have shown how to modify the
 function determining allocation to reproduction.
@@
 takes a given strategy and light environment and grows the plant,
-producing a trajectory of size over time (Fig. \ref{fig:plant}a). This
-is achieved by integrating an ordinary differential equation of
+producing a trajectory of plant size over time (Fig. \ref{fig:plant}a). This
+is achieved by integrating an ordinary differential equation with a
 size-dependent growth rate (see Appendices \ref{sec:demography} and
-\ref{sec:code} for more details). As the plant grows, allocation to
-different tissue types varies and the composition of the plant
+\ref{sec:code} for details). As the plant grows, allocation to
+different tissue types varies; specifically, the composition of the plant
 changes to include more stem and less leaf
-(Fig. \ref{fig:plant}b). This in turn affects maintenance and turnover
-costs and the growth rate of the plant.  This
-shift in allocation by itself can generate the widely-observed hump-shaped pattern
-of absolute height growth with size \citep[Fig. \ref{fig:plant}c;][]{King-2011} and the decline in relative mass growth rate with size from birth onwards \citep{Enquist-2007}.
+(Fig. \ref{fig:plant}b). This, in turn, affects maintenance and turnover
+costs and the growth rate of the plant. By itself, the
+shift in allocation can generate the widely observed hump-shaped dependence
+of absolute height growth on plant height \citep[Fig. \ref{fig:plant}c;][]{King-2011}, as well as the decline in relative mass growth rate with plant size from birth onwards \citep{Enquist-2007}.
 
 By varying the light environment and measuring growth rate, the whole-plant light compensation point (\textsc{wplcp}) can be computed (Fig.
-\ref{fig:plant}d). The \textsc{wplcp} is the light level at which the plant
-stops growing and is increasingly regarded as the most useful
+\ref{fig:plant}d). The \textsc{wplcp} is the light level where the plant
+stops growing, increasingly regarded as the most useful
 measure of a species' shade tolerance
-\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, \textsc{wplcp}
+\citep{Givnish-1988, Baltzer-2007, Lusk-2013}. As expected, the \textsc{wplcp}
 increases with plant size (Fig. \ref{fig:plant}d), due to increased costs of building and
-maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, \textsc{wplcp}
-decreases with \textsc{lma} (Fig. \ref{fig:plant}d), because high \textsc{lma} species have slower leaf turnover \citep{Baltzer-2007, Lusk-2013}.
+maintaining stem and leaf tissues \citep{Givnish-1988}. Likewise, the \textsc{wplcp}
+decreases with \textsc{lma} (Fig. \ref{fig:plant}d), because high-\textsc{lma} species have slower leaf turnover \citep{Baltzer-2007, Lusk-2013}.
 
@@
 Within patches of competing plants, competition for light generates
-strong non-linear feedbacks on growth, survival and reproduction. In the
-\texttt{FFW16} physiological model, we consider only the effect of
-shading on rates of biomass production, although competition for other
-resources such as nitrogen or water could also be considered by extending the
-physiological model. We assume that patches are vertically but not horizontally
-structured; in other words, we account for size differences but not spatial
-layout within the patch. Similar assumptions are found in other \textsc{TSPMs}
+strong nonlinear feedbacks on growth, survival and reproduction. In the
+\texttt{FF16} physiological model, we consider only the effects of
+shading on rates of biomass production. Competition for other
+resources such as nitrogen or water could be considered by extending the
+physiological model. We assume that patches are vertically -- but not horizontally --
+structured; in other words, we account for size differences, but not for the spatial
+layout within patches. Similar assumptions are made in other \textsc{TSPMs}
 \citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
-While ideally one would also consider spatial interactions within patches,
-such models are very computationally demanding and also
-impossible to solve in a deterministic fashion \citep{Shugart-1980,
-Pacala-1996}. This assumption can be motivated from the observation
+While one would ideally also consider spatial interactions within patches,
+such models are very computationally demanding \citep{Shugart-1980,
+Pacala-1996}. Simplifying spatial interactions within patches can be motivated from the observation
 that competitive thinning tends to break down spatial clusters
@@
 
-When modelling a patch of competing plants, the focus is on changes to the size-
-density-distribution \(N(x, H, a)\) with patch age \(a\), i.e. the density \(N\) of
+When modelling a patch of competing plants, the focus is on how the size-density-distribution \(N(x, H, a)\) changes with patch age \(a\), with the latter defined as the time that has passed since the last disturbance.
+This distribution describes the density \(N\) of
 individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
-Density is measured as individuals per unit height and per unit ground area.
-Modelling the development of  \(N(x, H, a)\) requires that both the initial size
-distribution and the inflow of new recruits be specified. We have primarily
-been interested in dynamics of a patch recovering from disturbance and have
-therefore started with an empty patch and constant flow of seeds from a global
-seed rain (Fig. \ref{fig:schematic}b). A non-trivial extension could be added to allow some plants to survive disturbance \citep[following][]{Kohyama-1993}.
+The density \(N\) is measured as individuals per unit height and per unit ground area.
+Modelling the dynamics of \(N(x, H, a)\) requires that both the initial size-density-distribution and the inflow of new recruits be specified. We have primarily
+been interested in the dynamics of a patch recovering from disturbance and have
+therefore started with an empty patch and a constant flow of seeds from a global
+seed rain (Fig. \ref{fig:schematic}b). A (non-trivial) extension would result from allowing some plants to survive disturbance \citep[following][]{Kohyama-1993}.
 
-The {\plant} model offers two methods for modelling the dynamics of \(N(x,
-H, a)\) in a competing population. In the first ``stochastic'' mode, users
+{\plant} offers two methods for modelling the dynamics of \(N(x,
+H, a)\) in a competing population. In the first stochastic mode, users
 specify an average seed rain and patch size, and {\plant} then generates a
-vector of seed arrival events and simulates stochastic
+vector of seed arrival events and simulates the stochastic
 development of the resulting finite-sized population.
@@
 dynamics within a finite-sized stochastic patch, it will be preferable to use {\plant}'s
-``deterministic'' approximation. The deterministic mode assumes that
-patches are very large, so that the population dynamics within the
+deterministic mode. This assumes that
+patches are sufficiently large that population dynamics within a
 patch approach their deterministic limit and can
-be approximated via a Partial Differential Equation (\textsc{pde})
-\citep{Deroos-1997, Kohyama-1993}.  The same \textsc{pde} is also thought to
+be approximated via a physiologically structured population model \citep{Metz-1986, Deroos-1997, Kohyama-1993}.
+These models are often formulated as partial differential equations where boundary conditions and coefficients may depend on the population state.
+Such structured population models are also thought to
 capture the average behaviour across a large number of small patches
-\citep{Moorcroft-2001}. Importantly, the ``deterministic'' mode is
-much faster than the stochastic model and without the demographic noise
-that arises in small populations (more details in Appendix \ref{sec:demography}).
+\citep{Moorcroft-2001}. Importantly, the deterministic mode is
+much faster than the stochastic mode, and eliminates the demographic noise
+that inevitably arises in finite-sized populations (see Appendix \ref{sec:demography} for details).
 
-Our approach for modelling deterministic size-structured population dynamics is
-based on the Escalator Boxcar Train (\textsc{ebt}) technique
-\citep{Deroos-1988, Deroos-1992, Deroos-1997, Brannstrom-2013}. The \textsc{ebt} solves
-the \textsc{pde} describing development of \(N\) by approximating the
-density function with a collection of cohorts spanning the size
-spectrum. Following a disturbance, a series of cohorts are introduced
-into each patch. These cohorts are then transported along the
-characteristic equations of the \textsc{pde}. Biologically these are
-the growth trajectories of individuals, conditioned on them surviving
-(e.g. Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
-to alter the density of individuals along the trajectories
-(Fig. \ref{fig:patch}b).
+Our approach for numerically solving deterministic size-structured population dynamics is
+based on the characteristic method \citep{Angulo-2004}. This method describes the development of \(N\) by approximating the
+density function along a collection of individual growth trajectories spanning the size
+spectrum. Following a disturbance, a series of such trajectories are introduced
+into each patch. These trajectories then change according to the growth
+of the corresponding individuals, conditioned on their survival
+(Fig. \ref{fig:patch}a). Meanwhile, growth and mortality combine
+to alter the density of individuals along these trajectories
+(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{ebt}) method
+\citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
+\citet{Falster-2015}, but not identical.
+An advantage of the characteristic method is that it allows direct approximation of the size-density-distribution.
+For the same reason, however, the characteristic method may be less suitable for resolving size-density-distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).
 
-We implement two extensions to the \textsc{ebt}: i) a new method for
-estimating the size-density distribution and ii) a technique for
-handling strong size-asymmetric competitive feedbacks, such as occur
+We implement a novel technique for
+handling strongly size-asymmetric competitive feedbacks, such as those occurring
 under strong competition for light (see Appendix \ref{sec:demography}
-for details). The original \textsc{ebt} proceeds by approximating the
-first and second moments of the density distribution \(N\) within each
-cohort \citep{Deroos-1997}.
-%
-In {\plant} we use a different approach. Instead of tracking first and
-second moments of the size-distribution \(N\) within cohorts, we track a
-point estimate of \(N\) along the characteristic for the boundary of
-each cohort. We found this approach preferable to the approach of
-approximating the first and second moments of \(N\) for each cohort
-because it does not require us to maintain a separate set of equations for the smallest
-cohort.
-
-The extension for handling strong size-asymmetric competition involves
-adaptive refining of time points at which new cohorts are introduced
+for details). This technique works for both the characteristic method and the Escalator Boxcar Train method. It involves an
+adaptive refinement of the time points where new growth trajectories or cohorts are introduced
 into the population, an idea first applied in \citet{Falster-2011} and
 described further in \citet{Falster-2015}.
-In the original \textsc{ebt}, cohorts were introduced at a fixed time
-interval. Under competition, however, the growth
-trajectories of individuals born at similar times can diverge
+Normally, growth trajectories are introduced evenly, spaced out according to a fixed time
+interval. Under size-asymmetric competition, however, the growth
+trajectories of individuals born at nearby times can diverge
 substantially over time (Fig. \ref{fig:patch}a; Appendix
-\ref{sec:demography} and \ref{sec:code}). This can become a problem later during patch
-development because much of the population can be located between
-widely-spaced cohorts, causing low numerical accuracy. To maintain
+\ref{sec:demography} and \ref{sec:code}). This can become a problem during patch
+development, because much of the population can be located in between
+widely-spaced trajectories, causing low numerical accuracy. To maintain
 numerical accuracy, we use an iterative algorithm to adaptively refine
-the cohort spacing until the trajectories of adjacent cohorts remain
-adequately resolved throughout development of the patch (see Appendix
-\ref{sec:demography} for details). This refinement gives rise to uneven
+the cohort spacing until adjacent trajectories remain
+adequately resolved throughout patch development (see Appendix
+\ref{sec:demography} for details). This refinement gives rise to an uneven
 distribution of cohort introduction times across patch age (Fig.
-\ref{fig:patch}a), with considerably tighter spacing of cohorts in
+\ref{fig:patch}a), with a considerably tighter spacing of trajectories in
 younger patches.
 
-Modelling of size-structured dynamics via solving a deterministic
-\textsc{pde} enables multiple demographic phenomena to be
+Modelling the full size-structured dynamics within patches enables multiple demographic phenomena to be
 investigated. First, {\plant} can simulate demographic
@@
 self-thinning are limited to a single species
-\citep[e.g.][]{Barnes-2004, Coomes-2007}, \textsc{TSPMs} can accommodate
+\citep[e.g.,][]{Barnes-2004, Coomes-2007}, \textsc{TSPMs} can accommodate
 multiple species with different traits (Fig. \ref{fig:schematic}b;
 Fig. \ref{fig:patch}). Second, the self-thinning and
-successional replacement emerge from the
+successional replacement emerges from the
 combined effects of physiology and competition for resources, rather
-than being prescribed in the model.
-As a consequence, it is possible to ask how alternative trait values would translate into demographic fitness (see further below).
-Third,  \textsc{TSPMs} allow for the combined effects of physiology and succession on emergent
-characteristics of the entire patch to be investigated
-\citep{Moorcroft-2001, Falster-2011}. Leaf area cover and rates of biomass
-production tend to vary  non-linearly with patch
+than being prescribed by the model.
+As a consequence, it is possible to ask how alternative trait values affect fitness (see below).
+Third, \textsc{TSPMs} allow the combined effects of physiology and succession on a patch's emergent
+properties to be investigated
+\citep{Moorcroft-2001, Falster-2011}. In particular, leaf area cover and rates of biomass
+production tend to vary nonlinearly with patch
 age \citep{Smith-2001, Binkley-2002, Ogawa-2010};
-in {\plant} such outcomes arise naturally as emergent properties,
+in {\plant}, such outcomes naturally arise as emergent properties,
 driven by demography (Fig. \ref{fig:patch}c).
 
-\section{Trait, size and patch structured vegetation}
+\section{Trait-, size- and patch-structured vegetation}
 
-All vegetation in {\plant} is subject to a disturbance regime, such that patches
-of the landscape are cleared at some rate, by fire, cyclone or
+All vegetation in {\plant} is subject to a disturbance regime, implying that patches
+are cleared at a given rate. Such clearance can be interpreted, for example
+as resulting from a fire, cyclone, clear-cutting or
 landslide. The vegetation thus comprises a series of patches
-differing in time since last disturbance and linked via seed dispersal
+differing in the time that has passed since their last disturbance, linked via seed dispersal
 (Fig. \ref{fig:schematic}b). Such a system is often referred to as a
@@
 
-In {\plant}, and following an approach established in previous TSPMs
+In {\plant}, following an approach established in previous \textsc{TSPMs}
 \citep{Kohyama-1993, Moorcroft-2001, Falster-2011},
 we assume an infinite number of patches, all experiencing
-the same disturbance regime and sharing a common seed dispersal pool
-(i.e., the island model; Fig. \ref{fig:schematic}b). We also assume
-that the disturbances remove all established vegetation.  With this
-approach we can scale up from deterministically solving the dynamics of
+the same disturbance regime and sharing a common seed dispersal pool, or global seed rain
+(as in the island model; Fig. \ref{fig:schematic}b). As mentioned before, we also assume
+that disturbances remove all vegetation. With this
+approach, we can scale up from deterministically solving the dynamics of
 a single patch to solving the dynamics of an entire metapopulation.
-The frequency-density \(P(a)\) of patches aged
-\(a\) in the landscape is needed. Assuming there are a large number of
-patches means the dynamics of \(P\) behave deterministically and can
-be approximated via a second \textsc{pde} (see Appendix
-\ref{sec:demography} for details). Scaling from patch to
+Under these assumptions, the frequency distribution \(P(a)\) of patches aged
+\(a\) changes deterministically according to a second partial differential equation (see Appendix
+\ref{sec:demography} for details). The scaling from patches to the
 metapopulation is then achieved by weighting the temporal dynamics of
-a single patch by \(P(a)\), the relative abundance of patches aged
+a single patch by \(P(a)\), i.e., by the relative abundance of patches aged
 \(a\) in the metapopulation (Fig. \ref{fig:patch}c).
@@
 metapopulation is to identify the equilibrium seed rain of the
-metapopulation. An ``input'' seed rain is required to simulate the
-demography of the metapopulation, and this in turn produces an ``output''
-number of seeds. Equilibrium occurs when, for each species,
-a given input seed rain produces an equivalent output seed rain. Finding this equilibrium
+metapopulation. An input seed rain is required to simulate the
+demography of the metapopulation, and this in turn produces an output
+seed rain. Equilibrium occurs when, for each species,
+the input seed rain equals the output seed rain. Identifying this equilibrium
 requires finding the root of a multi-dimensional function, subject to
-the constraint that this root is stable (the trivial equilibrium
+the constraint that a unique positive root exists and the root is dynamically stable (the trivial equilibrium
 of zero seed rain always exists, but is often not stable; Appendix
-\ref{sec:code}).  Real seed rains are of course quite noisy,
-but this approach still allows for the mean tendencies to be estimated (which is required for computing fitness; see below).
+\ref{sec:code}). Solving for multi-dimensional roots can be computationally challenging,
+with no single technique guaranteed to find a solution. In {\plant}, we sequentially
+applying rounds of iteration and nonlinear root finding algorithms, while also
+using some heuristics to try to ensure that the root returned is an attracting point
+(see Appendix \ref{sec:demography} for details). Real seed rains are of course noisy,
+but the aforementioned approach allows for mean tendencies to be estimated (which is required for computing fitness; see below).
 
 An attractive feature of the metapopulation concept implemented in \textsc{TSPMs}
-is that it integrates both equilibrium and non-equilibrium approaches to
+is that it integrates equilibrium and non-equilibrium approaches to
 modelling ecological dynamics \citep{Kohyama-1993, Moorcroft-2001,
-  Falster-2011}. An equilibrium may be approached at the level of
-metapopulation, meaning the seed rain and size structure across the
-entire metapopulation is approximately stable. Yet the structure of
-vegetation within individual patches is constantly in flux. Patches
-continue to age and their residents continue to grow until they are
-disturbed.
+ Falster-2011}. An equilibrium being approached at the level of
+the metapopulation implies that seed rain and size structure across the
+metapopulation are approximately stable. Yet, even under this condition, the structure of
+vegetation within individual patches remains constantly in flux: patches
+continue to age, and their residents continue to grow, until the next disturbance occurs.
 
-This dynamic equilibrium of the \textsc{TSPM} potentially resolves several
+Such dynamic equilibria arising in \textsc{TSPMs} potentially resolve several
 challenges faced by unstructured models. First, by scaling up from
-patch to metapopulation, we close the demographic feedback loop to
+patches to a metapopulation, we close the demographic feedback loop and
 create a self-regulating system.
-The equilibrium seed rains and size-density distributions  arise for each species present as outputs from the model, based on the combined effects of the model's physiological rules, species traits, competitive interactions
-and the prevailing disturbance regime.
+{\plant} yields as outputs the equilibrium seed rains and size-density-distributions for each considered species, based on the combined effects of the model's physiological rules, species traits, competitive interactions
+and disturbance regime.
 
@@
 regimes on vegetation structure and dynamics to be properly incorporated. In
-{\plant} characteristics for the entire vegetation (metapopulation) -- such
+{\plant}, characteristics for the entire vegetation (metapopulation) -- such
 as total leaf area, biomass and productivity -- are obtained by averaging
-patch-level outcomes over the frequency-distribution of patch ages (e.g.
+patch-level outcomes over the frequency distribution of patch ages (e.g.,
 Fig. \ref{fig:patch}c). Similarly, the distributions of abundance and growth rate
-observed in large forest  plots \citep[e.g.][]{Muller-2006, Coomes-2007} arise
-as  emergent phenomena. Moreover, the predictions from {\plant}
+observed in large forest plots \citep[e.g.,][]{Muller-2006, Coomes-2007} arise
+as emergent properties. Moreover, the predictions from {\plant}
 are for a frequency distribution of patch states across the landscape
-(Fig. \ref{fig:emergent}, see more detail in Appendix \ref{sec:code}). As such,
+(Fig. \ref{fig:emergent}, see Appendix \ref{sec:code} for details). As such,
 size- and patch-structured models provide unique opportunities for linking with
-satellite and forest survey data \citep{Moorcroft-2001, Purves-2008}.
+satellite data and forest survey data \citep{Moorcroft-2001, Purves-2008}.
 
 Finally, with the extension to the metapopulation level, we are
-modelling demography across the entire life cycle and are thus in a
-position to estimate invasion fitness -- the ability of a rare
-individual with particular traits to establish in a community of
+modelling demography across the entire plant life cycle and are thus in a
+position to estimate invasion fitness -- the rate at which a rare
+individual with particular traits can establish (positive invasion fitness) or cannot establish (negative invasion fitness) in a community of
 established residents at their equilibrium densities
-\citep{Metz-1992, Chesson-2000}.  While ideally calculated as the
+\citep{Metz-1992}. While often calculated as the
 long-term per capita growth rate, a more convenient indicator of
@@
 So a new type can
-invade when \(\log R > 0\). The calculation of \(R\) available in
-{\plant} follows equations outlined by \citet{Falster-2015} (
+invade when \(R > 1\). The calculation of \(R\) available in
+{\plant} follows equations given by \citet{Falster-2015} (
 see Appendix \ref{sec:demography} for details).
@@
 With this measure of fitness, we can model trait-based adaptive
-dynamics and community assembly of forest communities, following
- \citep{Falster-2015}. This is best thought of as occurring on a fitness landscape, a surface with fitness plotted against one or more species traits
+dynamics, and community assembly \citep{Metz-1992, Geritz-1998,
+Chesson-2000, Dieckmann-2004, Brannstrom-2013b}. An example
+for forests was introduced by
+\citet{Falster-2015}. This is best thought of as occurring on a fitness landscape,
+a surface with fitness plotted against one or more traits
 (Fig. \ref{fig:fitness}). The reproductive success of individuals
-depends both on their own traits and on densities of other occupants in the current resident community. Fitness landscapes will be positive in regions that allow invasion, and zero
-for species that are at equilibrium. The slope of the fitness
-landscape with respect to traits indicates the direction of selection
-(dashed line in Fig. \ref{fig:fitness}a) indicates selection for
+depends both on their own traits and on the densities of other occupants in the current resident community. Fitness landscapes are positive in trait regions that allow invasion, and zero
+for resident species that are at equilibrium. The slope of the fitness
+landscape with respect to traits determines the direction of selection
+(dashed line in Fig. \ref{fig:fitness}a), i.e., a negative slope indicates selection for
 smaller trait values (see Appendix \ref{sec:code} for
 details). Fitness landscapes are continually reshaped as species are
-introduced and their trait values evolve, reflecting the non-linear demographic feedbacks
+introduced and their trait values evolve, reflecting the nonlinear demographic feedbacks
 that occur under frequency-dependent evolution. Stable coexistence is
-possible when species with different traits can mutually invade
-\citep{Geritz-1998, Chesson-2000}. Following the fitness gradient may lead to an
-evolutionary branching point \citep{Geritz-1998}; at
-such a point there is no directional selection (solid line in
-Fig. \ref{fig:fitness}a) but invasion is possible both above and below
-it in trait space. It is often possible to find a configuration where
+possible when species with different traits can mutually invade.
+Following the fitness gradient may lead to an
+evolutionary branching point; at
+such a point, there is no directional selection (solid line in
+Fig. \ref{fig:fitness}a), but both smaller and larger trait values can invade.
+It is often possible to find a species configuration where
 fitness is negative everywhere (solid line in Fig. \ref{fig:fitness}b)
-except for the zero points where species are present. This community
-is immune to invasion from any other strategies and represents a
-prediction for the trait mixture that might be favoured by
-natural selection \citep{Geritz-1998}.
+except for the zero points where species are present. Such communities
+are immune to invasions by species with any other trait values and represent
+predictions for the trait mixtures favoured by
+natural selection .
 
 Identifying conditions promoting trait diversity is a valuable role
-for {\plant}. The simple example here can be extended to multiple dimensions
+for {\plant}. The simple example shown in the present paper can be extended to multiple dimensions
 and can be used to investigate how trade-offs facilitate coexistence
@@
 unknown \citep{Adler-2013}.
-A variety of algorithms from the area of adaptive dynamics could be
-applied for modelling community assembly \citep[e.g.][Fig. \ref{fig:schematic}]{Geritz-1998, Falster-2015}, using the estimate of invasion fitness provided by {\plant}.
+A variety of algorithms from the area of adaptive dynamics theory can be
+applied for modelling community assembly \citep[e.g.,][Fig. \ref{fig:schematic}]{
+Geritz-1998, Dieckmann-2004, Brannstrom-2013b, Falster-2015}, using the estimate of invasion fitness provided by {\plant}.
 
@@
 computationally intensive nature of the model, the core physiological
-and demographic components are in C++ to maximise speed. We used
+and demographic components are written in C++ to maximise speed. We used
 templated types that allow the demographic and evolutionary components
@@
 \citep{Eddelbuettel-2011, Eddelbuettel-2013}, \texttt{R6}
-\citep{Chang-2014}; and the Boost Library for C++
+\citep{Chang-2014}, and the Boost Library for C++
 \citep{Schaling-2014} via the R package \texttt{BH}
@@
 \href{https://github.com/traitecoevo/plant}{github.com/traitecoevo/plant}.
-Installation instructions are available at this webpage, binary
-releases will also be available there.  {\plant} does not
+Installation instructions are available at this webpage, and binary
+releases will also be available there. ({\plant} does not
 currently compile on Windows, but this will change as soon as R core
-upgrade the Windows C++ compiler.
+upgrade the Windows C++ compiler.)
 
-\section{Comparison to existing software}
+\section{Comparison with existing software}
 
-Increased interest in capturing the behaviour of size-structured  populations
+Increased interest in capturing the dynamics of size-structured populations
 has prompted development of several tools, each with its own distinct
@@
 Compared to the \textsc{EBT}, {\plant} is customised for plants, and
-to that end offers refinements needed to handle modelling of the strong
+to that end offers refinements needed to handle modelling of the strongly
 size-asymmetric feedbacks that occur in plant communities, as well as the
-extension to structured metapopulation. Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
-\citep[ver 1 and 2][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
-\citep{Smith-2014} and a\textsc{LPJmL-FIT}  \citep{Sakschewski-2015}. These
-models were designed for simulation of bio-geochemical cycles and vegetation
+extension to a structured metapopulation. Other \textsc{TSPMs} also exist for modelling of vegetation, like \textsc{ED}
+\citep[ver 1 and 2;][]{Moorcroft-2001, Medvigy-2009}, \textsc {LPJ-GUESS}
+\citep{Smith-2014} and \textsc{LPJmL-FIT} \citep{Sakschewski-2015}. These
+models were designed for simulating bio-geochemical cycles and vegetation
 structure across broad geographic areas, and their architectures are
-optimised for such use. By contrast, {\plant} was  designed so that the
+optimised for such use. By contrast, {\plant} was is designed so that the
 physiological rules driving the dynamics of the system could be easily
-manipulated and emergent demographic behaviours (including properties of individual plants, patches , metapopulations of patches and evolutionary fitness) studied in
-detail. Finally, like {\plant} the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
-to scale from individual-level behaviours to emergent properties of
+manipulated and emergent demographic properties (including those of individual plants, patches, metapopulations of patches and fitness) studied in
+detail. Finally, like {\plant}, the \textsc{LM3-PPA} \cite{Weng-2015} model also aims
+to scale from individual-level processes to emergent properties of
 vegetation, including estimating invasion fitness. Yet the approaches taken
 for solving the dynamics are very different. The \textsc{LM3-PPA} relies on
-complex analytical solutions to the size- structured interactions, whose
-derivation required simplifying elements of the physiology and ecology in the
-model. By contrast in {\plant} we have retained the full range of
+complex analytical solutions to the size-structured interactions, whose
+derivation required simplifying elements of the physiology and ecology in that
+model. By contrast, in {\plant} we have retained the full range of
 physiological and ecological feedbacks, and instead relied on using robust
 numerical methods to solve the system dynamics (see Appendix
-\ref{sec:demography} for further details).
+\ref{sec:demography} for details).
 
@@
 The {\plant} package is designed to make it easier for researchers and forest
-practitioners alike to investigate a variety of non-linear demographic
+practitioners alike to investigate a variety of nonlinear demographic, ecological and evolutionary
 phenomena. Rather than provide a model with a few entry points, we
@@
 applications highlighted here arise from our interest in the assembly of
-trait mixtures in communities. However a variety of other uses would
-be possible to address other aspects of community assembly. For
+trait mixtures in communities. A variety of other uses are
+possible, addressing other aspects of community assembly. For
 example, we expect the package will prove useful for i) investigating
 how morphological and physiological traits influence growth rate and
-shade tolerance, ii) emergent phenomena such as multi-species self-
-thinning, successional transitions and age-related changes in
-vegetation productivity, and iii) as a bridge between empirical data,
+shade tolerance, ii) studying emergent phenomena such as multi-species self-thinning, successional transitions and age-related changes in
+vegetation productivity, and iii) bridging between empirical data,
 large-scale simulators of global biogeochemical cycles and simple
@@
 
-Falster was supported by an Australian Research Council (ARC)
-discovery grant (DP110102086). FitzJohn was supported by the
-Science and Industry Endowment Fund (RP04-174). Westoby was
-supported by a fellowship from the ARC. Falster thanks {\AA}
-Br{\"a}nnstr{\"o}m and U Dieckmann for introducing him to the \textsc{ebt}
-framework and the eco-evolutionary framework implemented in {\plant}, and
-for discussions around prior software. We also thank J Camac, A De Roos,
-R Duursma, J Johansson and C Prentice for helpful discussions. The authors
-have no conflicts of interest to declare.
+We thank J Camac, A De Roos, R Duursma, J Johansson and C Prentice for helpful
+discussions and three anonymous reviewers for comments on an earlier draft. Falster
+was supported by an Australian Research Council (ARC) discovery grant (DP110102086).
+FitzJohn was supported by the Science and Industry Endowment Fund (RP04-174).
+Br{\"a}nnstr{\"o}m was supported by the Swedish Research Council Formas (2012-2008).
+Dieckmann was supported by the European Commission, the European Science Foundation,
+the Austrian Science Fund, the Austrian Ministry of Science and Research, and the
+Vienna Science and Technology Fund. Westoby was supported by a fellowship from the ARC.
+The authors have no conflicts of interest to declare.
 
@@
 \caption{Processes modelled within {\plant} include physiological, population and
-evolutionary dynamics.
-Dynamics across all three levels are driven by the
+adaptive dynamics.
+(a) Dynamics across all three levels are driven by the
 physiological sub-model determining an individual's physiological and
-demographic rates on the basis of its traits, size and light environment (a).
+demographic rates on the basis of its traits, size and light environment.
 (b) Competitive hierarchies are modelled by tracking the height distribution of
-individuals within a ``patch''. The intensity
+individuals within a patch. The intensity
 of shading indicates the density of individuals at a given height for
 different species (distinguished by colours). A size- and patch-structured
-metapopulation consists of a distribution of patches linked by seed dispersal.
-Disturbances occasionally remove all vegetation within a patch, resetting the
-patch. (c) The traits of the resident species determine the shading environment
-across the metapopulation, which in turn determines the invasion fitness.
-The dashed line indicates that users must supply their own algorithms for
+metapopulation consists of a distribution of patches linked by a global seed rain.
+Disturbances occasionally remove all vegetation within a patch, resetting its vegetation. (c) The traits of the resident species determine the light environment
+in all patches across the metapopulation, which in turn determines the invasion fitness.
+The dashed outline indicates that users must supply their own algorithms for
 modelling trait evolution and community assembly, using the estimate of
@@
 \caption{Physiological dynamics for individual plants varying in size,
-  trait and light environment. Solid lines are for the high
-  \textsc{lma} (leaf mass per unit area) strategy (\textsc{lma} =
-  0.267), dashed lines are for the low \textsc{lma} strategy
+ trait and light environment. Solid lines refer to the high-\textsc{lma} (leaf mass per unit leaf area) strategy (\textsc{lma} =
+ 0.267), while dashed lines refer to the low-\textsc{lma} strategy
  (\textsc{lma} = 0.08). (a) Growth trajectories are influenced by
  both light environment and traits. (b) Over time, the fraction of
-  living tissue switches from leaf towards sapwood, with high
-  \textsc{lma} species having relatively more mass in leaf than low
-  \textsc{lma} species. (c)
-  Size-dependent growth rates (derivative of panel (a)) peak at around
-  5m high, but location of the peak varies with both trait and light
-  level.  (d) Declines in growth rate with light vary with plant size
-  and traits.  Whole plant light compensation points (indicated by
-  dots) emerge as the point where growth rate is zero (x intercept).
-  The light level used in panels (a) and (c) is indicated with
+ living tissue switches from leaf towards sapwood, with high-\textsc{lma} species having relatively more mass in leaf than low-\textsc{lma} species. (c)
+ Size-dependent height growth rates, being the derivatives of the functions in (a), peak at a height of around
+ 5m, but the location of this peak varies with both trait and light
+ level. (d) Declines in height growth rate with light level vary with plant size
+ and traits. Whole-plant light compensation points (indicated by
+ circles) emerge at zero growth rate (intercepts on horizontal axis).
+ The light level used in (a) and (c) is indicated by
  coloured ticks. See Appendix \ref{sec:code} for code
@@
 \caption{Population dynamics for two species competing within a patch.
-  Red and blue lines are low and high \textsc{lma} species from
-  Fig. \ref{fig:plant}.  Lines in panel (a) show growth trajectories
-  for individuals on the boundary of successive ``cohorts'' for each
+ Red and blue lines refer to the low- and high-\textsc{lma} species from
+ Fig. \ref{fig:plant}. (a) Growth trajectories
+ for individuals germinating in successive ``cohorts'' for each
  species, using a schedule of introduction times generated by
  {\plant}'s adaptive algorithm. The initial growth rate advantage of
-  the low \textsc{lma} species (see Figure \ref{fig:plant}c) means it
-  quickly overtops the high \textsc{lma} species, suppressing its
-  growth.  Self-thinning creates space for the high \textsc{lma}
-  species to establish below the canopy of the faster species. Panel
-  (b) shows the same trajectories as in panel (a), but with shading
-  indicating the density $N$ of individuals at given size and patch
-  age (light = low density, dark = high density).  (c) The leaf area
+ the low-\textsc{lma} species (Figure \ref{fig:plant}c) means it
+ quickly overtops the high-\textsc{lma} species, suppressing its
+ growth. Self-thinning then creates space for the high-\textsc{lma}
+ species to establish below the canopy of the faster-growing species.
+ (b) Same trajectories as in (a), but shaded
+ according to the density $N$ of individuals at given size and patch
+ age (light = low density, dark = high density). (c) Leaf area
  index at ground level for each species. The dotted line is the
-  total leaf area index (summed across both species), and the lines on
-  the right side are averages when integrated over the whole
-  metapopulation structure.  See Appendix \ref{sec:code} for code
+ total leaf area index (summed across both species), while the lines on
+ the right-hand side are averages integrated over the whole
+ metapopulation. See Appendix \ref{sec:code} for code
  reproducing this figure.}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/emergent.pdf}
-\caption{Emergent size-distributions and demography across a metapopulation.
-Grey lines show patterns within individual patches differing in time
-since disturbance (age), with blue lines indicating patches of a
-specific age. Red lines show average relationship across the metapopulation,
-obtained by weighting the grey lines by patch and individual abundance.
-Panels show distributions of (a) individual density, (b) leaf area density
-and (c) height growth rate against size. Units of panels (a) and
-(b) are amount per unit height per unit ground area.
+\caption{Emergent size distributions and demography across a metapopulation.
+Grey lines show patterns within individual patches differing in patch age, i.e., in time
+since disturbance, with blue lines highlighting five selected
+patch ages. Red lines show the average relationship across the metapopulation,
+obtained by weighting the grey lines by patch frequency and individual abundance.
+(a) Density of individuals per unit height per unit ground area (b) Density of leaf area
+per unit height per unit ground area (c) Height growth rate.
 See Appendix \ref{sec:code} for code reproducing this figure.}
@@
 \includegraphics[width=15cm,height=15cm,keepaspectratio]{figures/fitness.pdf}
-\caption{Example fitness landscape for \textsc{lma} (leaf mass per
-  unit leaf area), showing potential for stable coexistence of
-  multiple types \citep[adapted from][]{Falster-2015}.  Fitness is the log of the population growth rate --
-  at zero species are at equilibrium, above zero they will increase in
-  density.  (a) Landscape generated by a single species (indicated by
-  dots).  At \textsc{lma} of 0.1 (the dashed line, open dot) there is
-  directional selection for smaller \textsc{lma}.  At \textsc{lma} of
-  0.08 (solid line, filled dot) there is a branching point (see inset
-  for more detail).  (b) Landscape generated by two species, holding
-  the first at \textsc{lma} of 0.08.  Introducing a species at the
-  maximum fitness value from panel (a) reduces the region of positive
-  fitness considerably (dashed line, open dot).  Dotted lines show
-  subsequent invasions and replacements of the second species, the
-  solid line shows a stable fitness landscape where both species exist
-  on a peak in the fitness landscape.  At this point no further
+\caption{Example of a fitness landscape for the \textsc{lma} (leaf mass per
+ unit leaf area) trait, showing potential for the stable coexistence of
+ multiple plant types \citep[adapted from][]{Falster-2015}. Invasion fitness is the logarithm of the population growth rate --
+ at zero fitness, species are at equilibrium, at positive fitness, they increase in
+ density. (a) Fitness landscape generated by a single species (indicated by
+ a circle). For an \textsc{lma} of 0.08 (solid line, filled circle), there is
+ directional selection for larger \textsc{lma}. For an \textsc{lma} of
+ 0.1 (dashed line, open circle), there is a branching point (see inset). (b) Fitness landscape generated by two species, holding
+ the first at an \textsc{lma} of 0.08. Introducing a species at the
+ maximum of the solid line in (a) reduces the region of positive
+ fitness considerably (dashed line, open circle). Dotted lines show
+ subsequent invasions and replacements of the second species. The
+ solid line shows the fitness landscape of the resultant evolutionarily stable two-species community, with both species coexisting
+ on separate peaks. At this point, no further
  invasion is possible. See Appendix \ref{sec:code} for code
@@
 
-\section{Modelling demography of plants, patches and metapopulations}\label{sec:demography}
+\section{Modelling the demography of plants, patches and metapopulations}\label{sec:demography}
 
-Describes the system of equations being solved in {\plant} when modelling demography of plants, patches and metapopulations. See attached file \url{vignettes/demography.pdf}
+Describes the system of equations being solved in {\plant} when modelling the demography of plants, patches and metapopulations. See attached file \url{vignettes/demography.pdf}
 
-\section{Description of the \texttt{FFW16} physiological model}\label{sec:FFW16}
+\section{Description of the \texttt{FF16} physiological model}\label{sec:FF16}
 
-Describes the core physiological model embedded in plant (\texttt{FFW16}). See attached file \url{vignettes/physiology.pdf}
+Describes the core physiological model embedded in plant (\texttt{FF16}). See attached file \url{vignettes/physiology.pdf}
 
@@
 
-A compendium of worked code examples showing how to interact with {\plant} from R
+A compendium of worked code examples showing how to interact with {\plant} from R.
 See attached file \url{vignettes/code.pdf}.
@@
 % LocalWords: Cramer Sitch Dekauwe focussed successional Haxeltine
-%  LocalWords:  Moorcroft lpj FFW lma wplcp Baltzer Lusk Kohyama pde
+% LocalWords: Moorcroft lpj FFBDW lma wplcp Baltzer Lusk Kohyama pde
 % LocalWords: Deroos Pacala Strigul ebt adaptively Coomes Binkley

### 2015-11-09 R/figures_emergent.R

@@ -91,7 +91,7 @@ figure_emergent <- function(data) {
   xlim <- c(0, max(height, na.rm=TRUE) * 1.05)
   matplot(height[, j], density[, j], type="l", lty=1, col="lightgrey",
           xlim=xlim, las=1, xaxt="n", log="y", yaxt="n",
-          xlab="Height (m)", ylab=expression("Density"~~("#"~m^-1~m^-2)))
+          xlab="Height (m)", ylab=expression("Density"~~(m^-1~m^-2)))
   axis(1, labels=FALSE)
   aty <- zapsmall(log10(axTicks(2)))
   labels <- sapply(aty,function(i) as.expression(bquote(10^ .(i))))

### 2015-11-09 R/figures_fitness.R

@@ -73,7 +73,7 @@ figure_fitness_landscape <- function(data) {
   plot(lma[i], w$second1[i], type="l", lty=2,
        log="x", xlim=xlim, ylim=ylim, ylab="Fitness", las=1)
   label_panel(2)
-  mtext(expression("Leaf mass per unit area"~("LMA;"~kg~m^-2)), 1, 3, cex=1)
+  mtext(expression("Leaf mass per unit leaf area"~("LMA;"~kg~m^-2)), 1, 3, cex=1)
   lines(lma[i], w$second2[i], lty=3)
   lines(lma[i], w$second3[i], lty=3)
   lines(lma, w$attractor2)

### 2015-11-09 R/figures_patch.R

@@ -42,7 +42,7 @@ figure_patch <- function(data) {
             mgp=c(2.3, 1, 0))
   on.exit(par(op))
   matplot(t, h1, type="l", col=make_transparent(cols[[1]], 0.5), lty=1,
-          las=1, xlab="Time (years)", ylab="Plant height (m)", xaxt="n")
+          las=1, xlab="Time (years)", ylab="Height (m)", xaxt="n")
   matlines(t, h2, col=make_transparent(cols[[2]], 0.5), lty=1)
   label_panel(1)
   axis(1, labels=FALSE)
@@ -53,7 +53,7 @@ figure_patch <- function(data) {
   n <- length(t)
   x <- matrix(rep(t, ncol(h1)), nrow(h1))
   plot(NA, xlim=range(t), ylim=range(h1, na.rm=TRUE),
-       las=1, xlab="Time (years)", ylab="Plant height (m)", xaxt="n")
+       las=1, xlab="Time (years)", ylab="Height (m)", xaxt="n")
   segments(x[-1, ], h2[-1, ], x[-n, ], h2[-n, ], col=col2[-n, ], lend="butt")
   segments(x[-1, ], h1[-1, ], x[-n, ], h1[-n, ], col=col1[-n, ], lend="butt")
   label_panel(2)
@@ -73,5 +73,5 @@ figure_patch <- function(data) {
   for(i in seq_len(3)) {
     axis(4, at=y_av[i], tck=0.1, col.ticks=cols[i], lty = lty[i])
   }
-  axis(1, at=108, labels="Av")
+  axis(1, at=108, labels="Average")
 }

### 2015-11-09 R/figures_plant.R

@@ -109,7 +109,7 @@ figure_plant <- function() {
   }
   alloc_h <- lapply(data, f)
   plot(NA, xlim=c(0, s1$hmat), ylim=c(0, 1), type="l",
-       ylab="Fractional of biomass", xlab="Height (m)", las=1)
+       ylab="Fraction of biomass", xlab="Height (m)", las=1)
   label_panel(2)
   matlines(alloc_h[["1_1"]][, 1],
            alloc_h[["1_1"]][, -1],

### 2015-11-09 vignettes/src/fitness.R

@@ -54,7 +54,7 @@ p1b <- add_eq(lma_b, p0)
 
 w1b <- fitness_landscape(lma, p1b)
 plot(lma, w1b, log="x", type="l", las=1,
-     xlab="Leaf mass per unit area", ylab="Fitness")
+     xlab="Leaf mass per unit leaf area", ylab="Fitness")
 abline(h=0, col="grey")
 points(lma_b, 0, pch=19)
 
@@ -64,7 +64,7 @@ points(lma_b, 0, pch=19)
 lma_detail <- trait_matrix(seq_log(lma_b * 0.95, lma_b * 1.05, 51), "lma")
 w1b_detail <- fitness_landscape(lma_detail, p1b)
 plot(lma_detail, w1b_detail, log="x", type="l", las=1,
-     xlab="Leaf mass per unit area", ylab="Fitness")
+     xlab="Leaf mass per unit leaf area", ylab="Fitness")
 abline(h=0, col="grey")
 points(lma_b, 0, pch=19)
 
@@ -77,7 +77,7 @@ lma_new <- lma[which.max(w1b)]
 lma_new
 
 plot(lma, w1b, log="x", type="l", las=1,
-     xlab="Leaf mass per unit area", ylab="Fitness")
+     xlab="Leaf mass per unit leaf area", ylab="Fitness")
 abline(h=0, col="grey")
 points(lma_b, 0, pch=19)
 abline(v=lma_new, col="red")
@@ -89,7 +89,7 @@ w2 <- fitness_landscape(lma, p2)
 ## down* around the second species, with a fitness gradient that
 ## points towards increased lma.
 plot(lma, w1b, log="x", type="l", las=1,
-     xlab="Leaf mass per unit area", ylab="Fitness", col="grey")
+     xlab="Leaf mass per unit leaf area", ylab="Fitness", col="grey")
 lines(lma, w2)
 abline(h=0, col="grey")
 points(lma_b, 0, pch=19)
@@ -106,7 +106,7 @@ p2_2 <- add_eq(lma_new2, p1b)
 w2_2 <- fitness_landscape(lma, p2_2)
 
 plot(lma, w1b, log="x", type="l", las=1,
-     xlab="Leaf mass per unit area", ylab="Fitness", col="grey")
+     xlab="Leaf mass per unit leaf area", ylab="Fitness", col="grey")
 lines(lma, w2, lty=2)
[…]

### 2015-11-09 vignettes/src/parameters.R

@@ -21,7 +21,7 @@ names(s)
 unlist(s[names(s) != "control"])
 
 ## All of these parameters can be directly changed.  For example, we
-## can double the leaf mass per unit area (lma) value:
+## can double the leaf mass per unit leaf area (lma) value:
 s$lma <- s$lma * 2
 
 ## and then from this construct a plant:

### 2015-11-09 R/figures_plant.R
Change strategy name from FFW16 to FF16
With expansion of author list we needed to find something
easier to type than first initials of all authors.

Suggesting letters of first two authors and year of paper.

Matches changes in traitecoevo/plant#0a14be82b5d580f962398e3fa48a1111dd907279

@@ -2,7 +2,7 @@
 default_parameters <- function() {
   p <- ebt_base_parameters()
   p$disturbance_mean_interval <- 30.0
-  p$strategy_default <- FFW16_Strategy(hmat = 30.0)
+  p$strategy_default <- FF16_Strategy(hmat = 30.0)
   p
 }
 
@@ -24,7 +24,7 @@ extract_plant_info <- function(p, env) {
 ## about plants.
 run_plant_to_heights <- function(heights, strategy, env,
                                  time_max=100) {
-  res <- grow_plant_to_size(FFW16_PlantPlus(strategy),
+  res <- grow_plant_to_size(FF16_PlantPlus(strategy),
                             heights, "height", env,
                             time_max, warn=FALSE, filter=TRUE)
   res$info <- do.call("rbind", lapply(res$plant, extract_plant_info, env))
@@ -33,7 +33,7 @@ run_plant_to_heights <- function(heights, strategy, env,
 
 ## Given vectors of heights and openness, lets do it all.
 height_dt_openness <- function(height, openness, strategy) {
-  pl <- FFW16_PlantPlus(strategy)
+  pl <- FF16_PlantPlus(strategy)
   pl$height <- height
   min <- lcp_whole_plant(pl)
   openness <- c(min, openness[openness > min])
@@ -60,8 +60,8 @@ figure_plant <- function() {
   env2 <- fixed_environment(0.5)
 
   ## Generate most of the data here:
-  heights1 <- seq(FFW16_PlantPlus(s1)$height, s1$hmat, length.out=100L)
-  heights2 <- seq(FFW16_PlantPlus(s2)$height, s2$hmat, length.out=100L)
+  heights1 <- seq(FF16_PlantPlus(s1)$height, s1$hmat, length.out=100L)
+  heights2 <- seq(FF16_PlantPlus(s2)$height, s2$hmat, length.out=100L)
   data <- list("1_1"=run_plant_to_heights(heights1, s1, env1),
                "2_1"=run_plant_to_heights(heights2, s2, env1),
                "1_2"=run_plant_to_heights(heights1, s1, env2),
@@ -69,8 +69,8 @@ figure_plant <- function() {
 
   ## And then the LCP data
   hmax <- s1$hmat / 2
-  h1 <- seq_log(FFW16_PlantPlus(s1)$height, hmax, 2)
[…]

### 2015-11-09 R/tables.R

@@ -15,8 +15,8 @@ table_plant_parameters <- function(filename, dest){
   tab$Units[to_math] <-
             gsub(" ", "\\,", tab$Units[to_math], fixed =TRUE)
 
-  s <- c(FFW16_Strategy(),
-         lapply(as.list(args(make_FFW16_hyperpar)), eval))
+  s <- c(FF16_Strategy(),
+         lapply(as.list(args(make_FF16_hyperpar)), eval))
   tab$Value <- NA_real_
 
   # Now extract actual values from plant

### 2015-11-09 vignettes/src/cohort_spacing.R

@@ -91,7 +91,7 @@ abline(h=0, col="grey")
 plot(tm, res - sr_1, xlab="Time (years)", ylab="Seed rain", las=1)
 abline(h=0, col="grey")
 
-ebt <- plant:::FFW16_EBT(p)
+ebt <- plant:::FF16_EBT(p)
 ebt$run()
 
 ## Now look at the contribution of different cohorts to see rain (x axis

### 2015-11-09 vignettes/src/parameters.R

@@ -7,7 +7,7 @@ library(plant)
 ## There are a large number of parameters to the physiological model,
 ## but all are changeable.  The default strategy is detailed in the
 ## "physiology" vignette:
-s <- FFW16_Strategy()
+s <- FF16_Strategy()
 names(s)
 
 ## The strategy object here is a special object of class `r class(s)`.
@@ -25,11 +25,11 @@ unlist(s[names(s) != "control"])
 s$lma <- s$lma * 2
 
 ## and then from this construct a plant:
-pl <- FFW16_Plant(s)
+pl <- FF16_Plant(s)
 pl$strategy$lma
 
 ## `lma` affects a few places in the model; see [the source
-## code](https://github.com/traitecoevo/plant/blob/master/src/ffw16_strategy.cpp),
+## code](https://github.com/traitecoevo/plant/blob/master/src/ff16_strategy.cpp),
 ## but only really for converting from leaf area to leaf mass (leaf
 ## mass being leaf area multiplied by leaf mass per leaf area).
 ## However, as a component of the leaf economic spectrum we imagine
@@ -39,12 +39,12 @@ pl$strategy$lma
 ## additional parameters and functions that mean that changing one
 ## parameter might affect a number of other lower-level parameters.
 ## Our default hyper-parameterisation is via the
-## `FFW16_hyperpar` function:
-FFW16_hyperpar
+## `FF16_hyperpar` function:
+FF16_hyperpar
 
 ## You will rarely need to call this function directly (see below) but
 ## note how setting of lma affects parameters `k_l`, `a_p1` and `r_l`
-FFW16_hyperpar(trait_matrix(0.1, "lma"), s)
+FF16_hyperpar(trait_matrix(0.1, "lma"), s)
 
 ## These are:
 ##   * `k_l`: Turnover rate for leaves
@@ -64,13 +64,13 @@ trait_matrix(c(0.1, 500), "rho")
 ## To make use of the hyper-parameterisation, the preferred way of
 ## setting parameters is through the utility functions `strategy` and
 ## `strategy_list`.  These take a `Parameters` object:
-p <- FFW16_Parameters()
+p <- FF16_Parameters()
 
 ## The Parameters object mostly contains information about the patch:
 names(p)
 
 ## and it comes pre-set with the hyper-parameterisation function:
-identical(p$hyperpar, FFW16_hyperpar)
[…]

### 2015-11-09 vignettes/src/plant.R

@@ -8,13 +8,13 @@
 ## with plant components of the model.
 library(plant)
 
-## Plants are constructed with the `FFW16_Plant` function.  That
+## Plants are constructed with the `FF16_Plant` function.  That
 ## function takes as its only argument a "strategy" object; the
-## default is `FFW16_Strategy`, but alternative strategies can be
+## default is `FF16_Strategy`, but alternative strategies can be
 ## provided (see below). The "strategy" object contains all
 ## the physiological underpinning the dynamics of individual plants
 ## and entire metapopulations.
-pl <- FFW16_Plant()
+pl <- FF16_Plant()
 
 ## Plants are an [R6](https://github.com/wch/R6) class, and have a
 ## number of elements and fields that can be accessed:
@@ -62,7 +62,7 @@ pl$internals$area_leaf
 ##
 ## Because of this, we have a "PlantPlus" object that exposes more of
 ## a strategy, and in addition stem diameter growth:
-pp <- FFW16_PlantPlus(pl$strategy)
+pp <- FF16_PlantPlus(pl$strategy)
 pp$compute_vars_phys(env)
 names(pp$internals)
 
@@ -103,7 +103,7 @@ derivs <- function(t, y, plant, env) {
   plant$compute_vars_phys(env)
   list(plant$ode_rates)
 }
-pl <- FFW16_Plant()
+pl <- FF16_Plant()
 tt <- seq(0, 50, length.out=101)
 y0 <- setNames(pl$ode_state, pl$ode_names)
 yy <- deSolve::lsoda(y0, tt, derivs, pl, env=env)
@@ -120,7 +120,7 @@ t <- spline(yy[, "height"], yy[, "time"], xout=h)$y
 ## Interpolate to find the state:
 y <- apply(yy[, -1], 2, function(y) spline(yy[, "time"], y, xout=t)$y)
 
-pl2 <- FFW16_Plant()
+pl2 <- FF16_Plant()
 pl2$ode_state <- y
 pl2$compute_vars_phys(env)
 ## Plant is expected height:
@@ -131,7 +131,7 @@ pl2$fecundity
 ## To make this type of operation easier, we provide the function
 ## `grow_plant_to_time`
 
-res <- grow_plant_to_time(FFW16_PlantPlus(FFW16_Strategy()), tt, env)
[…]

### 2015-11-09 ms.tex
Change strategy name from FFW16 to FF16
With expansion of author list we needed to find something
easier to type than first initials of all authors.

Suggesting letters of first two authors and year of paper.

Matches changes in traitecoevo/plant#0a14be82b5d580f962398e3fa48a1111dd907279
Add author contribution statement

@@
 model in a new C++ file and compiling that into the package. As an
-example, in the \texttt{FFBDWdev} physiology, we have shown how to modify the
+example, in the \texttt{FFdev} physiology, we have shown how to modify the
 function determining allocation to reproduction.
@@
 
+\section{Author contribution statement}
+
+RGF and DSF wrote and developed the software, including numerical algorithms.
+AB and UD helped develop algorithms included in prototype software.
+DSF, RGF and MW wrote the the paper, with input from AB and UD.
+All authors contributed to material presented in Supplementary material.
+
 \clearpage

### 2015-11-09 ms.tex
Incorporate suggestions from Ulf and Ake on SuppMatt

@@
 trait \textsc{lma} (leaf mass per unit leaf area) is used by default to estimate
-the rate of leaf turnover ($k_l$), based on an observed scaling
+the rate of leaf turnover ($k_{\rm l}$), based on an observed scaling
 relationship spanning across diverse vegetation and plant functional
 types \citep{Wright-2004},
-$$k_l = \beta_{kl1} \, \left(\phi / \phi_0\right)^{-\beta_{kl2}},$$
-where $k_l$ and $\phi$ (\textsc{lma}) are both parameters of the core physiological
+$$k_{\rm l} = \beta_{kl1} \, \left(\phi / \phi_0\right)^{-\beta_{kl2}},$$
+where $k_{\rm l}$ and $\phi$ (\textsc{lma}) are both parameters of the core physiological
 model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such

### 2015-11-09 R/figures_emergent.R
Rename EBT to SCM
This changed is motivated by the fact we are no longer using the EBT method, but rather "characteristic method".

Acronym stands for "Solver Chacteristic Method"

Corresponds to changes implemented in traitecoevo/plant/commit/b2ab54283b95e186c511720d71672ac2f0e57607

@@ -1,5 +1,5 @@
 make_emergent_data <- function(hires=TRUE) {
-  p0 <- ebt_base_parameters()
+  p0 <- scm_base_parameters()
   if(hires) {
     p0$control$schedule_eps <- 2e-6
     p0$control$ode_tol_rel <- 1e-6
@@ -8,7 +8,7 @@ make_emergent_data <- function(hires=TRUE) {
   p0$disturbance_mean_interval <- 30.0
   p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
   p1 <- build_schedule(p1)
-  data <- run_ebt_collect(p1)
+  data <- run_scm_collect(p1)
 
   height <- t(data$species[[1]]["height", , ])
   density <- exp(t(data$species[[1]]["log_density", , ]))
@@ -17,7 +17,7 @@ make_emergent_data <- function(hires=TRUE) {
 
   ## Then leaf area.  That requires a little more work because we don't
   ## have leaf areas for all the species.
-  tmp <- lapply(seq_along(data$time), ebt_patch, data)
+  tmp <- lapply(seq_along(data$time), scm_patch, data)
 
   xxx <- lapply(tmp, function(x) x$species[[1]]$area_leafs)
   data$leaf_area <- do.call("cbind", plant:::pad_matrix(xxx))

### 2015-11-09 R/figures_fitness.R

@@ -1,5 +1,5 @@
 make_fitness_data <- function() {
-  p0 <- ebt_base_parameters()
+  p0 <- scm_base_parameters()
   p0$control$equilibrium_nsteps <- 30
   p0$control$equilibrium_solver_name <- "hybrid"
   p0$disturbance_mean_interval <- 30.0

### 2015-11-09 R/figures_patch.R

@@ -1,5 +1,5 @@
 make_patch_data <- function() {
-  p0 <- ebt_base_parameters()
+  p0 <- scm_base_parameters()
   p0$control$equilibrium_nsteps <- 30
   p0$control$equilibrium_solver_name <- "hybrid"
   p0$disturbance_mean_interval <- 30.0
@@ -7,7 +7,7 @@ make_patch_data <- function() {
   ## in the fitness figure (see attractor2).
   p2 <- expand_parameters(trait_matrix(c(0.08, 0.267), "lma"), p0, FALSE)
   p2_eq <- equilibrium_seed_rain(p2)
-  run_ebt_collect(p2_eq, TRUE)
+  run_scm_collect(p2_eq, TRUE)
 }
 
 figure_patch <- function(data) {

### 2015-11-09 R/figures_plant.R

@@ -1,6 +1,6 @@
 ## First, growth trajectories of plants
 default_parameters <- function() {
-  p <- ebt_base_parameters()
+  p <- scm_base_parameters()
   p$disturbance_mean_interval <- 30.0
   p$strategy_default <- FF16_Strategy(hmat = 30.0)
   p
@@ -50,7 +50,7 @@ figure_plant <- function() {
   ## Everything will be done against two strategies (differing in LMA)
   ## and against two light environments (with the exception of the LCP
   ## plot).
-  p <- ebt_base_parameters()
+  p <- scm_base_parameters()
   p$disturbance_mean_interval <- 30.0
   ## NOTE: These lma values; 0.08 and 0.267 are the equilibrium values
   ## in the fitness figure (see attractor2).

### 2015-11-09 vignettes/src/cohort_spacing.R

@@ -4,14 +4,14 @@
 
 ## This vignette shows some details of cohort splitting.  It's
 ## probably not very interesting to most people, only those
-## interested in knowing how the EBT technique works in detail.  It also uses a lot
+## interested in knowing how the SCM technique works in detail.  It also uses a lot
 ## of non-exported, non-documented functions from plant so you'll see
 ## a lot of `plant:::` prefixes.
 
 library(plant)
 library(parallel)
 
-p0 <- ebt_base_parameters()
+p0 <- scm_base_parameters()
 p <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
 p$seed_rain <- 20 # close to equilibrium
 
@@ -38,12 +38,12 @@ interleave <- function(x) {
 t2 <- interleave(t1)
 t3 <- interleave(t2)
 
-## Consider running the EBT and computing seed rain at the end; this
+## Consider running the SCM and computing seed rain at the end; this
 ## is one of the key outputs from the model so a reasonable one to
 ## look for differences in.
 run_with_times <- function(p, t) {
   p$cohort_schedule_times[[1]] <- t
-  run_ebt(p)$seed_rains
+  run_scm(p)$seed_rains
 }
 
 sr_1 <- run_with_times(p, t1)
@@ -74,7 +74,7 @@ run_with_insert <- function(i, p, t) {
   run_with_times(p, insert_time(i, t))
 }
 
-## The internal function `plant:::run_ebt_error` runs the EBT and
+## The internal function `plant:::run_scm_error` runs the SCM and
 ## computes errors as the integration proceeds; this helps shed some
 ## light.
 i <- seq_len(length(t1) - 1)
@@ -91,7 +91,7 @@ abline(h=0, col="grey")
 plot(tm, res - sr_1, xlab="Time (years)", ylab="Seed rain", las=1)
 abline(h=0, col="grey")
 
-ebt <- plant:::FF16_EBT(p)
+ebt <- plant:::FF16_SCM(p)
 ebt$run()
 
 ## Now look at the contribution of different cohorts to see rain (x axis
@@ -109,10 +109,10 @@ abline(v=tm[j], col=c("red", rep("grey", 4)))
 
 ## Next up, need to work out what the fitness contribution of each
 ## cohort is.
-dat1 <- run_ebt_collect(p)
+dat1 <- run_scm_collect(p)
 p2 <- p
 p2$cohort_schedule_times[[1]] <- insert_time(j[[1]], t1)
[…]

### 2015-11-09 vignettes/src/emergent.R

@@ -8,7 +8,7 @@
 
 library(plant)
 
-p0 <- ebt_base_parameters()
+p0 <- scm_base_parameters()
 p0$control$equilibrium_nsteps <- 30
 p0$control$equilibrium_solver_name <- "hybrid"
 p0$disturbance_mean_interval <- 30.0
@@ -16,7 +16,7 @@ p0$disturbance_mean_interval <- 30.0
 ## We'll work with a single species at equilibrium
 p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
 p1_eq <- equilibrium_seed_rain(p1)
-data1 <- run_ebt_collect(p1_eq)
+data1 <- run_scm_collect(p1_eq)
 
 ## There was a bit of hassle in vignette:patch in plotting all
 ## trajectories along with a couple of lines corresponding to focal
@@ -70,7 +70,7 @@ text(height[1, i] + strwidth("x"), density[1, i],
 ## Early on there is a low density "dip" caused by self thinning (5
 ## years).  As the stand develops that dip broadens and deepens and
 ## moves forward in height (these very low density cohorts are still
-## travelling the characteristic equations of the EBT).  Later on (20
+## travelling the characteristic equations of the SCM).  Later on (20
 ## years) additional an second wave of recruitment gives a second
 ## pulse of high density (around 4 m tall), which can be seen
 ## travelling along in the 40 year profile to about 13 m.  When the
@@ -82,7 +82,7 @@ text(height[1, i] + strwidth("x"), density[1, i],
 ## from; a profile of leaf area with respect to height.  Again, this
 ## requires reconstructing the patches, and using an unexported
 ## function from `plant` to put this into a matrix:
-patch <- lapply(seq_along(data1$time), ebt_patch, data1)
+patch <- lapply(seq_along(data1$time), scm_patch, data1)
 leaf_area <- lapply(patch, function(x) x$species[[1]]$area_leafs)
 leaf_area <- do.call("cbind", plant:::pad_matrix(leaf_area))
 

### 2015-11-09 vignettes/src/equilibrium.R

@@ -7,7 +7,7 @@ library(parallel)
 
 run <- function(seed_rain_in, p) {
   p$seed_rain <- seed_rain_in
-  run_ebt(p)$seed_rains
+  run_scm(p)$seed_rains
 }
 
 run_new_schedule <- function(seed_rain_in, p) {
@@ -23,7 +23,7 @@ cobweb <- function(m, ...) {
 ## Try to establish what the equilibrium seed rain is.
 
 ## Set model parameters
-p0 <- ebt_base_parameters()
+p0 <- scm_base_parameters()
 p <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
 
 ## Some output seed rains, given an input seed rain:

### 2015-11-09 vignettes/src/fitness.R

@@ -6,7 +6,7 @@ library(plant)
 
 ## Start by setting a few parameters; this is the base set of
 ## parameters we'll use.
-p0 <- ebt_base_parameters()
+p0 <- scm_base_parameters()
 p0$control$equilibrium_nsteps <- 30
 p0$control$equilibrium_solver_name <- "hybrid"
 p0$disturbance_mean_interval <- 30.0

### 2015-11-09 vignettes/src/parameters.R

@@ -111,8 +111,8 @@ sapply(pp, function(p) p$area_leaf_above(0))
 p$control
 
 ## The defaults are rather too slow for many uses, so
-## `ebt_base_parameters` provides a faster set by using `fast_control`
+## `scm_base_parameters` provides a faster set by using `fast_control`
 ## to set many of these to less accurate values.
 
-p2 <- ebt_base_parameters()
+p2 <- scm_base_parameters()
 p2$control[unlist(p$control) != unlist(p2$control)]

### 2015-11-09 vignettes/src/patch.R

@@ -6,7 +6,7 @@
 ## within a single patch.
 library(plant)
 
-p0 <- ebt_base_parameters()
+p0 <- scm_base_parameters()
 p0$control$equilibrium_nsteps <- 30
 p0$control$equilibrium_solver_name <- "hybrid"
 p0$disturbance_mean_interval <- 30.0
@@ -19,7 +19,7 @@ p1_eq <- equilibrium_seed_rain(p1)
 
 ## This collects information about the state of the system at every
 ## ODE step:
-data1 <- run_ebt_collect(p1_eq)
+data1 <- run_scm_collect(p1_eq)
 
 ## Entries are:
 ##   * time: time of the step
@@ -96,9 +96,9 @@ plot(data1$time, y, type="l", las=1,
 ## Leaf area index is the driver that controls the canopy openness
 ## (via the light extinction coefficient `p1$k_I`, following
 ## exponential extinction).  This is not returned by
-## `run_ebt_collect` so instead we need to rebuild patches using
-## `ebt_patch`.
-patches1 <- lapply(seq_along(data1$time), ebt_patch, data1)
+## `run_scm_collect` so instead we need to rebuild patches using
+## `scm_patch`.
+patches1 <- lapply(seq_along(data1$time), scm_patch, data1)
 
 ## Each element of the resulting list is a Patch object, the same as
 ## was observed when running the model.
@@ -114,7 +114,7 @@ p2 <- expand_parameters(trait_matrix(0.267, "lma"), p1, FALSE)
 p2_eq <- equilibrium_seed_rain(p2)
 
 ## Then collect the patch-level dynamics:
-data2 <- run_ebt_collect(p2_eq)
+data2 <- run_scm_collect(p2_eq)
 
 t2 <- data2$time
 h1 <- data2$species[[1]]["height", , ]
@@ -166,7 +166,7 @@ segments(x[-1, ], h1[-1, ], x[-n, ], h1[-n, ], col=col1[-n, ], lend="butt")
 
 ## Each element of the resulting list is a Patch object, the same as
 ## was observed when running the model.
-patches2 <- lapply(seq_along(data2$time), ebt_patch, data2)
+patches2 <- lapply(seq_along(data2$time), scm_patch, data2)
 lai2 <- sapply(patches2, function(x) x$area_leaf_above(0.0))
 
 lai2_1 <- sapply(patches2, function(x) x$species[[1]]$area_leaf_above(0.0))

### 2015-11-09 vignettes/src/plant.R

@@ -179,7 +179,7 @@ plot(res_mass$time, mass, type="o", pch=19, las=1, xlab="Time (years)")
 ## two species that differ in their LMA values. This what is presented
 ## In Fig. 2a of the paper.
 
-p <- ebt_base_parameters()
+p <- scm_base_parameters()
 ## Low LMA ("fast growth") species
 s1 <- strategy(trait_matrix(0.08,  "lma"), p)
 ## High LMA ("low growth") species

### 2015-11-09 ms.tex
Rename EBT to SCM
This changed is motivated by the fact we are no longer using the EBT method, but rather "characteristic method".

Acronym stands for "Solver Chacteristic Method"

Corresponds to changes implemented in traitecoevo/plant/commit/b2ab54283b95e186c511720d71672ac2f0e57607
Resolve issue around naming and symbols for size distribution
Suggestion from Ulf. Previous approach was misleading, because N(x,H,a)
would imply density of plants per unit height per unit trait per unit age,
whereas as we were only recording density per unit height.

Instead use N(H|x,a)

@@
 
-When modelling a patch of competing plants, the focus is on how the size-density-distribution \(N(x, H, a)\) changes with patch age \(a\), with the latter defined as the time that has passed since the last disturbance.
+When modelling a patch of competing plants, the focus is on how the size distribution \(N(H | x, a)\) changes with patch age \(a\), with the latter defined as the time that has passed since the last disturbance.
 This distribution describes the density \(N\) of
-individuals with traits \(x\) and height \(H\) within a patch of age \(a\).
+individuals with height \(H\) for given traits \(x\) and patch age \(a\).
 The density \(N\) is measured as individuals per unit height and per unit ground area.
-Modelling the dynamics of \(N(x, H, a)\) requires that both the initial size-density-distribution and the inflow of new recruits be specified. We have primarily
+Modelling the dynamics of \(N(H | x, a)\) requires that both the initial size distribution and the inflow of new recruits be specified. We have primarily
 been interested in the dynamics of a patch recovering from disturbance and have
@@
 
-{\plant} offers two methods for modelling the dynamics of \(N(x,
-H, a)\) in a competing population. In the first stochastic mode, users
+{\plant} offers two methods for modelling the dynamics of \(N(H | x, a)\)
+in a competing population. In the first stochastic mode, users
 specify an average seed rain and patch size, and {\plant} then generates a
@@
 to alter the density of individuals along these trajectories
-(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{ebt}) method
+(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train method
 \citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
 \citet{Falster-2015}, but not identical.
-An advantage of the characteristic method is that it allows direct approximation of the size-density-distribution.
-For the same reason, however, the characteristic method may be less suitable for resolving size-density-distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).
+An advantage of the characteristic method is that it allows direct approximation of the size distribution.
+For the same reason, however, the characteristic method may be less suitable for resolving size distributions in models where those distributions naturally develop sharp peaks (which happens, e.g., when plants completely cease to grow at a given height).
 
@@
 create a self-regulating system.
-{\plant} yields as outputs the equilibrium seed rains and size-density-distributions for each considered species, based on the combined effects of the model's physiological rules, species traits, competitive interactions
+{\plant} yields as outputs the equilibrium seed rains and size distributions for each considered species, based on the combined effects of the model's physiological rules, species traits, competitive interactions
 and disturbance regime.

### 2015-11-09 ms.tex
Some final minor edits

@@
  in patches of competing plants, or in metapopulations under a
- disturbance regime. Patches can be modelled either as a stochastic
- realisation of a finite-sized population, or as their deterministic limit.
+ disturbance regime.
  These dynamics can be integrated into metapopulation--level
@@
 On top of the physiological model, {\plant} implements methods for
-population dynamics and evolutionary dynamics (Fig. \ref{fig:schematic}), following methods
+population dynamics and adaptive dynamics (Fig. \ref{fig:schematic}), following methods
 described by \citet{Falster-2011} and \citet{Falster-2015}. Demographic
@@
 {\plant} offers the capacity to model emergent phenomena in patches with
-specified size, and also through a deterministic approximation
+specified area, and also through a deterministic approximation
 where specifying a patch size is not required. Trait evolution and community
@@
 additional physiologies to be introduced. Users are able to modify the
-physiological model in two different ways. First, one can vary the
+physiological model in three different ways. First, one can vary the
 parameters of the \texttt{FF16} model directly (see Appendix
@@
 types \citep{Wright-2004},
-$$k_{\rm l} = \beta_{kl1} \, \left(\phi / \phi_0\right)^{-\beta_{kl2}},$$
+$$k_{\rm l} = \beta_{{\rm kl}1} \, \left(\phi / \phi_0\right)^{-\beta_{{\rm kl}2}},$$
 where $k_{\rm l}$ and $\phi$ (\textsc{lma}) are both parameters of the core physiological
-model, and $\beta_{kl1}, \beta_{kl2}, \phi_0$ are hyper-parameters. Such
+model, and $\beta_{{\rm kl}1}, \beta_{{\rm kl}2}, \phi_0$ are hyper-parameters. Such
 linkages are defined within a user-supplied hyper-parameterisation function
@@
 capture the average behaviour across a large number of small patches
-\citep{Moorcroft-2001}. Importantly, the deterministic mode is
+\citep{Moorcroft-2001}. Importantly, the deterministic mode in {\plant} is
 much faster than the stochastic mode, and eliminates the demographic noise
@@
 Our approach for numerically solving deterministic size-structured population dynamics is
-based on the characteristic method \citep{Angulo-2004}. This method describes the development of \(N\) by approximating the
-density function along a collection of individual growth trajectories spanning the size
+based on the characteristic method \citep{Angulo-2004}. This method describes the development of
+the size distribution\(N\) by approximating it along a collection of individual growth trajectories spanning the size
 spectrum. Following a disturbance, a series of such trajectories are introduced
@@
 to alter the density of individuals along these trajectories
-(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train method
+(Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{EBT})method
 \citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
@@
 under strong competition for light (see Appendix \ref{sec:demography}
-for details). This technique works for both the characteristic method and the Escalator Boxcar Train method. It involves an
+for details). This technique works for both the characteristic method and the \textsc{EBT} method. It involves an
 adaptive refinement of the time points where new growth trajectories or cohorts are introduced
@@
 in {\plant}, such outcomes naturally arise as emergent properties,
-driven by demography (Fig. \ref{fig:patch}c).
+driven by underlying physiology and demography (Fig. \ref{fig:patch}c).
 
@@
 with no single technique guaranteed to find a solution. In {\plant}, we sequentially
-applying rounds of iteration and nonlinear root finding algorithms, while also
+apply rounds of iteration and nonlinear root finding, while also
 using some heuristics to try to ensure that the root returned is an attracting point
-(see Appendix \ref{sec:demography} for details). Real seed rains are of course noisy,
+(see Appendix \ref{sec:demography}, \ref{sec:code} for details). Real seed rains are of course noisy,
 but the aforementioned approach allows for mean tendencies to be estimated (which is required for computing fitness; see below).
@@
 invade when \(R > 1\). The calculation of \(R\) available in
-{\plant} follows equations given by \citet{Falster-2015} (
-see Appendix \ref{sec:demography} for details).
+{\plant} follows equations given by \citet{Falster-2015}
+(see Appendix \ref{sec:demography} for details).
 
@@
 for resident species that are at equilibrium. The slope of the fitness
-landscape with respect to traits determines the direction of selection
-(dashed line in Fig. \ref{fig:fitness}a), i.e., a negative slope indicates selection for
+landscape with respect to traits determines the direction of selection,
+for example the negative slope of the dashed line around the resident
+in Fig. \ref{fig:fitness}a)indicates selection for
 smaller trait values (see Appendix \ref{sec:code} for
@@
 strengths. The \textsc{EBT} software (see De Roos 2016 in this special feature) provides
-a flexible framework that has been used to study a range of marine organisms.
+a flexible framework that has been used to study a range of organisms.
 Compared to the \textsc{EBT}, {\plant} is customised for plants, and
@@
 structure across broad geographic areas, and their architectures are
-optimised for such use. By contrast, {\plant} was is designed so that the
+optimised for such use. By contrast, {\plant} was designed so that the
 physiological rules driving the dynamics of the system could be easily
@@
 trait mixtures in communities. A variety of other uses are
-possible, addressing other aspects of community assembly. For
+possible, addressing other aspects of community assembly and function. For
 example, we expect the package will prove useful for i) investigating
@@
  Fig. \ref{fig:plant}. (a) Growth trajectories
- for individuals germinating in successive ``cohorts'' for each
+ for individuals germinating across a range of patch ages for each
  species, using a schedule of introduction times generated by
@@
  unit leaf area) trait, showing potential for the stable coexistence of
- multiple plant types \citep[adapted from][]{Falster-2015}. Invasion fitness is the logarithm of the population growth rate --
+ multiple plant types \citep[adapted from][]{Falster-2015}. Invasion fitness is the
+ logarithm of the basic reproduction ratio --
  at zero fitness, species are at equilibrium, at positive fitness, they increase in
  density. (a) Fitness landscape generated by a single species (indicated by
- a circle). For an \textsc{lma} of 0.08 (solid line, filled circle), there is
- directional selection for larger \textsc{lma}. For an \textsc{lma} of
- 0.1 (dashed line, open circle), there is a branching point (see inset). (b) Fitness landscape generated by two species, holding
- the first at an \textsc{lma} of 0.08. Introducing a species at the
+ a circle). For an \textsc{lma} of 0.2 (dashed line, open circle), there is
+ directional selection for smaller \textsc{lma}. For an \textsc{lma} of
+  0.0825 (solid line, filled circle), there is a branching point (see also inset). (b) Fitness landscape generated by two species, holding
+ the first at an \textsc{lma} of 0.0825. Introducing a species at the
  maximum of the solid line in (a) reduces the region of positive

### 2015-11-09 ms.tex
Resolve final issues with figures
- changed location of two species in figure 5 and thus other figures two. Between our first and second submission to MEE the location
of the ESS moved slightly and we failed to notice. Have now updated.
- Figure patch needed more space on right, and formatting of numbers.

@@
  trait and light environment. Solid lines refer to the high-\textsc{lma} (leaf mass per unit leaf area) strategy (\textsc{lma} =
- 0.267), while dashed lines refer to the low-\textsc{lma} strategy
- (\textsc{lma} = 0.08). (a) Growth trajectories are influenced by
+ 0.2625), while dashed lines refer to the low-\textsc{lma} strategy
+ (\textsc{lma} = 0.0825). (a) Growth trajectories are influenced by
  both light environment and traits. (b) Over time, the fraction of

### 2015-11-09 R/figures_emergent.R

@@ -6,7 +6,7 @@ make_emergent_data <- function(hires=TRUE) {
     p0$control$ode_tol_abs <- 1e-6
   }
   p0$disturbance_mean_interval <- 30.0
-  p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
+  p1 <- expand_parameters(trait_matrix(0.0825, "lma"), p0, FALSE)
   p1 <- build_schedule(p1)
   data <- run_scm_collect(p1)
 

### 2015-11-09 R/figures_fitness.R

@@ -14,12 +14,11 @@ make_fitness_data <- function() {
   ## figures_growth.R:make_growth_data() and
   ## figures_plant.R:figure_plant()
   communities <-
-    list(first      = 0.10,           # first individual
-         attractor1 = 0.08,           # first branching point
-         second1    = c(0.08, 0.22),  # ~= the max fitness
-         second2    = c(0.08, 0.25),  # ~= the max fitness
-         second3    = c(0.08, 0.30),  # ~= the max fitness
-         attractor2 = c(0.08, 0.267)) # stable
+    list(first      = 0.2,           # first individual
+         attractor1 = 0.0825,           # first branching point
+         second1    = c(0.0825, 0.355),  # ~= the max fitness
+         second2    = c(0.0825, 0.2878),  # ~= the max fitness
+         attractor2 = c(0.0825, 0.2625)) # stable
 
   lma_mutant <- seq_log_range(bounds, 101)
 
@@ -75,14 +74,12 @@ figure_fitness_landscape <- function(data) {
   label_panel(2)
   mtext(expression("Leaf mass per unit leaf area"~("LMA;"~kg~m^-2)), 1, 3, cex=1)
   lines(lma[i], w$second2[i], lty=3)
-  lines(lma[i], w$second3[i], lty=3)
   lines(lma, w$attractor2)
   abline(h=0)
   points(communities$attractor2[[1]], 0, pch=19)
   points(communities$attractor2[[2]], 0, pch=19)
   points(communities$second1[[2]], 0)
   points(communities$second2[[2]], 0, col="grey")
-  points(communities$second3[[2]], 0, col="grey")
 
   ## TODO: Would be nice to scale the spacing automatically.
   par(fig=c(px, py), mar=rep(0, 4), new=TRUE)

### 2015-11-09 R/figures_patch.R

@@ -3,9 +3,9 @@ make_patch_data <- function() {
   p0$control$equilibrium_nsteps <- 30
   p0$control$equilibrium_solver_name <- "hybrid"
   p0$disturbance_mean_interval <- 30.0
-  ## NOTE: These lma values; 0.08 and 0.267 are the equilibrium values
+  ## NOTE: These lma values; 0.0825 and 0.2625 are the equilibrium values
   ## in the fitness figure (see attractor2).
-  p2 <- expand_parameters(trait_matrix(c(0.08, 0.267), "lma"), p0, FALSE)
+  p2 <- expand_parameters(trait_matrix(c(0.0825, 0.2625), "lma"), p0, FALSE)
   p2_eq <- equilibrium_seed_rain(p2)
   run_scm_collect(p2_eq, TRUE)
 }
@@ -38,7 +38,7 @@ figure_patch <- function(data) {
   ## col1[rd1 < 0.05] <- NA
   ## col2[rd2 < 0.05] <- NA
 
-  op <- par(mfrow=c(3, 1), mar=c(1.0, 3.6, .5, .5), oma=c(2.6, 0, 0, 0),
+  op <- par(mfrow=c(3, 1), mar=c(1.0, 3.6, .5, 3), oma=c(2.6, 0, 0, 0),
             mgp=c(2.3, 1, 0))
   on.exit(par(op))
   matplot(t, h1, type="l", col=make_transparent(cols[[1]], 0.5), lty=1,
@@ -71,7 +71,7 @@ figure_patch <- function(data) {
   # find average value over metapopulation
   y_av <- apply(y, 2, function(x) plant:::trapezium(t, x*data$patch_density))
   for(i in seq_len(3)) {
-    axis(4, at=y_av[i], tck=0.1, col.ticks=cols[i], lty = lty[i])
+    axis(4, at=y_av[i], labels=sprintf("%0.2f",y_av[i]), tck=0.1, col.ticks=cols[i], lty = lty[i], las=1)
   }
   axis(1, at=108, labels="Average")
 }

### 2015-11-09 R/figures_plant.R

@@ -52,13 +52,12 @@ figure_plant <- function() {
   ## plot).
   p <- scm_base_parameters()
   p$disturbance_mean_interval <- 30.0
-  ## NOTE: These lma values; 0.08 and 0.267 are the equilibrium values
+  ## NOTE: These lma values; 0.0825 and 0.2625 are the equilibrium values
   ## in the fitness figure (see attractor2).
-  s1 <- strategy(trait_matrix(0.08,  "lma"), p)
-  s2 <- strategy(trait_matrix(0.267, "lma"), p)
+  s1 <- strategy(trait_matrix(0.0825,  "lma"), p)
+  s2 <- strategy(trait_matrix(0.2625, "lma"), p)
   env1 <- fixed_environment(1.0)
   env2 <- fixed_environment(0.5)
-
   ## Generate most of the data here:
   heights1 <- seq(FF16_PlantPlus(s1)$height, s1$hmat, length.out=100L)
   heights2 <- seq(FF16_PlantPlus(s2)$height, s2$hmat, length.out=100L)

### 2015-11-09 vignettes/src/cohort_spacing.R

@@ -12,7 +12,7 @@ library(plant)
 library(parallel)
 
 p0 <- scm_base_parameters()
-p <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
+p <- expand_parameters(trait_matrix(0.0825, "lma"), p0, FALSE)
 p$seed_rain <- 20 # close to equilibrium
 
 ## The default cohort introduction times are designed to concentrate

### 2015-11-09 vignettes/src/emergent.R

@@ -14,7 +14,7 @@ p0$control$equilibrium_solver_name <- "hybrid"
 p0$disturbance_mean_interval <- 30.0
 
 ## We'll work with a single species at equilibrium
-p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
+p1 <- expand_parameters(trait_matrix(0.0825, "lma"), p0, FALSE)
 p1_eq <- equilibrium_seed_rain(p1)
 data1 <- run_scm_collect(p1_eq)
 

### 2015-11-09 vignettes/src/equilibrium.R

@@ -24,7 +24,7 @@ cobweb <- function(m, ...) {
 
 ## Set model parameters
 p0 <- scm_base_parameters()
-p <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
+p <- expand_parameters(trait_matrix(0.0825, "lma"), p0, FALSE)
 
 ## Some output seed rains, given an input seed rain:
 run(1.0, p)
@@ -85,7 +85,7 @@ abline(0, 1, lty=2, col="grey")
 cobweb(approach, lty=3)
 
 ## # 5. Multiple species at once:
-lma <- c(0.08, 0.15)
+lma <- c(0.0825, 0.15)
 p2 <- expand_parameters(trait_matrix(lma, "lma"), p0, FALSE)
 
 p2_eq <- equilibrium_seed_rain(p2)

### 2015-11-09 vignettes/src/fitness.R

@@ -48,8 +48,8 @@ abline(h=0, col="grey")
 points(lma1, rep(0, 4), col=1:4, pch=19)
 
 ## For this system, there is an evolutionary attractor around
-## lma 0.08:
-lma_b <- 0.08
+## lma 0.0825:
+lma_b <- 0.0825
 p1b <- add_eq(lma_b, p0)
 
 w1b <- fitness_landscape(lma, p1b)
@@ -68,7 +68,7 @@ plot(lma_detail, w1b_detail, log="x", type="l", las=1,
 abline(h=0, col="grey")
 points(lma_b, 0, pch=19)
 
-## Holding the first species at 0.08 we can introduce additional
+## Holding the first species at 0.0825 we can introduce additional
 ## species (it's close enough to the optimum here, though in general
 ## this point might move substantially as new species are introduced).
 

### 2015-11-09 vignettes/src/patch.R

@@ -12,7 +12,7 @@ p0$control$equilibrium_solver_name <- "hybrid"
 p0$disturbance_mean_interval <- 30.0
 
 ## First, with a single species:
-p1 <- expand_parameters(trait_matrix(0.08, "lma"), p0, FALSE)
+p1 <- expand_parameters(trait_matrix(0.0825, "lma"), p0, FALSE)
 
 ## Run everything out to equilibrium:
 p1_eq <- equilibrium_seed_rain(p1)
@@ -110,7 +110,7 @@ plot(data1$time, lai1, type="l", las=1, xlab="Time (years)",
 ## another.  This adds a second species -- with a higher lma value
 ## than the first species -- to the population and runs them until
 ## they reach equilibrium seed rain at the metapopulation level.
-p2 <- expand_parameters(trait_matrix(0.267, "lma"), p1, FALSE)
+p2 <- expand_parameters(trait_matrix(0.2625, "lma"), p1, FALSE)
 p2_eq <- equilibrium_seed_rain(p2)
 
 ## Then collect the patch-level dynamics:

### 2015-11-09 vignettes/src/plant.R

@@ -181,9 +181,9 @@ plot(res_mass$time, mass, type="o", pch=19, las=1, xlab="Time (years)")
 
 p <- scm_base_parameters()
 ## Low LMA ("fast growth") species
-s1 <- strategy(trait_matrix(0.08,  "lma"), p)
+s1 <- strategy(trait_matrix(0.0825,  "lma"), p)
 ## High LMA ("low growth") species
-s2 <- strategy(trait_matrix(0.267, "lma"), p)
+s2 <- strategy(trait_matrix(0.2625, "lma"), p)
 
 ## Note that we're using an alternative way of specifying strategies
 ## here, to trigger our "hyper-parametrisation" approach.  This may be

### 2015-11-11 ms.tex
Add references to other papers in same issue (requested by editors)

@@
 instantaneous physiological effects to long-term evolutionary outcomes
-\citep{Harper-1977, Westoby-2002}.
+\citep{Harper-1977, Westoby-2002, Griffith-2016, Rees-2016}.
 %
@@
 leaf and wood construction costs influence growth rates
-\citep{Wright-2010}, while seed size and height
+\citep{Falster-2011, Visser-2016}, while seed size and height
 at maturation determine start and end points of ontogenetic
@@
 size structure is rarely considered, at least in plant models
-\cite[for animal examples, see][]{Deroos-1988, Deroos-1992}.
+\cite[for animal examples, see][]{Deroos-1997}.
 %
@@
 increasingly recognised in both evolutionary ecology
-\citep[e.g.,][]{Falster-2015} and vegetation dynamics
+\citep[e.g.,][]{Falster-2015, Rees-2016} and vegetation dynamics
 \citep[e.g.,][]{Moorcroft-2001, Purves-2008, Smith-2014,
@@
 provide unique opportunities for studying coexistence via differentiation in
-successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001}.
+successional strategy \citep[see also][]{Huston-1987, Moorcroft-2001, Uriate-2016}.
 At the same time, we expect {\plant} to be useful for modelling other demographic and evolutionary phenomena.
@@
 layout within patches. Similar assumptions are made in other \textsc{TSPMs}
-\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}.
+\citep{Shugart-1980, Kohyama-1993, Huston-1987, Moorcroft-2001, Smith-2014}, and also
+in related approaches, such as Integral Projection Models \citep[e.g.,][]{Rees-2016}.
 While one would ideally also consider spatial interactions within patches,
@@
 (Fig. \ref{fig:patch}b). The characteristic method is similar to the Escalator Boxcar Train (\textsc{EBT})method
-\citep{Deroos-1988, Deroos-1997} used by \citet{Falster-2011} and
+\citep{Deroos-1997} used by \citet{Falster-2011} and
 \citet{Falster-2015}, but not identical.
@@
 All vegetation in {\plant} is subject to a disturbance regime, implying that patches
-are cleared at a given rate. Such clearance can be interpreted, for example
+are cleared at a given rate \citep[e.g.,][this issue]{Treurnicht-2016}.
+Such clearance can be interpreted, for example
 as resulting from a fire, cyclone, clear-cutting or
@@
 Increased interest in capturing the dynamics of size-structured populations
+\citep[e.g.][]{Griffith-2016}
 has prompted development of several tools, each with its own distinct
-strengths. The \textsc{EBT} software (see De Roos 2016 in this special feature) provides
-a flexible framework that has been used to study a range of organisms.
-Compared to the \textsc{EBT}, {\plant} is customised for plants, and
+strengths. The \textsc{EBT} software \citep{Deroos-1997}, {\plant} is customised for plants, and
 to that end offers refinements needed to handle modelling of the strongly

### 2016-01-15 R/tables.R
Incorporate suggestions from Ulf and Ake on supplmentary materials

@@ -20,8 +20,8 @@ table_plant_parameters <- function(filename, dest){
   tab$Value <- NA_real_
 
   # Now extract actual values from plant
-  to_code <- tab[["\\plant"]] != "" & tab[["\\plant"]] %in% names(s)
-  code <- tab[["\\plant"]][to_code]
+  to_code <- tab[["Code"]] != "" & tab[["Code"]] %in% names(s)
+  code <- tab[["Code"]][to_code]
 
   val <- s[code]
   err <- sapply(val, is.null)
@@ -34,10 +34,10 @@ table_plant_parameters <- function(filename, dest){
   tab$Value[to_code] <- unlist(lapply(val, prettyNum))
 
   # Display names in latex
-  to_code <- tab[["\\plant"]] != ""
+  to_code <- tab[["Code"]] != ""
 
-  tab[["\\plant"]][to_code] <- sprintf("\\texttt{%s}",
-                                      gsub("_", "\\\\_", tab[["\\plant"]][to_code]))
+  tab[["Code"]][to_code] <- sprintf("\\texttt{%s}",
+                                      gsub("_", "\\\\_", tab[["Code"]][to_code]))
 
   x <- xtable(tab,
               hline.after=c(1),

### 2016-01-18 ms.tex
Add section on data (and code) accessibility (as requested)
They ask for data, but i decided to add link to github repository
reproducing paper as well
Rename strategy object from FFDev to FF16r
    - triggered by final submission of paper to MEE
    - better to have date in alternative strategy name
    - reflects changes made in traitecoevo/plant/0fe3ada1d85720821ec9a096d309295cf04ef176

@@
 model in a new C++ file and compiling that into the package. As an
-example, in the \texttt{FFdev} physiology, we have shown how to modify the
+example, in the \texttt{FF16R} physiology, we have shown how to modify the
 function determining allocation to reproduction.
@@
 
+\section{Data Accessibility and Reproducibility}
+
+Source code for reproducing the entire contents of this paper is available at
+\href{https://github.com/traitecoevo/plant\_paper}{github.com/traitecoevo/plant\_paper};
+see also Appendix \ref{sec:code} for worked examples reproducing the figures.
+This paper does use any data.
+
 \section{Author contribution statement}

### 2016-01-20 vignettes/src/parameters.R
Incorporate extra corrections on Supp Matt from Ulf and Ake

@@ -31,7 +31,7 @@ pl$strategy$lma
 ## `lma` affects a few places in the model; see [the source
 ## code](https://github.com/traitecoevo/plant/blob/master/src/ff16_strategy.cpp),
 ## but only really for converting from leaf area to leaf mass (leaf
-## mass being leaf area multiplied by leaf mass per leaf area).
+## mass being leaf area multiplied by leaf mass per unit leaf area).
 ## However, as a component of the leaf economic spectrum we imagine
 ## LMA as affecting a number of other components of the model.
 
